### 2014-12-02 paper.tex

\batchmode
\makeatletter
\def\input@path{{/accounts/gen/vis/paciorek/research/jmac/composition_paper//}}
\makeatother
\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage{array}
\usepackage{amsmath}
\usepackage{graphicx}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{times}
\usepackage{graphics,natbib}

\newcommand{\matern}{Mat\'{e}rn }
\newcommand{\N}{\mathcal{N}}

\makeatother
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\title{Statistically-estimated tree composition for the northeastern United
States \\
at the time of European settlement}
\maketitle
\begin{quote}
Christopher J. Paciorek, Andrew Thurman, Simon J. Goring, Charlie
V. Cogbill, \\
John W. Williams, David J. Mladenoff, Jody Peters, Jun Zhu, \\
and Jason S. McLachlan {[}author order to be discussed{]}\\
{[}please tell me if you do/don't want your middle initial{]}
\end{quote}
TODO:

look through code and labbook to make sure have all details

properly cite Goring stuff and reference details of data collection

focus on PLOS One, but Nature's Scientific Data is a possibility

\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the United States and the statistical
methodology used to produce the product. Composition is defined as
the proportion of stems larger than approximately X cm diameter at
breast height {[}Charlie? Simon? is X roughly 10?{]} in 22 taxonomic
groupings, generally at the genus level. The data come from settlement
survey records that provide raw data that are transcribed and then
aggregated spatially, giving count data. For the western portion of
the domain (Indiana through Minnesota) we use data at survey points
that are aggregated to a regular 8 km grid, while for the eastern
portion (Ohio through Maine), we use data aggregated into townships.
The product is based on a Bayesian statistical model fit to the count
data that estimates composition on a regular 8 km grid. The statistical
model allows us to estimate composition at locations with no data
and to smooth over noise caused by limited counts in locations with
data. Critically, it also allows us to quantify uncertainty in our
composition estimates. We expect this data product to be useful for
understanding the state of vegetation in the northeastern United States
prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

\section{Introduction}

{[}insert paragraph (Jack has volunteered) on context and importance
of historical datasets{]}

\section{Data}

The raw data are obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. For
the states of Ohio, Pennsylvania, New Jersey, New York and the six
New England states (the eastern subdomain), data are aggregated at
the township level. There is also data from a single township in Quebec
and a single township in northern Delaware. Data are essentially complete
in Minnesota, Wisconsin and Michigan but data in Illinois and Indiana
represent a sample of the full set of grid cells, with transcription
ongoing. Data for the remaining states are a subset of the full set
of townships. Fig. \ref{fig:domain} shows the domain, indicating
the grid cells and townships with data.

\begin{figure}
\label{fig:domain}

\includegraphics{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}

\caption{Spatial domain, with locations with data shown in gray. Locations
are grid cells in western portion and townships in eastern portion.}
\end{figure}

Note that surveys occurred over a period of more than 200 years as
European colonists (before U.S. independence) and the United States
settled what is now the northeastern United States. Our estimates
are for the period of settlement represented by the survey data and
therefore are time-transgressive; they do not represent any single
point in time across the domain, but rather the state of the landscape
at the time just before settlement occurred {[}get ref from Charlie
or Jason's historian pal{]}

Extensive details on the data are available in \cite{Gori:etal:2015}
and the raw data are available at XXX. The aggregation into taxonomic
groups is primarily at the genus level, but in some cases of monospecific
genera, is at the species level. In some other cases, because of ambiguity
in the common tree names used by surveyors, a group represents trees
from different families (orders?). We model the following taxa: {[}list
all{]}. {[}include description of size cutoff{]}

There are approximately 520,000 trees from the western subdomain and
420,000 trees from the eastern subdomain. In each subdomain, oak is
the most common taxon and pine the second most common.

Our domain is a rectangle covering all of the states, in the Albers
projection, with the rectangle split into 8 km cells, arranged in
a 296 by 180 grid of cells, with the centroid of the cell in the southeast
corner located at (-71000, 58000). For the modeling of the western
subdomain we use the western-most 146 cells. For the modeling of the
western subdomain we use the eastern-most 180 cells and omit 23 cells
in the north and 17 cells in the south. {[}check 23 in N vs 17 in
N; also check what is in netCDF product - should be for the full domain
with missing values?{]}

\section{Statistical model}

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
based on the full population of all the trees in an areal region.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition of the taxa at all of the grid cells. These samples
are the data product and can then be used in analyses. We also provide
the mean and standard deviation of the samples, which represent our
best estimate of composition and a Bayesian ``standard error'' for
the estimate.

\subsection{Data model}

The statistical model treats the observations as having a multinomial
distribution with a (latent) vector of proportions for each grid cell.
\[
y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i}))
\]
where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
location, $n_{i}$ is the number of trees counted in the location,
and $\theta(s_{i})$ is the vector of unknown proportions for those
taxa at that location. Note that we use a standard multinomial distribution
without overdispersion as the set of trees in the dataset is roughly
uniformly sampled across the cells or townships \citep{Gori:etal:2015}.

In what follows we describe the basic model for those states for which
we have raw data on the 8 km grid and in Section \ref{sub:Model-for-township}
we describe the extension of the model to accommodate data aggregated
to the township level.

The proportions, $\theta_{p}(s_{i}),\, p=1,\ldots,P$, are modeled
spatially by a set of $P$ Gaussian spatial processes, one per taxon,
$\alpha_{p}(s_{i}),\, p=1,\ldots,P$. This collection of processes
defines a multivariate spatial process for composition. The $\alpha_{p}(s)$
processes are defined on the 8 km grid, $\alpha_{p}=\{\alpha_{p}(s_{1}),\ldots,\alpha_{p}(s_{m})\}$
for the $m$ grid cells. In Section \ref{sub:Latent-Variable-Model}
we introduce a multinomial probit model that relates the $\alpha_{p}(s)$
processes to the proportion processes, $\theta_{p}(s)$, via the introduction
of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.

The critical component of the statistical model is the representation
of $\alpha_{p}(s)$ as a spatial process. This process is a prior
structure that serves to smooth across noise in the observations and
allows for interpolation to locations with no data. Apart from the
sum to one constraints, the taxa are considered to be independent
in the prior, as we did not want to impose any structure that ties
the different taxa together, as any correlation will likely vary across
space.

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \citep{Bane:etal:2003} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with Matern
covariance \citep{Lind:etal:2011}.

\subsection{Spatial process models}

MRF models work directly with the precision matrix of the values of
the spatial process, so calculation of the prior density of $\alpha_{p}(s)$
is computationally simple \citep{Rue:Held:2005}, but in situations
where the likelihood is not normal, it can be difficult to set up
effective MCMC algorithms that are able to move in the high-dimensional
space of $\alpha_{p}$. The latent variable representation of Section
\ref{sub:Model-for-township} helps to alleviate this problem.

\subsubsection{Standard conditional autoregressive models}

Our first model is a standard conditional autoregressive (CAR) model
\citep{Bane:etal:2003}. We use a standard form of this model, which
treats the four cardinal neighbors of each grid cell as the neighbors
of the grid cell. The corresponding precision matrix has diagonal
elements, $Q_{ii}$, equal to the number of neighbors for the $i$th
area (i.e., four except for cells on the boundary of the domain),
while $Q_{ik}=-1$ (the negative of a weight of one) when areas $i$
and $k$ are neighbors and $Q_{ik}=0$ when they are not. This gives
the following model for the values of $\alpha_{p}(s_{i})$ collected
as a vector across all of the grid cells, $i=1,\ldots,m$: 
\[
\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-})
\]
The use of the generalized inverse notation indicates that $Q$ is
not full-rank, but is of rank $m-1$; this gives an improper prior
on an implicit overall mean for the process values. This specification
is called an \textit{intrinsic conditional autoregression (ICAR)}
and $Q$ can also be expressed as $D-C$ where $C$ is the $m\times m$
adjacency matrix defining the neighborhood relation of the locations;
that is $(C)_{ik}=1$ if locations $i$ and $k$ are neighbors. The
matrix $D$ is an $m\times m$ diagonal matrix containing the row
sums of matrix $C$ as the diagonal entries $(D)_{ii}={\displaystyle \sum_{k=1}^{m}(C)_{ik}}.$

We refer to this as the \emph{CAR model}.

\subsubsection{Gaussian process approximation}

Gaussian processes (GP) are also standard models for spatial processes \citep{Bane:etal:2003}.
GP models are computationally challenging for large datasets because
of manipulations involving large covariance matrices. Given this,
\cite{Lind:etal:2011} proposed a new framework for using Gaussian
MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
partial differential equations (SPDE).

We consider Gaussian processes in the \matern class, using the following
parameterization of the \matern correlation function as 
\begin{equation}
R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
\end{equation}
where $d$ is Euclidean distance, $\rho$ is the spatial range parameter,
and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
the second kind, whose order is the smoothness (differentiability)
parameter, $\nu>0$. $\nu=0.5$ gives the exponential covariance.
For any pair of locations, $R(d)$ defines the correlation of the
process, e.g., $\alpha_{p}(s)$, as a function of the distance between
the locations. Considering all pairs of locations, this defines a
correlation matrix for all locations of interest.

The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
to the Matern-based GPs for $\nu=1$ and $\nu=2$. Our second model
is the Lindgren approximation for Matern-based GPs with $\nu=1$.
To implement the Lindgren model, one modifies the $Q$ matrix defined
previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$ and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

The primary difference between the CAR and Lindgren models is that
the Lindgren model provides an additional degree of freedom by estimating
$\rho$. In particular $\rho$ allows us to estimate the locality
of the smoothing. As $\rho$ decreases, the model uses more and more
local data to estimate the compositional proportions at a given location,
effectively averaging the empirical proportions over smaller neighborhoods.
In general, the \cite{Lind:etal:2011} model will generally provide
for a smoother estimate than the CAR model \citep{Paci:2013}.

To ensure that the $\sigma^{2}$ parameter is equivalent between the
two models, we reparameterize, producing our second model:
\[
\alpha_{p}\sim\mbox{N}(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1})
\]

We refer to this model as the \emph{SPDE model}.

\subsection{Prior Distributions}

\noindent The ICAR specification contains a set of hyperparameters
$\sigma_{p}^{2}$ for $p$ = $1,...,P$. Following \citep{Gelm:2006}
we use a uniform distribution on each $\sigma_{p}$ parameter, with
upper bound of {[}check max values{]}. For the SPDE model we also
have parameters $\mu_{p}$, which we give flat, non-informative priors
(truncated at $\pm10$), and $\rho_{p}$ which we give uniform priors
on the interval $(0.1,\exp(5))$.

\subsection{Latent Variable Model\label{sub:Latent-Variable-Model}}

It is well-known that devising an effective MCMC algorithm for models
with latent Gaussian process(es) and a non-Gaussian likelihood is
difficult (cite recent cent/noncent papers). To develop an algorithm,
we make use of a latent variable representation for the multinomial
probit model \citep{McCu:Ross:1994}. The representation introduces
latent variables that allow one to develop a MCMC sampling strategy
that takes advantage of closed form full conditional distributions
(so-called Gibbs sampling steps) for $\alpha_{p}$.

Suppose that compositional counts are available at a number of locations.
At location $i$, a sample size of $n_{i}$ observations are collected,
and each observation (i.e., each tree) can be classified into $P$
distinct categories. For a given tree $j$ at location $i$, let $Y_{ij}$
denote the response variable indicating the category. Let $Y_{ij}$
be associated with $P$ latent variables $W_{ij1},...,W_{ijP}$ such
that $Y_{ij}$ = $p$ if and only if $W_{ijp}={\displaystyle \max_{p'}\big\{ W_{ijp'}\big\}}$;
in other words, the maximum of the set of latent variables $\{W_{ijp}\}{\displaystyle _{p=1}^{P}}$
determines the category of observation $j$ at location $i$.

\noindent The final piece of the latent variable representation is
the relationship between the $W$ variables and the $\alpha_{p}(s)$
processes. We have that

\noindent 
\[
W_{ijp}\sim\mbox{N}(\alpha_{p}(s_{i}),1)
\]
independently for all of the $W_{ijp}$ values.

\noindent Consider the following example with two locations that are
neighbors and $P=2$ categories. Each tree $j$ at location $i$ is
associated with two variables $W_{ij1}$ and $W_{ij2}$, governed
by the latent variables $\alpha_{i1}$ and $\alpha_{i2}$, respectively.
Suppose that $\alpha_{i1}>\alpha_{i2}$ for a given location $i$.
Then this model implies that any tree $j$ is more likely to be labeled
1 than 2 at location $i$. The difference between $\alpha_{i1}$ and
$\alpha_{i2}$ explains the \textit{difference} in probability of
\textit{categories} 1 and 2 at location $i$, and the similarity between
$\alpha_{1p}$ and $\alpha_{2p}$ explains the \textit{correlation}
between the probabilities at \textit{locations} 1 and 2 for category
$p$.

\subsection{Model for township data\label{sub:Model-for-township}}

We developed an extension of the model described in previous sections
to account for data at a different aggregation than our core 8 km
grid. This extension introduces a new set of latent variables, one
per tree, that indicate the grid cell in which the tree is located.
These latent 'membership' variables, $c_{tj}$, for tree $j$ in township
$t$ can be sampled within the MCMC as additional unknown parameters.
The prior for $c_{tj}$ is a discrete distribution that puts mass
proportional to the overlap between the township in which the tree
is located and the grid cells that intersect the township, $\psi_{1(t)},\ldots,\psi_{T(t)}$,
with the priors across different trees in a township being independent.
\[
c_{tj}\sim\mbox{Multinom}(1,\{\psi_{1(t)},\ldots\psi_{T(t)}\}).
\]

In updating the other parameters in the model, we condition on the
current values, $c_{tj}$, which provides a ``soft'' assignment
of trees to grid cells that respects both the township in which the
tree was sampled and the uncertainty in which grid cell the tree was
sampled.

Note that this prior has the unrealistic feature that it does not
represent our knowledge that the trees in a township would be distributed
more regularly across the area of the township than expected by such
an independence prior.

\subsection{Computation}

The \cite{McCu:Ross:1994} representation is convenient for MCMC sampling,
particularly in this high-dimensional spatial context, as it allows
us to draw from the posterior conditional distribution of the $W_{ijp}$
variables (these distributions are truncated normal) in closed form
and to draw the entire vector of latent process values, $\alpha_{p}$,
as a single sample that respects the spatial dependence structure
for each taxon.

While the latent variable representation provides great advantages
in the MCMC sampling for each $\alpha_{p}$ compared to joint Metropolis
updates or single location at a time updates, there is still strong
dependence between the hyperparameters, $\{\sigma_{p}^{2},\mu_{p},\rho_{p}\}$
and the the latent process values (as well as between the latent process
values and the latent $W_{ijp}$ variables). To address the first,
we developed a 'cross-level' joint updating strategy in which we propose
$\phi\in\{\sigma_{p}\},p=1,\ldots,P$ and for the SPDE model, $\phi\in\{\{\sigma_{p}\},\{\mu_{p}\},\{\rho_{p}\}\},$
via a Metropolis-style random walk and then conditional on the proposed
value of $\phi$ propose $\alpha_{p}$ from its full conditional distribution
\textbackslash{}$\alpha_{p}$ and sampling from the marginalized distribution
$\phi|W$. Note that in sampling $\sigma_{p}$and $\rho_{p}$ in the
SPDE model, for each $p$, we jointly propose $\{\sigma_{p},\rho_{p}\}$
using an adaptive Metropolis proposal and then use the cross-level
strategy to also propose $\alpha_{p}$ as part of a single accept-reject
step. For these various joint samples of hyperparameters and $\alpha_{p}$,
we use adaptive Metropolis sampling \citep{Shab:Well:2011}.

The full description of the MCMC sampling steps is provided in the
Appendix. In addition, in the latent variable representation, $\theta_{p}(s)$
never appears explicitly and cannot be calculated in close form. Instead
we use Monte Carlo integration over $W_{ijp},\, p=1,\ldots,P$ to
estimate $\theta_{p}(s_{i})$, also described in the Appendix.

The model is implemented in R with core computational calculations
coded in C++ using the Rcpp package \citep{Edde:Fran:2011}. We also
make extensive use of sparse matrix representations and algorithms,
using the spam package in R \citep{Furr:Sain:2010}. All code is available
on Github, including pre- and post-processing code.

\section{Model comparison}

\subsection{Design}

We compared the CAR and SPDE models using cross-validation by holding
out data from the fitting process and assessing the fit of the model
on the held-out data. We used two types of test data:
\begin{enumerate}
\item We held out all the data from 95\% of the cells in Minnesota, with
cells selected at random. This was meant to assess the ability of
the model to interpolate from a sparse set of cells/townships and
mimics the limited data in Illinois and Indiana.
\item We held out 5\% of the trees from all of the trees in the dataset
(leaving aside the held-out Minnesota cells). This was meant to assess
the ability of the model to estimate the composition in cells in which
data were available. 
\end{enumerate}
Finally, in a separate sensitivity analysis we instead left out 80\%
of the cells in Minnesota at random. This variation on the first test
above was meant to indicate whether our model comparison conclusions
would be robust as the digitization process for Illinois and Indiana
progresses and provides us with increasingly dense data.

There has been extensive work in the statistical literature on good
metrics to use to compare the predictive ability of models; these
metrics are referred to as scoring rules. In particular it is well-recognized
that predictive distributions should maximize sharpness subject to
calibration. That is the predictive distribution should be as narrow
as possible while being calibrated such that the observations are
consistent with the distribution \citep{Gnei:etal:2007}. When thinking
in terms of prediction intervals, we seek intervals that are as narrow
as possible while still covering the truth the expected proportion
of the time. Following the suggestions in \citep{Gnei:etal:2007},
we consider these metrics.

We used the following criteria to compare models. For experiment 1,
$Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$ is the count of all trees in the
held-out cell and for experiment 2, $Y_{i}$ is the count of held-out
individual trees in the location.
\begin{enumerate}
\item Brier score: \citep{Gnei:etal:2007} suggest this measure, which has
been in use for decades. For multi-category as opposed to binary outcomes,
this takes the form
\[
\frac{1}{n_{i}}\sum_{j=1}^{n_{i}}\sum_{p=1}^{P}(y_{ijp}-\theta_{p}(s_{i}))^{2}
\]
for each cell. The full Brier score for the complete set of held-out
locations sums over all locations.
\item Log predictive density: This measure takes the log of the density
of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\theta_{1}(s_{i}),\ldots,\theta_{P}(s_{i})\})$,
summing on the log scale across all of the held-out data.

While in principle, this metric should be optimal \citep{Krnj:Drap:2014},
it suffers from a lack of robustness in being very sensitive to small
predictions near zero \citep{Gnei:etal:2007}. Even worse, our Monte
Carlo estimation of $\theta$ used 10000 samples, so in some cases
$\theta_{p}(s)=0$. When a tree is present in a cell but its corresponding
proportion is 0 this gives a log density of $-\infty$, preventing
use of the metric. As an informal solution to this we set $\theta_{p}(s)=0.5\frac{1}{10000}$
in such cases, but given these issues we treat the log predictive
density as a secondary measure.

\item Weighted root mean square prediction error (RMSPE) and mean absolute
error (MAE) for individual cells (only for experiment 1), where we
weight by the number of held-out trees to account for the greater
variability in the empirical proportions in locations with few held-out
trees. 
\item Coverage and length of 95\% prediction intervals for $Y_{ip}$. 
\end{enumerate}
For each of these metrics, we calculated the metric using the posterior
mean composition estimates (as a measure of our core predictions)
and averaging the metric over the posterior samples (as a measure
of our full data product, including uncertainty). In other words in
the first we first average over the composition posterior samples
and in the second we average of the posterior values of the cross-validation
metric.

In our fitting we noticed that the SPDE model produced boundary effects
in the predicted composition near the edges of the convex hull of
the observations. To attempt to alleviate this, we added a buffer
zone of 6 grid cells around our entire original domain, though the
boundary effects were still evident even after inclusion of the buffer.
For the model comparison, for comparability, we included this buffer
for both the SPDE and CAR models.

For the model comparison, we used an earlier version of the dataset
than that used for the final data product as the model comparison
runs involved substantial computation. This earlier version had incomplete
data for southern Michigan and was missing a small number of trees
(39 individual trees) classified as ``other hardwood'' due to an
update to our taxonomic aggregations of common names used by surveyors.
Given that part of the goal of the exercise is to understand the ability
of the models to interpolate to areas without data and the ongoing
nature of digitization, we believe this difference in datasets is
not an issue.

We ran each model for XXX iterations, retaining YYY after burnin and
subsampling to reduce storage needs.

\subsection{Results}

Here we summarize the results of our cross-validation analyses that
inform the choice between the CAR and SPDE models.

\subsubsection{Full cell hold-out experiment}

Table \ref{tab:score_cell_fivepercent} shows predictive ability for
Experiment 1: those cells in Minnesota held out of the fitting process,
for both the posterior mean predictions and the full posterior sample.
Table \ref{tab:coverage_cell_fivepercent} shows performance of the
uncertainty estimates. For the posterior distribution over the predictive
score values, the CAR model generally outperforms the SPDE model,
while for the posterior mean predictions, the SPDE model appears to
outperform the CAR model, but we do not have any uncertainty estimates
for this comparison. Coverage and interval lengths are similar between
the two models. From a practical perspective, based on the difference
in mean absolute error, which is on a readily interpretable scale,
the differences between the models are small.

\begin{table}

\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 95\% of entire cells
of data in Minnesota. Smaller values are better.}

\begin{tabular}{|c|c|c|>{\centering}p{3cm}|>{\centering}p{2.5cm}|>{\centering}p{2.5cm}|}
\hline 
 &
\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
\hline 
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}} &
{\small{Posterior Prob. CAR < SPDE}} &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Brier}} &
{\small{84997}} &
{\small{86803}} &
{\small{0.92}} &
{\small{76068}} &
{\small{75330}}\tabularnewline
\hline 
{\small{Negative Log Density}} &
{\small{222144}} &
{\small{242676}} &
{\small{1.00}} &
{\small{185748}} &
{\small{185843}}\tabularnewline
\hline 
{\small{Mean Absolute Error}} &
{\small{0.0382}} &
{\small{0.0390}} &
{\small{0.92}} &
{\small{0.0303}} &
{\small{0.0293}}\tabularnewline
\hline 
{\small{Root Mean Square Error}} &
{\small{0.0944}} &
{\small{0.0986}} &
{\small{0.77}} &
{\small{0.0701}} &
{\small{0.0677}}\tabularnewline
\hline 
\end{tabular}

\label{tab:score_cell_fivepercent}
\end{table}

\begin{table}
\caption{Coverage and length of prediction intervals for the CAR and SPDE spatial
models when holding out 95\% of entire cells of data in Minnesota. }

\begin{tabular}{|c|c|c|}
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Coverage}} &
{\small{0.977}} &
{\small{0.978}}\tabularnewline
\hline 
{\small{Mean Interval Length}} &
{\small{0.135}} &
{\small{0.142}}\tabularnewline
\hline 
{\small{Median Interval Length}} &
{\small{0.040}} &
{\small{0.031}}\tabularnewline
\hline 
\end{tabular}

\label{tab:coverage_cell_fivepercent}
\end{table}

Table \ref{tab:score_cell_20percent} shows the results when the proportion
of cells that are held out decreases from 95\% to 80\%, indicating
performance on less sparse data. Table \ref{tab:coverage_cell_fivepercent}
shows performance of the uncertainty estimates. For denser data, unlike
the results when data are more sparse, the SPDE model generally outperforms
the CAR model, but again differences from a practical perspective,
based on mean absolute error, are limited.

\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 80\% of entire cells
of data in Minnesota. Smaller values are better, except for coverage,
where values near 0.95 are optimal.}

\begin{tabular}{|c|c|c|>{\centering}p{3cm}|c|c|}
\hline 
 &
\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
\hline 
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}} &
{\small{Posterior Prob. CAR < SPDE}} &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Brier}} &
{\small{65870}} &
{\small{65004}} &
{\small{0.08}} &
{\small{60169}} &
{\small{60020}}\tabularnewline
\hline 
{\small{Negative Log Density}} &
{\small{167947}} &
{\small{167536}} &
{\small{0.43}} &
{\small{145404}} &
{\small{145750}}\tabularnewline
\hline 
{\small{Mean Absolute Error}} &
{\small{0.0322}} &
{\small{0.0307}} &
{\small{0.08}} &
{\small{0.0246}} &
{\small{0.0240}}\tabularnewline
\hline 
{\small{Root Mean Square Error}} &
{\small{0.0800}} &
{\small{0.0771}} &
{\small{0.01}} &
{\small{0.0578}} &
{\small{0.0571}}\tabularnewline
\hline 
\end{tabular}

\label{tab:score_cell_20percent}
\end{table}

\begin{table}
\caption{Coverage and length of prediction intervals for the CAR and SPDE spatial
models when holding out 80\% of entire cells of data in Minnesota. }

\begin{tabular}{|c|c|c|}
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Coverage}} &
{\small{0.982}} &
{\small{0.974}}\tabularnewline
\hline 
{\small{Mean Interval Length}} &
{\small{0.117}} &
{\small{0.107}}\tabularnewline
\hline 
{\small{Median Interval Length}} &
{\small{0.030}} &
{\small{0.021}}\tabularnewline
\hline 
\end{tabular}

\label{tab:coverage_cell_20percent}
\end{table}

\subsubsection{Individual tree hold-out experiment}

Table \ref{tab:score_tree} shows the predictive ability for Experiment
2. Here we have limited evidence (posterior probability of 0.84) that
the SPDE model is better based on the Brier score.

\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 5\% of trees. Smaller
values are better.}

\begin{tabular}{|c|c|c|>{\centering}p{3cm}|c|c|}
\hline 
 &
\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
\hline 
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}} &
{\small{Posterior Prob. CAR < SPDE}} &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Brier}} &
{\small{13158}} &
{\small{13143}} &
{\small{0.16}} &
{\small{13040}} &
{\small{13039}}\tabularnewline
\hline 
{\small{Negative Log Density}} &
{\small{31984}} &
{\small{31870}} &
{\small{0.02}} &
{\small{31212}} &
{\small{31229}}\tabularnewline
\hline 
\end{tabular}

\label{tab:score_tree}
\end{table}

\subsubsection{Choice of spatial model}

The differences between models are not consistent across the various
comparisons, so there is not a clear choice. In our final data product
we use the CAR model, for three reasons. First, the CAR model has
modestly better performance when data are sparse, as is still the
case for Illinois and Indiana. Second, the model is simpler and easier
to explain and computations can be done more quickly. Third, predictions
from the SPDE model showed boundary effects, with taxa showing non-negligible
posterior mean values at the edges of the domain, well away from where
the taxon was present in the empirical data. This included non-negligible
values within (but near the edge of) the convex hull of locations
with data.

\section{Data product}

For the final data product, we ran the model using the CAR specification
for X iterations, discarding the first Z iterations for burn-in and
retaining Y to limit storage needs and limit the size of the data
product. Mixing was generally reasonable, but for some of the hyperparameters
was relatively slow, particularly for less common taxa. Despite this,
mixing for the variables of substantive interest, the proportions
was good. {[}Report ESS values{]}

In Figure \ref{fig:select_maps}, we show maps of estimated composition
for the full domain for several taxa of substantive interest to illustrate
the results: beech, cherry, chestnut, elm, hemlock, oak, and pine.
These maps contrast the raw data proportions, the posterior means
and posterior standard deviations as pointwise estimates of uncertainty.
Fig. \ref{fig:all_predictions} shows posterior means for all 22 (???)
taxa.

\begin{figure}
\label{fig:select_maps}

\includegraphics[scale=0.75]{1_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig2.pdf}

\hspace{3mm}\includegraphics[scale=0.17]{2_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendRaw.pdf}\includegraphics[scale=0.17]{3_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendProp.pdf}\includegraphics[scale=0.17]{4_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendSD.pdf}

\caption{{[}THIS IS AN OLD VERSION FOR ILLUSTRATION ONLY - TO BE REPLACED WHEN
VERSION 0.3 FITS ARE DONE{]} Empirical proportions from raw data (column
1), posterior mean predictions (column 2) and posterior standard deviations
-- representing standard errors of prediction (column 3) for select
taxa.}
\end{figure}

\begin{figure}
\label{fig:all_predictions}

\includegraphics[scale=0.35]{5_var_tmp_paciorek_paleon_comp_0_2_output_PLScomposition_full_0_1_fits.pdf}

\caption{{[}THIS IS AN OLD VERSION FOR ILLUSTRATION ONLY - TO BE REPLACED AND
PUT IN 8 ROW X 3 COLUMN FORM WHEN VERSION 0.3 FITS ARE DONE{]} Posterior
mean predictions for all taxa over the entire domain.}

\end{figure}

While our data product is given for the full rectangular domain, including
hypothetical vegetation over water, uncertainty is large outside of
the core states with data and any analysis and inference should only
be done within those core states.

The data product is publically available at XX under YY license as
version 1.0 as of January 2015. The product is in the form of a netCDF-4
file, with dimensions x, y, and MCMC iteration. There is one variable
per taxon. The PalEON project will continue to maintain this product,
releasing new versions as additional data in Illinois, Indiana and
Ohio are digitized.

\section{Discussion}

Note that digitization of data from Illinois and Indiana is ongoing
and digitization of additional data from Ohio is planned as well.
As a result, at some point we expect to have complete data for the
western half of the domain. There will remain some missing townships
in the eastern half of the domain.

Given the density of data and the limited differences seen between
the ICAR and SPDE models, we expect the data product to be reasonably
robust to the choice of spatial model, particularly in those areas
with complete data. However, additional investigation of other statistical
representations is of interest, in particular nonstationary spatial
models and use of covariates. The biggest shortcoming of the current
model is its inability to account for local features such as rivers
(note the Minnesota River floodplain in evidence in the raw data).
The current model, by using a simple stationary spatial model that
smooths as a function of Euclidean distance, does not account for
topographic, soil, or other features.

Other related data products that are under development include:
\begin{itemize}
\item The gridded raw count data for the western subdomain and data aggregated
to township for the eastern subdomain. The raw data product is available
at XXX and that product provides the input data for the product described
in this work. Note that we cannot provide the raw data at point locations
because of limitations specified in the data use agreements under
which we have access to the data.
\item Gridded raw biomass estimates for Minnesota, Wisconsin, and Michigan
{[}right Simon, or also for IL and IN{]} based on the PLS data.
\item Statistically estimated biomass for Minnesota, Wisconsin, and Michigan
{[}also IL and IN?{]} using a statistical model applied to the raw
biomass estimates.
\end{itemize}
An additional drawback of the product is its focus on composition,
which does not directly tell us about vegetation structure, in particular
does not distinguish between closed forest, savanna, and prairie,
of particular note in Minnesota, Wisconsin, Illinois, and into Indiana.
The statistically-estimated biomass product mentioned just above will
directly inform questions about vegetation structure. Extensions of
that product will also estimate basal area and stem density.

\section*{Acknowledgments}

The authors are deeply indebted to all of the researchers over the
years who have preserved, collected, and digitized survey records,
in particular Ed Schools (Michigan DNR, \{ Brugam, ??? - Jody/Charlie/David/Jason/Simon
please add names here\}. This work was carried out by the PalEON Project
with support from the National Science Foundation MacroSystems Program
through grants EF-1065656, EF-1241868, DEB-1241874 and DEB-1241868.

\bibliographystyle{/accounts/gen/vis/paciorek/latex/RSSstylefile/Chicago}

\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology}

\section{Appendix}

\subsection{MCMC details}

Define $\bar{w}_{i\cdot p}=n_{i}^{-1}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
be the average of the $W$ values for the $p$th taxon in the $i$th
grid cell. Let $A$ be a diagonal matrix where $A_{ii}$ is the number
of trees in the $i$th grid cell. When there are no trees in a grid
cell, $\bar{w}_{i\cdot p}=0$ and $A_{ii}=0$. For the township data,
at each iteration, based on the current values of the grid cell membership
variables, $c_{tj}$, trees are aggregated into grid cells and the
calculations above can then be carried out.

The conditional distribution for $W_{mjp}$ given the other unknowns
in the model and the data is as follows. Let $\mbox{TN}(a,b,\mu,\tau^{2})$
denote the truncated normal distribution with mean parameter $\mu$
and variance parameter $\tau^{2}$, truncated below by $a$ and above
by $b$.

\begin{align}
W_{ijp} & \sim\begin{cases}
\mbox{TN}\big({\displaystyle \max_{p^{*}\neq y_{ij}}w_{ijp^{*}},\infty,\alpha_{y_{ij}}(s_{i}),1\big),} & \mbox{if }p\mbox{ = \ensuremath{y_{ij}}}\\
\mbox{TN}\big(-\infty,w_{ijy_{ij}},\alpha_{p}(s_{i}),1\big), & \mbox{if }p\mbox{ \ensuremath{\neq y_{ij}}}
\end{cases}\label{sampW}
\end{align}
In essence, the truncation value is determined by the taxon of the
$j$th tree. For a given $p$, the $W$ values for all trees in all
cells can be sampled in parallel.

The conditional distribution of $\alpha_{p}$ is 
\begin{align}
\alpha_{p} & \sim\mbox{N}\bigg(\Big(A+Q_{p}\Big)^{-1}A\bar{w}_{p},\Big(A+Q_{p}\Big)^{-1}\bigg).\label{sampalpha}
\end{align}
where $Q_{p}=(\sigma_{p}^{2})^{-1}Q$ for the CAR model and $\left(\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}\right)^{-1}Q(\rho_{p})$
for the SPDE model. For each hyperparameter, $\phi\in\{\{\mu_{p}\},\{\log\sigma_{p},\rho_{p}\}\}$,
we sampled $\{\phi,\alpha_{p}\}$ jointly, proposing $\phi$ as a
random walk and, conditional on the proposed value of $\phi$, sampling
$\alpha_{p}$ from the distribution just above. The joint proposal
is accepted or rejected as a standard Metropolis-Hastings proposal,
with adaptation of the proposal (co)variance \citep{Shab:Well:2011}.
As mentioned previously, for the SPDE model, we jointly proposed $\phi=\{\log\sigma_{p},\rho_{p}\}$
from a bivariate normal distribution to account for the posterior
dependence of these parameters, while for the CAR model, $\phi\in\{\{\log\sigma_{p}\}\}$.

\noindent

For the township-level data, for a given tree, we draw the latent
tree membership variable, $c_{tj}$, from a discrete distribution
by normalizing the weights, $\{\psi_{1(t)}L_{1(t)},\ldots,\psi_{T(t)}L_{T(t)}\}$
where $L_{k}$ is density of $W_{tj\cdot}$ if $c_{tj}=k$, namely
the product of independent normal densities, $W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{k}),1)$,
over $p=1,\ldots,P$. Thus the posterior weights the prior based on
how consistent the $W$ values are with the $\alpha$ values for the
candidate grid cells.

\subsection{Estimating $\theta_{p}(s)$ via Monte Carlo integration}

In the latent variable representation, $\theta_{p}(s)$ never appears
explicitly and cannot be calculated in close form. Instead we use
Monte Carlo integration over $W_{ijp},\, p=1,\ldots,P$ to estimate
$\theta_{p}(s_{i})$. The quantity $\theta_{p}(s_{i})=P(W_{ijp}={\displaystyle \max_{p^{*}}W_{ijp^{*}})}$
defines the probability of taxon $p$ at grid cell $i$. This requires
one to choose the number of Monte Carlo samples, which we set at 10000.
For each of the saved MCMC samples, $k=1,\ldots K$, we estimate $\theta_{p}^{(k)}(s_{i})$
numerically. Specifically, for $t=1,\ldots,10000$, we independently
draw
\[
W_{ijpt}^{(k)}\sim\mbox{N}(\alpha_{p}^{(k)}(s_{i}),1),\, p=1,\ldots,P
\]
and estimate 
\[
\hat{\theta}_{p}^{(k)}(s_{i})=\frac{1}{10000}{\displaystyle \sum_{t=1}^{10000}1(W_{ijpt}^{(k)}={\displaystyle \max_{p^{*}}W_{ijp^{*}t}^{(k)})}}
\]
where $1(\cdot)$ is the indicator function that evaluates to 1 if
the expression is true and 0 if false. In other words, we calculate
the proportion of times that the maximum of $W_{ijp},\, p=1,\ldots,P$
corresponds to taxon $p$. Considering $\hat{\theta}_{p}^{(k)}(s_{i}),\, k=1,\ldots,K$,
we have a sample from the posterior of $\theta_{p}(s_{i})$.

\end{document}

### 2014-12-11 paper.tex
added simon intro material

@@
 \usepackage{graphicx}
+\usepackage[authoryear]{natbib}
 
@@
 V. Cogbill, \\
-John W. Williams, David J. Mladenoff, Jody Peters, Jun Zhu, \\
+John W. Williams, David J. Mladenoff, Jody A. Peters, Jun Zhu, \\
 and Jason S. McLachlan {[}author order to be discussed{]}\\
@@
 
+%
 \begin{abstract}
@@
 survey records that provide raw data that are transcribed and then
-aggregated spatially, giving count data. For the western portion of
-the domain (Indiana through Minnesota) we use data at survey points
-that are aggregated to a regular 8 km grid, while for the eastern
-portion (Ohio through Maine), we use data aggregated into townships.
+aggregated spatially, giving count data. The domain is divided into
+two regions, western (Indiana through Minnesota) and eastern (Ohio
+to Maine). Public Land Survey point data in the western region is
+aggregated to a regular 8 km grid, while data in the eastern region,
+from Town Proprietor Surveys is aggregated at the township level.
 The product is based on a Bayesian statistical model fit to the count
@@
 
-{[}insert paragraph (Jack has volunteered) on context and importance
-of historical datasets{]}
+Historical datasets provide critical context to understand forest
+ecology. Historical datasets allow researchers to define 'baseline'
+conditions in conservation management, to understand ecosystem processes
+at decadal and centennial scales, and, particularly in regions with
+widespread land use change, to understand the extent to which forest
+conversion and regeneration differ from the original forest cover.
+
+Euro-American settlement and subsequent land use change occured in
+a time transient fashion across North America, and surveys of vegetation
+followed a similar pattern. Early surveys (from 1620 until 1825) in
+the northeastern United States were areal surveys at a township level
+\citep{cogbill2002forests,thompson2013four}, which we call the Town
+Proprietor Survey (TPS). These surveys reporting percent composition
+of major tree species using sometimes inconsistent common names. Later
+surveys, after the establishment of the Public Land Survey System
+(PLSS) - from 1819 to 1904 - were point estimates along a regular
+grid, with 1 mile spacing \citep{schulte2001original,bourdo1956review,goring2015composition}.
+Survey instructions during the PLSS varied through time, and by point
+type. This required the application of spatially varying correction
+factors \citep{cogbill2015corrections,goring2015composition} to accurately
+assess stem density, basal area and biomass from the early settlement
+records.
+
+Logging, agriculture and abandonment has left an indelible mark on
+forests in the northeastern United States \citep{foster1998land,goring2015composition,rhemtulla2009legacies,thompson2013four},
+however most studies have studied these effects at the state level
+or below \citep{rhemtulla2009historical,friedman2005regional}, and
+with varying spatial resolution, from townships (36 square miles)
+to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
+provide a dataset across the upper Midwest that is resolved to an
+8 x 8 km grid size, providing broad spatial coverage at a meso-spatial
+scale that can be compared to the Forest Inventory and Analysis products
+\citet{gray2012forest}. The limitation of this dataset is that it
+provides only estimates of variance within cell for PLSS data in the
+western region, and no township estimates of error for regions surveyed
+as part of the Town Proprietor Survey.
+
+Properly assessing uncertainty in ecological data is imperative to
+understanding and modelling ecological processes \citep{cressie2009accounting}.
+In this way, a model that can account for the spatial structure of
+the underlying PLSS and TPS data, and provide reliable estimates of
+uncertainty across the northeastern United States provides an enormously
+valuable tool for researchers interested in the ecological structure
+and function of forests at longer time scales.
 
@@
 in Minnesota, Wisconsin and Michigan but data in Illinois and Indiana
-represent a sample of the full set of grid cells, with transcription
-ongoing. Data for the remaining states are a subset of the full set
-of townships. Fig. \ref{fig:domain} shows the domain, indicating
-the grid cells and townships with data.
+represent a sample of the full set of grid cells, with survey record
+transcription ongoing. Data for the remaining states are a subset
+of the full set of townships. Fig. \ref{fig:domain} {[}Fig. 1 - Lyx
+is having a numbering problem{]} shows the domain, indicating the
+grid cells and townships with data.
 
 \begin{figure}
-\label{fig:domain}
-
-\includegraphics{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}
+\label{fig:domain}\includegraphics{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}
 
@@
 groups is primarily at the genus level, but in some cases of monospecific
-genera, is at the species level. In some other cases, because of ambiguity
-in the common tree names used by surveyors, a group represents trees
-from different families (orders?). We model the following taxa: {[}list
-all{]}. {[}include description of size cutoff{]}
+genera, is at the species level. We model the following 22 taxa plus
+an ``other hardwood'' category: Atlantic white cedar (Chamaecyparis
+thyoides), Ash (Fraxinus spp.), Basswood (Tilia americana), Beech
+(Fagus grandifolia), Birch (Betula spp.), Black gum/sweet gum (Nyssa
+sylvatica and Liquidambar styraciflua), Cedar/juniper (what is in
+here other than Juniperus and is the Juniperus only Juniperus virginiana),
+Cherry (Prunus spp.), Chestnut (Castanea dentata), Dogwood (Cornus
+spp.), Elm (Ulmus spp. or Ulmus americana???) , Fir (Abies spp.),
+Hemlock (Tsuga canadensis), Hickory (Carya spp.), Ironwood (Carpinus
+caroliniana and Ostrya virginiana (anything else?)), Maple (Acer spp.),
+Oak (Quercus spp.), Pine (Pinus spp.), Poplar/tulip poplar (Populus
+spp. and Liriodendron tulipifera), Spruce (Picea spp.), Tamarack (Larix
+laricina), Walnut (Juglans nigra). {[}Simon/Charlie to check my Latin
+and whether I've left out anything that falls into a given common
+name{]} Note that in several cases (e.g., black gum/sweet gum, ironwood,
+poplar/tulip poplar, cedar/juniper), because of ambiguity in the common
+tree names used by surveyors, a group represents trees from different
+families and even orders. For the western subdomain we do not fit
+statistical models for Atlantic white cedar and chestnut as these
+have 0 and 5 trees present in the dataset, respectively.
+
+{[}include description of size cutoff{]} {[}Simon, Jody mentioned
+that ther are 402 trees in IN below \textasciitilde{}10 cm and at
+least 43 in IL. Do you screen those out or do we just live with it?{]}
 
@@
 Our domain is a rectangle covering all of the states, in the Albers
-projection, with the rectangle split into 8 km cells, arranged in
-a 296 by 180 grid of cells, with the centroid of the cell in the southeast
-corner located at (-71000, 58000). For the modeling of the western
-subdomain we use the western-most 146 cells. For the modeling of the
-western subdomain we use the eastern-most 180 cells and omit 23 cells
-in the north and 17 cells in the south. {[}check 23 in N vs 17 in
-N; also check what is in netCDF product - should be for the full domain
-with missing values?{]}
+projection (NAD 1983 Great Lakes and St. Lawrence Albers) {[}Simon,
+please confirm this language is right{]}, with the rectangle split
+into 8 km cells, arranged in a 296 by 180 grid of cells, with the
+centroid of the cell in the southeast corner located at (-71000, 58000).
+For the modeling of the western subdomain we use the western-most
+146 by 180 grid of cells. For the modeling of the eastern subdomain
+we use the eastern-most 180 by 180 grid of cells and then omit 23
+cells in the north and 17 cells in the south. {[}check 23 in N vs
+17 in N; also check what is in netCDF product - should be for the
+full domain with missing values?{]}
 
@@
 
+%
 \begin{table}
@@
 of cells that are held out decreases from 95\% to 80\%, indicating
-performance on less sparse data. Table \ref{tab:coverage_cell_fivepercent}
+performance on less sparse data. Table \ref{tab:coverage_cell_20percent}
 shows performance of the uncertainty estimates. For denser data, unlike
@@
 years who have preserved, collected, and digitized survey records,
-in particular Ed Schools (Michigan DNR, \{ Brugam, ??? - Jody/Charlie/David/Jason/Simon
+in particular Ed Schools (Michigan DNR), \{ Brugam, ??? - Jody/Charlie/David/Jason/Simon
 please add names here\}. This work was carried out by the PalEON Project
@@
 
-\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology}
+\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,goring_intro}
 
@@
 
-The conditional distribution for $W_{mjp}$ given the other unknowns
+The conditional distribution for $W_{ijp}$ given the other unknowns
 in the model and the data is as follows. Let $\mbox{TN}(a,b,\mu,\tau^{2})$

### 2015-03-18 paper.tex
varisou minor edits to paper text and figures

@@
 
-look through code and labbook to make sure have all details
-
-properly cite Goring stuff and reference details of data collection
+full edit
+
+properly cite Goring stuff
 
@@
 
-%
+check journal requirements for archiving and presentation; talk to
+Jody/Ann about archiving/DOI
+
 \begin{abstract}
@@
 two regions, western (Indiana through Minnesota) and eastern (Ohio
-to Maine). Public Land Survey point data in the western region is
+to Maine). Public Land Survey point data in the western region are
 aggregated to a regular 8 km grid, while data in the eastern region,
@@
 of major tree species using sometimes inconsistent common names. Later
-surveys, after the establishment of the Public Land Survey System
-(PLSS) - from 1819 to 1904 - were point estimates along a regular
-grid, with 1 mile spacing \citep{schulte2001original,bourdo1956review,goring2015composition}.
-Survey instructions during the PLSS varied through time, and by point
+surveys, after the establishment of the U.S. Public Land Survey System
+(PLS) - from 1832 to 1907 - were point estimates along a regular grid,
+with 1 mile spacing \citep{schulte2001original,bourdo1956review,goring2015composition}.
+Survey instructions during the PLS varied through time, and by point
 type. This required the application of spatially varying correction
@@
 \citet{gray2012forest}. The limitation of this dataset is that it
-provides only estimates of variance within cell for PLSS data in the
+provides only estimates of variance within cell for PLS data in the
 western region, and no township estimates of error for regions surveyed
-as part of the Town Proprietor Survey.
+as part of the TPS.
 
@@
 subdomain), data are available at survey point locations and have
-been aggregated to a regular 8 km grid in the Albers projection. For
-the states of Ohio, Pennsylvania, New Jersey, New York and the six
-New England states (the eastern subdomain), data are aggregated at
-the township level. There is also data from a single township in Quebec
-and a single township in northern Delaware. Data are essentially complete
-in Minnesota, Wisconsin and Michigan but data in Illinois and Indiana
-represent a sample of the full set of grid cells, with survey record
-transcription ongoing. Data for the remaining states are a subset
-of the full set of townships. Fig. \ref{fig:domain} {[}Fig. 1 - Lyx
-is having a numbering problem{]} shows the domain, indicating the
-grid cells and townships with data.
+been aggregated to a regular 8 km grid in the Albers projection. (Note
+that for Indiana and Illinois, at the moment trees are associated
+with township centroids and then assigned to 8 km grid cells based
+on the centroid but in the near future we will have point locations
+available for each tree.) For the states of Ohio, Pennsylvania, New
+Jersey, New York and the six New England states (the eastern subdomain),
+data are aggregated at the township level. There is also data from
+a single township in Quebec and a single township in northern Delaware.
+Data are essentially complete in Minnesota, Wisconsin and Michigan
+but data in Illinois and Indiana represent a sample of the full set
+of grid cells, with survey record transcription ongoing. Data for
+the remaining states are a subset of the full set of townships. Fig.
+\ref{fig:domain} {[}Fig. 1 - Lyx is having a numbering problem{]}
+shows the domain, indicating the grid cells and townships with data.
 
@@
 \caption{Spatial domain, with locations with data shown in gray. Locations
-are grid cells in western portion and townships in eastern portion.}
+are grid cells in western portion and townships in eastern portion.
+Grid cells completely covered in water are white (e.g., a few locations
+in Minnesota and Wisconsin).}
 \end{figure}
@@
 sylvatica and Liquidambar styraciflua), Cedar/juniper (what is in
-here other than Juniperus and is the Juniperus only Juniperus virginiana),
+here other than Juniperus and is the Juniperus only Juniperus virginiana?),
 Cherry (Prunus spp.), Chestnut (Castanea dentata), Dogwood (Cornus
@@
 statistical models for Atlantic white cedar and chestnut as these
-have 0 and 5 trees present in the dataset, respectively.
+have 0 and 6 trees present, respectively. The taxa grouped into other
+hardwood are those for which fewer than roughly 2000 trees were present
+in the dataset; however we include Atlantic white cedar explicitly
+despite it only having 336 trees in the dataset.
 
@@
 
-There are approximately 520,000 trees from the western subdomain and
-420,000 trees from the eastern subdomain. In each subdomain, oak is
-the most common taxon and pine the second most common.
+There are approximately 860,000 trees from the western subdomain and
+420,000 trees from the eastern subdomain. In the western subdomain,
+oak is the most common taxon and pine the second most common, while
+in the eastern subdomain oak is the most common and beech the second
+most common.
 
 Our domain is a rectangle covering all of the states, in the Albers
-projection (NAD 1983 Great Lakes and St. Lawrence Albers) {[}Simon,
-please confirm this language is right{]}, with the rectangle split
-into 8 km cells, arranged in a 296 by 180 grid of cells, with the
-centroid of the cell in the southeast corner located at (-71000, 58000).
-For the modeling of the western subdomain we use the western-most
-146 by 180 grid of cells. For the modeling of the eastern subdomain
-we use the eastern-most 180 by 180 grid of cells and then omit 23
-cells in the north and 17 cells in the south. {[}check 23 in N vs
-17 in N; also check what is in netCDF product - should be for the
-full domain with missing values?{]}
+projection (NAD 1983 Albers Great Lakes and St. Lawrence), with the
+rectangle split into 8 km cells, arranged in a 296 by 180 grid of
+cells, with the centroid of the cell in the southeast corner located
+at (-71000, 58000). For the modeling of the western subdomain we use
+the western-most 146 by 180 grid of cells. For the modeling of the
+eastern subdomain we use the eastern-most 180 by 180 grid of cells
+and then omit 23 cells in the north and 17 cells in the south.
 
@@
 out data from the fitting process and assessing the fit of the model
-on the held-out data. We used two types of test data:
+on the held-out data. This cross-validation used a subregion containing
+most of Minnesota and a small amount of western Wisconsin, defined
+to be the cells whose x-coordinate was less than 300,000 (this defines
+a north-south line that approximately goes through Duluth, Minnesota)
+and hereafter referred to as the ``Minnesota subregion'' . We used
+two types of test data:
 \begin{enumerate}
-\item We held out all the data from 95\% of the cells in Minnesota, with
-cells selected at random. This was meant to assess the ability of
-the model to interpolate from a sparse set of cells/townships and
+\item We held out all the data from 95\% of the cells in the Minnesota subregion,
+with cells selected at random. This was meant to assess the ability
+of the model to interpolate from a sparse set of cells/townships and
 mimics the limited data in Illinois and Indiana.
 \item We held out 5\% of the trees from all of the trees in the dataset
-(leaving aside the held-out Minnesota cells). This was meant to assess
-the ability of the model to estimate the composition in cells in which
-data were available. 
+(leaving aside the held-out Minnesota subregion cells). This was meant
+to assess the ability of the model to estimate the composition in
+cells in which data were available. 
 \end{enumerate}
 Finally, in a separate sensitivity analysis we instead left out 80\%
-of the cells in Minnesota at random. This variation on the first test
-above was meant to indicate whether our model comparison conclusions
-would be robust as the digitization process for Illinois and Indiana
-progresses and provides us with increasingly dense data.
+of the cells in Minnesota subregion at random. This variation on the
+first test above was meant to indicate whether our model comparison
+conclusions would be robust as the digitization process for Illinois
+and Indiana progresses and provides us with increasingly dense data.
 
@@
 
-We ran each model for XXX iterations, retaining YYY after burnin and
-subsampling to reduce storage needs.
+We ran each model for 150,000 iterations, retaining 250 after a burnin
+(25,000 iterations) and subsampling to reduce storage needs.
 
@@
 Table \ref{tab:score_cell_fivepercent} shows predictive ability for
-Experiment 1: those cells in Minnesota held out of the fitting process,
-for both the posterior mean predictions and the full posterior sample.
-Table \ref{tab:coverage_cell_fivepercent} shows performance of the
-uncertainty estimates. For the posterior distribution over the predictive
-score values, the CAR model generally outperforms the SPDE model,
-while for the posterior mean predictions, the SPDE model appears to
-outperform the CAR model, but we do not have any uncertainty estimates
+Experiment 1: those cells in the Minnesota subregion held out of the
+fitting process, for both the posterior mean predictions and the full
+posterior sample. Table \ref{tab:coverage_cell_fivepercent} shows
+performance of the uncertainty estimates. For the posterior distribution
+over the predictive score values, the CAR model outperforms the SPDE
+model, while for the posterior mean predictions, the SPDE model appears
+to outperform the CAR model, but we do not have any uncertainty estimates
 for this comparison. Coverage and interval lengths are similar between
@@
 
-%
 \begin{table}
-
 \caption{Predictive ability based on several predictive score criteria for
@@
 {\small{Brier}} &
-{\small{84997}} &
-{\small{86803}} &
-{\small{0.92}} &
-{\small{76068}} &
-{\small{75330}}\tabularnewline
+{\small{180688}} &
+{\small{186045}} &
+{\small{0.98}} &
+{\small{162813}} &
+{\small{161559}}\tabularnewline
 \hline 
 {\small{Negative Log Density}} &
-{\small{222144}} &
-{\small{242676}} &
+{\small{468397}} &
+{\small{513144}} &
 {\small{1.00}} &
-{\small{185748}} &
-{\small{185843}}\tabularnewline
+{\small{395685}} &
+{\small{395998}}\tabularnewline
 \hline 
 {\small{Mean Absolute Error}} &
-{\small{0.0382}} &
-{\small{0.0390}} &
-{\small{0.92}} &
-{\small{0.0303}} &
-{\small{0.0293}}\tabularnewline
+{\small{0.0364}} &
+{\small{0.0383}} &
+{\small{0.98}} &
+{\small{0.0275}} &
+{\small{0.0269}}\tabularnewline
 \hline 
 {\small{Root Mean Square Error}} &
-{\small{0.0944}} &
-{\small{0.0986}} &
-{\small{0.77}} &
-{\small{0.0701}} &
-{\small{0.0677}}\tabularnewline
+{\small{0.0897}} &
+{\small{0.0959}} &
+{\small{0.96}} &
+{\small{0.0647}} &
+{\small{0.0626}}\tabularnewline
 \hline 
@@
 {\small{Coverage}} &
-{\small{0.977}} &
+{\small{0.976}} &
 {\small{0.978}}\tabularnewline
@@
 {\small{Mean Interval Length}} &
-{\small{0.135}} &
+{\small{0.129}} &
 {\small{0.142}}\tabularnewline
@@
 {\small{Median Interval Length}} &
-{\small{0.040}} &
-{\small{0.031}}\tabularnewline
+{\small{0.037}} &
+{\small{0.033}}\tabularnewline
 \hline 
@@
 
-%
 \begin{table}
@@
 {\small{Brier}} &
-{\small{65870}} &
-{\small{65004}} &
-{\small{0.08}} &
-{\small{60169}} &
-{\small{60020}}\tabularnewline
+{\small{141546}} &
+{\small{140211}} &
+{\small{0.10}} &
+{\small{130006}} &
+{\small{129910}}\tabularnewline
 \hline 
 {\small{Negative Log Density}} &
-{\small{167947}} &
-{\small{167536}} &
-{\small{0.43}} &
-{\small{145404}} &
-{\small{145750}}\tabularnewline
+{\small{356985}} &
+{\small{355640}} &
+{\small{0.31}} &
+{\small{312509}} &
+{\small{312950}}\tabularnewline
 \hline 
 {\small{Mean Absolute Error}} &
-{\small{0.0322}} &
-{\small{0.0307}} &
-{\small{0.08}} &
-{\small{0.0246}} &
-{\small{0.0240}}\tabularnewline
+{\small{0.0309}} &
+{\small{0.0297}} &
+{\small{0.10}} &
+{\small{0.0225}} &
+{\small{0.0222}}\tabularnewline
 \hline 
 {\small{Root Mean Square Error}} &
-{\small{0.0800}} &
-{\small{0.0771}} &
-{\small{0.01}} &
-{\small{0.0578}} &
-{\small{0.0571}}\tabularnewline
+{\small{0.0763}} &
+{\small{0.0740}} &
+{\small{0.04}} &
+{\small{0.0531}} &
+{\small{0.0528}}\tabularnewline
 \hline 
@@
 {\small{Coverage}} &
-{\small{0.982}} &
-{\small{0.974}}\tabularnewline
+{\small{0.981}} &
+{\small{0.971}}\tabularnewline
 \hline 
 {\small{Mean Interval Length}} &
-{\small{0.117}} &
-{\small{0.107}}\tabularnewline
+{\small{0.112}} &
+{\small{0.103}}\tabularnewline
 \hline 
 {\small{Median Interval Length}} &
-{\small{0.030}} &
+{\small{0.028}} &
 {\small{0.021}}\tabularnewline
@@
 Table \ref{tab:score_tree} shows the predictive ability for Experiment
-2. Here we have limited evidence (posterior probability of 0.84) that
-the SPDE model is better based on the Brier score.
-
-%
+2. Here we have evidence (posterior probability of 0.94) that the
+SPDE model is better based on the Brier score.
+
 \begin{table}
@@
 {\small{Brier}} &
-{\small{13158}} &
-{\small{13143}} &
-{\small{0.16}} &
-{\small{13040}} &
-{\small{13039}}\tabularnewline
+{\small{21296}} &
+{\small{21277}} &
+{\small{0.06}} &
+{\small{21146}} &
+{\small{21143}}\tabularnewline
 \hline 
 {\small{Negative Log Density}} &
-{\small{31984}} &
-{\small{31870}} &
-{\small{0.02}} &
-{\small{31212}} &
-{\small{31229}}\tabularnewline
+{\small{52199}} &
+{\small{52052}} &
+{\small{0.01}} &
+{\small{51141}} &
+{\small{51156}}\tabularnewline
 \hline 
@@
 For the final data product, we ran the model using the CAR specification
-for X iterations, discarding the first Z iterations for burn-in and
-retaining Y to limit storage needs and limit the size of the data
-product. Mixing was generally reasonable, but for some of the hyperparameters
-was relatively slow, particularly for less common taxa. Despite this,
-mixing for the variables of substantive interest, the proportions
-was good. {[}Report ESS values{]}
+for 150,000 iterations, discarding the first 25,000 iterations for
+burn-in and retaining 250 to limit storage needs and limit the size
+of the data product. Mixing was generally reasonable, but for some
+of the hyperparameters was relatively slow, particularly for less
+common taxa. Despite this, mixing for the variables of substantive
+interest, the proportions was good. {[}Report ESS values{]}
 
@@
 and posterior standard deviations as pointwise estimates of uncertainty.
-Fig. \ref{fig:all_predictions} shows posterior means for all 22 (???)
-taxa.
+Fig. \ref{fig:all_predictions} shows posterior means for all 23 taxa.
 
@@
 
-\hspace{3mm}\includegraphics[scale=0.17]{2_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendRaw.pdf}\includegraphics[scale=0.17]{3_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendProp.pdf}\includegraphics[scale=0.17]{4_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendSD.pdf}
-
-\caption{{[}THIS IS AN OLD VERSION FOR ILLUSTRATION ONLY - TO BE REPLACED WHEN
-VERSION 0.3 FITS ARE DONE{]} Empirical proportions from raw data (column
-1), posterior mean predictions (column 2) and posterior standard deviations
--- representing standard errors of prediction (column 3) for select
-taxa.}
+\hspace{4mm}\includegraphics[scale=0.16]{2_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendRaw.pdf}\hspace{4mm}\includegraphics[scale=0.16]{3_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendProp.pdf}\hspace{3.5mm}\includegraphics[scale=0.16]{4_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendSD.pdf}
+
+\caption{Empirical proportions from raw data (column 1), posterior mean predictions
+(column 2) and posterior standard deviations -- representing standard
+errors of prediction (column 3) for select taxa.}
 \end{figure}
@@
 
-\includegraphics[scale=0.35]{5_var_tmp_paciorek_paleon_comp_0_2_output_PLScomposition_full_0_1_fits.pdf}
-
-\caption{{[}THIS IS AN OLD VERSION FOR ILLUSTRATION ONLY - TO BE REPLACED AND
-PUT IN 8 ROW X 3 COLUMN FORM WHEN VERSION 0.3 FITS ARE DONE{]} Posterior
-mean predictions for all taxa over the entire domain.}
+\includegraphics[scale=0.9]{5_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig3.pdf}
+
+\caption{Posterior mean predictions for all taxa over the entire domain.}
 
@@
 
-While our data product is given for the full rectangular domain, including
-hypothetical vegetation over water, uncertainty is large outside of
-the core states with data and any analysis and inference should only
-be done within those core states.
-
-The data product is publically available at XX under YY license as
-version 1.0 as of January 2015. The product is in the form of a netCDF-4
-file, with dimensions x, y, and MCMC iteration. There is one variable
-per taxon. The PalEON project will continue to maintain this product,
+The data product is publicly available at XX under YY license as version
+0.3 as of April 2015. The product is in the form of a netCDF-4 file,
+with dimensions x, y, and MCMC iteration. There is one variable per
+taxon. The PalEON project will continue to maintain this product,
 releasing new versions as additional data in Illinois, Indiana and
@@
 \item Gridded raw biomass estimates for Minnesota, Wisconsin, and Michigan
-{[}right Simon, or also for IL and IN{]} based on the PLS data.
+based on the PLS data, with extension to Illinois and Indiana planned.
 \item Statistically estimated biomass for Minnesota, Wisconsin, and Michigan
-{[}also IL and IN?{]} using a statistical model applied to the raw
-biomass estimates.
+using a statistical model applied to the raw biomass estimates, with
+extension to Illinois and Indiana planned.
 \end{itemize}

### 2015-03-18 paper.R

@@ -49,6 +49,8 @@ d <- add_map_albers(plot_obj = d, map_data = usFortified, dat = presence_long)
 d <- d + scale_y_continuous(limits = c(70000, 1490000)) 
 d <- theme_clean(d)
 
+setwd(paperDir)
+
 pdf('fig1.pdf', width=4.85, height=3)
 print(d)
 dev.off()

### 2015-04-13 coverLetter.tex
minor edits and addition of cover letter

\documentclass[12pt]{letterhead_stat}
\usepackage{graphicx}
\special{papersize=8.5in,11in}
\textwidth=6.5in
\textheight=8.5in
\parindent=0.0cm
\parskip=0.4cm
\oddsidemargin=0.0in
\evensidemargin=0.0in
\topmargin=-0.594in
\begin{document}
\telephone{(510) 642-6190\\
E-mail: paciorek@stat.berkeley.edu\\
fax: (510) 642-7892}
\name{Christopher Paciorek}
\signature{\vspace{-2em}\hspace{-.3cm}\includegraphics[height=.6in,width=2in]{goodSignature.jpg}\\Christopher Paciorek\\Associate Research Statistician\\Department of Statistics\\University of California, Berkeley}
\begin{letter}{\mbox{}}
\opening{To whom it may concern,}

Please consider our submission as a research article.

This submission presents a new statistically-derived data product and the methods used to create the product. The product provides an estimate of forest composition, with uncertainty, for much of the northeastern United States at the time of European settlement. As such, it is valuable resource for ecologists and for climate change work as it provides baseline information about forests before widespread European land use change.

The raw data underlying this product have not been collected and made available in one place to date as they represent the work of a number of researchers and in some cases data are still unpublished. Our product provides a publicly available resource.

With regard to possible editors, either someone with expertise in statistics applied to environmental applications or someone with expertise in biogeography or landscape-scale ecology would be appropriate. There are a number listed on your website with expertise in biogeography, community ecology, global change ecology, or spatial and landscape ecology. In terms of editors with expertise in statistics, potential editors include Alex Cook, Thomas Nichols, Nicholas Reich.

\closing{Sincerely,}
%\encl{names, bio}
\end{letter}
\end{document}

### 2015-04-19 paper.tex
added plos_one latex files and some minor edits to the paper

@@
 \begin{quote}
-Christopher J. Paciorek, Andrew Thurman, Simon J. Goring, Charlie
-V. Cogbill, \\
-John W. Williams, David J. Mladenoff, Jody A. Peters, Jun Zhu, \\
-and Jason S. McLachlan {[}author order to be discussed{]}\\
+Christopher J. Paciorek 1{*}, Simon J. Goring 2\&, Andrew Thurman
+3\&, Charles V. Cogbill 4, \\
+John W. Williams 3,5, David J. Mladenoff 6, Jody A. Peters 7, Jun
+Zhu 8, \\
+and Jason S. McLachlan 7 \\
+{[}author order to be discussed{]}\\
 {[}please tell me if you do/don't want your middle initial{]}
+
+{[}let me know if there are any issues with how I've listed your affiliation{]}
+
+1 Department of Statistics, University of California, Berkeley, California,
+USA
+
+2 Department of Geography, University of Wisconsin, Madison, Wisconsin,
+USA
+
+3 Department of Statistics, University of Iowa, Iowa City, Iowa, USA
+
+4 Harvard Forest, Harvard University, Petersham, Massachusetts, USA
+
+5 Center for Climatic Research, University of Wisconsin, Madison,
+Wisconsin, USA
+
+6 Department of Forestry and Wildlife Ecology, University of Wisconsin,
+Madison, Wisconsin, USA
+
+7 Department of Biological Sciences, University of Notre Dame, Notre
+Dame, Indiana, USA
+
+8 Department of Statistics, University of Wisconsin, Madison, Wisconsin,
+USA
+
+{*} Corresponding author
+
+E-mail paciorek@stat.berkeley.edu (CJP)
+
+\& These authors contributed equally to this work. 
 \end{quote}
@@
 
-full edit
-
-properly cite Goring stuff
-
 focus on PLOS One, but Nature's Scientific Data is a possibility
@@
 We present a data product of the estimated composition of tree taxa
-at the time of European settlement of the United States and the statistical
-methodology used to produce the product. Composition is defined as
-the proportion of stems larger than approximately X cm diameter at
-breast height {[}Charlie? Simon? is X roughly 10?{]} in 22 taxonomic
-groupings, generally at the genus level. The data come from settlement
-survey records that provide raw data that are transcribed and then
-aggregated spatially, giving count data. The domain is divided into
-two regions, western (Indiana through Minnesota) and eastern (Ohio
-to Maine). Public Land Survey point data in the western region are
-aggregated to a regular 8 km grid, while data in the eastern region,
-from Town Proprietor Surveys is aggregated at the township level.
-The product is based on a Bayesian statistical model fit to the count
-data that estimates composition on a regular 8 km grid. The statistical
-model allows us to estimate composition at locations with no data
-and to smooth over noise caused by limited counts in locations with
-data. Critically, it also allows us to quantify uncertainty in our
-composition estimates. We expect this data product to be useful for
-understanding the state of vegetation in the northeastern United States
-prior to large-scale European settlement. In addition to specific
+at the time of European settlement of the northeastern United States
+and the statistical methodology used to produce the product. Composition
+is defined as the proportion of stems larger than approximately 20
+cm diameter at breast height in 22 taxonomic groupings, generally
+at the genus level. The data come from settlement survey records that
+provide raw data that are transcribed and then aggregated spatially,
+giving count data. The domain is divided into two regions, eastern
+(Maine to Ohio) and western (Indian to Minnesota). Public Land Survey
+point data in the western region are aggregated to a regular 8 km
+grid, while data in the eastern region, from Town Proprietor Surveys,
+is aggregated at the township level in irregularly-shaped local administrative
+units. The product is based on a Bayesian statistical model fit to
+the count data that estimates composition on a regular 8 km grid.
+The statistical model allows us to estimate composition at locations
+with no data and to smooth over noise caused by limited counts in
+locations with data. Critically, it also allows us to quantify uncertainty
+in our composition estimates. We expect this data product to be useful
+for understanding the state of vegetation in the northeastern United
+States prior to large-scale European settlement. In addition to specific
 regional questions, the data product can also serve as a baseline
@@
 Historical datasets provide critical context to understand forest
-ecology. Historical datasets allow researchers to define 'baseline'
-conditions in conservation management, to understand ecosystem processes
-at decadal and centennial scales, and, particularly in regions with
-widespread land use change, to understand the extent to which forest
-conversion and regeneration differ from the original forest cover.
+ecology. They allow researchers to define 'baseline' conditions for
+conservation management, to understand ecosystem processes at decadal
+and centennial scales, and, particularly in regions with widespread
+land use change, to understand the extent to which forests after conversion
+and regeneration differ from the original forest cover.
 
 Euro-American settlement and subsequent land use change occured in
-a time transient fashion across North America, and surveys of vegetation
-followed a similar pattern. Early surveys (from 1620 until 1825) in
-the northeastern United States were areal surveys at a township level
-\citep{cogbill2002forests,thompson2013four}, which we call the Town
-Proprietor Survey (TPS). These surveys reporting percent composition
-of major tree species using sometimes inconsistent common names. Later
-surveys, after the establishment of the U.S. Public Land Survey System
-(PLS) - from 1832 to 1907 - were point estimates along a regular grid,
-with 1 mile spacing \citep{schulte2001original,bourdo1956review,goring2015composition}.
-Survey instructions during the PLS varied through time, and by point
-type. This required the application of spatially varying correction
-factors \citep{cogbill2015corrections,goring2015composition} to accurately
-assess stem density, basal area and biomass from the early settlement
-records.
-
-Logging, agriculture and abandonment has left an indelible mark on
-forests in the northeastern United States \citep{foster1998land,goring2015composition,rhemtulla2009legacies,thompson2013four},
-however most studies have studied these effects at the state level
-or below \citep{rhemtulla2009historical,friedman2005regional}, and
-with varying spatial resolution, from townships (36 square miles)
+a time transient fashion across North America, and land surveys that
+provide vegetation information followed a similar pattern. Early surveys
+(from 1620 until 1825) in the northeastern United States provide areally-aggregated
+data at the township level \citep{cogbill2002forests,thompson2013four};
+we refer to these as the Town Proprietor Survey (TPS). These surveys
+provide composition of major tree species using sometimes inconsistent
+common names. Later surveys, after the establishment of the U.S. Public
+Land Survey System (PLS) - from 1832 to 1907 - provide point-level
+data along a regular grid, with one mile spacing \citet{bourdo1956review,schulte2001original,goring2015composition},
+again with sometimes inconsistent common names. Survey instructions
+during the PLS varied through time and by point type. This requires
+the application of spatially-varying correction factors \citep{cogbill2015corrections,goring2015composition}
+to accurately assess stem density, basal area and biomass from the
+early settlement records but has limit impact when estimating composition.
+
+Logging, agriculture, and abandonment have left an indelible mark
+on forests in the northeastern United States \citet{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
+However most studies have assessed these effects in individual states
+or smaller domains \citep{rhemtulla2009historical,friedman2005regional}
+and with varying spatial resolution, from townships (36 square miles)
 to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
-provide a dataset across the upper Midwest that is resolved to an
-8 x 8 km grid size, providing broad spatial coverage at a meso-spatial
-scale that can be compared to the Forest Inventory and Analysis products
-\citet{gray2012forest}. The limitation of this dataset is that it
-provides only estimates of variance within cell for PLS data in the
-western region, and no township estimates of error for regions surveyed
-as part of the TPS.
+provide a dataset for the upper Midwest that is resolved to an 8 x
+8 km grid size, providing broad spatial coverage at a spatial scale
+that can be compared to the Forest Inventory and Analysis products
+\citet{gray2012forest}. Combined with additional, coarsely-sampled
+PLS data from Illinois and Indiana and with the TPS data, this gives
+us raw data for much of the northeastern United States. However, there
+are several limitations of using the raw data that can be alleviated
+by the use of a statistical model to develop a statistically-estimated
+data product. First, the PLS and TPS data provide only estimates of
+within-cell variance that do not account for information from nearby
+locations. Second, the available digitized data from Illinois and
+Indiana represent a small fraction of those states and missing townships
+are common in the TPS data.
 
@@
 In this way, a model that can account for the spatial structure of
-the underlying PLSS and TPS data, and provide reliable estimates of
-uncertainty across the northeastern United States provides an enormously
-valuable tool for researchers interested in the ecological structure
-and function of forests at longer time scales.
+the underlying PLS and TPS data, and provide reliable estimates of
+uncertainty across the northeastern United States, provides a valuable
+tool for researchers interested in the ecological structure and function
+of forests at longer time scales.
 
@@
 
-The raw data are obtained from survey records collated from across
+The raw data were obtained from survey records collated from across
 the northeastern U.S. by a number of researchers. For the states of
 Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
-subdomain), data are available at survey point locations and have
+subdomain), data are available at PLS survey point locations and have
 been aggregated to a regular 8 km grid in the Albers projection. (Note
@@
 Jersey, New York and the six New England states (the eastern subdomain),
-data are aggregated at the township level. There is also data from
+data are aggregated at the township level. There are also data from
 a single township in Quebec and a single township in northern Delaware.
-Data are essentially complete in Minnesota, Wisconsin and Michigan
+Data are essentially complete in Minnesota, Wisconsin and Michigan,
 but data in Illinois and Indiana represent a sample of the full set
 of grid cells, with survey record transcription ongoing. Data for
-the remaining states are a subset of the full set of townships. Fig.
-\ref{fig:domain} {[}Fig. 1 - Lyx is having a numbering problem{]}
-shows the domain, indicating the grid cells and townships with data.
+the remaining states are available for a subset of the full set of
+townships covering the domain. Fig. \ref{fig:domain} {[}Fig. 1 -
+Lyx is having a numbering problem{]} shows the domain, indicating
+the grid cells and townships with data.
 
@@
 are grid cells in western portion and townships in eastern portion.
-Grid cells completely covered in water are white (e.g., a few locations
+In addition to locations without data being indicated in white, grid
+cells completely covered in water are white (e.g., a few locations
 in Minnesota and Wisconsin).}
@@
 point in time across the domain, but rather the state of the landscape
-at the time just before settlement occurred {[}get ref from Charlie
-or Jason's historian pal{]}
+at the time just before settlement occurred \citep{Whit:1996,Cogb:etal:2002}.
 
 Extensive details on the data are available in \cite{Gori:etal:2015}
-and the raw data are available at XXX. The aggregation into taxonomic
-groups is primarily at the genus level, but in some cases of monospecific
-genera, is at the species level. We model the following 22 taxa plus
-an ``other hardwood'' category: Atlantic white cedar (Chamaecyparis
-thyoides), Ash (Fraxinus spp.), Basswood (Tilia americana), Beech
-(Fagus grandifolia), Birch (Betula spp.), Black gum/sweet gum (Nyssa
-sylvatica and Liquidambar styraciflua), Cedar/juniper (what is in
-here other than Juniperus and is the Juniperus only Juniperus virginiana?),
-Cherry (Prunus spp.), Chestnut (Castanea dentata), Dogwood (Cornus
-spp.), Elm (Ulmus spp. or Ulmus americana???) , Fir (Abies spp.),
-Hemlock (Tsuga canadensis), Hickory (Carya spp.), Ironwood (Carpinus
-caroliniana and Ostrya virginiana (anything else?)), Maple (Acer spp.),
-Oak (Quercus spp.), Pine (Pinus spp.), Poplar/tulip poplar (Populus
-spp. and Liriodendron tulipifera), Spruce (Picea spp.), Tamarack (Larix
-laricina), Walnut (Juglans nigra). {[}Simon/Charlie to check my Latin
-and whether I've left out anything that falls into a given common
-name{]} Note that in several cases (e.g., black gum/sweet gum, ironwood,
-poplar/tulip poplar, cedar/juniper), because of ambiguity in the common
-tree names used by surveyors, a group represents trees from different
-families and even orders. For the western subdomain we do not fit
-statistical models for Atlantic white cedar and chestnut as these
-have 0 and 6 trees present, respectively. The taxa grouped into other
-hardwood are those for which fewer than roughly 2000 trees were present
-in the dataset; however we include Atlantic white cedar explicitly
-despite it only having 336 trees in the dataset.
-
-{[}include description of size cutoff{]} {[}Simon, Jody mentioned
-that ther are 402 trees in IN below \textasciitilde{}10 cm and at
-least 43 in IL. Do you screen those out or do we just live with it?{]}
+and the raw data are available at XXX {[}ask Simon/Charlie about public
+access to Paleon composition data{]}. The aggregation into taxonomic
+groups is primarily at the genus level, but is at the species level
+in some cases of monospecific genera. We model the following 22 taxa
+plus an ``other hardwood'' category: Atlantic white cedar (\emph{Chamaecyparis
+thyoides}), Ash (\emph{Fraxinus spp.}), Basswood (\emph{Tilia americana}),
+Beech (\emph{Fagus grandifolia)}, Birch (\emph{Betula spp.}), Black
+gum/sweet gum (\emph{Nyssa sylvatica} and \emph{Liquidambar styraciflua}),
+Cedar/juniper (\emph{Juniperus virginiana} and \emph{Thuja occidentalis}),
+Cherry (\emph{Prunus spp.}), Chestnut (\emph{Castanea dentata}), Dogwood
+(\emph{Cornus spp.}), Elm (\emph{Ulmus spp.}), Fir (\emph{Abies spp.}),
+Hemlock (\emph{Tsuga canadensis}), Hickory (\emph{Carya spp.}), Ironwood
+(\emph{Carpinus caroliniana} and \emph{Ostrya virginiana}), Maple
+(\emph{Acer spp.}), Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}),
+Poplar/tulip poplar (\emph{Populus spp.} and \emph{Liriodendron tulipifera}),
+Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), Walnut
+(\emph{Juglans nigra}). Note that in several cases (e.g., black gum/sweet
+gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
+in the common tree names used by surveyors, a group represents trees
+from different families and even orders. For the western subdomain
+we do not fit statistical models for Atlantic white cedar and chestnut
+as these have 0 and 6 trees present, respectively. The taxa grouped
+into the other hardwood category are those for which fewer than roughly
+2000 trees were present in the dataset; however we include Atlantic
+white cedar explicitly despite it only having 336 trees in the dataset.
+Given the nature of the data, there is not a firm cutoff on the sizes
+of trees included, but the data roughly represent trees greater than
+8 inches (\textasciitilde{}20 cm) diameter at breast height (dbh),
+but with some trees as small as 4 inches. {[}Charlie,Simon, please
+check this wording{]} In Indiana and Illinois there are also 402 trees
+and 29 trees, respectively, that are between 1 and 4 inches dbh, but
+these will be excluded in the next version of the product.
 
@@
 eastern subdomain we use the eastern-most 180 by 180 grid of cells
-and then omit 23 cells in the north and 17 cells in the south.
+and then omit 23 cells in the north and 17 cells in the south outside
+the states containing data.
 
@@
 the empirical proportions represent estimates of the true proportions
-based on the full population of all the trees in an areal region.
+that could be calculated using full population of all the trees in
+an areal region.
 \end{enumerate}
@@
 (MCMC) is a set of representative samples from the posterior distribution
-for the composition of the taxa at all of the grid cells. These samples
-are the data product and can then be used in analyses. We also provide
-the mean and standard deviation of the samples, which represent our
-best estimate of composition and a Bayesian ``standard error'' for
+for the composition in the 23 taxonomic groupings at each of the grid
+cells. These samples are the data product (described further in the
+Data Product section), and can then be used in subsequent analyses.
+The mean and standard deviation of the samples for each cell by taxon
+pair represent our best estimate (i.e., prediction) of composition
+and a Bayesian ``standard error'' quantifying the uncertainty in
 the estimate.
@@
 
-The statistical model treats the observations as having a multinomial
+We start by describing the basic model for those states for which
+we have raw data on the 8 km grid and in Section \ref{sub:Model-for-township}
+we describe the extension of the model to accommodate data aggregated
+at the township level.
+
+The statistical model treats the observations as coming from a multinomial
 distribution with a (latent) vector of proportions for each grid cell.
@@
 where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
-location, $n_{i}$ is the number of trees counted in the location,
-and $\theta(s_{i})$ is the vector of unknown proportions for those
-taxa at that location. Note that we use a standard multinomial distribution
-without overdispersion as the set of trees in the dataset is roughly
-uniformly sampled across the cells or townships \citep{Gori:etal:2015}.
-
-In what follows we describe the basic model for those states for which
-we have raw data on the 8 km grid and in Section \ref{sub:Model-for-township}
-we describe the extension of the model to accommodate data aggregated
-to the township level.
+cell, $n_{i}$ is the number of trees counted in the cell, and $\theta(s_{i})$
+is the vector of unknown proportions for those taxa at that cell.
+Note that we use a standard multinomial distribution without overdispersion
+as the set of trees in the dataset is roughly uniformly sampled across
+the cells or townships \citep{Gori:etal:2015}.
 
@@
 sum to one constraints, the taxa are considered to be independent
-in the prior, as we did not want to impose any structure that ties
-the different taxa together, as any correlation will likely vary across
+in the prior. We did not want to impose any structure that ties the
+different taxa together, as any correlation will likely vary across
 space.
@@
 is called an \textit{intrinsic conditional autoregression (ICAR)}
-and $Q$ can also be expressed as $D-C$ where $C$ is the $m\times m$
-adjacency matrix defining the neighborhood relation of the locations;
-that is $(C)_{ik}=1$ if locations $i$ and $k$ are neighbors. The
-matrix $D$ is an $m\times m$ diagonal matrix containing the row
-sums of matrix $C$ as the diagonal entries $(D)_{ii}={\displaystyle \sum_{k=1}^{m}(C)_{ik}}.$
+and we can write $Q=D-C$ where $C$ is the $m\times m$ adjacency
+matrix defining the neighborhood relation of the locations; that is
+$(C)_{ik}=1$ if locations $i$ and $k$ are neighbors. The matrix
+$D$ is an $m\times m$ diagonal matrix containing the row sums of
+matrix $C$ as the diagonal entries $(D)_{ii}={\displaystyle \sum_{k=1}^{m}(C)_{ik}}.$
 
@@
 For any pair of locations, $R(d)$ defines the correlation of the
-process, e.g., $\alpha_{p}(s)$, as a function of the distance between
-the locations. Considering all pairs of locations, this defines a
-correlation matrix for all locations of interest.
+process, (i.e., $\alpha_{p}(s)$ in our context), as a function of
+the distance between the locations. Considering all pairs of locations,
+this defines a correlation matrix for all locations of interest.
 
 The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
-to the Matern-based GPs for $\nu=1$ and $\nu=2$. Our second model
+to the Matern-based GP for $\nu=1$ and $\nu=2$. Our second model
 is the Lindgren approximation for Matern-based GPs with $\nu=1$.
@@
 elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
-neighbors are $-2a$. Those for diagonal neighbors are $2$ and those
+neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
 for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
@@
 
-To ensure that the $\sigma^{2}$ parameter is equivalent between the
-two models, we reparameterize, producing our second model:
+To ensure that the $\sigma^{2}$ parameter is mathematically equivalent
+between the two models, we reparameterize, producing our second model:
 \[
-\alpha_{p}\sim\mbox{N}(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1})
+\alpha_{p}\sim\mbox{N}\left(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1}\right)
 \]
@@
 \noindent The ICAR specification contains a set of hyperparameters
-$\sigma_{p}^{2}$ for $p$ = $1,...,P$. Following \citep{Gelm:2006}
+$\sigma_{p}^{2}$ for $p$ = $1,...,P$. Following \cite{Gelm:2006}
 we use a uniform distribution on each $\sigma_{p}$ parameter, with
@@
 with latent Gaussian process(es) and a non-Gaussian likelihood is
-difficult (cite recent cent/noncent papers). To develop an algorithm,
-we make use of a latent variable representation for the multinomial
-probit model \citep{McCu:Ross:1994}. The representation introduces
-latent variables that allow one to develop a MCMC sampling strategy
-that takes advantage of closed form full conditional distributions
+difficult (cite recent cent/noncent papers \citep{Rue:Held:2005,Chri:etal:2006,Tan:Nott:2013}.
+To develop an algorithm, we make use of a latent variable representation
+for the multinomial probit model \citep{McCu:Ross:1994}. The representation
+introduces latent variables that allow one to develop a MCMC sampling
+strategy that takes advantage of closed form full conditional distributions
 (so-called Gibbs sampling steps) for $\alpha_{p}$.
@@
 associated with two variables $W_{ij1}$ and $W_{ij2}$, governed
-by the latent variables $\alpha_{i1}$ and $\alpha_{i2}$, respectively.
-Suppose that $\alpha_{i1}>\alpha_{i2}$ for a given location $i$.
+by the latent variables $\alpha_{1i}$ and $\alpha_{2i}$, respectively.
+Suppose that $\alpha_{1i}>\alpha_{2i}$ for a given location $i$.
 Then this model implies that any tree $j$ is more likely to be labeled
-1 than 2 at location $i$. The difference between $\alpha_{i1}$ and
-$\alpha_{i2}$ explains the \textit{difference} in probability of
+1 than 2 at location $i$. The difference between $\alpha_{1i}$ and
+$\alpha_{2i}$ explains the \textit{difference} in probability of
 \textit{categories} 1 and 2 at location $i$, and the similarity between
-$\alpha_{1p}$ and $\alpha_{2p}$ explains the \textit{correlation}
+$\alpha_{p1}$ and $\alpha_{p2}$ explains the \textit{correlation}
 between the probabilities at \textit{locations} 1 and 2 for category
@@
 These latent 'membership' variables, $c_{tj}$, for tree $j$ in township
-$t$ can be sampled within the MCMC as additional unknown parameters.
-The prior for $c_{tj}$ is a discrete distribution that puts mass
-proportional to the overlap between the township in which the tree
-is located and the grid cells that intersect the township, $\psi_{1(t)},\ldots,\psi_{T(t)}$,
-with the priors across different trees in a township being independent.
+$t$, $t=1,\ldots,T$ can be sampled within the MCMC as additional
+unknown parameters. The prior for $c_{tj}$ is a discrete distribution
+that puts mass proportional to the areal overlap between the township
+in which the tree is located and the $m$ grid cells, with the priors
+across different trees in a township being independent and with most
+of the . 
 \[
-c_{tj}\sim\mbox{Multinom}(1,\{\psi_{1(t)},\ldots\psi_{T(t)}\}).
+c_{tj}\sim\mbox{Multinom}(1,\{\psi_{1},\ldots,\psi_{m}\}).
 \]
-
-In updating the other parameters in the model, we condition on the
-current values, $c_{tj}$, which provides a ``soft'' assignment
-of trees to grid cells that respects both the township in which the
-tree was sampled and the uncertainty in which grid cell the tree was
-sampled.
+Because the townships overlap a limited number of grid cells, most
+of the $\psi_{1},\ldots,\psi_{m}$ values are zero.
+
+Using the latent variable representation, we have that $W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{c_{tj}},1)$
+for tree $j$ in township $t$. In updating the other parameters in
+the model during the MCMC (specifically the $\alpha$ values), we
+condition on the current values, $c_{tj}$, which provides a ``soft''
+assignment of trees to grid cells that respects both the known township
+in which the tree occurred and the uncertainty in which grid cell
+the tree occurred.
 
@@
 particularly in this high-dimensional spatial context, as it allows
-us to draw from the posterior conditional distribution of the $W_{ijp}$
+us to draw from the posterior conditional distributions of the $W_{ijp}$
 variables (these distributions are truncated normal) in closed form
-and to draw the entire vector of latent process values, $\alpha_{p}$,
-as a single sample that respects the spatial dependence structure
-for each taxon.
+and to draw the entire vector of latent process values for each taxon,
+$\alpha_{p}$, as a single sample that respects the spatial dependence
+structure for each taxon.
 
@@
 in the MCMC sampling for each $\alpha_{p}$ compared to joint Metropolis
-updates or single location at a time updates, there is still strong
+updates or updating each location individually, there is still strong
 dependence between the hyperparameters, $\{\sigma_{p}^{2},\mu_{p},\rho_{p}\}$
-and the the latent process values (as well as between the latent process
+and the latent process values (as well as between the latent process
 values and the latent $W_{ijp}$ variables). To address the first,
 we developed a 'cross-level' joint updating strategy in which we propose
-$\phi\in\{\sigma_{p}\},p=1,\ldots,P$ and for the SPDE model, $\phi\in\{\{\sigma_{p}\},\{\mu_{p}\},\{\rho_{p}\}\},$
+$\phi\in\{\sigma_{p}\},p=1,\ldots,P$ (and for the SPDE model, $\phi\in\{\{\sigma_{p}\},\{\mu_{p}\},\{\rho_{p}\}\}),$
 via a Metropolis-style random walk and then conditional on the proposed
-value of $\phi$ propose $\alpha_{p}$ from its full conditional distribution
-\textbackslash{}$\alpha_{p}$ and sampling from the marginalized distribution
-$\phi|W$. Note that in sampling $\sigma_{p}$and $\rho_{p}$ in the
-SPDE model, for each $p$, we jointly propose $\{\sigma_{p},\rho_{p}\}$
-using an adaptive Metropolis proposal and then use the cross-level
-strategy to also propose $\alpha_{p}$ as part of a single accept-reject
-step. For these various joint samples of hyperparameters and $\alpha_{p}$,
-we use adaptive Metropolis sampling \citep{Shab:Well:2011}.
+value of $\phi^{*}$ propose $\alpha_{p}$ from its full conditional
+distribution given $\phi^{*}$and the latent $W$ variables. This
+is equivalent to sampling from the marginalized (with respect to $\alpha_{p}$)
+distribution of $\phi$ conditional on $W_{p}$. Note that in sampling
+$\sigma_{p}$and $\rho_{p}$ in the SPDE model, for each $p$, we
+jointly propose $\{\sigma_{p},\rho_{p}\}$ using an adaptive Metropolis
+proposal and then use the cross-level strategy to also propose $\alpha_{p}$
+as part of a single accept-reject step. For these various joint samples
+of hyperparameters and $\alpha_{p}$, we use adaptive Metropolis sampling
+\citep{Shab:Well:2011}.
 
@@
 
-The model is implemented in R with core computational calculations
-coded in C++ using the Rcpp package \citep{Edde:Fran:2011}. We also
-make extensive use of sparse matrix representations and algorithms,
-using the spam package in R \citep{Furr:Sain:2010}. All code is available
-on Github, including pre- and post-processing code.
+The model is implemented in R \cite{R:2014} with core computational
+calculations coded in C++ using the \emph{Rcpp} package \citep{Edde:Fran:2011}.
+We also make extensive use of sparse matrix representations and algorithms,
+using the \emph{spam} package in R \citep{Furr:Sain:2010}. All code
+is available on Github, including pre- and post-processing code, at
+https://github.com/Paleon-Project/composition.
 
@@
 and hereafter referred to as the ``Minnesota subregion'' . We used
-two types of test data:
+two types of experiments:
 \begin{enumerate}
@@
 consistent with the distribution \citep{Gnei:etal:2007}. When thinking
-in terms of prediction intervals, we seek intervals that are as narrow
-as possible while still covering the truth the expected proportion
-of the time. Following the suggestions in \citep{Gnei:etal:2007},
-we consider these metrics.
-
-We used the following criteria to compare models. For experiment 1,
-$Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$ is the count of all trees in the
-held-out cell and for experiment 2, $Y_{i}$ is the count of held-out
-individual trees in the location.
+in terms of prediction intervals as summaries of the predictive distribution,
+we seek intervals that are as narrow as possible while still covering
+the truth the expected proportion of the time.
+
+Following the suggestions in \cite{Gnei:etal:2007}, we considered
+the following metrics. For experiment 1, $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$
+is the count of all trees in the held-out cell and for experiment
+2, $Y_{i}$ is the count of held-out individual trees in the cell.
+For each of these metrics, we calculated the metric in two ways. First,
+we used the posterior mean composition estimates (as a measure of
+our core predictions), with $\tilde{\theta}_{p}(s)$ being the posterior
+mean. Second, we averaged the metric over the posterior samples (as
+a measure of our full data product, including uncertainty), taking
+$\tilde{\theta}_{p}(s)$ to be an individual MCMC sample and then
+averaging the metric over all the posterior samples. 
 \begin{enumerate}
-\item Brier score: \citep{Gnei:etal:2007} suggest this measure, which has
+\item Brier score: \cite{Gnei:etal:2007} suggest this measure, which has
 been in use for decades. For multi-category as opposed to binary outcomes,
@@
 \[
-\frac{1}{n_{i}}\sum_{j=1}^{n_{i}}\sum_{p=1}^{P}(y_{ijp}-\theta_{p}(s_{i}))^{2}
+\frac{1}{n_{i}}\sum_{j=1}^{n_{i}}\sum_{p=1}^{P}(y_{ijp}-\tilde{\theta}_{p}(s_{i}))^{2}
 \]
@@
 \item Log predictive density: This measure takes the log of the density
-of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\theta_{1}(s_{i}),\ldots,\theta_{P}(s_{i})\})$,
+of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
 summing on the log scale across all of the held-out data.
@@
 Carlo estimation of $\theta$ used 10000 samples, so in some cases
-$\theta_{p}(s)=0$. When a tree is present in a cell but its corresponding
-proportion is 0 this gives a log density of $-\infty$, preventing
-use of the metric. As an informal solution to this we set $\theta_{p}(s)=0.5\frac{1}{10000}$
-in such cases, but given these issues we treat the log predictive
-density as a secondary measure.
-
-\item Weighted root mean square prediction error (RMSPE) and mean absolute
-error (MAE) for individual cells (only for experiment 1), where we
-weight by the number of held-out trees to account for the greater
+$\tilde{\theta}_{p}(s)=0$. When a tree is present in a cell but its
+corresponding proportion is 0, this gives a log density of $-\infty$,
+preventing use of the metric. As an informal solution to this we set
+$\tilde{\theta}_{p}(s)=\frac{1}{100000}$ in such cases, but given
+these issues we treat the log predictive density as a secondary measure.
+
+\item (Experiment 1 only) Weighted root mean square prediction error (RMSPE)
+and mean absolute error (MAE), averaging over cells and taxa. We weight
+by the number of held-out trees in each cell to account for the greater
 variability in the empirical proportions in locations with few held-out
 trees. 
-\item Coverage and length of 95\% prediction intervals for $Y_{ip}$. 
+\item (Experiment 1 only) Coverage and length of 95\% prediction intervals
+for $Y_{ip}$. We considered only cells with at least 50 trees to
+focus our assessment on cases where the raw estimates, $\hat{\theta}_{p}(s)$,
+were reasonably certain and avoid being strongly influenced by predictive
+inference for cells where observational variability dominates.
 \end{enumerate}
-For each of these metrics, we calculated the metric using the posterior
-mean composition estimates (as a measure of our core predictions)
-and averaging the metric over the posterior samples (as a measure
-of our full data product, including uncertainty). In other words in
-the first we first average over the composition posterior samples
-and in the second we average of the posterior values of the cross-validation
-metric.
-
-In our fitting we noticed that the SPDE model produced boundary effects
-in the predicted composition near the edges of the convex hull of
-the observations. To attempt to alleviate this, we added a buffer
-zone of 6 grid cells around our entire original domain, though the
-boundary effects were still evident even after inclusion of the buffer.
-For the model comparison, for comparability, we included this buffer
-for both the SPDE and CAR models.
+Note that for the calculation of the metric applied to the individual
+posterior samples, we can estimate the posterior probability that
+one model has a lower (better) value of the metric than the other
+model by simply calculating the proportion of samples for which each
+model has a lower value of the metric.
+
+In our initial exploratory fitting, we noticed that the SPDE model
+produced boundary effects in the predicted composition near the edges
+of the convex hull of the observations. To attempt to alleviate this,
+we added a buffer zone of six grid cells around our entire original
+domain, although the boundary effects were still evident even after
+inclusion of the buffer. For the model comparison, for comparability,
+we included this buffer for both the SPDE and CAR models.
 
 For the model comparison, we used an earlier version of the dataset
-than that used for the final data product as the model comparison
-runs involved substantial computation. This earlier version had incomplete
-data for southern Michigan and was missing a small number of trees
-(39 individual trees) classified as ``other hardwood'' due to an
-update to our taxonomic aggregations of common names used by surveyors.
-Given that part of the goal of the exercise is to understand the ability
-of the models to interpolate to areas without data and the ongoing
-nature of digitization, we believe this difference in datasets is
-not an issue.
-
-We ran each model for 150,000 iterations, retaining 250 after a burnin
-(25,000 iterations) and subsampling to reduce storage needs.
+{[}include text if we stick with 0.5 for CV{]}. Given that part of
+the goal of the exercise is to understand the ability of the models
+to interpolate to areas without data and the ongoing nature of digitization,
+we believe this difference in datasets is not an issue.
+
+We ran each model for 150,000 iterations. After discarding 25,000
+iterations for burn-in, we retained a posterior sample of 250 subsampled
+iterations to reduce post-processing computations and storage needs.
 
@@
 Experiment 1: those cells in the Minnesota subregion held out of the
-fitting process, for both the posterior mean predictions and the full
-posterior sample. Table \ref{tab:coverage_cell_fivepercent} shows
-performance of the uncertainty estimates. For the posterior distribution
-over the predictive score values, the CAR model outperforms the SPDE
-model, while for the posterior mean predictions, the SPDE model appears
-to outperform the CAR model, but we do not have any uncertainty estimates
+fitting process, for both the mean of the metric applied to the individual
+posterior samples and for the metric applied to the posterior mean
+predictions. Table \ref{tab:coverage_cell_fivepercent} shows performance
+of the uncertainty estimates. For the posterior distribution over
+the predictive metric values, the CAR model outperforms the SPDE model,
+while for the posterior mean predictions, the SPDE model appears to
+outperform the CAR model, but we do not have any uncertainty estimates
 for this comparison. Coverage and interval lengths are similar between
@@
 \begin{table}
-\caption{Predictive ability based on several predictive score criteria for
+\caption{Predictive ability based on several predictive metric criteria for
 the CAR and SPDE spatial models when holding out 95\% of entire cells
@@
  &
-\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
-\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
+\multicolumn{3}{c|}{{\small{Posterior mean of metric}}} &
+\multicolumn{2}{c|}{{\small{Metric of posterior mean predictions}}}\tabularnewline
 \hline 
@@
 to explain and computations can be done more quickly. Third, predictions
-from the SPDE model showed boundary effects, with taxa showing non-negligible
-posterior mean values at the edges of the domain, well away from where
-the taxon was present in the empirical data. This included non-negligible
-values within (but near the edge of) the convex hull of locations
-with data.
+from the SPDE model showed boundary effects, with some taxa showing
+non-negligible posterior mean values at the edges of the domain, well
+away from where the taxa were present in the empirical data. This
+included non-negligible values within (but near the edge of) the convex
+hull of locations with data.
 
@@
 for 150,000 iterations, discarding the first 25,000 iterations for
-burn-in and retaining 250 to limit storage needs and limit the size
-of the data product. Mixing was generally reasonable, but for some
-of the hyperparameters was relatively slow, particularly for less
-common taxa. Despite this, mixing for the variables of substantive
+burn-in and retaining 250 subsampled iterations to limit storage needs
+and limit the size of the data product. Mixing was generally reasonable,
+but for some of the hyperparameters was relatively slow, particularly
+for less common taxa. Despite this, mixing for the variables of substantive
 interest, the proportions was good. {[}Report ESS values{]}
@@
 
-\caption{Empirical proportions from raw data (column 1), posterior mean predictions
-(column 2) and posterior standard deviations -- representing standard
-errors of prediction (column 3) for select taxa.}
+\caption{Empirical proportions from raw data (column 1), predictions in the
+form of posterior means (column 2) and uncertainty estimates in the
+form of posterior standard deviations -- representing standard errors
+of prediction (column 3) for select taxa.}
 \end{figure}
@@
 
-\caption{Posterior mean predictions for all taxa over the entire domain.}
+\caption{Predictions (posterior means) for all taxa over the entire domain.}
 
@@
 
-The data product is publicly available at XX under YY license as version
-0.3 as of April 2015. The product is in the form of a netCDF-4 file,
-with dimensions x, y, and MCMC iteration. There is one variable per
-taxon. The PalEON project will continue to maintain this product,
-releasing new versions as additional data in Illinois, Indiana and
-Ohio are digitized.
+The data product is publicly available at XX (insert accession number)
+under YY license as version 0.3 as of April 2015. The product is in
+the form of a netCDF-4 file, with dimensions x, y, and MCMC iteration.
+There is one variable per taxon. The PalEON project will continue
+to maintain this product, releasing new versions as additional data
+in Illinois, Indiana and Ohio are digitized.
+
+{[}comment on stability of hosting once get better feel for this{]}
 
@@
 
-Note that digitization of data from Illinois and Indiana is ongoing
+Note that digitization of data from Illinois and Indiana is ongoing,
 and digitization of additional data from Ohio is planned as well.
@@
 Given the density of data and the limited differences seen between
-the ICAR and SPDE models, we expect the data product to be reasonably
+the CAR and SPDE models, we expect the data product to be reasonably
 robust to the choice of spatial model, particularly in those areas
@@
 years who have preserved, collected, and digitized survey records,
-in particular Ed Schools (Michigan DNR), \{ Brugam, ??? - Jody/Charlie/David/Jason/Simon
-please add names here\}. This work was carried out by the PalEON Project
-with support from the National Science Foundation MacroSystems Program
-through grants EF-1065656, EF-1241868, DEB-1241874 and DEB-1241868.
-
-\bibliographystyle{/accounts/gen/vis/paciorek/latex/RSSstylefile/Chicago}
-
-\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,goring_intro}
+in particular Jim Dyer, Peter Marks, Robert McIntosh, and Ed Schools
+\{David/Simon please add names here\}\{Jason/Jody, should we mention
+Bowles and Stuckey; if so I need full names\}. We thank Madeline Ruid,
+Ben Seliger, Morgan Ross and Daniel Handel for processing of the southern
+Michigan data. Indiana and Illinois data were made possible through
+the hard work of many Notre Dame undergraduates in the McLachlan lab.
+This work was carried out by the PalEON Project with support from
+the National Science Foundation MacroSystems Program through grants
+EF-1065656, EF-1241868, DEB-1241874 and DEB-1241868 and from the Notre
+Dame Environmental Change Initiative.
+
+%\bibliographystyle{/accounts/gen/vis/paciorek/latex/RSSstylefile/Chicago}
+\bibliographystyle{/accounts/gen/vis/paciorek/latex/PLOS_ONE/plos215}
+
+\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
 
@@
 
-For the township-level data, for a given tree, we draw the latent
-tree membership variable, $c_{tj}$, from a discrete distribution
-by normalizing the weights, $\{\psi_{1(t)}L_{1(t)},\ldots,\psi_{T(t)}L_{T(t)}\}$
-where $L_{k}$ is density of $W_{tj\cdot}$ if $c_{tj}=k$, namely
+For the township-level data, for a given tree $j$ in township $t$,
+we draw the latent tree membership variable, $c_{tj}\in\{1,\ldots,m\}$,
+from a discrete distribution by normalizing posterior weights, $\{\psi_{1}L_{1},\ldots,\psi_{m}L_{m}\}$,
+produced by multiplying the prior weights by a likelihood contribution.
+For cell $k$, $L_{k}$ is the density of the latent $W_{tj1},\ldots,W_{tjP}$
+values for the given tree under the condition that $c_{tj}=k$, namely
 the product of independent normal densities, $W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{k}),1)$,
-over $p=1,\ldots,P$. Thus the posterior weights the prior based on
-how consistent the $W$ values are with the $\alpha$ values for the
-candidate grid cells.
+over $p=1,\ldots,P$. Thus the posterior reweights the prior based
+on how consistent the current $W$ values for a tree are with the
+$\alpha$ values for the candidate grid cells.
 
@@
 \[
-W_{ijpt}^{(k)}\sim\mbox{N}(\alpha_{p}^{(k)}(s_{i}),1),\, p=1,\ldots,P
+W_{itp}^{(k)}\sim\mbox{N}(\alpha_{p}^{(k)}(s_{i}),1),\, p=1,\ldots,P
 \]
@@
 \[
-\hat{\theta}_{p}^{(k)}(s_{i})=\frac{1}{10000}{\displaystyle \sum_{t=1}^{10000}1(W_{ijpt}^{(k)}={\displaystyle \max_{p^{*}}W_{ijp^{*}t}^{(k)})}}
+\theta_{p}^{(k)}(s_{i})\approx\frac{1}{10000}{\displaystyle \sum_{t=1}^{10000}1(W_{itp}^{(k)}={\displaystyle \max_{p^{*}}W_{itp^{*}}^{(k)})}}
 \]
@@
 the expression is true and 0 if false. In other words, we calculate
-the proportion of times that the maximum of $W_{ijp},\, p=1,\ldots,P$
-corresponds to taxon $p$. Considering $\hat{\theta}_{p}^{(k)}(s_{i}),\, k=1,\ldots,K$,
+the proportion of times that the maximum of $W_{itp},\, p=1,\ldots,P$
+corresponds to taxon $p$. Considering $\theta_{p}^{(k)}(s_{i}),\, k=1,\ldots,K$,
 we have a sample from the posterior of $\theta_{p}(s_{i})$.

### 2015-08-20 paper.tex
added final edits from coauthors and cleaned up some formatting

@@
 \& These authors contributed equally to this work.
+
+Short title: Settlement-era northeast U.S. tree composition
 \end{quote}
-TODO:
-
-focus on PLOS One, but Nature's Scientific Data is a possibility
-
-check journal requirements for archiving and presentation; talk to
-Jody/Ann about archiving/DOI
 
@@
 giving count data. The domain is divided into two regions, eastern
-(Maine to Ohio) and western (Indian to Minnesota). Public Land Survey
+(Maine to Ohio) and western (Indiana to Minnesota). Public Land Survey
 point data in the western region are aggregated to a regular 8 km
 grid, while data in the eastern region, from Town Proprietor Surveys,
-is aggregated at the township level in irregularly-shaped local administrative
+are aggregated at the township level in irregularly-shaped local administrative
 units. The product is based on a Bayesian statistical model fit to
@@
 Historical datasets provide critical context to understand forest
-ecology. They allow researchers to define 'baseline' conditions for
+ecology. They allow researchers to define `baseline' conditions for
 conservation management, to understand ecosystem processes at decadal
@@
 
-Euro-American settlement and subsequent land use change occured in
-a time transient fashion across North America, and land surveys that
+Euro-American settlement and subsequent land use change occurred in
+a time-transient fashion across North America, and land surveys that
 provide vegetation information followed a similar pattern. Early surveys
@@
 data at the township level \citep{cogbill2002forests,thompson2013four};
-we refer to these as the Town Proprietor Survey (TPS). These surveys
-provide composition of major tree species using sometimes inconsistent
-common names. Later surveys, after the establishment of the U.S. Public
-Land Survey System (PLS) - from 1832 to 1907 - provide point-level
-data along a regular grid, with one mile spacing \citet{bourdo1956review,schulte2001original,goring2015composition},
-again with sometimes inconsistent common names. Survey instructions
-during the PLS varied through time and by point type. This requires
-the application of spatially-varying correction factors \citep{cogbill2015corrections,goring2015composition}
+we refer to these as the Town Proprietor Survey (TPS). Later surveys,
+after the establishment of the U.S. Public Land Survey System (PLS)
+provide point-level data along a regular grid, with one mile spacing
+during the period 1832 to 1907 \citep{bourdo1956review,schulte2001original,goring2015composition}.
+Survey instructions during the PLS varied through time and by point
+type. Accounting for this requires the application of spatially-varying
+correction factors \citep{cogbill2015corrections,goring2015composition}
 to accurately assess stem density, basal area and biomass from the
-early settlement records but has limit impact when estimating composition.
-
-Logging, agriculture, and abandonment have left an indelible mark
-on forests in the northeastern United States \citet{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
+early settlement records, but the impact on composition estimates
+is limited. Both types of surveys provide composition of major tree
+species but using sometimes inconsistent common names.
+
+Logging, agriculture, and land abandonment have left an indelible
+mark on forests in the northeastern United States \citep{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
 However most studies have assessed these effects in individual states
-or smaller domains \citep{rhemtulla2009historical,friedman2005regional}
+or smaller domains \citep{friedman2005regional,rhemtulla2009historical}
 and with varying spatial resolution, from townships (36 square miles)
@@
 8 km grid size, providing broad spatial coverage at a spatial scale
-that can be compared to the Forest Inventory and Analysis products
-\citet{gray2012forest}. Combined with additional, coarsely-sampled
-PLS data from Illinois and Indiana and with the TPS data, this gives
-us raw data for much of the northeastern United States. However, there
-are several limitations of using the raw data that can be alleviated
-by the use of a statistical model to develop a statistically-estimated
+that can be compared to the modern forests using Forest Inventory
+and Analysis products \citep{gray2012forest}. Combined with additional,
+coarsely-sampled PLS data from Illinois and Indiana and with the TPS
+data, this gives us raw data for much of the northeastern United States.
+However, there are several limitations of using the raw data that
+can be alleviated by the use of a statistical model to develop a statistically-estimated
 data product. First, the PLS and TPS data provide only estimates of
@@
 Indiana represent a small fraction of those states and missing townships
-are common in the TPS data.
-
-Properly assessing uncertainty in ecological data is imperative to
-understanding and modelling ecological processes \citep{cressie2009accounting}.
-In this way, a model that can account for the spatial structure of
-the underlying PLS and TPS data, and provide reliable estimates of
-uncertainty across the northeastern United States, provides a valuable
-tool for researchers interested in the ecological structure and function
-of forests at longer time scales.
-
-\section{Data}
+are common in the TPS data. Our statistical model allows us to provide
+a spatially-complete data product of settlement-era tree composition
+with uncertainty across the northeastern U.S.
+
+In Section \ref{sec:Data} we describe the data sources, while Section
+\ref{sec:Statistical-model} describes our statistical models. In
+Section \ref{sec:Model-comparison} we quantitatively compare competing
+statistical specifications, and in Section \ref{sec:Data-product}
+we describe the final data product. In Section \ref{sec:Discussion}
+we discuss limitations of the statistical model, and we list related
+data products under development.
+
+\section{Data\label{sec:Data}}
 
@@
 the remaining states are available for a subset of the full set of
-townships covering the domain. Fig. \ref{fig:domain} {[}Fig. 1 -
-Lyx is having a numbering problem{]} shows the domain, indicating
-the grid cells and townships with data.
+townships covering the domain. Fig. \ref{fig:Spatial-domain} shows
+the domain, indicating the grid cells and townships with data.
 
 \begin{figure}
-\label{fig:domain}\includegraphics{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}
-
-\caption{Spatial domain, with locations with data shown in gray. Locations
-are grid cells in western portion and townships in eastern portion.
-In addition to locations without data being indicated in white, grid
-cells completely covered in water are white (e.g., a few locations
-in Minnesota and Wisconsin).}
+\label{fig:domain}\includegraphics[scale=1.3]{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}
+
+\caption{Spatial domain of the northeastern United States, with locations with
+data shown in gray. Locations are grid cells in western portion and
+townships in eastern portion. In addition to locations without data
+being indicated in white, grid cells completely covered in water are
+white (e.g., a few locations in the northwestern portion of the domain
+in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
 \end{figure}
@@
 
-Extensive details on the data are available in \cite{Gori:etal:2015}
-and the raw data are available at XXX {[}ask Simon/Charlie about public
-access to Paleon composition data{]}. The aggregation into taxonomic
-groups is primarily at the genus level, but is at the species level
-in some cases of monospecific genera. We model the following 22 taxa
-plus an ``other hardwood'' category: Atlantic white cedar (\emph{Chamaecyparis
-thyoides}), Ash (\emph{Fraxinus spp.}), Basswood (\emph{Tilia americana}),
-Beech (\emph{Fagus grandifolia)}, Birch (\emph{Betula spp.}), Black
-gum/sweet gum (\emph{Nyssa sylvatica} and \emph{Liquidambar styraciflua}),
-Cedar/juniper (\emph{Juniperus virginiana} and \emph{Thuja occidentalis}),
-Cherry (\emph{Prunus spp.}), Chestnut (\emph{Castanea dentata}), Dogwood
-(\emph{Cornus spp.}), Elm (\emph{Ulmus spp.}), Fir (\emph{Abies spp.}),
-Hemlock (\emph{Tsuga canadensis}), Hickory (\emph{Carya spp.}), Ironwood
-(\emph{Carpinus caroliniana} and \emph{Ostrya virginiana}), Maple
-(\emph{Acer spp.}), Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}),
-Poplar/tulip poplar (\emph{Populus spp.} and \emph{Liriodendron tulipifera}),
-Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), Walnut
-(\emph{Juglans nigra}). Note that in several cases (e.g., black gum/sweet
+Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
+data and processing steps are available in \cite{Gori:etal:2015}.
+The lower Midwest (Illinois, Indiana) data were purchased from the
+Indiana State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
+Illinois) and processed using similar steps as for the upper Midwest
+data. The TPS data were compiled by C.V. Cogbill from a myriad of
+archival sources representing land division surveys conducted in connection
+with local settlement. The complete raw vegetation data are available
+through the NIS Data Portal (accession number and version to be inserted
+when available). The aggregation into taxonomic groups is primarily
+at the genus level, but is at the species level in some cases of monospecific
+genera. We model the following 22 taxa plus an ``other hardwood''
+category: Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash
+(\emph{Fraxinus spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus
+grandifolia)}, Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa
+sylvatica} and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
+virginiana} and \emph{Thuja occidentalis}), Cherry (\emph{Prunus spp.}),
+Chestnut (\emph{Castanea dentata}), Dogwood (\emph{Cornus spp.}),
+Elm (\emph{Ulmus spp.}), Fir (\emph{Abies spp.}), Hemlock (\emph{Tsuga
+canadensis}), Hickory (\emph{Carya spp.}), Ironwood (\emph{Carpinus
+caroliniana} and \emph{Ostrya virginiana}), Maple (\emph{Acer spp.}),
+Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}), Poplar/tulip
+poplar (\emph{Populus spp.}~and \emph{Liriodendron tulipifera}),
+Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), and
+Walnut (\emph{Juglans nigra}). Note that in several cases (black gum/sweet
 gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
@@
 we do not fit statistical models for Atlantic white cedar and chestnut
-as these have 0 and 6 trees present, respectively. The taxa grouped
+as these have 0 and 7 trees present, respectively. The taxa grouped
 into the other hardwood category are those for which fewer than roughly
@@
 white cedar explicitly despite it only having 336 trees in the dataset.
-Given the nature of the data, there is not a firm cutoff on the sizes
-of trees included, but the data roughly represent trees greater than
-8 inches (\textasciitilde{}20 cm) diameter at breast height (dbh),
-but with some trees as small as 4 inches. {[}Charlie,Simon, please
-check this wording{]} In Indiana and Illinois there are also 402 trees
-and 29 trees, respectively, that are between 1 and 4 inches dbh, but
-these will be excluded in the next version of the product.
+Although surveyors avoided using small trees, there was no consistent
+lower diameter limit. The data, however, roughly represent trees greater
+than 8 inches (\textasciitilde{}20 cm) diameter at breast height (dbh),
+but with some trees as small as 3 inches dbh. In Indiana and Illinois
+there are also 402 trees and 29 trees, respectively, that are between
+1 and 4 inches dbh, but these will be excluded in the next version
+of the product {[}waiting on Simon to see if he stripped them out{]}.
 
@@
 rectangle split into 8 km cells, arranged in a 296 by 180 grid of
-cells, with the centroid of the cell in the southeast corner located
+cells, with the centroid of the cell in the southwest corner located
 at (-71000, 58000). For the modeling of the western subdomain we use
@@
 
-\section{Statistical model}
+\section{Statistical model\label{sec:Statistical-model}}
 
@@
 cells. These samples are the data product (described further in the
-Data Product section), and can then be used in subsequent analyses.
-The mean and standard deviation of the samples for each cell by taxon
-pair represent our best estimate (i.e., prediction) of composition
+Data Product section) and can then be used in subsequent analyses.
+The mean and standard deviation of the samples for each pair of cell
+and taxon represent our best estimate (i.e., prediction) of composition
 and a Bayesian ``standard error'' quantifying the uncertainty in
@@
 We start by describing the basic model for those states for which
-we have raw data on the 8 km grid and in Section \ref{sub:Model-for-township}
+we have raw data on the 8 km grid, and in Section \ref{sub:Model-for-township}
 we describe the extension of the model to accommodate data aggregated
@@
 The statistical model treats the observations as coming from a multinomial
-distribution with a (latent) vector of proportions for each grid cell.
+distribution with a (latent) vector of proportions for each grid cell,
 \[
-y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i}))
+y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i})),
 \]
@@
 MRF models work directly with the precision matrix of the values of
-the spatial process, so calculation of the prior density of $\alpha_{p}(s)$
+the spatial process, so calculation of the prior density of $\alpha_{p}$
 is computationally simple \citep{Rue:Held:2005}, but in situations
@@
 
-\subsubsection{Standard conditional autoregressive models}
+\paragraph{Standard conditional autoregressive models}
 
@@
 \[
-\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-})
+\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-}).
 \]
@@
 matrix defining the neighborhood relation of the locations; that is
-$(C)_{ik}=1$ if locations $i$ and $k$ are neighbors. The matrix
+$C{}_{ik}=1$ if locations $i$ and $k$ are neighbors. The matrix
 $D$ is an $m\times m$ diagonal matrix containing the row sums of
-matrix $C$ as the diagonal entries $(D)_{ii}={\displaystyle \sum_{k=1}^{m}(C)_{ik}}.$
+matrix $C$ as the diagonal entries $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$
 
@@
 
-\subsubsection{Gaussian process approximation}
+\paragraph{Gaussian process approximation}
 
@@
 The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
-to the Matern-based GP for $\nu=1$ and $\nu=2$. Our second model
-is the Lindgren approximation for Matern-based GPs with $\nu=1$.
+to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second model
+is the Lindgren approximation for \matern-based GPs with $\nu=1$.
 To implement the Lindgren model, one modifies the $Q$ matrix defined
@@
 \noindent The ICAR specification contains a set of hyperparameters
-$\sigma_{p}^{2}$ for $p$ = $1,...,P$. Following \cite{Gelm:2006}
+$\{\sigma_{p}^{2}\},$ $p=1,\ldots P$. Following \cite{Gelm:2006}
 we use a uniform distribution on each $\sigma_{p}$ parameter, with
-upper bound of {[}check max values{]}. For the SPDE model we also
-have parameters $\mu_{p}$, which we give flat, non-informative priors
-(truncated at $\pm10$), and $\rho_{p}$ which we give uniform priors
-on the interval $(0.1,\exp(5))$.
+upper bound of 1000. For the SPDE model we also have parameters $\mu_{p}$,
+which we give flat, non-informative priors (truncated at $\pm10$),
+and $\rho_{p}$ which we give uniform priors on the interval $(0.1,\exp(5))$.
 
@@
 with latent Gaussian process(es) and a non-Gaussian likelihood is
-difficult (cite recent cent/noncent papers \citep{Rue:Held:2005,Chri:etal:2006,Tan:Nott:2013}.
-To develop an algorithm, we make use of a latent variable representation
+difficult \citep{Rue:Held:2005,Chri:etal:2006,Tan:Nott:2013}. To
+develop an algorithm, we make use of a latent variable representation
 for the multinomial probit model \citep{McCu:Ross:1994}. The representation
@@
 Suppose that compositional counts are available at a number of locations.
-At location $i$, a sample size of $n_{i}$ observations are collected,
+At location $i$, a sample of size $n_{i}$ observations is collected,
 and each observation (i.e., each tree) can be classified into $P$
@@
 in other words, the maximum of the set of latent variables $\{W_{ijp}\}{\displaystyle _{p=1}^{P}}$
-determines the category of observation $j$ at location $i$.
-
-\noindent The final piece of the latent variable representation is
-the relationship between the $W$ variables and the $\alpha_{p}(s)$
-processes. We have that
+determines the category of observation $j$ at location $i$. The
+final piece of the latent variable representation is the relationship
+between the $W$ variables and the $\alpha_{p}(s)$ processes. We
+have that
 
@@
 \]
-independently for all of the $W_{ijp}$ values.
-
-\noindent Consider the following example with two locations that are
-neighbors and $P=2$ categories. Each tree $j$ at location $i$ is
-associated with two variables $W_{ij1}$ and $W_{ij2}$, governed
-by the latent variables $\alpha_{1i}$ and $\alpha_{2i}$, respectively.
-Suppose that $\alpha_{1i}>\alpha_{2i}$ for a given location $i$.
-Then this model implies that any tree $j$ is more likely to be labeled
-1 than 2 at location $i$. The difference between $\alpha_{1i}$ and
-$\alpha_{2i}$ explains the \textit{difference} in probability of
-\textit{categories} 1 and 2 at location $i$, and the similarity between
-$\alpha_{p1}$ and $\alpha_{p2}$ explains the \textit{correlation}
-between the probabilities at \textit{locations} 1 and 2 for category
-$p$.
+independently for all of the $W_{ijp}$ values. Consider the following
+example with two locations that are neighbors and $P=2$ categories.
+Each tree $j$ at location $i$ is associated with two variables $W_{ij1}$
+and $W_{ij2}$, governed by the latent variables $\alpha_{1i}$ and
+$\alpha_{2i}$, respectively. Suppose that $\alpha_{1i}>\alpha_{2i}$
+for a given location $i$. Then this model implies that any tree $j$
+is more likely to be labeled 1 than 2 at location $i$. The difference
+between $\alpha_{1i}$ and $\alpha_{2i}$ explains the \textit{difference}
+in probability of \textit{categories} 1 and 2 at location $i$, and
+the similarity between $\alpha_{p1}$ and $\alpha_{p2}$ explains
+the \textit{correlation} between the probabilities at \textit{locations}
+1 and 2 for category $p$.
+
+\noindent
+
+\noindent
 
@@
 grid. This extension introduces a new set of latent variables, one
-per tree, that indicate the grid cell in which the tree is located.
-These latent 'membership' variables, $c_{tj}$, for tree $j$ in township
-$t$, $t=1,\ldots,T$ can be sampled within the MCMC as additional
-unknown parameters. The prior for $c_{tj}$ is a discrete distribution
-that puts mass proportional to the areal overlap between the township
-in which the tree is located and the $m$ grid cells, with the priors
-across different trees in a township being independent and with most
-of the . 
+per tree, that indicate the grid cell in which the tree is located
+and can be sampled within the MCMC as additional unknown parameters.
+Specifically, $c_{tj}$, is the latent `membership' variable for tree
+$j$ in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$ is
+a discrete distribution that puts mass, $\psi_{ti}$, $i=1,\ldots,m$,
+proportional to the areal overlap between the township in which the
+tree is located and the $m$ grid cells, giving 
 \[
-c_{tj}\sim\mbox{Multinom}(1,\{\psi_{1},\ldots,\psi_{m}\}).
+c_{tj}\sim\mbox{Multinom}(1,\{\psi_{t1},\ldots,\psi_{tm}\}),
 \]
-Because the townships overlap a limited number of grid cells, most
-of the $\psi_{1},\ldots,\psi_{m}$ values are zero.
+independently across all trees. Because the townships overlap a limited
+number of grid cells, most of the $\psi_{t1},\ldots,\psi_{tm}$ values
+are zero.
 
@@
 values and the latent $W_{ijp}$ variables). To address the first,
-we developed a 'cross-level' joint updating strategy in which we propose
-$\phi\in\{\sigma_{p}\},p=1,\ldots,P$ (and for the SPDE model, $\phi\in\{\{\sigma_{p}\},\{\mu_{p}\},\{\rho_{p}\}\}),$
+we developed a `cross-level' joint updating strategy for the CAR model
+in which we propose $\phi_{p}=\sigma_{p},p=1,\ldots,P,$ (and for
+the SPDE model, $\phi_{p}\in\{\{\sigma_{p},\rho_{p}\},\mu_{p}\}$)
 via a Metropolis-style random walk and then conditional on the proposed
-value of $\phi^{*}$ propose $\alpha_{p}$ from its full conditional
-distribution given $\phi^{*}$and the latent $W$ variables. This
-is equivalent to sampling from the marginalized (with respect to $\alpha_{p}$)
-distribution of $\phi$ conditional on $W_{p}$. Note that in sampling
-$\sigma_{p}$and $\rho_{p}$ in the SPDE model, for each $p$, we
-jointly propose $\{\sigma_{p},\rho_{p}\}$ using an adaptive Metropolis
-proposal and then use the cross-level strategy to also propose $\alpha_{p}$
-as part of a single accept-reject step. For these various joint samples
-of hyperparameters and $\alpha_{p}$, we use adaptive Metropolis sampling
-\citep{Shab:Well:2011}.
+value of $\phi_{p}^{*}$ propose $\alpha_{p}$ from its full conditional
+distribution given $\phi_{p}^{*}$ and the latent $W_{p}$ variables.
+This is equivalent to sampling from the marginalized (with respect
+to $\alpha_{p}$) distribution of $\phi_{p}$ conditional on $W_{p}$.
+For these various joint samples of hyperparameters and $\alpha_{p}$,
+we use adaptive Metropolis sampling \citep{Shab:Well:2011}.
 
@@
 
-The model is implemented in R \cite{R:2014} with core computational
+The model is implemented in R \citep{R:2014} with core computational
 calculations coded in C++ using the \emph{Rcpp} package \citep{Edde:Fran:2011}.
@@
 
-\section{Model comparison}
+\section{Model comparison\label{sec:Model-comparison}}
 
@@
 and hereafter referred to as the ``Minnesota subregion'' . We used
-two types of experiments:
+two cross-validation experiments:
 \begin{enumerate}
@@
 of the cells in Minnesota subregion at random. This variation on the
-first test above was meant to indicate whether our model comparison
+first experiment above was meant to indicate whether our model comparison
 conclusions would be robust as the digitization process for Illinois
@@
 that predictive distributions should maximize sharpness subject to
-calibration. That is the predictive distribution should be as narrow
+calibration. That is, the predictive distribution should be as narrow
 as possible while being calibrated such that the observations are
@@
 Following the suggestions in \cite{Gnei:etal:2007}, we considered
-the following metrics. For experiment 1, $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$
-is the count of all trees in the held-out cell and for experiment
-2, $Y_{i}$ is the count of held-out individual trees in the cell.
-For each of these metrics, we calculated the metric in two ways. First,
-we used the posterior mean composition estimates (as a measure of
-our core predictions), with $\tilde{\theta}_{p}(s)$ being the posterior
-mean. Second, we averaged the metric over the posterior samples (as
-a measure of our full data product, including uncertainty), taking
-$\tilde{\theta}_{p}(s)$ to be an individual MCMC sample and then
-averaging the metric over all the posterior samples. 
+the metrics listed below. For experiment 1, we define $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$
+as the count of all trees in held-out cell $i$ and for experiment
+2, $Y_{i}$ is the count of held-out individual trees in the cell,
+while $y_{ijp}$ is an indicator variable taking value either 0 or
+1 depending on whether the $j$th held-out tree in the $i$th cell
+is of taxon $p$. We calculated each of the metrics below in two ways.
+First, we used the posterior mean composition estimates (as an evaluation
+of our core predictions), with $\tilde{\theta}_{p}(s)$ being the
+posterior mean. Second, we averaged the metric over the posterior
+samples (as an evaluation of our full data product, including uncertainty),
+taking $\tilde{\theta}_{p}(s)$ to be an individual MCMC sample and
+then averaging the metric over all the posterior samples. 
 \begin{enumerate}
-\item Brier score: \cite{Gnei:etal:2007} suggest this measure, which has
+\item Brier score: \cite{Gnei:etal:2007} suggest this metric, which has
 been in use for decades. For multi-category as opposed to binary outcomes,
@@
 \[
-\frac{1}{n_{i}}\sum_{j=1}^{n_{i}}\sum_{p=1}^{P}(y_{ijp}-\tilde{\theta}_{p}(s_{i}))^{2}
+\frac{1}{n}\sum_{i=1}^{m}\sum_{j=1}^{n_{i}}\sum_{p=1}^{P}(y_{ijp}-\tilde{\theta}_{p}(s_{i}))^{2}
 \]
-for each cell. The full Brier score for the complete set of held-out
-locations sums over all locations.
-\item Log predictive density: This measure takes the log of the density
-of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
+where $n=\sum_{i=1}^{m}n_{i}$ is the total number of held-out trees
+for a given experiment and $j$ indexes across held-out trees in cell
+$i$. 
+\item Log predictive density: This metric takes the log of the density of
+held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
 summing on the log scale across all of the held-out data.
@@
 \item (Experiment 1 only) Weighted root mean square prediction error (RMSPE)
-and mean absolute error (MAE), averaging over cells and taxa. We weight
-by the number of held-out trees in each cell to account for the greater
-variability in the empirical proportions in locations with few held-out
-trees. 
+and mean absolute error (MAE) of the estimated proportions relative
+to the empirical proportions based on the held-out trees, averaging
+over cells and taxa. We weight by the number of held-out trees in
+each cell to account for the greater variability in the empirical
+proportions in locations with few held-out trees. 
 \item (Experiment 1 only) Coverage and length of 95\% prediction intervals
 for $Y_{ip}$. We considered only cells with at least 50 trees to
-focus our assessment on cases where the raw estimates, $\hat{\theta}_{p}(s)$,
-were reasonably certain and avoid being strongly influenced by predictive
-inference for cells where observational variability dominates.
+focus our assessment on cases where empirical proportions were reasonably
+certain and avoid being strongly influenced by predictive inference
+for cells where observational variability dominates.
 \end{enumerate}
@@
 one model has a lower (better) value of the metric than the other
-model by simply calculating the proportion of samples for which each
+model by simply calculating the proportion of samples for which the
 model has a lower value of the metric.
@@
 we added a buffer zone of six grid cells around our entire original
-domain, although the boundary effects were still evident even after
-inclusion of the buffer. For the model comparison, for comparability,
+domain, but note that the boundary effects were still evident even
+after inclusion of the buffer. For the model comparison, for comparability,
 we included this buffer for both the SPDE and CAR models.
-
-For the model comparison, we used an earlier version of the dataset
-{[}include text if we stick with 0.5 for CV{]}. Given that part of
-the goal of the exercise is to understand the ability of the models
-to interpolate to areas without data and the ongoing nature of digitization,
-we believe this difference in datasets is not an issue.
 
@@
 iterations for burn-in, we retained a posterior sample of 250 subsampled
-iterations to reduce post-processing computations and storage needs.
+iterations (subsampled to reduce post-processing computations and
+storage needs).
 
@@
 
-\subsubsection{Full cell hold-out experiment}
-
-Table \ref{tab:score_cell_fivepercent} shows predictive ability for
-Experiment 1: those cells in the Minnesota subregion held out of the
-fitting process, for both the mean of the metric applied to the individual
-posterior samples and for the metric applied to the posterior mean
-predictions. Table \ref{tab:coverage_cell_fivepercent} shows performance
-of the uncertainty estimates. For the posterior distribution over
-the predictive metric values, the CAR model outperforms the SPDE model,
-while for the posterior mean predictions, the SPDE model appears to
-outperform the CAR model, but we do not have any uncertainty estimates
-for this comparison. Coverage and interval lengths are similar between
-the two models. From a practical perspective, based on the difference
-in mean absolute error, which is on a readily interpretable scale,
-the differences between the models are small.
+\paragraph{Full cell hold-out experiment}
+
+Table \ref{tab:score_cell_fivepercent} shows predictive ability based
+on Experiment 1 for those cells in the Minnesota subregion held out
+of the fitting process, for both the mean of the metric applied to
+the individual posterior samples and for the metric applied to the
+posterior mean predictions. Table \ref{tab:coverage_cell_fivepercent}
+shows performance of the uncertainty estimates. For the posterior
+distribution over the predictive metric values, the CAR model outperforms
+the SPDE model, while for the posterior mean predictions, the SPDE
+model appears to outperform the CAR model to a lesser degree, but
+we do not have any uncertainty estimates for this comparison. Coverage
+and interval lengths are similar between the two models. From a practical
+perspective, based on the difference in mean absolute error, which
+is on a readily interpretable scale, the differences between the models
+are small.
 
@@
 {\small{Brier}} &
-{\small{180688}} &
-{\small{186045}} &
+{\small{0.819}} &
+{\small{0.844}} &
 {\small{0.98}} &
-{\small{162813}} &
-{\small{161559}}\tabularnewline
+{\small{0.738}} &
+{\small{0.733}}\tabularnewline
 \hline 
 {\small{Negative Log Density}} &
-{\small{468397}} &
-{\small{513144}} &
+{\small{466325}} &
+{\small{510383}} &
 {\small{1.00}} &
-{\small{395685}} &
-{\small{395998}}\tabularnewline
+{\small{394003}} &
+{\small{394554}}\tabularnewline
 \hline 
@@
 {\small{0.0897}} &
-{\small{0.0959}} &
-{\small{0.96}} &
+{\small{0.0960}} &
+{\small{0.97}} &
 {\small{0.0647}} &
-{\small{0.0626}}\tabularnewline
+{\small{0.0627}}\tabularnewline
 \hline 
@@
 \caption{Coverage and length of prediction intervals for the CAR and SPDE spatial
-models when holding out 95\% of entire cells of data in Minnesota. }
+models when holding out 95\% of entire cells of data in Minnesota.
+Coverage values near 0.95 are optimal, while shorter intervals are
+better.}
 
@@
 {\small{Coverage}} &
-{\small{0.976}} &
+{\small{0.977}} &
 {\small{0.978}}\tabularnewline
@@
 \begin{table}
-\caption{Predictive ability based on several predictive score criteria for
+\caption{Predictive ability based on several predictive metric criteria for
 the CAR and SPDE spatial models when holding out 80\% of entire cells
-of data in Minnesota. Smaller values are better, except for coverage,
-where values near 0.95 are optimal.}
+of data in Minnesota. Smaller values are better.}
 
@@
 {\small{Brier}} &
-{\small{141546}} &
-{\small{140211}} &
+{\small{0.773}} &
+{\small{0.765}} &
 {\small{0.10}} &
-{\small{130006}} &
-{\small{129910}}\tabularnewline
+{\small{0.710}} &
+{\small{0.710}}\tabularnewline
 \hline 
 {\small{Negative Log Density}} &
-{\small{356985}} &
-{\small{355640}} &
-{\small{0.31}} &
-{\small{312509}} &
-{\small{312950}}\tabularnewline
+{\small{355928}} &
+{\small{353987}} &
+{\small{0.25}} &
+{\small{311525}} &
+{\small{311902}}\tabularnewline
 \hline 
@@
 {\small{0.0309}} &
-{\small{0.0297}} &
+{\small{0.0296}} &
 {\small{0.10}} &
-{\small{0.0225}} &
-{\small{0.0222}}\tabularnewline
+{\small{0.0226}} &
+{\small{0.0223}}\tabularnewline
 \hline 
@@
 {\small{0.0763}} &
-{\small{0.0740}} &
-{\small{0.04}} &
-{\small{0.0531}} &
-{\small{0.0528}}\tabularnewline
+{\small{0.0739}} &
+{\small{0.02}} &
+{\small{0.0533}} &
+{\small{0.0530}}\tabularnewline
 \hline 
@@
 \caption{Coverage and length of prediction intervals for the CAR and SPDE spatial
-models when holding out 80\% of entire cells of data in Minnesota. }
+models when holding out 80\% of entire cells of data in Minnesota.
+Coverage values near 0.95 are optimal, while shorter intervals are
+better.}
 
@@
 {\small{0.981}} &
-{\small{0.971}}\tabularnewline
+{\small{0.972}}\tabularnewline
 \hline 
@@
 {\small{0.028}} &
-{\small{0.021}}\tabularnewline
+{\small{0.022}}\tabularnewline
 \hline 
@@
 
-\subsubsection{Individual tree hold-out experiment}
-
-Table \ref{tab:score_tree} shows the predictive ability for Experiment
-2. Here we have evidence (posterior probability of 0.94) that the
-SPDE model is better based on the Brier score.
+\paragraph{Individual tree hold-out experiment}
+
+Table \ref{tab:score_tree} shows the predictive ability based on
+Experiment 2. Here we have evidence (posterior probability of 0.93)
+that the SPDE model is better based on the Brier score, but the Brier
+score values for the two models are numerically almost the same.
 
 \begin{table}
-\caption{Predictive ability based on several predictive score criteria for
+\caption{Predictive ability based on several predictive metric criteria for
 the CAR and SPDE spatial models when holding out 5\% of trees. Smaller
@@
 {\small{Brier}} &
-{\small{21296}} &
-{\small{21277}} &
-{\small{0.06}} &
-{\small{21146}} &
-{\small{21143}}\tabularnewline
+{\small{0.662}} &
+{\small{0.661}} &
+{\small{0.07}} &
+{\small{0.657}} &
+{\small{0.657}}\tabularnewline
 \hline 
 {\small{Negative Log Density}} &
-{\small{52199}} &
-{\small{52052}} &
+{\small{51757}} &
+{\small{51626}} &
 {\small{0.01}} &
-{\small{51141}} &
-{\small{51156}}\tabularnewline
+{\small{50705}} &
+{\small{50736}}\tabularnewline
 \hline 
@@
 
-\subsubsection{Choice of spatial model}
+\paragraph{Choice of spatial model}
 
@@
 
-\section{Data product}
+\section{Data product\label{sec:Data-product}}
 
@@
 for 150,000 iterations, discarding the first 25,000 iterations for
-burn-in and retaining 250 subsampled iterations to limit storage needs
-and limit the size of the data product. Mixing was generally reasonable,
-but for some of the hyperparameters was relatively slow, particularly
-for less common taxa. Despite this, mixing for the variables of substantive
-interest, the proportions was good. {[}Report ESS values{]}
-
-In Figure \ref{fig:select_maps}, we show maps of estimated composition
+burn-in and retaining 250 subsampled iterations (subsampled to limit
+storage needs and the size of the data product). Mixing was generally
+reasonable, but for some of the hyperparameters was relatively slow,
+particularly for less common taxa. Despite this, mixing for the variables
+of substantive interest -- the proportions -- was good.
+
+In Fig. \ref{fig:select_maps}, we show maps of estimated composition
 for the full domain for several taxa of substantive interest to illustrate
-the results: beech, cherry, chestnut, elm, hemlock, oak, and pine.
-These maps contrast the raw data proportions, the posterior means
-and posterior standard deviations as pointwise estimates of uncertainty.
-Fig. \ref{fig:all_predictions} shows posterior means for all 23 taxa.
+the results. These maps contrast the raw data proportions, the posterior
+means, and posterior standard deviations as pointwise estimates of
+uncertainty. Fig. \ref{fig:all_predictions} shows posterior means
+for all 23 taxa.
 
 \begin{figure}
-\label{fig:select_maps}
-
 \includegraphics[scale=0.75]{1_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig2.pdf}
@@
 form of posterior standard deviations -- representing standard errors
-of prediction (column 3) for select taxa.}
+of prediction (column 3) for select taxa. \label{fig:select_maps}}
 \end{figure}
@@
 \begin{figure}
-\label{fig:all_predictions}
-
-\includegraphics[scale=0.9]{5_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig3.pdf}
-
-\caption{Predictions (posterior means) for all taxa over the entire domain.}
+\includegraphics[scale=0.8]{5_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig3.pdf}
+
+\caption{Predictions (posterior means) for all taxa over the entire domain.\label{fig:all_predictions}}
 
@@
 
-The data product is publicly available at XX (insert accession number)
-under YY license as version 0.3 as of April 2015. The product is in
-the form of a netCDF-4 file, with dimensions x, y, and MCMC iteration.
-There is one variable per taxon. The PalEON project will continue
+The data product is publicly available at the NIS Data Portal (accession
+number to be inserted when available) under YY license as version
+0.4 as of August 2015. The product is in the form of a netCDF-4 file,
+with dimensions x, y, and MCMC iteration. There is one variable per
+taxon. The PalEON project (in particular the first author) will continue
 to maintain this product, releasing new versions as additional data
-in Illinois, Indiana and Ohio are digitized.
-
-{[}comment on stability of hosting once get better feel for this{]}
-
-\section{Discussion}
-
-Note that digitization of data from Illinois and Indiana is ongoing,
-and digitization of additional data from Ohio is planned as well.
-As a result, at some point we expect to have complete data for the
-western half of the domain. There will remain some missing townships
-in the eastern half of the domain.
+in Illinois, Indiana and Ohio are digitized. Note that digitization
+of data from Illinois and Indiana is ongoing, and digitization of
+additional data from Ohio is planned as well. As a result, at some
+point we expect to have complete data for the western half of the
+domain. There will remain some missing townships in the eastern half
+of the domain.
+
+\section{Discussion\label{sec:Discussion}}
 
@@
 smooths as a function of Euclidean distance, does not account for
-topographic, soil, or other features.
-
-Other related data products that are under development include:
+topographic, soil, or other features. An additional drawback of the
+product is its focus on composition, which does not directly tell
+us about vegetation structure, in particular does not distinguish
+between closed forest, savanna, and prairie, of particular note in
+Minnesota, Wisconsin, Illinois, and into Indiana. The statistically-estimated
+biomass product mentioned next will directly inform questions about
+vegetation structure.
+
+In addition to the existing raw composition count data product discussed
+in Section \ref{sec:Data}, we have related data products that are
+under development:
 \begin{itemize}
-\item The gridded raw count data for the western subdomain and data aggregated
-to township for the eastern subdomain. The raw data product is available
-at XXX and that product provides the input data for the product described
-in this work. Note that we cannot provide the raw data at point locations
-because of limitations specified in the data use agreements under
-which we have access to the data.
 \item Gridded raw biomass estimates for Minnesota, Wisconsin, and Michigan
@@
 \end{itemize}
-An additional drawback of the product is its focus on composition,
-which does not directly tell us about vegetation structure, in particular
-does not distinguish between closed forest, savanna, and prairie,
-of particular note in Minnesota, Wisconsin, Illinois, and into Indiana.
-The statistically-estimated biomass product mentioned just above will
-directly inform questions about vegetation structure. Extensions of
-that product will also estimate basal area and stem density.
+We also plan to estimate basal area and stem density using a similar
+approach to that used for biomass.
+
+\section*{Author contributions}
+
+CJP, AT, and JZ developed the statistical model and code. CJP carried
+out the model comparison and created the data product. CJP wrote the
+paper based on an initial draft by AT and with feedback from SJG,
+CVC, JWW, DJM, JAP, JZ and JSM. CVC, DJM, SJG, JWW, and JAP provided
+data.
 
@@
 years who have preserved, collected, and digitized survey records,
-in particular Jim Dyer, Peter Marks, Robert McIntosh, and Ed Schools
-\{David/Simon please add names here\}\{Jason/Jody, should we mention
-Bowles and Stuckey; if so I need full names\}. We thank Madeline Ruid,
-Ben Seliger, Morgan Ross and Daniel Handel for processing of the southern
-Michigan data. Indiana and Illinois data were made possible through
-the hard work of many Notre Dame undergraduates in the McLachlan lab.
-This work was carried out by the PalEON Project with support from
-the National Science Foundation MacroSystems Program through grants
-EF-1065656, EF-1241868, DEB-1241874 and DEB-1241868 and from the Notre
-Dame Environmental Change Initiative.
-
-%\bibliographystyle{/accounts/gen/vis/paciorek/latex/RSSstylefile/Chicago}
-\bibliographystyle{/accounts/gen/vis/paciorek/latex/PLOS_ONE/plos215}
+in particular Jim Dyer, Peter Marks, Robert McIntosh, Ronald Stuckey,
+Ed Schools, and the Ohio Biological Survey \{David/Simon please add
+names here\} We thank Madeline Ruid, Ben Seliger, Morgan Ross and
+Daniel Handel for processing of the southern Michigan data. Indiana
+and Illinois data were made possible through the hard work of many
+Notre Dame undergraduates in the McLachlan lab. This work was carried
+out by the PalEON Project with support from the National Science Foundation
+MacroSystems Program through grants EF-1065656, EF-1241868, DEB-1241874
+and DEB-1241868 and from the Notre Dame Environmental Change Initiative.
+
+\bibliographystyle{/accounts/gen/vis/paciorek/latex/RSSstylefile/Chicago}
 
@@
 
-Define $\bar{w}_{i\cdot p}=n_{i}^{-1}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
+Define $\bar{w}_{ip}=\frac{1}{n_{i}}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
 be the average of the $W$ values for the $p$th taxon in the $i$th
-grid cell. Let $A$ be a diagonal matrix where $A_{ii}$ is the number
-of trees in the $i$th grid cell. When there are no trees in a grid
-cell, $\bar{w}_{i\cdot p}=0$ and $A_{ii}=0$. For the township data,
-at each iteration, based on the current values of the grid cell membership
-variables, $c_{tj}$, trees are aggregated into grid cells and the
-calculations above can then be carried out.
+grid cell and $\bar{w}_{p}=\{\bar{w}_{ip}\},$ $i=1,\ldots,m$. Let
+$A$ be a diagonal matrix where $A_{ii}$ is the number of trees in
+the $i$th grid cell. When there are no trees in a grid cell, $\bar{w}_{ip}=0$
+and $A_{ii}=0$. For the township data, at each iteration, based on
+the current values of the grid cell membership variables, $c_{tj}$,
+trees are aggregated into grid cells and the calculations above can
+then be carried out.
 
@@
 where $Q_{p}=(\sigma_{p}^{2})^{-1}Q$ for the CAR model and $\left(\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}\right)^{-1}Q(\rho_{p})$
-for the SPDE model. For each hyperparameter, $\phi\in\{\{\mu_{p}\},\{\log\sigma_{p},\rho_{p}\}\}$,
-we sampled $\{\phi,\alpha_{p}\}$ jointly, proposing $\phi$ as a
-random walk and, conditional on the proposed value of $\phi$, sampling
-$\alpha_{p}$ from the distribution just above. The joint proposal
-is accepted or rejected as a standard Metropolis-Hastings proposal,
-with adaptation of the proposal (co)variance \citep{Shab:Well:2011}.
-As mentioned previously, for the SPDE model, we jointly proposed $\phi=\{\log\sigma_{p},\rho_{p}\}$
-from a bivariate normal distribution to account for the posterior
-dependence of these parameters, while for the CAR model, $\phi\in\{\{\log\sigma_{p}\}\}$.
+for the SPDE model. For each hyperparameter, $\phi_{p}=\log\sigma_{p}$
+for the CAR model and $\phi_{p}\in\{\mu_{p},\{\log\sigma_{p},\log\rho_{p}\}\}$
+for the SPDE model, we sampled $\{\phi_{p},\alpha_{p}\}$ jointly,
+proposing $\phi_{p}$ as a random walk and, conditional on the proposed
+value of $\phi_{p}$, sampling $\alpha_{p}$ from the distribution
+just above. The joint proposal is accepted or rejected as a standard
+Metropolis-Hastings proposal, with adaptation of the proposal (co)variance
+\citep{Shab:Well:2011}. The proposal distribution for $\phi_{p}$
+is a normal distribution (bivariate for $\phi_{p}=(\log\sigma_{p},\log\rho_{p})$).
 
@@
 we draw the latent tree membership variable, $c_{tj}\in\{1,\ldots,m\}$,
-from a discrete distribution by normalizing posterior weights, $\{\psi_{1}L_{1},\ldots,\psi_{m}L_{m}\}$,
-produced by multiplying the prior weights by a likelihood contribution.
-For cell $k$, $L_{k}$ is the density of the latent $W_{tj1},\ldots,W_{tjP}$
-values for the given tree under the condition that $c_{tj}=k$, namely
-the product of independent normal densities, $W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{k}),1)$,
-over $p=1,\ldots,P$. Thus the posterior reweights the prior based
-on how consistent the current $W$ values for a tree are with the
-$\alpha$ values for the candidate grid cells.
+from a discrete distribution by normalizing posterior weights, $\{\psi_{1}L_{tj1},\ldots,\psi_{m}L_{tjm}\}$,
+produced by multiplying the prior weights by a likelihood contribution,
+$L_{tji}$, $i=1,\ldots,m$. $L_{tji}$ is the density of the latent
+$W_{tj1},\ldots,W_{tjP}$ values for the given tree under the condition
+that $c_{tj}=i$, namely the product of independent normal densities,
+$W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{i}),1)$, over $p=1,\ldots,P$.
+Thus the posterior reweights the prior based on how consistent the
+current $W_{tj}$ values for a tree are with the $\alpha$ values
+for the candidate grid cells.
 
@@
 In the latent variable representation, $\theta_{p}(s)$ never appears
-explicitly and cannot be calculated in close form. Instead we use
+explicitly and cannot be calculated in closed form. Instead we use
 Monte Carlo integration over $W_{ijp},\, p=1,\ldots,P$ to estimate
@@
 For each of the saved MCMC samples, $k=1,\ldots K$, we estimate $\theta_{p}^{(k)}(s_{i})$
-numerically. Specifically, for $t=1,\ldots,10000$, we independently
-draw
+numerically. Specifically, for $t=1,\ldots,10000$ samples (i.e.,
+hypothetical trees), we independently draw
 \[
@@
 \]
-and estimate 
+and estimate using 
 \[
-\theta_{p}^{(k)}(s_{i})\approx\frac{1}{10000}{\displaystyle \sum_{t=1}^{10000}1(W_{itp}^{(k)}={\displaystyle \max_{p^{*}}W_{itp^{*}}^{(k)})}}
+\theta_{p}^{(k)}(s_{i})\approx\frac{1}{10000}{\displaystyle \sum_{t=1}^{10000}1(W_{itp}^{(k)}={\displaystyle \max_{p^{*}}W_{itp^{*}}^{(k)})}},\, p=1,\ldots,P
 \]

### 2015-08-20 paper.R

@@ -1,8 +1,8 @@
 # code to generate figures for paper
+# run this code from ~/research/jmac/composition so use packrat controlled packages
 
-paperDir <- getwd()
+paperDir <- '~/research/jmac/composition_paper'
 
-setwd('~/research/jmac/composition')
 source("config")
 
 source(file.path(codeDir, "plot.R"))
@@ -15,6 +15,7 @@ require(maptools)
 require(rgdal)
 require(raster)
 require(gridExtra)
+
 
 # for distinguishing no trees from no data
 west_presence <- read.csv(file.path(dataDir, 'western.csv'))

### 2015-09-05 arxiv/paperArxiv.tex
added arxiv and plos versions of tex

--- 2015-08-20 paper.tex
@@
-\batchmode
-\makeatletter
-\def\input@path{{/accounts/gen/vis/paciorek/research/jmac/composition_paper//}}
-\makeatother
-\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[12pt]{article}
+\usepackage[]{graphicx}
+\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
 %% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+%\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+%\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+%\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+%\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+%\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+%\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+%\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+%\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+%\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+%\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+%\usepackage{framed}
+%\makeatletter
+%\newenvironment{kframe}{%
+% \def\at@end@of@kframe{}%
+% \ifinner\ifhmode%
+%  \def\at@end@of@kframe{\end{minipage}}%
+%  \begin{minipage}{\columnwidth}%
+% \fi\fi%
+% \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+% \colorbox{shadecolor}{##1}\hskip-\fboxsep
+%     % There is no \\@totalrightmargin, so:
+%     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+% \MakeFramed {\advance\hsize-\width
+%   \@totalleftmargin\z@ \linewidth\hsize
+%   \@setminipage}}%
+% {\par\unskip\endMakeFramed%
+% \at@end@of@kframe}
+%\makeatother
+
+%\definecolor{shadecolor}{rgb}{.97, .97, .97}
+%\definecolor{messagecolor}{rgb}{0, 0, 0}
+%\definecolor{warningcolor}{rgb}{1, 0, 1}
+%\definecolor{errorcolor}{rgb}{1, 0, 0}
+%\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
+\newcommand{\lyxaddress}[1]{
+\par {\raggedright #1
+\vspace{1.4em}
+\noindent\par}
+}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
@@
 \makeatother
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
 \begin{document}
@@
 States \\
-at the time of European settlement}
+at the time of Euro-American settlement}
+
+\author{Christopher J. Paciorek\textsuperscript{1,{*}}, Simon J. Goring\textsuperscript{2,\&},
+Andrew L. Thurman\textsuperscript{3,\&}, \\
+Charles V. Cogbill\textsuperscript{4}, John W. Williams\textsuperscript{2,3},
+David J. Mladenoff\textsuperscript{6}, \\
+Jody A. Peters\textsuperscript{7}, Jun Zhu\textsuperscript{8}, and
+Jason S. McLachlan\textsuperscript{7}}
+
 \maketitle
-\begin{quote}
-Christopher J. Paciorek 1{*}, Simon J. Goring 2\&, Andrew Thurman
-3\&, Charles V. Cogbill 4, \\
-John W. Williams 3,5, David J. Mladenoff 6, Jody A. Peters 7, Jun
-Zhu 8, \\
-and Jason S. McLachlan 7 \\
-{[}author order to be discussed{]}\\
-{[}please tell me if you do/don't want your middle initial{]}
-
-{[}let me know if there are any issues with how I've listed your affiliation{]}
-
-1 Department of Statistics, University of California, Berkeley, California,
-USA
-
-2 Department of Geography, University of Wisconsin, Madison, Wisconsin,
-USA
-
-3 Department of Statistics, University of Iowa, Iowa City, Iowa, USA
-
-4 Harvard Forest, Harvard University, Petersham, Massachusetts, USA
-
-5 Center for Climatic Research, University of Wisconsin, Madison,
-Wisconsin, USA
-
-6 Department of Forestry and Wildlife Ecology, University of Wisconsin,
-Madison, Wisconsin, USA
-
-7 Department of Biological Sciences, University of Notre Dame, Notre
-Dame, Indiana, USA
-
-8 Department of Statistics, University of Wisconsin, Madison, Wisconsin,
-USA
-
-{*} Corresponding author
-
-E-mail paciorek@stat.berkeley.edu (CJP)
-
-\& These authors contributed equally to this work.
-
-Short title: Settlement-era northeast U.S. tree composition
-\end{quote}
+
+\lyxaddress{\textsuperscript{1}Department of Statistics, University of California, Berkeley, California, USA}
+
+\lyxaddress{\textsuperscript{2}Department of Geography, University of Wisconsin, Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{3}Department of Statistics, University of Iowa, Iowa City, Iowa, USA}
+
+\lyxaddress{\textsuperscript{4}Harvard Forest, Harvard University, Petersham, Massachusetts, USA}
+
+\lyxaddress{\textsuperscript{5}Center for Climatic Research, University of Wisconsin,
+Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{6}Department of Forestry and Wildlife Ecology,
+University of Wisconsin, Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{7}Department of Biological Sciences, University
+of Notre Dame, Notre Dame, Indiana, USA}
+
+\lyxaddress{\textsuperscript{8}Department of Statistics, University of Wisconsin,
+Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{{*}}Corresponding author; E-mail: paciorek@stat.berkeley.edu}
+
+\lyxaddress{\textsuperscript{\&}These authors contributed equally to this work.
+\newpage}
 
 \begin{abstract}
-We present a data product of the estimated composition of tree taxa
-at the time of European settlement of the northeastern United States
-and the statistical methodology used to produce the product. Composition
+We present a gridded 8 km-resolution data product of the estimated
+composition of tree taxa at the time of Euro-American settlement of
+the northeastern United States and the statistical methodology used
+to produce the product from trees recorded by land surveyors. Composition
 is defined as the proportion of stems larger than approximately 20
-cm diameter at breast height in 22 taxonomic groupings, generally
-at the genus level. The data come from settlement survey records that
-provide raw data that are transcribed and then aggregated spatially,
-giving count data. The domain is divided into two regions, eastern
-(Maine to Ohio) and western (Indiana to Minnesota). Public Land Survey
-point data in the western region are aggregated to a regular 8 km
-grid, while data in the eastern region, from Town Proprietor Surveys,
-are aggregated at the township level in irregularly-shaped local administrative
-units. The product is based on a Bayesian statistical model fit to
-the count data that estimates composition on a regular 8 km grid.
-The statistical model allows us to estimate composition at locations
-with no data and to smooth over noise caused by limited counts in
-locations with data. Critically, it also allows us to quantify uncertainty
-in our composition estimates. We expect this data product to be useful
+cm diameter at breast height for 22 tree taxa, generally at the genus
+level. The data come from settlement-era public survey records that
+are transcribed and then aggregated spatially, giving count data.
+The domain is divided into two regions, eastern (Maine to Ohio) and
+midwestern (Indiana to Minnesota). Public Land Survey point data in
+the midwestern region (ca. 0.8-km resolution) are aggregated to a
+regular 8 km grid, while data in the eastern region, from Town Proprietor
+Surveys, are aggregated at the township level in irregularly-shaped
+local administrative units. The product is based on a Bayesian statistical
+model fit to the count data that estimates composition on a regular
+8 km grid across the entire domain. The statistical model is designed
+to handle data from both the regular grid and the irregularly-sized
+townships and allows us to estimate composition at locations with
+no data and to smooth over noise caused by limited counts in locations
+with data. Critically, the model also allows us to quantify uncertainty
+in our composition estimates, making the product suitable for applications
+employing data assimilation. We expect this data product to be useful
 for understanding the state of vegetation in the northeastern United
-States prior to large-scale European settlement. In addition to specific
-regional questions, the data product can also serve as a baseline
-against which to investigate how forests and ecosystems change after
-intensive settlement.
+States prior to large-scale Euro-American settlement. In addition
+to specific regional questions, the data product can also serve as
+a baseline against which to investigate how forests and ecosystems
+change after intensive settlement. The data product is being made
+available at the NIS data portal as version 0.4 under the CC BY license.
+
+Keywords: biogeography, species composition, old-growth forests, spatial
+modeling, Bayesian statistical model, vegetation mapping
 \end{abstract}
@@
 conservation management, to understand ecosystem processes at decadal
-and centennial scales, and, particularly in regions with widespread
-land use change, to understand the extent to which forests after conversion
-and regeneration differ from the original forest cover.
+and centennial scales, to track forest responses to shifting climates,
+and, particularly in regions with widespread land use change, to understand
+the extent to which forests after conversion and regeneration differ
+from the original forest cover.
 
 Euro-American settlement and subsequent land use change occurred in
-a time-transient fashion across North America, and land surveys that
-provide vegetation information followed a similar pattern. Early surveys
-(from 1620 until 1825) in the northeastern United States provide areally-aggregated
-data at the township level \citep{cogbill2002forests,thompson2013four};
-we refer to these as the Town Proprietor Survey (TPS). Later surveys,
-after the establishment of the U.S. Public Land Survey System (PLS)
-provide point-level data along a regular grid, with one mile spacing
-during the period 1832 to 1907 \citep{bourdo1956review,schulte2001original,goring2015composition}.
+a time-transient fashion across North America and were accompanied
+by land surveys needed to demarcate land for land tenure and use.
+Various systems were used by surveyors to locate legal boundary markers,
+usually by recording and marking trees adjacent to survey markers.
+These data provide vegetation information that can be mapped and used
+quantitatively to represent the period of settlement. Early surveys
+(from 1620 until 1825) in the northeastern United States provide spatially-aggregated
+data at the township level \citep{Cogb:etal:2002,thompson2013four},
+with typical township size on the order of 200 $\mbox{km}^{2}$ and
+no information about the locations of individual trees; we refer to
+these as the Town Proprietor Survey (TPS). Later surveys after the
+establishment of the U.S. Public Land Survey System (PLS) by the General
+Land Office (GLO) provide point-level data along a regular grid, with
+one-half mile (800 m) spacing, for Ohio and westward during the period
+1785 to 1907 \citep{bourdo1956review,pattison1957beginnings,schulte2001original,goring2015composition}.
+At each point 2-4 trees were identified, and the common name, diameter
+at breast height, and distance and bearing from the point were recorded.
 Survey instructions during the PLS varied through time and by point
-type. Accounting for this requires the application of spatially-varying
+type. Accounting for this variation requires data screening to maximize
+consistency among points and the application of spatially-varying
 correction factors \citep{cogbill2015corrections,goring2015composition}
-to accurately assess stem density, basal area and biomass from the
-early settlement records, but the impact on composition estimates
-is limited. Both types of surveys provide composition of major tree
-species but using sometimes inconsistent common names.
+to accurately assess tree stem density, basal area and biomass from
+the early settlement records, but the impact on composition estimates
+is limited \citep{liu2011broadscale}. Surveyors sometimes used ambiguous
+common names, which requires matching to scientific names and standardization
+\citep{mladenoff2002narrowing,goring2015composition}.
 
@@
 or smaller domains \citep{friedman2005regional,rhemtulla2009historical}
-and with varying spatial resolution, from townships (36 square miles)
+and with various spatial resolutions, from townships (36 square miles)
 to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
-provide a dataset for the upper Midwest that is resolved to an 8 x
-8 km grid size, providing broad spatial coverage at a spatial scale
-that can be compared to the modern forests using Forest Inventory
+provide a new dataset of forest composition, biomass, and stem density
+based on PLS data for the upper Midwest that is resolved to an 8 km
+by 8 km grid cell scale, providing broad spatial coverage at a spatial
+scale that can be compared to modern forests using Forest Inventory
 and Analysis products \citep{gray2012forest}. Combined with additional,
-coarsely-sampled PLS data from Illinois and Indiana and with the TPS
-data, this gives us raw data for much of the northeastern United States.
-However, there are several limitations of using the raw data that
-can be alleviated by the use of a statistical model to develop a statistically-estimated
-data product. First, the PLS and TPS data provide only estimates of
+coarsely-sampled PLS data from Illinois and Indiana, newly-digitized
+data from southern Michigan, and with the TPS data, this gives us
+raw data for much of the northeastern United States. However, there
+are several limitations of using the raw data that can be alleviated
+by the use of a statistical model to develop a statistically-estimated
+data product. First, the PLS and TPS data only provide estimates of
 within-cell variance that do not account for information from nearby
-locations. Second, the available digitized data from Illinois and
-Indiana represent a small fraction of those states and missing townships
-are common in the TPS data. Our statistical model allows us to provide
-a spatially-complete data product of settlement-era tree composition
-with uncertainty across the northeastern U.S.
+locations. Second, there are data gaps: the available digitized data
+from Illinois and Indiana represent a small fraction of those states
+and missing townships are common in the TPS data. Third, the TPS and
+PLS data have fundamentally different sampling design and spatial
+resolution. Our statistical model allows us to provide a spatially-complete
+data product of settlement-era tree composition for a common 8 km
+grid with uncertainty across the northeastern U.S.
 
@@
 
-The raw data were obtained from survey records collated from across
-the northeastern U.S. by a number of researchers. For the states of
-Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
-subdomain), data are available at PLS survey point locations and have
-been aggregated to a regular 8 km grid in the Albers projection. (Note
-that for Indiana and Illinois, at the moment trees are associated
-with township centroids and then assigned to 8 km grid cells based
-on the centroid but in the near future we will have point locations
-available for each tree.) For the states of Ohio, Pennsylvania, New
-Jersey, New York and the six New England states (the eastern subdomain),
-data are aggregated at the township level. There are also data from
-a single township in Quebec and a single township in northern Delaware.
-Data are essentially complete in Minnesota, Wisconsin and Michigan,
-but data in Illinois and Indiana represent a sample of the full set
-of grid cells, with survey record transcription ongoing. Data for
-the remaining states are available for a subset of the full set of
-townships covering the domain. Fig. \ref{fig:Spatial-domain} shows
-the domain, indicating the grid cells and townships with data.
+The raw data were obtained from land division survey records collated
+and digitized from across the northeastern U.S. by a number of researchers
+(Fig. \ref{fig:Spatial-domain}). For the states of Minnesota, Wisconsin,
+Illinois, Indiana, and Michigan (the midwestern subdomain), digitized
+data are available at PLS survey point locations and have been aggregated
+to a regular 8 km grid in the Albers projection. (Note that for Indiana
+and Illinois, at the moment trees are associated with township centroids
+and then assigned to 8 km grid cells based on the centroid but in
+the near future we will have point locations available for each tree.)
+For the states of Ohio, Pennsylvania, New Jersey, New York and the
+six New England states (the eastern subdomain), data are aggregated
+at the township level. There are also data from a single township
+in Quebec and a single township in northern Delaware. Digitization
+of PLS data in Minnesota, Wisconsin and Michigan is essentially complete,
+with PLS data for nearly all 8 km grid cells, but data in Illinois
+and Indiana represent a sample of the full set of grid cells, with
+survey record transcription ongoing. Data for the eastern states are
+available for a subset of the full set of townships covering the domain;
+the TPS data for some townships were lost, incomplete, or have not
+been located \citep{Cogb:etal:2002}.
 
 \begin{figure}
-\label{fig:domain}\includegraphics[scale=1.3]{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}
+\label{fig:domain}\includegraphics[scale=1.3]{fig1}
 
 \caption{Spatial domain of the northeastern United States, with locations with
-data shown in gray. Locations are grid cells in western portion and
-townships in eastern portion. In addition to locations without data
-being indicated in white, grid cells completely covered in water are
-white (e.g., a few locations in the northwestern portion of the domain
-in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
+data shown in gray. Locations are grid cells in midwestern portion
+and townships in eastern portion. In addition to locations without
+data being indicated in white, grid cells completely covered in water
+are white (e.g., a few locations in the northwestern portion of the
+domain in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
 \end{figure}
@@
 European colonists (before U.S. independence) and the United States
-settled what is now the northeastern United States. Our estimates
-are for the period of settlement represented by the survey data and
-therefore are time-transgressive; they do not represent any single
-point in time across the domain, but rather the state of the landscape
-at the time just before settlement occurred \citep{Whit:1996,Cogb:etal:2002}.
+settled what is now the northeastern and midwestern United States.
+Our estimates are for the period of settlement represented by the
+survey data and therefore are time-transgressive; they do not represent
+any single point in time across the domain, but rather the state of
+the landscape at the time just prior to widespread Euro-American settlement
+and land use \citep{Whit:1996,Cogb:etal:2002}. These forest composition
+datasets do include the effects of Native American land use and early
+Euro-American settlement activities \citep[e.g.,][]{Blac:etal:2006},
+but it is likely that the imprint of this earlier land use is highly
+concentrated rather than spatially extensive \citep{munoz2014defining}.
 
 Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
-data and processing steps are available in \cite{Gori:etal:2015}.
-The lower Midwest (Illinois, Indiana) data were purchased from the
-Indiana State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
+data and processing steps are available in \cite{goring2015composition};
+key elements include the use of only corner points, the use of only
+the two closest trees at each corner point, spatially-varying correction
+factors for sampling effort, and a standardized taxonomy table. The
+lower Midwest (Illinois, Indiana) data were purchased from the Indiana
+State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
 Illinois) and processed using similar steps as for the upper Midwest
-data. The TPS data were compiled by C.V. Cogbill from a myriad of
-archival sources representing land division surveys conducted in connection
-with local settlement. The complete raw vegetation data are available
-through the NIS Data Portal (accession number and version to be inserted
-when available). The aggregation into taxonomic groups is primarily
-at the genus level, but is at the species level in some cases of monospecific
-genera. We model the following 22 taxa plus an ``other hardwood''
-category: Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash
-(\emph{Fraxinus spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus
-grandifolia)}, Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa
-sylvatica} and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
+data. Digitization of the Illinois and Indiana data is still underway,
+so many grid cells contained no data at the time the statistical model
+was fit. Note that the number of trees per grid cell varies depending
+on the number of survey points in a cell, with an average of 124 trees
+per cell. The gridded data at the 8 km resolution for the upper Midwest
+are available through the NIS Data portal (accession number and version
+to be inserted when available). The TPS data were compiled by C.V.
+Cogbill from a myriad of archival sources representing land division
+surveys conducted in connection with local settlement.
+
+The aggregation into taxonomic groups is primarily at the genus level
+but is at the species level in some cases of monospecific genera.
+We model the following 22 taxa plus an ``other hardwood'' category:
+Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash (\emph{Fraxinus
+spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus grandifolia)},
+Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa sylvatica}
+and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
 virginiana} and \emph{Thuja occidentalis}), Cherry (\emph{Prunus spp.}),
 Chestnut (\emph{Castanea dentata}), Dogwood (\emph{Cornus spp.}),
-Elm (\emph{Ulmus spp.}), Fir (\emph{Abies spp.}), Hemlock (\emph{Tsuga
+Elm (\emph{Ulmus spp.}), Fir (\emph{Abies balsamea}), Hemlock (\emph{Tsuga
 canadensis}), Hickory (\emph{Carya spp.}), Ironwood (\emph{Carpinus
@@
 Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), and
-Walnut (\emph{Juglans nigra}). Note that in several cases (black gum/sweet
+Walnut (\emph{Juglans spp.}). Note that in several cases (black gum/sweet
 gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
 in the common tree names used by surveyors, a group represents trees
-from different families and even orders. For the western subdomain
+from different genera or even families. For the midwestern subdomain
 we do not fit statistical models for Atlantic white cedar and chestnut
-as these have 0 and 7 trees present, respectively. The taxa grouped
-into the other hardwood category are those for which fewer than roughly
-2000 trees were present in the dataset; however we include Atlantic
-white cedar explicitly despite it only having 336 trees in the dataset.
-Although surveyors avoided using small trees, there was no consistent
-lower diameter limit. The data, however, roughly represent trees greater
-than 8 inches (\textasciitilde{}20 cm) diameter at breast height (dbh),
-but with some trees as small as 3 inches dbh. In Indiana and Illinois
-there are also 402 trees and 29 trees, respectively, that are between
-1 and 4 inches dbh, but these will be excluded in the next version
-of the product {[}waiting on Simon to see if he stripped them out{]}.
-
-There are approximately 860,000 trees from the western subdomain and
-420,000 trees from the eastern subdomain. In the western subdomain,
+as these species have 0 and 7 trees present, respectively. The taxa
+grouped into the other hardwood category are those for which fewer
+than roughly 2000 trees were present in the dataset; however we include
+Atlantic white cedar explicitly despite it only having 336 trees in
+the dataset because of specific ecological interest in Atlantic white
+cedar wetlands.
+
+Diameters are only recorded in the PLS data. Although surveyors avoided
+using small trees, there was no consistent lower diameter limit. The
+PLS data generally represent trees greater than 8 inches (\textasciitilde{}20
+cm) diameter at breast height (dbh), but with some trees as small
+as 1 inch dbh (smaller trees were much more common in far northern
+Minnesota). TPS data have no information about dbh, but the trees
+were large enough to blaze and are presumed to be relatively large
+trees useful for marking property boundaries.
+
+There are approximately 860,000 trees in the midwestern subdomain
+and 420,000 trees in the eastern subdomain. In the midwestern subdomain,
 oak is the most common taxon and pine the second most common, while
@@
 
-Our domain is a rectangle covering all of the states, in the Albers
-projection (NAD 1983 Albers Great Lakes and St. Lawrence), with the
-rectangle split into 8 km cells, arranged in a 296 by 180 grid of
-cells, with the centroid of the cell in the southwest corner located
-at (-71000, 58000). For the modeling of the western subdomain we use
-the western-most 146 by 180 grid of cells. For the modeling of the
-eastern subdomain we use the eastern-most 180 by 180 grid of cells
-and then omit 23 cells in the north and 17 cells in the south outside
-the states containing data.
+Our domain is a rectangle covering all of the states using a metric
+Albers (Great Lakes and St. Lawrence) projection (PROJ4: EPSG:3175),
+with the rectangle split into 8 km cells, arranged in a 296 by 180
+grid of cells, with the centroid of the cell in the southwest corner
+located at (-71000 m, 58000 m). For the midwestern subdomain we use
+the western-most 146 by 180 grid of cells. For the eastern subdomain
+we use the eastern-most 180 by 180 grid of cells and then omit 23
+rows of cells in the north and 17 rows of cells in the south, as these
+grid cells are outside of the states containing data.
 
@@
 the empirical proportions represent estimates of the true proportions
-that could be calculated using full population of all the trees in
-an areal region.
+that could be calculated using the full population of all the trees
+in grid cell or township.
 \end{enumerate}
@@
 cells. These samples are the data product (described further in the
-Data Product section) and can then be used in subsequent analyses.
-The mean and standard deviation of the samples for each pair of cell
-and taxon represent our best estimate (i.e., prediction) of composition
+Data Product section) and can be used in subsequent analyses. The
+mean and standard deviation of the samples for each pair of cell and
+taxon represent our best estimate (i.e., prediction) of composition
 and a Bayesian ``standard error'' quantifying the uncertainty in
@@
 is the vector of unknown proportions for those taxa at that cell.
-Note that we use a standard multinomial distribution without overdispersion
-as the set of trees in the dataset is roughly uniformly sampled across
-the cells or townships \citep{Gori:etal:2015}.
+Note that we use a standard multinomial distribution without overdispersion,
+because the set of trees in the dataset is roughly uniformly sampled
+across the cells or townships \citep{goring2015composition}.
 
@@
 allows for interpolation to locations with no data. Apart from the
-sum to one constraints, the taxa are considered to be independent
-in the prior. We did not want to impose any structure that ties the
-different taxa together, as any correlation will likely vary across
-space.
+sum-to-one constraint, the taxa are considered to be independent in
+the prior. We did not want to impose any structure that ties the different
+taxa together, as any correlation will likely vary across space.
 
@@
 structure of the $\alpha_{p}(s)$ processes, a standard conditional
-autoregressive model \citep{Bane:etal:2003} and a Gaussian Markov
+autoregressive model \citep{Bane:etal:2004} and a Gaussian Markov
 random field (MRF) approximation to a Gaussian process with Matern
@@
 space of $\alpha_{p}$. The latent variable representation of Section
-\ref{sub:Model-for-township} helps to alleviate this problem.
+\ref{sub:Latent-Variable-Model} helps to alleviate this problem.
+Next we describe two alternative spatial models that we considered;
+in Section \ref{sec:Model-comparison}, we use cross-validation to
+choose between these two representations.
 
@@
 Our first model is a standard conditional autoregressive (CAR) model
-\citep{Bane:etal:2003}. We use a standard form of this model, which
+\citep{Bane:etal:2004}. We use a standard form of this model, which
 treats the four cardinal neighbors of each grid cell as the neighbors
@@
 matrix defining the neighborhood relation of the locations; that is
-$C{}_{ik}=1$ if locations $i$ and $k$ are neighbors. The matrix
-$D$ is an $m\times m$ diagonal matrix containing the row sums of
-matrix $C$ as the diagonal entries $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$
+$C{}_{ik}=1$ if locations $i$ and $k$ are neighbors and zero otherwise.
+The matrix $D$ is an $m\times m$ diagonal matrix containing the
+row sums of matrix $C$ as the diagonal entries, $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$
 
@@
 
-Gaussian processes (GP) are also standard models for spatial processes \citep{Bane:etal:2003}.
+Gaussian processes (GP) are also standard models for spatial processes \citep{Bane:etal:2004}.
 GP models are computationally challenging for large datasets because
@@
 MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
-partial differential equations (SPDE).
+partial differential equations (SPDEs).
 
 We consider Gaussian processes in the \matern class, using the following
-parameterization of the \matern correlation function as 
+parameterization of the \matern correlation function, 
 \begin{equation}
@@
 The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
-to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second model
-is the Lindgren approximation for \matern-based GPs with $\nu=1$.
+to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second spatial
+model is the Lindgren approximation for \matern-based GPs with $\nu=1$.
 To implement the Lindgren model, one modifies the $Q$ matrix defined
@@
 $\rho$. In particular $\rho$ allows us to estimate the locality
-of the smoothing. As $\rho$ decreases, the model uses more and more
-local data to estimate the compositional proportions at a given location,
-effectively averaging the empirical proportions over smaller neighborhoods.
-In general, the \cite{Lind:etal:2011} model will generally provide
-for a smoother estimate than the CAR model \citep{Paci:2013}.
+of the smoothing. As $\rho$ decreases, the model uses increasingly
+localized data to estimate the compositional proportions at a given
+location, effectively averaging the empirical proportions over smaller
+neighborhoods. In general, the \cite{Lind:etal:2011} model will generally
+provide for a smoother estimate than the CAR model \citep{Paci:2013}.
 
@@
 we use a uniform distribution on each $\sigma_{p}$ parameter, with
-upper bound of 1000. For the SPDE model we also have parameters $\mu_{p}$,
+upper bound of 1000. For the SPDE model we also have parameters $\{\mu_{p}\}$,
 which we give flat, non-informative priors (truncated at $\pm10$),
-and $\rho_{p}$ which we give uniform priors on the interval $(0.1,\exp(5))$.
+and $\{\rho_{p}\}$, which we give uniform priors on the interval
+$(0.1,\exp(5))$.
 
@@
 and can be sampled within the MCMC as additional unknown parameters.
-Specifically, $c_{tj}$, is the latent `membership' variable for tree
-$j$ in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$ is
-a discrete distribution that puts mass, $\psi_{ti}$, $i=1,\ldots,m$,
+Specifically, $c_{tj}$ is the latent ``membership'' variable for
+tree $j$ in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$
+is a discrete distribution that puts mass, $\psi_{ti}$, $i=1,\ldots,m$,
 proportional to the areal overlap between the township in which the
@@
 
-Using the latent variable representation, we have that $W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{c_{tj}},1)$
+Using the latent variable representation, we have that $W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{c_{tj}}),1)$
 for tree $j$ in township $t$. In updating the other parameters in
 the model during the MCMC (specifically the $\alpha$ values), we
-condition on the current values, $c_{tj}$, which provides a ``soft''
-assignment of trees to grid cells that respects both the known township
-in which the tree occurred and the uncertainty in which grid cell
-the tree occurred.
-
-Note that this prior has the unrealistic feature that it does not
-represent our knowledge that the trees in a township would be distributed
-more regularly across the area of the township than expected by such
-an independence prior.
+condition on the current values, $\{c_{tj}\}$, which provides a ``soft''
+(i.e., probabilistic) assignment of trees to grid cells that respects
+both the known township in which the trees occurred and the uncertainty
+in which grid cells the trees occurred.
+
+Note that this prior represents the location of each tree in a township
+as being independent of the other trees; this is somewhat unrealistic
+because it does not represent our knowledge that the trees in a township
+would be distributed somewhat regularly across the area of the township
+because the witness trees were used to indicate property boundaries.
 
@@
 values and the latent $W_{ijp}$ variables). To address the first,
-we developed a `cross-level' joint updating strategy for the CAR model
-in which we propose $\phi_{p}=\sigma_{p},p=1,\ldots,P,$ (and for
-the SPDE model, $\phi_{p}\in\{\{\sigma_{p},\rho_{p}\},\mu_{p}\}$)
-via a Metropolis-style random walk and then conditional on the proposed
-value of $\phi_{p}^{*}$ propose $\alpha_{p}$ from its full conditional
-distribution given $\phi_{p}^{*}$ and the latent $W_{p}$ variables.
+we developed a ``cross-level'' joint updating strategy for the CAR
+model in which we propose $\phi_{p}=\sigma_{p},p=1,\ldots,P,$ (and
+for the SPDE model, $\phi_{p}\in\{\mu_{p},(\sigma_{p},\rho_{p})\}$)
+via a Metropolis-style random walk and then given the proposed value,
+$\phi_{p}^{*}$, propose $\alpha_{p}$ from its full conditional distribution
+given $\phi_{p}^{*}$ and the latent $W_{p}$ variables, where $W_{p}$
+is the vector of all $W_{ijp}$ values for taxon $p$: $W_{p}=\{W_{ijp}\},i=1,\ldots,m;j=1,\ldots,n_{i}$.
 This is equivalent to sampling from the marginalized (with respect
@@
 out data from the fitting process and assessing the fit of the model
-on the held-out data. This cross-validation used a subregion containing
-most of Minnesota and a small amount of western Wisconsin, defined
-to be the cells whose x-coordinate was less than 300,000 (this defines
-a north-south line that approximately goes through Duluth, Minnesota)
-and hereafter referred to as the ``Minnesota subregion'' . We used
-two cross-validation experiments:
+on the held-out data. We used two cross-validation experiments:
 \begin{enumerate}
-\item We held out all the data from 95\% of the cells in the Minnesota subregion,
-with cells selected at random. This was meant to assess the ability
-of the model to interpolate from a sparse set of cells/townships and
+\item The first experiment used a subregion containing most of Minnesota
+and a small amount of western Wisconsin, defined to be the cells whose
+x-coordinate was less than 300,000 m (this defines a north-south line
+that approximately goes through Duluth, Minnesota) and hereafter referred
+to as the ``Minnesota subregion''. We chose this subregion for cross-validation
+because of its high data density, allowing us to experiment with the
+effects of increasing data sparsity on model performance. We held
+out all data from 95\% of the cells in this Minnesota subregion, with
+cells selected at random. This was meant to assess the ability of
+the model to interpolate from a sparse set of cells/townships and
 mimics the limited data in Illinois and Indiana.
 \item We held out 5\% of the trees from all of the trees in the dataset
-(leaving aside the held-out Minnesota subregion cells). This was meant
-to assess the ability of the model to estimate the composition in
-cells in which data were available. 
+for the midwestern subdomain (leaving aside the held-out Minnesota
+subregion cells). This was meant to assess the ability of the model
+to estimate the composition in cells in which data were available. 
 \end{enumerate}
@@
 metrics to use to compare the predictive ability of models; these
-metrics are referred to as scoring rules. In particular it is well-recognized
-that predictive distributions should maximize sharpness subject to
-calibration. That is, the predictive distribution should be as narrow
-as possible while being calibrated such that the observations are
-consistent with the distribution \citep{Gnei:etal:2007}. When thinking
-in terms of prediction intervals as summaries of the predictive distribution,
-we seek intervals that are as narrow as possible while still covering
-the truth the expected proportion of the time.
+metrics are referred to as scoring rules. A general conclusion from
+this work is that predictive distributions should maximize sharpness
+subject to calibration. That is, the predictive distribution should
+be as narrow as possible while being calibrated such that the observations
+are consistent with the distribution \citep{Gnei:etal:2007}. When
+thinking in terms of prediction intervals as summaries of the predictive
+distribution, we seek intervals that are as narrow as possible while
+still covering the truth the expected proportion (e.g., 95\% for a
+95\% prediction interval) of the time.
 
 Following the suggestions in \cite{Gnei:etal:2007}, we considered
-the metrics listed below. For experiment 1, we define $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$
-as the count of all trees in held-out cell $i$ and for experiment
-2, $Y_{i}$ is the count of held-out individual trees in the cell,
-while $y_{ijp}$ is an indicator variable taking value either 0 or
-1 depending on whether the $j$th held-out tree in the $i$th cell
-is of taxon $p$. We calculated each of the metrics below in two ways.
-First, we used the posterior mean composition estimates (as an evaluation
-of our core predictions), with $\tilde{\theta}_{p}(s)$ being the
-posterior mean. Second, we averaged the metric over the posterior
-samples (as an evaluation of our full data product, including uncertainty),
-taking $\tilde{\theta}_{p}(s)$ to be an individual MCMC sample and
-then averaging the metric over all the posterior samples. 
+the following metrics: Brier score, log predictive density, mean square
+prediction error, mean absolute error, and coverage and length of
+prediction intervals. Further details on each are given below. For
+experiment 1, we define $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$ as the count
+of all trees in held-out cell $i$ and for experiment 2, $Y_{i}$
+is the count of held-out individual trees in the cell, while $y_{ijp}$
+is an indicator variable taking value either 0 or 1 depending on whether
+the $j$th held-out tree in the $i$th cell is of taxon $p$. We calculated
+each of the metrics in two ways. First, we used the posterior mean
+composition estimates (as an evaluation of our core predictions),
+with $\tilde{\theta}_{p}(s)$ being the posterior mean. Second, we
+averaged the metric over the posterior samples (as an evaluation of
+our full data product, including uncertainty), taking $\tilde{\theta}_{p}(s)$
+to be an individual MCMC sample and then averaging the metric over
+all the posterior samples. 
 \begin{enumerate}
@@
 While in principle, this metric should be optimal \citep{Krnj:Drap:2014},
-it suffers from a lack of robustness in being very sensitive to small
-predictions near zero \citep{Gnei:etal:2007}. Even worse, our Monte
-Carlo estimation of $\theta$ used 10000 samples, so in some cases
-$\tilde{\theta}_{p}(s)=0$. When a tree is present in a cell but its
-corresponding proportion is 0, this gives a log density of $-\infty$,
-preventing use of the metric. As an informal solution to this we set
-$\tilde{\theta}_{p}(s)=\frac{1}{100000}$ in such cases, but given
-these issues we treat the log predictive density as a secondary measure.
+it is very sensitive to small predictions near zero \citep{Gnei:etal:2007}.
+Even worse, our Monte Carlo estimation of $\theta$ used 10000 samples,
+so in some cases $\tilde{\theta}_{p}(s)=0$. When a tree is present
+in a cell but its corresponding proportion is 0, this gives a log
+density of $-\infty$, preventing use of the metric. As an informal
+solution to this we set $\tilde{\theta}_{p}(s)=\frac{1}{100000}$
+in such cases, but given these issues we treat the log predictive
+density as a secondary measure.
 
 \item (Experiment 1 only) Weighted root mean square prediction error (RMSPE)
-and mean absolute error (MAE) of the estimated proportions relative
-to the empirical proportions based on the held-out trees, averaging
-over cells and taxa. We weight by the number of held-out trees in
-each cell to account for the greater variability in the empirical
-proportions in locations with few held-out trees. 
+and mean absolute error (MAE): These metrics calculate the error of
+the estimated proportions relative to the empirical proportions based
+on the held-out trees, averaging over cells and taxa. We weight by
+the number of held-out trees in each cell to account for the greater
+variability in the empirical proportions in locations with few held-out
+trees. 
 \item (Experiment 1 only) Coverage and length of 95\% prediction intervals
@@
 \end{enumerate}
-Note that for the calculation of the metric applied to the individual
-posterior samples, we can estimate the posterior probability that
-one model has a lower (better) value of the metric than the other
-model by simply calculating the proportion of samples for which the
-model has a lower value of the metric.
+Note that all of the metrics except coverage and interval length can
+be applied to individual posterior samples and therefore allow us
+to estimate the posterior probability that one model has a lower (better)
+value of the metric than the other model by simply calculating the
+proportion of samples for which the model has a lower value of the
+metric.
 
@@
 of the convex hull of the observations. To attempt to alleviate this,
-we added a buffer zone of six grid cells around our entire original
-domain, but note that the boundary effects were still evident even
-after inclusion of the buffer. For the model comparison, for comparability,
-we included this buffer for both the SPDE and CAR models.
+we added a buffer zone with a width of six grid cells around our entire
+original domain, but note that the boundary effects were still evident
+even after inclusion of the buffer. For the model comparison, we included
+this buffer for both the SPDE and CAR models.
 
@@
 iterations for burn-in, we retained a posterior sample of 250 subsampled
-iterations (subsampled to reduce post-processing computations and
-storage needs).
+iterations -- we use a subsample instead of the full 125,000 post-burn-in
+iterations to reduce post-processing computations and storage needs.
 
@@
 
-Table \ref{tab:score_cell_fivepercent} shows predictive ability based
-on Experiment 1 for those cells in the Minnesota subregion held out
-of the fitting process, for both the mean of the metric applied to
-the individual posterior samples and for the metric applied to the
-posterior mean predictions. Table \ref{tab:coverage_cell_fivepercent}
-shows performance of the uncertainty estimates. For the posterior
-distribution over the predictive metric values, the CAR model outperforms
-the SPDE model, while for the posterior mean predictions, the SPDE
-model appears to outperform the CAR model to a lesser degree, but
-we do not have any uncertainty estimates for this comparison. Coverage
-and interval lengths are similar between the two models. From a practical
-perspective, based on the difference in mean absolute error, which
-is on a readily interpretable scale, the differences between the models
-are small.
+For Experiment 1 for cells in the Minnesota subregion held out of
+the fitting process, the CAR model outperforms the SPDE model based
+on the posterior distribution over the predictive metric values (Table
+\ref{tab:score_cell_fivepercent}). For the posterior mean predictions,
+the SPDE model appears to outperform the CAR model to a lesser degree,
+but we do not have any uncertainty estimates for this comparison.
+Coverage and interval lengths are similar between the two models (Table
+\ref{tab:coverage_cell_fivepercent}). From a practical perspective,
+based on the difference in mean absolute error, the differences between
+the models are small (Table \ref{tab:score_cell_fivepercent}).
 
@@
  &
-\multicolumn{3}{c|}{{\small{Posterior mean of metric}}} &
-\multicolumn{2}{c|}{{\small{Metric of posterior mean predictions}}}\tabularnewline
+\multicolumn{3}{c|}{{\small{}Posterior mean of metric}} &
+\multicolumn{2}{c|}{{\small{}Metric of posterior mean predictions}}\tabularnewline
 \hline 
@@
  &
-{\small{CAR model}} &
-{\small{SPDE model}} &
-{\small{Posterior Prob. CAR < SPDE}} &
-{\small{CAR model}} &
-{\small{SPDE model}}\tabularnewline
-\hline 
-{\small{Brier}} &
-{\small{0.819}} &
-{\small{0.844}} &
-{\small{0.98}} &
-{\small{0.738}} &
-{\small{0.733}}\tabularnewline
-\hline 
-{\small{Negative Log Density}} &
-{\small{466325}} &
-{\small{510383}} &
-{\small{1.00}} &
-{\small{394003}} &
-{\small{394554}}\tabularnewline
-\hline 
-{\small{Mean Absolute Error}} &
-{\small{0.0364}} &
-{\small{0.0383}} &
-{\small{0.98}} &
-{\small{0.0275}} &
-{\small{0.0269}}\tabularnewline
-\hline 
-{\small{Root Mean Square Error}} &
-{\small{0.0897}} &
-{\small{0.0960}} &
-{\small{0.97}} &
-{\small{0.0647}} &
-{\small{0.0627}}\tabularnewline
+{\small{}CAR model} &
+{\small{}SPDE model} &
+{\small{}Posterior Prob. CAR < SPDE} &
+{\small{}CAR model} &
+{\small{}SPDE model}\tabularnewline
+\hline 
+{\small{}Brier} &
+{\small{}0.819} &
+{\small{}0.844} &
+{\small{}0.98} &
+{\small{}0.738} &
+{\small{}0.733}\tabularnewline
+\hline 
+{\small{}Negative Log Density} &
+{\small{}466325} &
+{\small{}510383} &
+{\small{}1.00} &
+{\small{}394003} &
+{\small{}394554}\tabularnewline
+\hline 
+{\small{}Mean Absolute Error} &
+{\small{}0.0364} &
+{\small{}0.0383} &
+{\small{}0.98} &
+{\small{}0.0275} &
+{\small{}0.0269}\tabularnewline
+\hline 
+{\small{}Root Mean Square Error} &
+{\small{}0.0897} &
+{\small{}0.0960} &
+{\small{}0.97} &
+{\small{}0.0647} &
+{\small{}0.0627}\tabularnewline
 \hline 
@@
  &
-{\small{CAR model}} &
-{\small{SPDE model}}\tabularnewline
-\hline 
-{\small{Coverage}} &
-{\small{0.977}} &
-{\small{0.978}}\tabularnewline
-\hline 
-{\small{Mean Interval Length}} &
-{\small{0.129}} &
-{\small{0.142}}\tabularnewline
-\hline 
-{\small{Median Interval Length}} &
-{\small{0.037}} &
-{\small{0.033}}\tabularnewline
+{\small{}CAR model} &
+{\small{}SPDE model}\tabularnewline
+\hline 
+{\small{}Coverage} &
+{\small{}0.977} &
+{\small{}0.978}\tabularnewline
+\hline 
+{\small{}Mean Interval Length} &
+{\small{}0.129} &
+{\small{}0.142}\tabularnewline
+\hline 
+{\small{}Median Interval Length} &
+{\small{}0.037} &
+{\small{}0.033}\tabularnewline
 \hline 
@@
 
-Table \ref{tab:score_cell_20percent} shows the results when the proportion
-of cells that are held out decreases from 95\% to 80\%, indicating
-performance on less sparse data. Table \ref{tab:coverage_cell_20percent}
-shows performance of the uncertainty estimates. For denser data, unlike
-the results when data are more sparse, the SPDE model generally outperforms
-the CAR model, but again differences from a practical perspective,
-based on mean absolute error, are limited.
+The results for the variation on Experiment 1 in which the proportion
+of cells that are held out decreases from 95\% to 80\% show that the
+SPDE model generally outperforms the CAR model, but again differences
+from a practical perspective, based on mean absolute error, are limited
+(Tables \ref{tab:score_cell_20percent}-\ref{tab:coverage_cell_20percent}).
 
@@
  &
-\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
-\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
+\multicolumn{3}{c|}{{\small{}Posterior mean of score}} &
+\multicolumn{2}{c|}{{\small{}Score of posterior mean predictions}}\tabularnewline
 \hline 
@@
  &
-{\small{CAR model}} &
-{\small{SPDE model}} &
-{\small{Posterior Prob. CAR < SPDE}} &
-{\small{CAR model}} &
-{\small{SPDE model}}\tabularnewline
-\hline 
-{\small{Brier}} &
-{\small{0.773}} &
-{\small{0.765}} &
-{\small{0.10}} &
-{\small{0.710}} &
-{\small{0.710}}\tabularnewline
-\hline 
-{\small{Negative Log Density}} &
-{\small{355928}} &
-{\small{353987}} &
-{\small{0.25}} &
-{\small{311525}} &
-{\small{311902}}\tabularnewline
-\hline 
-{\small{Mean Absolute Error}} &
-{\small{0.0309}} &
-{\small{0.0296}} &
-{\small{0.10}} &
-{\small{0.0226}} &
-{\small{0.0223}}\tabularnewline
-\hline 
-{\small{Root Mean Square Error}} &
-{\small{0.0763}} &
-{\small{0.0739}} &
-{\small{0.02}} &
-{\small{0.0533}} &
-{\small{0.0530}}\tabularnewline
+{\small{}CAR model} &
+{\small{}SPDE model} &
+{\small{}Posterior Prob. CAR < SPDE} &
+{\small{}CAR model} &
+{\small{}SPDE model}\tabularnewline
+\hline 
+{\small{}Brier} &
+{\small{}0.773} &
+{\small{}0.765} &
+{\small{}0.10} &
+{\small{}0.710} &
+{\small{}0.710}\tabularnewline
+\hline 
+{\small{}Negative Log Density} &
+{\small{}355928} &
+{\small{}353987} &
+{\small{}0.25} &
+{\small{}311525} &
+{\small{}311902}\tabularnewline
+\hline 
+{\small{}Mean Absolute Error} &
+{\small{}0.0309} &
+{\small{}0.0296} &
+{\small{}0.10} &
+{\small{}0.0226} &
+{\small{}0.0223}\tabularnewline
+\hline 
+{\small{}Root Mean Square Error} &
+{\small{}0.0763} &
+{\small{}0.0739} &
+{\small{}0.02} &
+{\small{}0.0533} &
+{\small{}0.0530}\tabularnewline
 \hline 
@@
  &
-{\small{CAR model}} &
-{\small{SPDE model}}\tabularnewline
-\hline 
-{\small{Coverage}} &
-{\small{0.981}} &
-{\small{0.972}}\tabularnewline
-\hline 
-{\small{Mean Interval Length}} &
-{\small{0.112}} &
-{\small{0.103}}\tabularnewline
-\hline 
-{\small{Median Interval Length}} &
-{\small{0.028}} &
-{\small{0.022}}\tabularnewline
+{\small{}CAR model} &
+{\small{}SPDE model}\tabularnewline
+\hline 
+{\small{}Coverage} &
+{\small{}0.981} &
+{\small{}0.972}\tabularnewline
+\hline 
+{\small{}Mean Interval Length} &
+{\small{}0.112} &
+{\small{}0.103}\tabularnewline
+\hline 
+{\small{}Median Interval Length} &
+{\small{}0.028} &
+{\small{}0.022}\tabularnewline
 \hline 
@@
 
-Table \ref{tab:score_tree} shows the predictive ability based on
-Experiment 2. Here we have evidence (posterior probability of 0.93)
+In Experiment 2, we have evidence (posterior probability of 0.93)
 that the SPDE model is better based on the Brier score, but the Brier
-score values for the two models are numerically almost the same.
+score values for the two models are numerically almost the same (Table
+\ref{tab:score_tree}).
 
@@
  &
-{\small{CAR model}} &
-{\small{SPDE model}} &
-{\small{Posterior Prob. CAR < SPDE}} &
-{\small{CAR model}} &
-{\small{SPDE model}}\tabularnewline
-\hline 
-{\small{Brier}} &
-{\small{0.662}} &
-{\small{0.661}} &
-{\small{0.07}} &
-{\small{0.657}} &
-{\small{0.657}}\tabularnewline
-\hline 
-{\small{Negative Log Density}} &
-{\small{51757}} &
-{\small{51626}} &
-{\small{0.01}} &
-{\small{50705}} &
-{\small{50736}}\tabularnewline
+{\small{}CAR model} &
+{\small{}SPDE model} &
+{\small{}Posterior Prob. CAR < SPDE} &
+{\small{}CAR model} &
+{\small{}SPDE model}\tabularnewline
+\hline 
+{\small{}Brier} &
+{\small{}0.662} &
+{\small{}0.661} &
+{\small{}0.07} &
+{\small{}0.657} &
+{\small{}0.657}\tabularnewline
+\hline 
+{\small{}Negative Log Density} &
+{\small{}51757} &
+{\small{}51626} &
+{\small{}0.01} &
+{\small{}50705} &
+{\small{}50736}\tabularnewline
 \hline 
@@
 
-For the final data product, we ran the model using the CAR specification
+The final data product is a dataset that contains 250 posterior samples
+of the proportions of each of the 23 tree taxa at each grid cell in
+the states in our domain of the northeastern United States.
+
+For this final data product, we ran the model using the CAR specification
 for 150,000 iterations, discarding the first 25,000 iterations for
-burn-in and retaining 250 subsampled iterations (subsampled to limit
-storage needs and the size of the data product). Mixing was generally
-reasonable, but for some of the hyperparameters was relatively slow,
-particularly for less common taxa. Despite this, mixing for the variables
-of substantive interest -- the proportions -- was good.
-
-In Fig. \ref{fig:select_maps}, we show maps of estimated composition
-for the full domain for several taxa of substantive interest to illustrate
-the results. These maps contrast the raw data proportions, the posterior
-means, and posterior standard deviations as pointwise estimates of
-uncertainty. Fig. \ref{fig:all_predictions} shows posterior means
-for all 23 taxa.
+burn-in and retaining 250 subsampled iterations from the remaining
+125,000 iterations (subsampled to limit storage needs and the size
+of the data product). Based on graphical checks and calculation of
+effective sample size values, mixing was generally reasonable, but
+for some of the hyperparameters was relatively slow, particularly
+for less common taxa. Despite this, mixing for the variables of substantive
+interest -- the proportions -- was good, with effective sample sizes
+for the final product generally near 250.
+
+Maps of estimated composition for the full domain for several taxa
+of substantive interest illustrate the results, contrasting the raw
+data proportions, the posterior means, and posterior standard deviations
+as pointwise estimates of uncertainty (Fig. \ref{fig:select_maps}).
+We also present the posterior means for all 23 taxa (Fig. \ref{fig:all_predictions}).
 
 \begin{figure}
-\includegraphics[scale=0.75]{1_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig2.pdf}
-
-\hspace{4mm}\includegraphics[scale=0.16]{2_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendRaw.pdf}\hspace{4mm}\includegraphics[scale=0.16]{3_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendProp.pdf}\hspace{3.5mm}\includegraphics[scale=0.16]{4_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendSD.pdf}
+\includegraphics[scale=0.75]{fig2}
+
+\hspace{4mm}\includegraphics[scale=0.16]{legendRaw}\hspace{4mm}\includegraphics[scale=0.16]{legendProp}\hspace{3.5mm}\includegraphics[scale=0.16]{legendSD}
 
@@
 \begin{figure}
-\includegraphics[scale=0.8]{5_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig3.pdf}
+\includegraphics[scale=0.8]{fig3}
 
@@
 
-The data product is publicly available at the NIS Data Portal (accession
-number to be inserted when available) under YY license as version
-0.4 as of August 2015. The product is in the form of a netCDF-4 file,
-with dimensions x, y, and MCMC iteration. There is one variable per
-taxon. The PalEON project (in particular the first author) will continue
-to maintain this product, releasing new versions as additional data
-in Illinois, Indiana and Ohio are digitized. Note that digitization
-of data from Illinois and Indiana is ongoing, and digitization of
-additional data from Ohio is planned as well. As a result, at some
-point we expect to have complete data for the western half of the
-domain. There will remain some missing townships in the eastern half
-of the domain.
+The data product is being made publicly available at the NIS Data
+Portal (accession number to be inserted when available) under the
+CC BY license as version 0.4 as of August 2015. The product is in
+the form of a netCDF-4 file, with dimensions x-coordinate, y-coordinate,
+and MCMC iteration. There is one variable per taxon. In addition,
+dynamic visualizations of the product using the Shiny tool are available
+at http://gandalf.berkeley.edu:3838/paciorek/setVegComp {[}this URL
+is temporary and will be updated{]}. The PalEON Project (in particular
+the first author) will continue to maintain this product, releasing
+new versions as additional data in Illinois, Indiana and Ohio are
+digitized. Note that digitization of data from Illinois and Indiana
+is ongoing, and digitization of additional data from Ohio is planned
+as well. As a result, at some point we expect to have complete data
+for the midwestern half of the domain.
 
@@
 model is its inability to account for local features such as rivers
-(note the Minnesota River floodplain in evidence in the raw data).
-The current model, by using a simple stationary spatial model that
-smooths as a function of Euclidean distance, does not account for
-topographic, soil, or other features. An additional drawback of the
-product is its focus on composition, which does not directly tell
-us about vegetation structure, in particular does not distinguish
-between closed forest, savanna, and prairie, of particular note in
-Minnesota, Wisconsin, Illinois, and into Indiana. The statistically-estimated
-biomass product mentioned next will directly inform questions about
-vegetation structure.
-
-In addition to the existing raw composition count data product discussed
-in Section \ref{sec:Data}, we have related data products that are
-under development:
-\begin{itemize}
-\item Gridded raw biomass estimates for Minnesota, Wisconsin, and Michigan
-based on the PLS data, with extension to Illinois and Indiana planned. 
-\item Statistically estimated biomass for Minnesota, Wisconsin, and Michigan
-using a statistical model applied to the raw biomass estimates, with
-extension to Illinois and Indiana planned.
-\end{itemize}
-We also plan to estimate basal area and stem density using a similar
-approach to that used for biomass.
+(e.g., the Minnesota River riparian corridor can be seen in the raw
+data but is missing in the statistical estimates). The current model,
+by using a simple stationary spatial model that smooths as a function
+of Euclidean distance, does not account for topographic, soil, or
+other features. An additional drawback of the product is its focus
+on composition, which does not directly tell us about tree density
+or other aspects of vegetation structure, and in particular does not
+distinguish between closed forest, savanna, and prairie. This limitation
+is particularly critical at the prairie-forest transition in Minnesota,
+Wisconsin, Illinois, and into Indiana \citep{transeau1935prairie}.
+
+This latter limitation can be addressed by developing estimates of
+absolute abundance (e.g., biomass) rather than the relative abundance
+compositional data used here. A gridded dataset of biomass, stem density,
+and basal area is already available for Minnesota, Wisconsin, and
+northern Michigan \citep{goring2015composition}, based on the PLS
+data. An extension to southern Michigan, Illinois, and Indiana is
+planned. We are currently developing statistical estimates of biomass
+for Minnesota, Wisconsin, and Michigan using a statistical model applied
+to the gridded biomass dataset, with extension to Illinois and Indiana
+planned. We also plan to estimate stem density and basal area using
+a similar approach to that used for biomass.
 
@@
 out the model comparison and created the data product. CJP wrote the
-paper based on an initial draft by AT and with feedback from SJG,
-CVC, JWW, DJM, JAP, JZ and JSM. CVC, DJM, SJG, JWW, and JAP provided
-data.
+paper based on an initial draft by AT and with feedback and editing
+from SJG, CVC, JWW, DJM, JAP, JZ and JSM. SJG, JAP, CVC, DJM, JSM,
+and JWW led the processing and analysis of the PLS and TPS data and
+assisted with interpretation of results.
 
@@
 years who have preserved, collected, and digitized survey records,
-in particular Jim Dyer, Peter Marks, Robert McIntosh, Ronald Stuckey,
-Ed Schools, and the Ohio Biological Survey \{David/Simon please add
-names here\} We thank Madeline Ruid, Ben Seliger, Morgan Ross and
-Daniel Handel for processing of the southern Michigan data. Indiana
-and Illinois data were made possible through the hard work of many
-Notre Dame undergraduates in the McLachlan lab. This work was carried
-out by the PalEON Project with support from the National Science Foundation
-MacroSystems Program through grants EF-1065656, EF-1241868, DEB-1241874
-and DEB-1241868 and from the Notre Dame Environmental Change Initiative.
+in particular John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed
+Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
+We thank Madeline Ruid, Ben Seliger, Morgan Ross and Daniel Handel
+for processing of the southern Michigan data. Indiana and Illinois
+data were made possible through the hard work of many Notre Dame undergraduates
+in the McLachlan lab. This work was carried out by the PalEON Project
+with support from the National Science Foundation MacroSystems Program
+through grants EF-1065702, EF-1065656, DEB-1241874 and DEB-1241868
+and from the Notre Dame Environmental Change Initiative.
 
@@
 
-\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
+% \bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
+\bibliography{paperArxiv}
 
@@
 and $A_{ii}=0$. For the township data, at each iteration, based on
-the current values of the grid cell membership variables, $c_{tj}$,
+the current values of the grid cell membership variables, $\{c_{tj}\}$,
 trees are aggregated into grid cells and the calculations above can
@@
 for the SPDE model. For each hyperparameter, $\phi_{p}=\log\sigma_{p}$
-for the CAR model and $\phi_{p}\in\{\mu_{p},\{\log\sigma_{p},\log\rho_{p}\}\}$
-for the SPDE model, we sampled $\{\phi_{p},\alpha_{p}\}$ jointly,
+for the CAR model and $\phi_{p}\in\{\mu_{p},(\log\sigma_{p},\log\rho_{p})\}$
+for the SPDE model, we sample $\{\phi_{p},\alpha_{p}\}$ jointly,
 proposing $\phi_{p}$ as a random walk and, conditional on the proposed
@@
 Monte Carlo integration over $W_{ijp},\, p=1,\ldots,P$ to estimate
-$\theta_{p}(s_{i})$. The quantity $\theta_{p}(s_{i})=P(W_{ijp}={\displaystyle \max_{p^{*}}W_{ijp^{*}})}$
+$\theta_{p}(s_{i})$. The quantity $\theta_{p}(s_{i})=\mbox{Prob}(W_{ijp}={\displaystyle \max_{p^{*}}W_{ijp^{*}})}$
 defines the probability of taxon $p$ at grid cell $i$. This requires
-one to choose the number of Monte Carlo samples, which we set at 10000.
-For each of the saved MCMC samples, $k=1,\ldots K$, we estimate $\theta_{p}^{(k)}(s_{i})$
-numerically. Specifically, for $t=1,\ldots,10000$ samples (i.e.,
-hypothetical trees), we independently draw
+one to choose the number of Monte Carlo samples, which we set at 10000,
+effectively sampling 10000 hypothetical trees and estimating the probabilities
+of the different taxa in the population from the empirical proportions
+in this sample of trees. For each of the saved MCMC samples, $k=1,\ldots K$,
+we estimate $\theta_{p}^{(k)}(s_{i})$ numerically. Specifically,
+for $t=1,\ldots,10000$ samples (i.e., hypothetical trees), we independently
+draw
 \[

### 2015-09-05 paperPlos.tex
added arxiv and plos versions of tex

--- 2015-09-05 arxiv/paperArxiv.tex
@@
-\documentclass[12pt]{article}
-\usepackage[]{graphicx}
-\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-
-%\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-%\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-%\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-%\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-%\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-%\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-%\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-%\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-%\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-%\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-%\usepackage{framed}
-%\makeatletter
-%\newenvironment{kframe}{%
-% \def\at@end@of@kframe{}%
-% \ifinner\ifhmode%
-%  \def\at@end@of@kframe{\end{minipage}}%
-%  \begin{minipage}{\columnwidth}%
-% \fi\fi%
-% \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
-% \colorbox{shadecolor}{##1}\hskip-\fboxsep
-%     % There is no \\@totalrightmargin, so:
-%     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
-% \MakeFramed {\advance\hsize-\width
-%   \@totalleftmargin\z@ \linewidth\hsize
-%   \@setminipage}}%
-% {\par\unskip\endMakeFramed%
-% \at@end@of@kframe}
-%\makeatother
-
-%\definecolor{shadecolor}{rgb}{.97, .97, .97}
-%\definecolor{messagecolor}{rgb}{0, 0, 0}
-%\definecolor{warningcolor}{rgb}{1, 0, 1}
-%\definecolor{errorcolor}{rgb}{1, 0, 0}
-%\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\usepackage[T1]{fontenc}
-\usepackage[latin9]{inputenc}
-\usepackage{geometry}
-\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
+% Template for PLoS
+% Version 3.1 February 2015
+%
+% To compile to pdf, run:
+% latex plos_template
+% bibtex plos_template
+% latex plos_template
+% latex plos_template
+% dvipdf plos_template
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
+% the final text of your manuscript.
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig." instead of "Figure".
+% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - tabs/spacing/line breaks within cells to alter layout or alignment
+% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
+% - colors, shading, or graphic objects
+% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
+%
+% Please be sure to include all portions of an equation in the math environment.
+%
+% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% fixltx2e package for \textsubscript
+\usepackage{fixltx2e}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% rotating package for sideways tables
+\usepackage{rotating}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{latex/plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Leave date blank
+\date{}
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\lhead{\includegraphics[width=2.0in]{latex/PLOS-Submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\sf PLOS}
+
+%% user-provided packages
 \usepackage{array}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage[authoryear]{natbib}
-
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
-\newcommand{\lyxaddress}[1]{
-\par {\raggedright #1
-\vspace{1.4em}
-\noindent\par}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
-\usepackage{times}
-\usepackage{graphics,natbib}
+
+%% Include all macros below
 
@@
 
-\makeatother
+%% END MACROS SECTION
 
 \begin{document}
-
-\title{Statistically-estimated tree composition for the northeastern United
-States \\
-at the time of Euro-American settlement}
-
-\author{Christopher J. Paciorek\textsuperscript{1,{*}}, Simon J. Goring\textsuperscript{2,\&},
-Andrew L. Thurman\textsuperscript{3,\&}, \\
-Charles V. Cogbill\textsuperscript{4}, John W. Williams\textsuperscript{2,3},
-David J. Mladenoff\textsuperscript{6}, \\
-Jody A. Peters\textsuperscript{7}, Jun Zhu\textsuperscript{8}, and
-Jason S. McLachlan\textsuperscript{7}}
-
-\maketitle
-
-\lyxaddress{\textsuperscript{1}Department of Statistics, University of California, Berkeley, California, USA}
-
-\lyxaddress{\textsuperscript{2}Department of Geography, University of Wisconsin, Madison, Wisconsin, USA}
-
-\lyxaddress{\textsuperscript{3}Department of Statistics, University of Iowa, Iowa City, Iowa, USA}
-
-\lyxaddress{\textsuperscript{4}Harvard Forest, Harvard University, Petersham, Massachusetts, USA}
-
-\lyxaddress{\textsuperscript{5}Center for Climatic Research, University of Wisconsin,
-Madison, Wisconsin, USA}
-
-\lyxaddress{\textsuperscript{6}Department of Forestry and Wildlife Ecology,
-University of Wisconsin, Madison, Wisconsin, USA}
-
-\lyxaddress{\textsuperscript{7}Department of Biological Sciences, University
-of Notre Dame, Notre Dame, Indiana, USA}
-
-\lyxaddress{\textsuperscript{8}Department of Statistics, University of Wisconsin,
-Madison, Wisconsin, USA}
-
-\lyxaddress{\textsuperscript{{*}}Corresponding author; E-mail: paciorek@stat.berkeley.edu}
-
-\lyxaddress{\textsuperscript{\&}These authors contributed equally to this work.
-\newpage}
-
-\begin{abstract}
+\vspace*{0.35in}
+
+% Title must be 250 characters or less.
+% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
+\begin{flushleft}
+{\Large
+\textbf\newline{Statistically-estimated Tree Composition for the\\
+Northeastern United States at Euro-American Settlement}
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Christopher J. Paciorek\textsuperscript{1,*},
+Simon J. Goring\textsuperscript{2,\Yinyang},
+Andrew L. Thurman\textsuperscript{3,\Yinyang},
+Charles V. Cogbill\textsuperscript{4},
+John W. Williams\textsuperscript{2,5},
+David J. Mladenoff\textsuperscript{6},
+Jody A. Peters\textsuperscript{7},
+Jun Zhu\textsuperscript{8},
+Jason S. McLachlan\textsuperscript{7,\ddag}
+\\
+\bigskip
+\bf{1} Department of Statistics, University of California, Berkeley, California,
+USA
+\\
+\bf{2} Department of Geography, University of Wisconsin, Madison, Wisconsin,
+USA
+\\
+\bf{3} Department of Statistics, University of Iowa, Iowa City, Iowa, USA
+\\
+\bf{4} Harvard Forest, Harvard University, Petersham, Massachusetts, USA
+\\
+\bf{5} Center for Climatic Research, University of Wisconsin, Madison,
+Wisconsin, USA
+\\
+\bf{6} Department of Forestry and Wildlife Ecology, University of Wisconsin,
+Madison, Wisconsin, USA
+\\
+\bf{7} Department of Biological Sciences, University of Notre Dame, Notre
+Dame, Indiana, USA
+\\
+\bf{8} Department of Statistics, University of Wisconsin, Madison, Wisconsin,
+USA
+\\
+\bigskip
+
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+% 
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+\Yinyang These authors contributed equally to this work.
+
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+ \ddag PI of the PalEON Project
+
+% Current address notes
+% \textcurrency a Insert current address of first author with an address update
+% \textcurrency b Insert current address of second author with an address update
+% \textcurrency c Insert current address of third author with an address update
+
+% Deceased author note
+%\dag Deceased
+
+% Group/Consortium Author Note
+%\textpilcrow Membership list can be found in the Acknowledgments section.
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* paciorek@stat.berkeley.edu (CJP)
+
+%TODO for PLOS\_ONE
+
+%* coverletter
+%* general guideline
+
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
+
 We present a gridded 8 km-resolution data product of the estimated
@@
 local administrative units. The product is based on a Bayesian statistical
-model fit to the count data that estimates composition on a regular
+model fit to the count data that estimates composition on the
 8 km grid across the entire domain. The statistical model is designed
@@
 change after intensive settlement. The data product is being made
-available at the NIS data portal as version 0.4 under the CC BY license.
-
-Keywords: biogeography, species composition, old-growth forests, spatial
-modeling, Bayesian statistical model, vegetation mapping
-\end{abstract}
-
-\section{Introduction}
+available at the NIS data portal as version 0.4.
+
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step. 
+% Author Summary not valid for PLOS ONE submissions.   
+\linenumbers
+
+\section*{Introduction}
 
@@
 (from 1620 until 1825) in the northeastern United States provide spatially-aggregated
-data at the township level \citep{Cogb:etal:2002,thompson2013four},
-with typical township size on the order of 200 $\mbox{km}^{2}$ and
+data at the township level \cite{Cogb:etal:2002,thompson2013four},
+with typical township size on the order of 200 km\textsuperscript{2} and
 no information about the locations of individual trees; we refer to
@@
 one-half mile (800 m) spacing, for Ohio and westward during the period
-1785 to 1907 \citep{bourdo1956review,pattison1957beginnings,schulte2001original,goring2015composition}.
+1785 to 1907 \cite{bourdo1956review,pattison1957beginnings,schulte2001original,goring2015composition}.
 At each point 2-4 trees were identified, and the common name, diameter
@@
 consistency among points and the application of spatially-varying
-correction factors \citep{cogbill2015corrections,goring2015composition}
+correction factors \cite{goring2015composition}
 to accurately assess tree stem density, basal area and biomass from
 the early settlement records, but the impact on composition estimates
-is limited \citep{liu2011broadscale}. Surveyors sometimes used ambiguous
+is limited \cite{liu2011broadscale}. Surveyors sometimes used ambiguous
 common names, which requires matching to scientific names and standardization
-\citep{mladenoff2002narrowing,goring2015composition}.
+\cite{mladenoff2002narrowing,goring2015composition}.
 
 Logging, agriculture, and land abandonment have left an indelible
-mark on forests in the northeastern United States \citep{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
+mark on forests in the northeastern United States \cite{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
 However most studies have assessed these effects in individual states
-or smaller domains \citep{friedman2005regional,rhemtulla2009historical}
+or smaller domains \cite{friedman2005regional,rhemtulla2009historical}
 and with various spatial resolutions, from townships (36 square miles)
-to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
+to forest zones of hundreds or thousands of square miles. \cite{goring2015composition}
 provide a new dataset of forest composition, biomass, and stem density
@@
 scale that can be compared to modern forests using Forest Inventory
-and Analysis products \citep{gray2012forest}. Combined with additional,
+and Analysis products \cite{gray2012forest}. Combined with additional,
 coarsely-sampled PLS data from Illinois and Indiana, newly-digitized
@@
 
-In Section \ref{sec:Data} we describe the data sources, while Section
-\ref{sec:Statistical-model} describes our statistical models. In
-Section \ref{sec:Model-comparison} we quantitatively compare competing
-statistical specifications, and in Section \ref{sec:Data-product}
-we describe the final data product. In Section \ref{sec:Discussion}
+In \emph{Data} we describe the data sources, while in \emph{Statistical model}
+we describe the statistical model used to create the data product.
+In \emph{Model comparison} we quantitatively compare competing
+statistical specifications, and \emph{Data Product}
+we describe the final data product. In \emph{Discussion}
 we discuss limitations of the statistical model, and we list related
@@
 
-\section{Data\label{sec:Data}}
+\section*{Methods}
+
+\subsection*{Data\label{sec:Data}}
 
@@
 the TPS data for some townships were lost, incomplete, or have not
-been located \citep{Cogb:etal:2002}.
+been located \cite{Cogb:etal:2002}.
 
 \begin{figure}
-\label{fig:domain}\includegraphics[scale=1.3]{fig1}
-
-\caption{Spatial domain of the northeastern United States, with locations with
-data shown in gray. Locations are grid cells in midwestern portion
+% \label{fig:domain}\includegraphics[scale=1.3]{fig1.pdf}
+
+\caption{{\bf Spatial domain of the northeastern United States, with locations with
+data shown in gray.} \\ Locations are grid cells in midwestern portion
 and townships in eastern portion. In addition to locations without
@@
 are white (e.g., a few locations in the northwestern portion of the
-domain in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
+domain in the states of Minnesota and Wisconsin).}}
+\label{fig:Spatial-domain}
 \end{figure}
@@
 the landscape at the time just prior to widespread Euro-American settlement
-and land use \citep{Whit:1996,Cogb:etal:2002}. These forest composition
+and land use \cite{Whit:1996,Cogb:etal:2002}. These forest composition
 datasets do include the effects of Native American land use and early
-Euro-American settlement activities \citep[e.g.,][]{Blac:etal:2006},
+Euro-American settlement activities, e.g. \cite{Blac:etal:2006},
 but it is likely that the imprint of this earlier land use is highly
-concentrated rather than spatially extensive \citep{munoz2014defining}.
+concentrated rather than spatially extensive \cite{munoz2014defining}.
 
 Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
-data and processing steps are available in \cite{goring2015composition};
+data and processing steps are available \cite{goring2015composition};
 key elements include the use of only corner points, the use of only
@@
 using small trees, there was no consistent lower diameter limit. The
-PLS data generally represent trees greater than 8 inches (\textasciitilde{}20
+PLS data generally represent trees greater than 8 inches (ca. 20
 cm) diameter at breast height (dbh), but with some trees as small
@@
 
-\section{Statistical model\label{sec:Statistical-model}}
+\subsection*{Statistical model\label{sec:Statistical-model}}
 
@@
 
-\subsection{Data model}
+\paragraph*{Data model}
 
 We start by describing the basic model for those states for which
-we have raw data on the 8 km grid, and in Section \ref{sub:Model-for-township}
+we have raw data on the 8 km grid, and in \emph{Model for township data}
 we describe the extension of the model to accommodate data aggregated
@@
 \]
-where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
+where $y_{i}$ is the vector of counts for the \emph{P} taxa at the \emph{i}th
 cell, $n_{i}$ is the number of trees counted in the cell, and $\theta(s_{i})$
@@
 because the set of trees in the dataset is roughly uniformly sampled
-across the cells or townships \citep{goring2015composition}.
+across the cells or townships \cite{goring2015composition}.
 
@@
 processes are defined on the 8 km grid, $\alpha_{p}=\{\alpha_{p}(s_{1}),\ldots,\alpha_{p}(s_{m})\}$
-for the $m$ grid cells. In Section \ref{sub:Latent-Variable-Model}
+for the $m$ grid cells. In \emph{Latent variable model}
 we introduce a multinomial probit model that relates the $\alpha_{p}(s)$
@@
 structure of the $\alpha_{p}(s)$ processes, a standard conditional
-autoregressive model \citep{Bane:etal:2004} and a Gaussian Markov
+autoregressive model \cite{Bane:etal:2004} and a Gaussian Markov
 random field (MRF) approximation to a Gaussian process with Matern
-covariance \citep{Lind:etal:2011}.
-
-\subsection{Spatial process models}
+covariance \cite{Lind:etal:2011}.
+
+\paragraph*{Spatial process models}
 
@@
 the spatial process, so calculation of the prior density of $\alpha_{p}$
-is computationally simple \citep{Rue:Held:2005}, but in situations
+is computationally simple \cite{Rue:Held:2005}, but in situations
 where the likelihood is not normal, it can be difficult to set up
 effective MCMC algorithms that are able to move in the high-dimensional
-space of $\alpha_{p}$. The latent variable representation of Section
-\ref{sub:Latent-Variable-Model} helps to alleviate this problem.
+space of $\alpha_{p}$. The latent variable representation helps to alleviate this problem.
 Next we describe two alternative spatial models that we considered;
-in Section \ref{sec:Model-comparison}, we use cross-validation to
+in \emph{Model comparison}, we use cross-validation to
 choose between these two representations.
@@
 Our first model is a standard conditional autoregressive (CAR) model
-\citep{Bane:etal:2004}. We use a standard form of this model, which
+\cite{Bane:etal:2004}. We use a standard form of this model, which
 treats the four cardinal neighbors of each grid cell as the neighbors
 of the grid cell. The corresponding precision matrix has diagonal
-elements, $Q_{ii}$, equal to the number of neighbors for the $i$th
+elements, $Q_{ii}$, equal to the number of neighbors for the \emph{i}th
 area (i.e., four except for cells on the boundary of the domain),
-while $Q_{ik}=-1$ (the negative of a weight of one) when areas $i$
-and $k$ are neighbors and $Q_{ik}=0$ when they are not. This gives
+while $Q_{ik}=-1$ (the negative of a weight of one) when areas \emph{i}
+and \emph{k} are neighbors and $Q_{ik}=0$ when they are not. This gives
 the following model for the values of $\alpha_{p}(s_{i})$ collected
@@
 \[
-\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-}).
+\alpha_{p}\sim\N(0,\sigma_{p}^{2}Q^{-}).
 \]
-The use of the generalized inverse notation indicates that $Q$ is
+The use of the generalized inverse notation indicates that \emph{Q} is
 not full-rank, but is of rank $m-1$; this gives an improper prior
@@
 matrix defining the neighborhood relation of the locations; that is
-$C{}_{ik}=1$ if locations $i$ and $k$ are neighbors and zero otherwise.
-The matrix $D$ is an $m\times m$ diagonal matrix containing the
-row sums of matrix $C$ as the diagonal entries, $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$
+$C{}_{ik}=1$ if locations \emph{i} and \emph{k} are neighbors and zero otherwise.
+The matrix \emph{D} is an $m\times m$ diagonal matrix containing the
+row sums of matrix \emph{C} as the diagonal entries, $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$
 
@@
 
-Gaussian processes (GP) are also standard models for spatial processes \citep{Bane:etal:2004}.
+Gaussian processes (GP) are also standard models for spatial processes \cite{Bane:etal:2004}.
 GP models are computationally challenging for large datasets because
@@
 \begin{equation}
-R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
+R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),
 \end{equation}
-where $d$ is Euclidean distance, $\rho$ is the spatial range parameter,
+where \emph{d} is Euclidean distance, $\rho$ is the spatial range parameter,
 and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
@@
 
-The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
+The  approach of \cite{Lind:etal:2011} allows us to consider MRF approximations
 to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second spatial
 model is the Lindgren approximation for \matern-based GPs with $\nu=1$.
-To implement the Lindgren model, one modifies the $Q$ matrix defined
+To implement the Lindgren model, one modifies the \emph{Q} matrix defined
 previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
-elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
-neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
-for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
+elements of \emph{Q} are $4+a^{2}$. The entries corresponding to cardinal
+neighbors are -2. Those for diagonal neighbors are 2, and those
+for 2nd-order cardinal neighbors are 1. This extends the neighborhood
 structure relative to the CAR model and parameterizes it as a function
@@
 location, effectively averaging the empirical proportions over smaller
-neighborhoods. In general, the \cite{Lind:etal:2011} model will generally
-provide for a smoother estimate than the CAR model \citep{Paci:2013}.
+neighborhoods. In general, the model of \cite{Lind:etal:2011} will generally
+provide for a smoother estimate than the CAR model \cite{Paci:2013}.
 
@@
 \[
-\alpha_{p}\sim\mbox{N}\left(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1}\right)
+\alpha_{p}\sim\N\left(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1}\right)
 \]
@@
 
-\subsection{Prior Distributions}
+\paragraph*{Prior distributions}
 
@@
 upper bound of 1000. For the SPDE model we also have parameters $\{\mu_{p}\}$,
-which we give flat, non-informative priors (truncated at $\pm10$),
+which we give flat, non-informative priors (truncated at 10),
 and $\{\rho_{p}\}$, which we give uniform priors on the interval
@@
 
-\subsection{Latent Variable Model\label{sub:Latent-Variable-Model}}
+\paragraph*{Latent variable model\label{sub:Latent-Variable-Model}}
 
@@
 with latent Gaussian process(es) and a non-Gaussian likelihood is
-difficult \citep{Rue:Held:2005,Chri:etal:2006,Tan:Nott:2013}. To
+difficult \cite{Rue:Held:2005,Chri:etal:2006,Tan:Nott:2013}. To
 develop an algorithm, we make use of a latent variable representation
-for the multinomial probit model \citep{McCu:Ross:1994}. The representation
+for the multinomial probit model \cite{McCu:Ross:1994}. The representation
 introduces latent variables that allow one to develop a MCMC sampling
@@
 Suppose that compositional counts are available at a number of locations.
-At location $i$, a sample of size $n_{i}$ observations is collected,
-and each observation (i.e., each tree) can be classified into $P$
-distinct categories. For a given tree $j$ at location $i$, let $Y_{ij}$
+At location \emph{i}, a sample of size $n_{i}$ observations is collected,
+and each observation (i.e., each tree) can be classified into \emph{P}
+distinct categories. For a given tree \emph{j} at location \emph{i}, let $Y_{ij}$
 denote the response variable indicating the category. Let $Y_{ij}$
-be associated with $P$ latent variables $W_{ij1},...,W_{ijP}$ such
+be associated with \emph{P} latent variables $W_{ij1},...,W_{ijP}$ such
 that $Y_{ij}$ = $p$ if and only if $W_{ijp}={\displaystyle \max_{p'}\big\{ W_{ijp'}\big\}}$;
 in other words, the maximum of the set of latent variables $\{W_{ijp}\}{\displaystyle _{p=1}^{P}}$
-determines the category of observation $j$ at location $i$. The
+determines the category of observation \emph{j} at location \emph{i}. The
 final piece of the latent variable representation is the relationship
@@
 \[
-W_{ijp}\sim\mbox{N}(\alpha_{p}(s_{i}),1)
+W_{ijp}\sim\N(\alpha_{p}(s_{i}),1)
 \]
@@
 example with two locations that are neighbors and $P=2$ categories.
-Each tree $j$ at location $i$ is associated with two variables $W_{ij1}$
+Each tree \emph{j} at location \emph{i} is associated with two variables $W_{ij1}$
 and $W_{ij2}$, governed by the latent variables $\alpha_{1i}$ and
 $\alpha_{2i}$, respectively. Suppose that $\alpha_{1i}>\alpha_{2i}$
-for a given location $i$. Then this model implies that any tree $j$
-is more likely to be labeled 1 than 2 at location $i$. The difference
+for a given location \emph{i}. Then this model implies that any tree \emph{j}
+is more likely to be labeled 1 than 2 at location \emph{i}. The difference
 between $\alpha_{1i}$ and $\alpha_{2i}$ explains the \textit{difference}
-in probability of \textit{categories} 1 and 2 at location $i$, and
+in probability of \textit{categories} 1 and 2 at location \emph{i}, and
 the similarity between $\alpha_{p1}$ and $\alpha_{p2}$ explains
 the \textit{correlation} between the probabilities at \textit{locations}
-1 and 2 for category $p$.
-
-\noindent
-
-\noindent
-
-\subsection{Model for township data\label{sub:Model-for-township}}
+1 and 2 for category \emph{p}.
+
+\paragraph*{Model for township data}
+\label{sub:Model-for-township}
+
+\linenumbers
 
@@
 Specifically, $c_{tj}$ is the latent ``membership'' variable for
-tree $j$ in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$
+tree \emph{j} in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$
 is a discrete distribution that puts mass, $\psi_{ti}$, $i=1,\ldots,m$,
 proportional to the areal overlap between the township in which the
-tree is located and the $m$ grid cells, giving 
+tree is located and the \emph{m} grid cells, giving 
 \[
@@
 
-Using the latent variable representation, we have that $W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{c_{tj}}),1)$
-for tree $j$ in township $t$. In updating the other parameters in
+Using the latent variable representation, we have that $W_{tjp}\sim\N(\alpha_{p}(s_{c_{tj}}),1)$
+for tree \emph{j} in township \emph{t}. In updating the other parameters in
 the model during the MCMC (specifically the $\alpha$ values), we
@@
 
-\subsection{Computation}
-
-The \cite{McCu:Ross:1994} representation is convenient for MCMC sampling,
+\paragraph*{Computation}
+
+The representation of \cite{McCu:Ross:1994} is convenient for MCMC sampling,
 particularly in this high-dimensional spatial context, as it allows
@@
 given $\phi_{p}^{*}$ and the latent $W_{p}$ variables, where $W_{p}$
-is the vector of all $W_{ijp}$ values for taxon $p$: $W_{p}=\{W_{ijp}\},i=1,\ldots,m;j=1,\ldots,n_{i}$.
+is the vector of all $W_{ijp}$ values for taxon \emph{p}: $W_{p}=\{W_{ijp}\},i=1,\ldots,m;j=1,\ldots,n_{i}$.
 This is equivalent to sampling from the marginalized (with respect
@@
 For these various joint samples of hyperparameters and $\alpha_{p}$,
-we use adaptive Metropolis sampling \citep{Shab:Well:2011}.
-
-The full description of the MCMC sampling steps is provided in the
-Appendix. In addition, in the latent variable representation, $\theta_{p}(s)$
+we use adaptive Metropolis sampling \cite{Shab:Well:2011}.
+
+The full description of the MCMC sampling steps is provided in
+\nameref{SI}. In addition, in the latent variable representation, $\theta_{p}(s)$
 never appears explicitly and cannot be calculated in close form. Instead
 we use Monte Carlo integration over $W_{ijp},\, p=1,\ldots,P$ to
-estimate $\theta_{p}(s_{i})$, also described in the Appendix.
-
-The model is implemented in R \citep{R:2014} with core computational
-calculations coded in C++ using the \emph{Rcpp} package \citep{Edde:Fran:2011}.
+estimate $\theta_{p}(s_{i})$, also described in \nameref{SI}.
+
+The model is implemented in R \cite{R:2014} with core computational
+calculations coded in C++ using the \emph{Rcpp} package \cite{Edde:Fran:2011}.
 We also make extensive use of sparse matrix representations and algorithms,
-using the \emph{spam} package in R \citep{Furr:Sain:2010}. All code
+using the \emph{spam} package in R \cite{Furr:Sain:2010}. All code
 is available on Github, including pre- and post-processing code, at
@@
 
-\section{Model comparison\label{sec:Model-comparison}}
-
-\subsection{Design}
+\subsection*{Model comparison\label{sec:Model-comparison}}
+
+\paragraph*{Design}
 
@@
 be as narrow as possible while being calibrated such that the observations
-are consistent with the distribution \citep{Gnei:etal:2007}. When
+are consistent with the distribution \cite{Gnei:etal:2007}. When
 thinking in terms of prediction intervals as summaries of the predictive
@@
 experiment 1, we define $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$ as the count
-of all trees in held-out cell $i$ and for experiment 2, $Y_{i}$
+of all trees in held-out cell \emph{i} and for experiment 2, $Y_{i}$
 is the count of held-out individual trees in the cell, while $y_{ijp}$
 is an indicator variable taking value either 0 or 1 depending on whether
-the $j$th held-out tree in the $i$th cell is of taxon $p$. We calculated
+the \emph{j}th held-out tree in the \emph{i}th cell is of taxon \emph{p}. We calculated
 each of the metrics in two ways. First, we used the posterior mean
@@
 all the posterior samples.
+
 \begin{enumerate}
@@
 where $n=\sum_{i=1}^{m}n_{i}$ is the total number of held-out trees
-for a given experiment and $j$ indexes across held-out trees in cell
-$i$. 
+for a given experiment and \emph{j} indexes across held-out trees in cell
+\emph{i}. 
 \item Log predictive density: This metric takes the log of the density of
@@
 
-While in principle, this metric should be optimal \citep{Krnj:Drap:2014},
-it is very sensitive to small predictions near zero \citep{Gnei:etal:2007}.
+While in principle, this metric should be optimal \cite{Krnj:Drap:2014},
+it is very sensitive to small predictions near zero \cite{Gnei:etal:2007}.
 Even worse, our Monte Carlo estimation of $\theta$ used 10000 samples,
@@
 
-\subsection{Results}
+\paragraph*{Results}
 
@@
 
-\paragraph{Full cell hold-out experiment}
-
-For Experiment 1 for cells in the Minnesota subregion held out of
+For Experiment 1 (full cells held out) for cells in the Minnesota subregion held out of
 the fitting process, the CAR model outperforms the SPDE model based
@@
 \begin{table}
-\caption{Predictive ability based on several predictive metric criteria for
+\begin{adjustwidth}{-2.25in}{0in}
+\caption{{\bf Predictive ability based on several predictive metric criteria for
 the CAR and SPDE spatial models when holding out 95\% of entire cells
-of data in Minnesota. Smaller values are better.}
+of data in Minnesota.}}
 
@@
 {\small{}SPDE model} &
-{\small{}Posterior Prob. CAR < SPDE} &
+{\small{}Posterior Prob. CAR \textless  SPDE} &
 {\small{}CAR model} &
@@
 \end{tabular}
-
+\begin{flushleft}
+Smaller values are better for all metrics.
+\end{flushleft}
 \label{tab:score_cell_fivepercent}
+\end{adjustwidth}
 \end{table}
@@
 \begin{table}
-\caption{Coverage and length of prediction intervals for the CAR and SPDE spatial
+\begin{adjustwidth}{-2.25in}{0in}
+\caption{{\bf Coverage and length of prediction intervals for the CAR and SPDE spatial
 models when holding out 95\% of entire cells of data in Minnesota.
-Coverage values near 0.95 are optimal, while shorter intervals are
-better.}
+}}
 
@@
 \end{tabular}
-
+\begin{flushleft}
+Coverage values near 0.95 are optimal, while shorter intervals are
+better.
+\end{flushleft}
 \label{tab:coverage_cell_fivepercent}
+\end{adjustwidth}
 \end{table}
@@
 \begin{table}
-\caption{Predictive ability based on several predictive metric criteria for
+\begin{adjustwidth}{-2.25in}{0in}
+\caption{{\bf Predictive ability based on several predictive metric criteria for
 the CAR and SPDE spatial models when holding out 80\% of entire cells
-of data in Minnesota. Smaller values are better.}
+of data in Minnesota.}}
 
@@
 \end{tabular}
+\begin{flushleft}
+ Smaller values are better for all metrics.
+\end{flushleft}
 
 \label{tab:score_cell_20percent}
+\end{adjustwidth}
 \end{table}
@@
 \begin{table}
-\caption{Coverage and length of prediction intervals for the CAR and SPDE spatial
-models when holding out 80\% of entire cells of data in Minnesota.
-Coverage values near 0.95 are optimal, while shorter intervals are
-better.}
+\begin{adjustwidth}{-2.25in}{0in}
+\caption{{\bf Coverage and length of prediction intervals for the CAR and SPDE spatial
+models when holding out 80\% of entire cells of data in Minnesota.}}
 
@@
 \end{tabular}
+\begin{flushleft}
+Coverage values near 0.95 are optimal, while shorter intervals are
+better.
+\end{flushleft}
 
 \label{tab:coverage_cell_20percent}
+\end{adjustwidth}
 \end{table}
 
-\paragraph{Individual tree hold-out experiment}
-
-In Experiment 2, we have evidence (posterior probability of 0.93)
+In Experiment 2 (individual trees held out), we have evidence (posterior probability of 0.93)
 that the SPDE model is better based on the Brier score, but the Brier
@@
 \begin{table}
-\caption{Predictive ability based on several predictive metric criteria for
-the CAR and SPDE spatial models when holding out 5\% of trees. Smaller
-values are better.}
+\begin{adjustwidth}{-2.25in}{0in}
+\caption{{\bf Predictive ability based on several predictive metric criteria for
+the CAR and SPDE spatial models when holding out 5\% of trees.}}
 
@@
 \end{tabular}
-
+\begin{flushleft}
+ Smaller values are better for all metrics.
+\end{flushleft}
 \label{tab:score_tree}
+\end{adjustwidth}
 \end{table}
-
-\paragraph{Choice of spatial model}
 
@@
 
-\section{Data product\label{sec:Data-product}}
+\section*{Data Product\label{sec:Data-product}}
 
@@
 \begin{figure}
-\includegraphics[scale=0.75]{fig2}
-
-\hspace{4mm}\includegraphics[scale=0.16]{legendRaw}\hspace{4mm}\includegraphics[scale=0.16]{legendProp}\hspace{3.5mm}\includegraphics[scale=0.16]{legendSD}
-
-\caption{Empirical proportions from raw data (column 1), predictions in the
+% \includegraphics[scale=0.75]{fig2.pdf}
+
+% \hspace{4mm}\includegraphics[scale=0.16]{legendRaw.pdf}\hspace{4mm}\includegraphics[scale=0.16]{legendProp.pdf}\hspace{3.5mm}\includegraphics[scale=0.16]{legendSD.pdf}
+
+\caption{{\bf Raw data, predictions, and uncertainty for select taxa.} 
+\\Empirical proportions from raw data (column 1), predictions in the
 form of posterior means (column 2) and uncertainty estimates in the
 form of posterior standard deviations -- representing standard errors
-of prediction (column 3) for select taxa. \label{fig:select_maps}}
+of prediction (column 3). In raw data plots, white indicates no data.}
+\label{fig:select_maps}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[scale=0.8]{fig3}
-
-\caption{Predictions (posterior means) for all taxa over the entire domain.\label{fig:all_predictions}}
+% \includegraphics[scale=0.8]{fig3.pdf}
+
+\caption{{\bf Predictions (posterior means) for all taxa over the entire domain.}}
+\label{fig:all_predictions}
 
@@
 dynamic visualizations of the product using the Shiny tool are available
-at http://gandalf.berkeley.edu:3838/paciorek/setVegComp {[}this URL
-is temporary and will be updated{]}. The PalEON Project (in particular
+at http://gandalf.berkeley.edu:3838/paciorek/setVegComp [this URL
+is temporary and will be updated]. The PalEON Project (in particular
 the first author) will continue to maintain this product, releasing
@@
 
-\section{Discussion\label{sec:Discussion}}
+\section*{Discussion}\label{sec:Discussion}
 
@@
 is particularly critical at the prairie-forest transition in Minnesota,
-Wisconsin, Illinois, and into Indiana \citep{transeau1935prairie}.
+Wisconsin, Illinois, and into Indiana \cite{transeau1935prairie}.
 
@@
 and basal area is already available for Minnesota, Wisconsin, and
-northern Michigan \citep{goring2015composition}, based on the PLS
+northern Michigan \cite{goring2015composition}, based on the PLS
 data. An extension to southern Michigan, Illinois, and Indiana is
@@
 
-\section*{Author contributions}
-
-CJP, AT, and JZ developed the statistical model and code. CJP carried
-out the model comparison and created the data product. CJP wrote the
-paper based on an initial draft by AT and with feedback and editing
-from SJG, CVC, JWW, DJM, JAP, JZ and JSM. SJG, JAP, CVC, DJM, JSM,
-and JWW led the processing and analysis of the PLS and TPS data and
-assisted with interpretation of results.
-
-\section*{Acknowledgments}
-
-The authors are deeply indebted to all of the researchers over the
-years who have preserved, collected, and digitized survey records,
-in particular John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed
-Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
-We thank Madeline Ruid, Ben Seliger, Morgan Ross and Daniel Handel
-for processing of the southern Michigan data. Indiana and Illinois
-data were made possible through the hard work of many Notre Dame undergraduates
-in the McLachlan lab. This work was carried out by the PalEON Project
-with support from the National Science Foundation MacroSystems Program
-through grants EF-1065702, EF-1065656, DEB-1241874 and DEB-1241868
-and from the Notre Dame Environmental Change Initiative.
-
-\bibliographystyle{/accounts/gen/vis/paciorek/latex/RSSstylefile/Chicago}
-
-% \bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
-\bibliography{paperArxiv}
-
-\section{Appendix}
-
-\subsection{MCMC details}
-
-Define $\bar{w}_{ip}=\frac{1}{n_{i}}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
-be the average of the $W$ values for the $p$th taxon in the $i$th
-grid cell and $\bar{w}_{p}=\{\bar{w}_{ip}\},$ $i=1,\ldots,m$. Let
+\section*{S1 Appendix}
+\label{SI}
+
+\subsection*{MCMC details}
+
+{\bf Define $\bar{w}_{ip}=\frac{1}{n_{i}}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
+be the average of the $W$ values for the \emph{p}th taxon in the \emph{i}th
+grid cell and $\bar{w}_{p}=\{\bar{w}_{ip}\},$ $i=1,\ldots,m$.} Let
 $A$ be a diagonal matrix where $A_{ii}$ is the number of trees in
-the $i$th grid cell. When there are no trees in a grid cell, $\bar{w}_{ip}=0$
+the \emph{i}th grid cell. When there are no trees in a grid cell, $\bar{w}_{ip}=0$
 and $A_{ii}=0$. For the township data, at each iteration, based on
@@
 \mbox{TN}\big(-\infty,w_{ijy_{ij}},\alpha_{p}(s_{i}),1\big), & \mbox{if }p\mbox{ \ensuremath{\neq y_{ij}}}
-\end{cases}\label{sampW}
+\end{cases}
 \end{align}
 In essence, the truncation value is determined by the taxon of the
-$j$th tree. For a given $p$, the $W$ values for all trees in all
+\emph{j}th tree. For a given \emph{p}, the $W$ values for all trees in all
 cells can be sampled in parallel.
@@
 \begin{align}
-\alpha_{p} & \sim\mbox{N}\bigg(\Big(A+Q_{p}\Big)^{-1}A\bar{w}_{p},\Big(A+Q_{p}\Big)^{-1}\bigg).\label{sampalpha}
+\alpha_{p} & \sim\N\bigg(\Big(A+Q_{p}\Big)^{-1}A\bar{w}_{p},\Big(A+Q_{p}\Big)^{-1}\bigg).
 \end{align}
@@
 Metropolis-Hastings proposal, with adaptation of the proposal (co)variance
-\citep{Shab:Well:2011}. The proposal distribution for $\phi_{p}$
+\cite{Shab:Well:2011}. The proposal distribution for $\phi_{p}$
 is a normal distribution (bivariate for $\phi_{p}=(\log\sigma_{p},\log\rho_{p})$).
@@
 
-For the township-level data, for a given tree $j$ in township $t$,
+For the township-level data, for a given tree \emph{j} in township $t$,
 we draw the latent tree membership variable, $c_{tj}\in\{1,\ldots,m\}$,
@@
 that $c_{tj}=i$, namely the product of independent normal densities,
-$W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{i}),1)$, over $p=1,\ldots,P$.
+$W_{tjp}\sim\N(\alpha_{p}(s_{i}),1)$, over $p=1,\ldots,P$.
 Thus the posterior reweights the prior based on how consistent the
@@
 
-\subsection{Estimating $\theta_{p}(s)$ via Monte Carlo integration}
-
-In the latent variable representation, $\theta_{p}(s)$ never appears
-explicitly and cannot be calculated in closed form. Instead we use
+\subsection*{Estimating $\theta_{p}(s)$ via Monte Carlo integration}
+\label{SI1}
+{\bf In the latent variable representation, $\theta_{p}(s)$ never appears
+explicitly and cannot be calculated in closed form.} Instead we use
 Monte Carlo integration over $W_{ijp},\, p=1,\ldots,P$ to estimate
 $\theta_{p}(s_{i})$. The quantity $\theta_{p}(s_{i})=\mbox{Prob}(W_{ijp}={\displaystyle \max_{p^{*}}W_{ijp^{*}})}$
-defines the probability of taxon $p$ at grid cell $i$. This requires
+defines the probability of taxon \emph{p} at grid cell \emph{i}. This requires
 one to choose the number of Monte Carlo samples, which we set at 10000,
@@
 \[
-W_{itp}^{(k)}\sim\mbox{N}(\alpha_{p}^{(k)}(s_{i}),1),\, p=1,\ldots,P
+W_{itp}^{(k)}\sim\N(\alpha_{p}^{(k)}(s_{i}),1),\, p=1,\ldots,P
 \]
@@
 the proportion of times that the maximum of $W_{itp},\, p=1,\ldots,P$
-corresponds to taxon $p$. Considering $\theta_{p}^{(k)}(s_{i}),\, k=1,\ldots,K$,
+corresponds to taxon \emph{p}. Considering $\theta_{p}^{(k)}(s_{i}),\, k=1,\ldots,K$,
 we have a sample from the posterior of $\theta_{p}(s_{i})$.
 
+\section*{Acknowledgments}
+The authors are deeply indebted to all of the researchers over the
+years who have preserved, collected, and digitized survey records,
+in particular John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed
+Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
+We thank Madeline Ruid, Ben Seliger, Morgan Ross and Daniel Handel
+for processing of the southern Michigan data. Indiana and Illinois
+data were made possible through the hard work of many Notre Dame undergraduates
+in the McLachlan lab. 
+% This work was carried out by the PalEON Project
+%with support from the National Science Foundation MacroSystems Program
+%through grants EF-1065702, EF-1065656, DEB-1241874 and DEB-1241868
+%and from the Notre Dame Environmental Change Initiative.
+
+\nolinenumbers
+
+\section*{References}
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% OR
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here.
+%
+
+\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
+
+%\begin{thebibliography}{10}
+
+%\end{thebibliography}
+
 \end{document}

### 2015-09-05 paper.R

@@ -53,6 +53,7 @@ d <- theme_clean(d)
 setwd(paperDir)
 
 pdf('fig1.pdf', width=4.85, height=3)
+#postscript('fig1.eps', width=4.85, height=3)
 print(d)
 dev.off()
 
@@ -87,6 +88,7 @@ for( taxon in focalTaxa ) {
 setwd(paperDir)
 
 pdf('fig2.pdf', width=7, height=10)
+#postscript('fig2.eps', width=7, height=10)
 do.call(grid.arrange, c(figs, nrow = length(focalTaxa), ncol = 3))
 dev.off()
 
@@ -94,7 +96,8 @@ scaling = .5
 ix=1
 iy=seq(0,1,len=300)
 iz=matrix(iy,nc=300)
-pdf(file='legendRaw.pdf',width=5/scaling,height=.7/scaling,paper="special")
+#postscript(file='legendRaw.pdf',width=5/scaling,height=.7/scaling,paper="special")
+pdf(file='legendRaw.eps',width=5/scaling,height=.7/scaling,paper="special")
 par(fig=c(0,1,0,1),mai=c(0.6,.2,0,.3),mgp=c(1.8,.7,0),cex.axis=1.5)
 cols <- rev(terrain.colors(length(propBreaks)-1))
 cols[1] <- terrain.colors(40)[39]
@@ -109,6 +112,7 @@ ix=1
 iy=seq(0,1,len=300)
 iz=matrix(iy,nc=300)
 pdf(file='legendProp.pdf',width=5/scaling,height=.7/scaling,paper="special")
+#postscript(file='legendProp.eps',width=5/scaling,height=.7/scaling,paper="special")
 par(fig=c(0,1,0,1),mai=c(0.6,.2,0,.3),mgp=c(1.8,.7,0),cex.axis=1.5)
 image(iy,ix,t(iz), yaxt = "n", xlab = "", xaxt='n', 
             ylab = "", col = rev(terrain.colors(length(propBreaks)-1)))
@@ -121,6 +125,7 @@ ix=1
 iy=seq(0,1,len=300)
 iz=matrix(iy,nc=300)
 pdf(file='legendSD.pdf',width=5/scaling,height=.7/scaling,paper="special")
+pdf(file='legendSD.eps',width=5/scaling,height=.7/scaling,paper="special")
 par(fig=c(0,1,0,1),mai=c(0.6,.2,0,.3),mgp=c(1.8,.7,0),cex.axis=1.5)
[…]

### 2016-02-03 arxiv/paperArxiv.tex
revised and added for plos revision january 2016

@@
 
-\lyxaddress{\textsuperscript{6}Department of Forestry and Wildlife Ecology,
+\lyxaddress{\textsuperscript{6}Department of Forest and Wildlife Ecology,
 University of Wisconsin, Madison, Wisconsin, USA}
@@
 8 km grid across the entire domain. The statistical model is designed
-to handle data from both the regular grid and the irregularly-sized
+to handle data from both the regular grid and the irregularly-shaped
 townships and allows us to estimate composition at locations with
@@
 that could be calculated using the full population of all the trees
-in grid cell or township.
+in a grid cell or township.
 \end{enumerate}
@@
 autoregressive model \citep{Bane:etal:2004} and a Gaussian Markov
-random field (MRF) approximation to a Gaussian process with Matern
+random field (MRF) approximation to a Gaussian process with \matern
 covariance \citep{Lind:etal:2011}.
@@
 to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second spatial
-model is the Lindgren approximation for \matern-based GPs with $\nu=1$.
+model is this Lindgren approximation for \matern-based GPs with $\nu=1$.
 To implement the Lindgren model, one modifies the $Q$ matrix defined
@@
 grid. This extension introduces a new set of latent variables, one
-per tree, that indicate the grid cell in which the tree is located
+per tree,  that indicate  the grid cells in which the trees are located
 and can be sampled within the MCMC as additional unknown parameters.
@@
 is available on Github, including pre- and post-processing code, at
-https://github.com/Paleon-Project/composition.
+https://github.com/PalEON-Project/composition.
 
@@
 Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
-We thank Madeline Ruid, Ben Seliger, Morgan Ross and Daniel Handel
+We thank Madeline Ruid, Ben Seliger, Morgan Ripp and Daniel Handel
 for processing of the southern Michigan data. Indiana and Illinois
@@
 Define $\bar{w}_{ip}=\frac{1}{n_{i}}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
-be the average of the $W$ values for the $p$th taxon in the $i$th
+as the average of the $W$ values for the $p$th taxon in the $i$th
 grid cell and $\bar{w}_{p}=\{\bar{w}_{ip}\},$ $i=1,\ldots,m$. Let

### 2016-02-03 paper.tex
revised and added for plos revision january 2016

@@
 \geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
+\usepackage{color}
 \usepackage{array}
@@
 \usepackage[authoryear]{natbib}
+\usepackage[unicode=true]
+ {hyperref}
 
@@
 States \\
-at the time of European settlement}
+at the time of Euro-American settlement}
+
+\author{Christopher J. Paciorek\textsuperscript{1,{*}}, Simon J. Goring\textsuperscript{2,\&},
+Andrew L. Thurman\textsuperscript{3,\&}, \\
+Charles V. Cogbill\textsuperscript{4}, John W. Williams\textsuperscript{2,3},
+David J. Mladenoff\textsuperscript{6}, \\
+Jody A. Peters\textsuperscript{7}, Jun Zhu\textsuperscript{8}, and
+Jason S. McLachlan\textsuperscript{7}}
+
 \maketitle
+
+\lyxaddress{\textsuperscript{1}Department of Statistics, University of California,
+Berkeley, California, USA}
+
+\lyxaddress{\textsuperscript{2}Department of Geography, University of Wisconsin,
+Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{3}Department of Statistics, University of Iowa,
+Iowa City, Iowa, USA}
+
+\lyxaddress{\textsuperscript{4}Harvard Forest, Harvard University, Petersham,
+Massachusetts, USA}
+
+\lyxaddress{\textsuperscript{5}Center for Climatic Research, University of Wisconsin,
+Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{6}Department of Forest and Wildlife Ecology, University
+of Wisconsin, Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{7}Department of Biological Sciences, University
+of Notre Dame, Notre Dame, Indiana, USA}
+
+\lyxaddress{\textsuperscript{8}Department of Statistics, University of Wisconsin,
+Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{{*}}Corresponding author; E-mail: paciorek@stat.berkeley.edu}
+
+\lyxaddress{\textsuperscript{\&}These authors contributed equally to this work.
+\newpage}
 \begin{quote}
-Christopher J. Paciorek 1{*}, Simon J. Goring 2\&, Andrew Thurman
-3\&, Charles V. Cogbill 4, \\
-John W. Williams 3,5, David J. Mladenoff 6, Jody A. Peters 7, Jun
-Zhu 8, \\
-and Jason S. McLachlan 7 \\
-{[}author order to be discussed{]}\\
-{[}please tell me if you do/don't want your middle initial{]}
-
-{[}let me know if there are any issues with how I've listed your affiliation{]}
-
-1 Department of Statistics, University of California, Berkeley, California,
-USA
-
-2 Department of Geography, University of Wisconsin, Madison, Wisconsin,
-USA
-
-3 Department of Statistics, University of Iowa, Iowa City, Iowa, USA
-
-4 Harvard Forest, Harvard University, Petersham, Massachusetts, USA
-
-5 Center for Climatic Research, University of Wisconsin, Madison,
-Wisconsin, USA
-
-6 Department of Forestry and Wildlife Ecology, University of Wisconsin,
-Madison, Wisconsin, USA
-
-7 Department of Biological Sciences, University of Notre Dame, Notre
-Dame, Indiana, USA
-
-8 Department of Statistics, University of Wisconsin, Madison, Wisconsin,
-USA
-
-{*} Corresponding author
-
-E-mail paciorek@stat.berkeley.edu (CJP)
-
-\& These authors contributed equally to this work.
-
-Short title: Settlement-era northeast U.S. tree composition
+
 \end{quote}
 
+%
 \begin{abstract}
-We present a data product of the estimated composition of tree taxa
-at the time of European settlement of the northeastern United States
-and the statistical methodology used to produce the product. Composition
+We present a gridded 8 km-resolution data product of the estimated
+composition of tree taxa at the time of Euro-American settlement of
+the northeastern United States and the statistical methodology used
+to produce the product from trees recorded by land surveyors. Composition
 is defined as the proportion of stems larger than approximately 20
-cm diameter at breast height in 22 taxonomic groupings, generally
-at the genus level. The data come from settlement survey records that
-provide raw data that are transcribed and then aggregated spatially,
-giving count data. The domain is divided into two regions, eastern
-(Maine to Ohio) and western (Indiana to Minnesota). Public Land Survey
-point data in the western region are aggregated to a regular 8 km
-grid, while data in the eastern region, from Town Proprietor Surveys,
-are aggregated at the township level in irregularly-shaped local administrative
-units. The product is based on a Bayesian statistical model fit to
-the count data that estimates composition on a regular 8 km grid.
-The statistical model allows us to estimate composition at locations
-with no data and to smooth over noise caused by limited counts in
-locations with data. Critically, it also allows us to quantify uncertainty
-in our composition estimates. We expect this data product to be useful
+cm diameter at breast height for 22 tree taxa, generally at the genus
+level. The data come from settlement-era public survey records that
+are transcribed and then aggregated spatially, giving count data.
+The domain is divided into two regions, eastern (Maine to Ohio) and
+midwestern (Indiana to Minnesota). Public Land Survey point data in
+the midwestern region (ca. 0.8-km resolution) are aggregated to a
+regular 8 km grid, while data in the eastern region, from Town Proprietor
+Surveys, are aggregated at the township level in irregularly-shaped
+local administrative units. The product is based on a Bayesian statistical
+model fit to the count data that estimates composition on a regular
+8 km grid across the entire domain. The statistical model is designed
+to handle data from both the regular grid and the irregularly-shaped
+townships and allows us to estimate composition at locations with
+no data and to smooth over noise caused by limited counts in locations
+with data. Critically, the model also allows us to quantify uncertainty
+in our composition estimates, making the product suitable for applications
+employing data assimilation. We expect this data product to be useful
 for understanding the state of vegetation in the northeastern United
-States prior to large-scale European settlement. In addition to specific
-regional questions, the data product can also serve as a baseline
-against which to investigate how forests and ecosystems change after
-intensive settlement.
+States prior to large-scale Euro-American settlement. In addition
+to specific regional questions, the data product can also serve as
+a baseline against which to investigate how forests and ecosystems
+change after intensive settlement. The data product is being made
+available at the NIS data portal as version 0.4.
+
+Keywords: biogeography, species composition, old-growth forests, spatial
+modeling, Bayesian statistical model, vegetation mapping
 \end{abstract}
@@
 conservation management, to understand ecosystem processes at decadal
-and centennial scales, and, particularly in regions with widespread
-land use change, to understand the extent to which forests after conversion
-and regeneration differ from the original forest cover.
+and centennial scales, to track forest responses to shifting climates,
+and, particularly in regions with widespread land use change, to understand
+the extent to which forests after conversion and regeneration differ
+from the original forest cover.
 
 Euro-American settlement and subsequent land use change occurred in
-a time-transient fashion across North America, and land surveys that
-provide vegetation information followed a similar pattern. Early surveys
-(from 1620 until 1825) in the northeastern United States provide areally-aggregated
-data at the township level \citep{cogbill2002forests,thompson2013four};
-we refer to these as the Town Proprietor Survey (TPS). Later surveys,
-after the establishment of the U.S. Public Land Survey System (PLS)
-provide point-level data along a regular grid, with one mile spacing
-during the period 1832 to 1907 \citep{bourdo1956review,schulte2001original,goring2015composition}.
+a time-transient fashion across North America and were accompanied
+by land surveys needed to demarcate land for land tenure and use.
+Various systems were used by surveyors to locate legal boundary markers,
+usually by recording and marking trees adjacent to survey markers.
+These data provide vegetation information that can be mapped and used
+quantitatively to represent the period of settlement. Early surveys
+(from 1620 until 1825) in the northeastern United States provide spatially-aggregated
+data at the township level \citep{Cogb:etal:2002,thompson2013four},
+with typical township size on the order of 200 $\mbox{km}^{2}$ and
+no information about the locations of individual trees; we refer to
+these as the Town Proprietor Survey (TPS). Later surveys after the
+establishment of the U.S. Public Land Survey System (PLS) by the General
+Land Office (GLO) provide point-level data along a regular grid, with
+one-half mile (800 m) spacing, for Ohio and westward during the period
+1785 to 1907 \citep{bourdo1956review,pattison1957beginnings,schulte2001original,goring2015composition}.
+At each point 2-4 trees were identified, and the common name, diameter
+at breast height, and distance and bearing from the point were recorded.
 Survey instructions during the PLS varied through time and by point
-type. Accounting for this requires the application of spatially-varying
-correction factors \citep{cogbill2015corrections,goring2015composition}
-to accurately assess stem density, basal area and biomass from the
-early settlement records, but the impact on composition estimates
-is limited. Both types of surveys provide composition of major tree
-species but using sometimes inconsistent common names.
+type. Accounting for this variation requires data screening to maximize
+consistency among points and the application of spatially-varying
+correction factors \citet{goring2015composition} to accurately assess
+tree stem density, basal area and biomass from the early settlement
+records, but the impact on composition estimates is limited \citep{liu2011broadscale}.
+Surveyors sometimes used ambiguous common names, which requires matching
+to scientific names and standardization \citep{mladenoff2002narrowing,goring2015composition}.
 
@@
 or smaller domains \citep{friedman2005regional,rhemtulla2009historical}
-and with varying spatial resolution, from townships (36 square miles)
+and with various spatial resolutions, from townships (36 square miles)
 to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
-provide a dataset for the upper Midwest that is resolved to an 8 x
-8 km grid size, providing broad spatial coverage at a spatial scale
-that can be compared to the modern forests using Forest Inventory
+provide a new dataset of forest composition, biomass, and stem density
+based on PLS data for the upper Midwest that is resolved to an 8 km
+by 8 km grid cell scale, providing broad spatial coverage at a spatial
+scale that can be compared to modern forests using Forest Inventory
 and Analysis products \citep{gray2012forest}. Combined with additional,
-coarsely-sampled PLS data from Illinois and Indiana and with the TPS
-data, this gives us raw data for much of the northeastern United States.
-However, there are several limitations of using the raw data that
-can be alleviated by the use of a statistical model to develop a statistically-estimated
-data product. First, the PLS and TPS data provide only estimates of
+coarsely-sampled PLS data from Illinois and Indiana, newly-digitized
+data from southern Michigan, and with the TPS data, this gives us
+raw data for much of the northeastern United States. However, there
+are several limitations of using the raw data that can be alleviated
+by the use of a statistical model to develop a statistically-estimated
+data product. First, the PLS and TPS data only provide estimates of
 within-cell variance that do not account for information from nearby
-locations. Second, the available digitized data from Illinois and
-Indiana represent a small fraction of those states and missing townships
-are common in the TPS data. Our statistical model allows us to provide
-a spatially-complete data product of settlement-era tree composition
-with uncertainty across the northeastern U.S.
+locations. Second, there are data gaps: the available digitized data
+from Illinois and Indiana represent a small fraction of those states,
+and missing townships are common in the TPS data. Third, the TPS and
+PLS data have fundamentally different sampling design and spatial
+resolution. Our statistical model allows us to provide a spatially-complete
+data product of settlement-era tree composition for a common 8 km
+grid with uncertainty across the northeastern U.S.
 
@@
 we describe the final data product. In Section \ref{sec:Discussion}
-we discuss limitations of the statistical model, and we list related
-data products under development.
+we discuss \textcolor{red}{the uncertainties estimated by }and the
+limitations of the statistical model, and we list related data products
+under development.
 
@@
 
-The raw data were obtained from survey records collated from across
-the northeastern U.S. by a number of researchers. For the states of
-Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
-subdomain), data are available at PLS survey point locations and have
-been aggregated to a regular 8 km grid in the Albers projection. (Note
-that for Indiana and Illinois, at the moment trees are associated
-with township centroids and then assigned to 8 km grid cells based
-on the centroid but in the near future we will have point locations
-available for each tree.) For the states of Ohio, Pennsylvania, New
-Jersey, New York and the six New England states (the eastern subdomain),
-data are aggregated at the township level. There are also data from
-a single township in Quebec and a single township in northern Delaware.
-Data are essentially complete in Minnesota, Wisconsin and Michigan,
-but data in Illinois and Indiana represent a sample of the full set
-of grid cells, with survey record transcription ongoing. Data for
-the remaining states are available for a subset of the full set of
-townships covering the domain. Fig. \ref{fig:Spatial-domain} shows
-the domain, indicating the grid cells and townships with data.
+The raw data were obtained from land division survey records collated
+and digitized from across the northeastern U.S. by a number of researchers
+(Fig. \ref{fig:Spatial-domain}). For the states of Minnesota, Wisconsin,
+Illinois, Indiana, and Michigan (the midwestern subdomain), digitized
+data are available at PLS survey point locations and have been aggregated
+to a regular 8 km grid in the Albers projection. (Note that for Indiana
+and Illinois, at the moment trees are associated with township centroids
+and then assigned to 8 km grid cells based on the centroid, but in
+the near future we will have point locations available for each tree.)
+For the states of Ohio, Pennsylvania, New Jersey, New York and the
+six New England states (the eastern subdomain), data are aggregated
+at the township level. \textcolor{red}{We make predictions for all
+of the states listed above; these constitute our core domain. }There
+are also data from a single township in Quebec and a single township
+in northern Delaware\textcolor{red}{; these data help inform predictions
+in nearby locations within our core domain, but predictions are not
+made for Quebec or Delaware}. Digitization of PLS data in Minnesota,
+Wisconsin and Michigan is essentially complete, with PLS data for
+nearly all 8 km grid cells, but data in Illinois and Indiana represent
+a sample of the full set of grid cells, with survey record transcription
+ongoing. Data for the eastern states are available for a subset of
+the full set of townships covering the domain; the TPS data for some
+townships were lost, incomplete, or have not been located \citep{Cogb:etal:2002}.
 
@@
 \caption{Spatial domain of the northeastern United States, with locations with
-data shown in gray. Locations are grid cells in western portion and
-townships in eastern portion. In addition to locations without data
-being indicated in white, grid cells completely covered in water are
-white (e.g., a few locations in the northwestern portion of the domain
-in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
+data shown in gray. Locations are grid cells in midwestern portion
+and townships in eastern portion. In addition to locations without
+data being indicated in white, grid cells completely covered in water
+are white (e.g., a few locations in the northwestern portion of the
+domain in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
 \end{figure}
@@
 European colonists (before U.S. independence) and the United States
-settled what is now the northeastern United States. Our estimates
-are for the period of settlement represented by the survey data and
-therefore are time-transgressive; they do not represent any single
-point in time across the domain, but rather the state of the landscape
-at the time just before settlement occurred \citep{Whit:1996,Cogb:etal:2002}.
+settled what is now the northeastern and midwestern United States.
+Our estimates are for the period of settlement represented by the
+survey data and therefore are time-transgressive; they do not represent
+any single point in time across the domain, but rather the state of
+the landscape at the time just prior to widespread Euro-American settlement
+and land use \citep{Whit:1996,Cogb:etal:2002}. These forest composition
+datasets do include the effects of Native American land use and early
+Euro-American settlement activities \citep[e.g.,][]{Blac:etal:2006},
+but it is likely that the imprint of this earlier land use is highly
+concentrated rather than spatially extensive \citep{munoz2014defining}.
 
 Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
-data and processing steps are available in \cite{Gori:etal:2015}.
-The lower Midwest (Illinois, Indiana) data were purchased from the
-Indiana State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
+data and processing steps are available \cite{goring2015composition};
+key elements include the use of only corner points, the use of only
+the two closest trees at each corner point, spatially-varying correction
+factors for sampling effort, and a standardized taxonomy table. The
+lower Midwest (Illinois, Indiana) data were purchased from the Indiana
+State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
 Illinois) and processed using similar steps as for the upper Midwest
-data. The TPS data were compiled by C.V. Cogbill from a myriad of
-archival sources representing land division surveys conducted in connection
-with local settlement. The complete raw vegetation data are available
-through the NIS Data Portal (accession number and version to be inserted
-when available). The aggregation into taxonomic groups is primarily
-at the genus level, but is at the species level in some cases of monospecific
-genera. We model the following 22 taxa plus an ``other hardwood''
-category: Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash
-(\emph{Fraxinus spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus
-grandifolia)}, Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa
-sylvatica} and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
+data. Digitization of the Illinois and Indiana data is still underway,
+so many grid cells contained no data at the time the statistical model
+was fit. Note that the number of trees per grid cell varies depending
+on the number of survey points in a cell, with an average of 124 trees
+per cell. The gridded data at the 8 km resolution for the midwest
+subdomain are available through the NIS data portal \textcolor{red}{as
+DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5} \textcolor{red}{\citep{Gori:etal:data:2016}}.
+The TPS data were compiled by C.V. Cogbill from a myriad of archival
+sources representing land division surveys conducted in connection
+with local settlement\textcolor{red}{{} and are available through the
+NIS data portal as DOI:10.5072/FK2/4d44aebe71dbf65290d8ce1bde03598b
+and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \citep{Cogb:dataNE:2016,Cogb:dataOH:2016}}.
+
+The aggregation into taxonomic groups is primarily at the genus level
+but is at the species level in some cases of monospecific genera.
+We model the following 22 taxa plus an ``other hardwood'' category:
+Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash (\emph{Fraxinus
+spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus grandifolia)},
+Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa sylvatica}
+and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
 virginiana} and \emph{Thuja occidentalis}), Cherry (\emph{Prunus spp.}),
 Chestnut (\emph{Castanea dentata}), Dogwood (\emph{Cornus spp.}),
-Elm (\emph{Ulmus spp.}), Fir (\emph{Abies spp.}), Hemlock (\emph{Tsuga
+Elm (\emph{Ulmus spp.}), Fir (\emph{Abies balsamea}), Hemlock (\emph{Tsuga
 canadensis}), Hickory (\emph{Carya spp.}), Ironwood (\emph{Carpinus
@@
 Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), and
-Walnut (\emph{Juglans nigra}). Note that in several cases (black gum/sweet
+Walnut (\emph{Juglans spp.}). Note that in several cases (black gum/sweet
 gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
 in the common tree names used by surveyors, a group represents trees
-from different families and even orders. For the western subdomain
+from different genera or even families. For the midwestern subdomain
 we do not fit statistical models for Atlantic white cedar and chestnut
-as these have 0 and 7 trees present, respectively. The taxa grouped
-into the other hardwood category are those for which fewer than roughly
-2000 trees were present in the dataset; however we include Atlantic
-white cedar explicitly despite it only having 336 trees in the dataset.
-Although surveyors avoided using small trees, there was no consistent
-lower diameter limit. The data, however, roughly represent trees greater
-than 8 inches (\textasciitilde{}20 cm) diameter at breast height (dbh),
-but with some trees as small as 3 inches dbh. In Indiana and Illinois
-there are also 402 trees and 29 trees, respectively, that are between
-1 and 4 inches dbh, but these will be excluded in the next version
-of the product {[}waiting on Simon to see if he stripped them out{]}.
-
-There are approximately 860,000 trees from the western subdomain and
-420,000 trees from the eastern subdomain. In the western subdomain,
+as these species have 0 and 7 trees present, respectively. The taxa
+grouped into the other hardwood category are those for which fewer
+than roughly 2000 trees were present in the dataset; however, we include
+Atlantic white cedar explicitly despite it only having 336 trees in
+the dataset because of specific ecological interest in Atlantic white
+cedar wetlands.
+
+Diameters are only recorded in the PLS data. Although surveyors avoided
+using small trees, there was no consistent lower diameter limit. The
+PLS data generally represent trees greater than 8 inches (\textasciitilde{}20
+cm) diameter at breast height (dbh), but with some trees as small
+as 1 inch dbh (smaller trees were much more common in far northern
+Minnesota). TPS data have no information about dbh, but the trees
+were large enough to blaze and are presumed to be relatively large
+trees useful for marking property boundaries.
+
+There are approximately 860,000 trees in the midwestern subdomain
+and 420,000 trees in the eastern subdomain. In the midwestern subdomain,
 oak is the most common taxon and pine the second most common, while
@@
 
-Our domain is a rectangle covering all of the states, in the Albers
-projection (NAD 1983 Albers Great Lakes and St. Lawrence), with the
-rectangle split into 8 km cells, arranged in a 296 by 180 grid of
-cells, with the centroid of the cell in the southwest corner located
-at (-71000, 58000). For the modeling of the western subdomain we use
-the western-most 146 by 180 grid of cells. For the modeling of the
-eastern subdomain we use the eastern-most 180 by 180 grid of cells
-and then omit 23 cells in the north and 17 cells in the south outside
-the states containing data.
+Our domain is a rectangle covering all of the states using a metric
+Albers (Great Lakes and St. Lawrence) projection (PROJ4: EPSG:3175),
+with the rectangle split into 8 km cells, arranged in a 296 by 180
+grid of cells, with the centroid of the cell in the southwest corner
+located at (-71000 m, 58000 m). For the midwestern subdomain we use
+the western-most 146 by 180 grid of cells\textcolor{red}{{} when fitting
+the statistical models}. For the eastern subdomain we use the eastern-most
+180 by 180 grid of cells and then omit 23 rows of cells in the north
+and 17 rows of cells in the south, as these grid cells are outside
+of the states containing data.
 
@@
 the empirical proportions represent estimates of the true proportions
-that could be calculated using full population of all the trees in
-an areal region.
+that could be calculated using the full population of all the trees
+in a grid cell or township.
 \end{enumerate}
+\textcolor{red}{At a high level, the Bayesian statistical model estimates
+composition across the domain, even in locations with sparse or no
+data, by combining the raw composition data with the assumption that
+composition varies in a smooth spatial fashion across the domain.
+The information in the data is quantified by the data model, also
+known as the likelihood. The assumption of smoothness is built into
+the model by representing the true unknown spatially-varying composition
+using a statistical spatial process representation that induces smoothing
+of estimates across nearby locations. This spatial process representation
+is a form of prior distribution and is a function of model parameters
+called hyperparameters that determine the correlation structure of
+the process and are also estimated based on the data. }
+
 The result of fitting the Bayesian model via Markov chain Monte Carlo
@@
 cells. These samples are the data product (described further in the
-Data Product section) and can then be used in subsequent analyses.
+Section \ref{sec:Data-product}) and can be used in subsequent analyses.
 The mean and standard deviation of the samples for each pair of cell
@@
 the estimate.
+
+\textcolor{red}{In the remainder of this section we provide the technical
+specification of the model and of the computations involved in fitting
+the model.}
 
@@
 of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.
+\textcolor{red}{A multinomial probit model is similar to logistic
+regression, used for modeling a binary outcome based on an underlying
+probability the outcome will occur, but generalizes to modeling a
+categorical variable based on probabilities for each category. }
 
 The critical component of the statistical model is the representation
-of $\alpha_{p}(s)$ as a spatial process. This process is a prior
-structure that serves to smooth across noise in the observations and
-allows for interpolation to locations with no data. Apart from the
-sum to one constraints, the taxa are considered to be independent
-in the prior. We did not want to impose any structure that ties the
-different taxa together, as any correlation will likely vary across
-space.
+of $\alpha_{p}(s)$ as a spatial process. \textcolor{red}{A spatial
+process is a statistical representation that models spatially-correlated
+values. It provides} a prior structure that serves to smooth across
+noise in the observations and allows for prediction at locations based
+on information from nearby locations, including interpolation to locations
+with no data. Apart from the sum-to-one constraint, the taxa are considered
+to be independent in the prior. We did not want to impose any structure
+that ties the different taxa together, as any correlation will likely
+vary across space.
 
@@
 structure of the $\alpha_{p}(s)$ processes, a standard conditional
-autoregressive model \citep{Bane:etal:2003} and a Gaussian Markov
-random field (MRF) approximation to a Gaussian process with Matern
-covariance \citep{Lind:etal:2011}.
+autoregressive model \citep{Bane:etal:2004} and a Gaussian Markov
+random field (MRF) approximation to a Gaussian process with \matern
+covariance \citep{Lind:etal:2011}. \textcolor{red}{These models are
+specific statistical formulations of spatial processes that represent
+spatial correlation by defining neighborhoods around each location
+that are used to help inform predictions at the location. }
 
@@
 
-MRF models work directly with the precision matrix of the values of
-the spatial process, so calculation of the prior density of $\alpha_{p}$
-is computationally simple \citep{Rue:Held:2005}, but in situations
-where the likelihood is not normal, it can be difficult to set up
-effective MCMC algorithms that are able to move in the high-dimensional
-space of $\alpha_{p}$. The latent variable representation of Section
-\ref{sub:Model-for-township} helps to alleviate this problem.
+MRF models \textcolor{red}{represent the neighborhood information
+}by working directly with the precision matrix \textcolor{red}{(the
+inverse of the covariance matrix)} of the values of the spatial process,
+so calculation of the prior density of $\alpha_{p}$ is computationally
+simple \citep{Rue:Held:2005}. \textcolor{red}{However,} in situations
+\textcolor{red}{in which} the likelihood is not normal, \textcolor{red}{such
+as our multinomial likelihood}, it can be difficult to set up effective
+MCMC algorithms that are able to move in the high-dimensional space
+of $\alpha_{p}$. The latent variable representation of Section \ref{sub:Latent-Variable-Model}
+helps to alleviate this problem. Next we describe two alternative
+spatial models that we considered; in Section \ref{sec:Model-comparison},
+\textcolor{red}{we evaluate the models on held-out data} to choose
+between the two.
 
@@
 
-Our first model is a standard conditional autoregressive (CAR) model
-\citep{Bane:etal:2003}. We use a standard form of this model, which
-treats the four cardinal neighbors of each grid cell as the neighbors
-of the grid cell. The corresponding precision matrix has diagonal
-elements, $Q_{ii}$, equal to the number of neighbors for the $i$th
-area (i.e., four except for cells on the boundary of the domain),
-while $Q_{ik}=-1$ (the negative of a weight of one) when areas $i$
-and $k$ are neighbors and $Q_{ik}=0$ when they are not. This gives
-the following model for the values of $\alpha_{p}(s_{i})$ collected
-as a vector across all of the grid cells, $i=1,\ldots,m$: 
+Our first model is a standard conditional autoregressive (CAR) model\textcolor{red}{;
+technical details can be found in} \citep{Bane:etal:2004}. We use
+a standard form of this model \textcolor{red}{that} treats the four
+cardinal neighbors of each grid cell as the neighbors of the grid
+cell. The corresponding precision matrix has diagonal elements, $Q_{ii}$,
+equal to the number of neighbors for the $i$th area (i.e., four except
+for cells on the boundary of the domain), while $Q_{ik}=-1$ (the
+negative of a weight of one) when areas $i$ and $k$ are neighbors
+and $Q_{ik}=0$ when they are not. This gives the following model
+for the values of $\alpha_{p}(s_{i})$ collected as a vector across
+all of the grid cells, $i=1,\ldots,m$: 
 \[
@@
 not full-rank, but is of rank $m-1$; this gives an improper prior
-on an implicit overall mean for the process values. This specification
+on an implicit overall mean for the process values. \textcolor{red}{Note
+that we specify an explicit mean of zero because a non-zero mean would
+not be identifiable in light of the implicit mean.} This specification
 is called an \textit{intrinsic conditional autoregression (ICAR)}
@@
 
-Gaussian processes (GP) are also standard models for spatial processes \citep{Bane:etal:2003}.
+Gaussian processes (GP) are also standard models for spatial processes \citep{Bane:etal:2004}.
 GP models are computationally challenging for large datasets because
-of manipulations involving large covariance matrices. Given this,
-\cite{Lind:etal:2011} proposed a new framework for using Gaussian
-MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
-partial differential equations (SPDE).
-
-We consider Gaussian processes in the \matern class, using the following
-parameterization of the \matern correlation function as 
+of \textcolor{red}{computational} manipulations involving large covariance
+matrices. Given this, \cite{Lind:etal:2011} proposed a new framework
+for using Gaussian MRFs (GMRFs) as approximations to GPs, based on
+the use of stochastic partial differential equations (SPDEs).
+
+\textcolor{red}{Gaussian processes are generally constructed using
+one of a number of correlation functions that define how the strength
+of correlation between the values of the process at two locations
+decays as a function of the distance between the locations. }We consider
+Gaussian processes in the \textcolor{red}{commonly-used} \matern
+class, using the following parameterization of the \matern correlation
+function, 
 \begin{equation}
@@
 
-The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
-to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second model
-is the Lindgren approximation for \matern-based GPs with $\nu=1$.
-To implement the Lindgren model, one modifies the $Q$ matrix defined
-previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
+The approach of \citet{Lind:etal:2011} allows us to consider MRF
+approximations to the \matern-based GP for $\nu=1$ and $\nu=2$.
+Our second spatial model is this Lindgren approximation for \matern-based
+GPs with $\nu=1$. To implement this Lindgren model, one modifies
+the $Q$ matrix defined previously as follows \textcolor{red}{based
+on the technical specification of the precision matrix provided in
+\citet{Lind:etal:2011}}. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
 elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
@@
 $\rho$. In particular $\rho$ allows us to estimate the locality
-of the smoothing. As $\rho$ decreases, the model uses more and more
-local data to estimate the compositional proportions at a given location,
-effectively averaging the empirical proportions over smaller neighborhoods.
-In general, the \cite{Lind:etal:2011} model will generally provide
-for a smoother estimate than the CAR model \citep{Paci:2013}.
+of the \textcolor{red}{spatial} smoothing. As $\rho$ decreases, the
+model uses increasingly localized data to estimate the compositional
+proportions at a given location, effectively averaging the empirical
+proportions over smaller neighborhoods. In general, the \cite{Lind:etal:2011}
+model will generally provide for a smoother estimate than the CAR
+model \citep{Paci:2013}.
 
@@
 we use a uniform distribution on each $\sigma_{p}$ parameter, with
-upper bound of 1000. For the SPDE model we also have parameters $\mu_{p}$,
-which we give flat, non-informative priors (truncated at $\pm10$),
-and $\rho_{p}$ which we give uniform priors on the interval $(0.1,\exp(5))$.
+upper bound of 1000. For the SPDE model we also have hyperparameters
+$\{\mu_{p}\}$, which we give flat, non-informative priors (truncated
+at $\pm10$), and $\{\rho_{p}\}$, which we give uniform priors on
+the interval $(0.1,\exp(5))$. \textcolor{red}{These various hyperparameters
+are unknown parameters that control the spatial structure of the two
+spatial models and are estimated from both the data and the prior
+distributions just specified based on the Bayesian approach.}
 
@@
 for the multinomial probit model \citep{McCu:Ross:1994}. The representation
-introduces latent variables that allow one to develop a MCMC sampling
-strategy that takes advantage of closed form full conditional distributions
+introduces latent variables that allow one to develop an MCMC sampling
+strategy that takes advantage of closed-form full conditional distributions
 (so-called Gibbs sampling steps) for $\alpha_{p}$.
@@
 Each tree $j$ at location $i$ is associated with two variables $W_{ij1}$
-and $W_{ij2}$, governed by the latent variables $\alpha_{1i}$ and
-$\alpha_{2i}$, respectively. Suppose that $\alpha_{1i}>\alpha_{2i}$
+and $W_{ij2}$, governed by the latent variables $\alpha_{1}(s_{i})$
+and $\alpha_{2}(s_{i})$, respectively. Suppose that $\alpha_{1}(s_{i})>\alpha_{2}(s_{i})$
 for a given location $i$. Then this model implies that any tree $j$
 is more likely to be labeled 1 than 2 at location $i$. The difference
-between $\alpha_{1i}$ and $\alpha_{2i}$ explains the \textit{difference}
-in probability of \textit{categories} 1 and 2 at location $i$, and
-the similarity between $\alpha_{p1}$ and $\alpha_{p2}$ explains
-the \textit{correlation} between the probabilities at \textit{locations}
-1 and 2 for category $p$.
+between $\alpha_{1}(s_{i})$ and $\alpha_{2}(s_{i})$ explains the
+\textit{difference} in probability of \textit{categories} 1 and 2
+at location $i$, and the similarity between $\alpha_{p}(s_{1})$
+and $\alpha_{p}(s_{2})$ explains the \textit{correlation} between
+the probabilities at \textit{locations} 1 and 2 for category $p$.
 
@@
 grid. This extension introduces a new set of latent variables, one
-per tree, that indicate the grid cell in which the tree is located
-and can be sampled within the MCMC as additional unknown parameters.
-Specifically, $c_{tj}$, is the latent `membership' variable for tree
-$j$ in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$ is
-a discrete distribution that puts mass, $\psi_{ti}$, $i=1,\ldots,m$,
+per tree, that indicate the grid cells in which the trees are located
+and that can be sampled within the MCMC as additional unknown parameters.
+Specifically, $c_{tj}$ is the latent ``membership'' variable for
+tree $j$ in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$
+is a discrete distribution that puts mass, $\psi_{ti}$, $i=1,\ldots,m$,
 proportional to the areal overlap between the township in which the
@@
 
-We compared the CAR and SPDE models using cross-validation by holding
-out data from the fitting process and assessing the fit of the model
-on the held-out data. This cross-validation used a subregion containing
-most of Minnesota and a small amount of western Wisconsin, defined
-to be the cells whose x-coordinate was less than 300,000 (this defines
-a north-south line that approximately goes through Duluth, Minnesota)
-and hereafter referred to as the ``Minnesota subregion'' . We used
-two cross-validation experiments:
+We compared the CAR and SPDE models \textcolor{red}{by holding out
+data} from the fitting process and assessing the fit of the model
+on the held-out data. We used two experiments\textcolor{red}{{} with
+held-out data}:
 \begin{enumerate}
-\item We held out all the data from 95\% of the cells in the Minnesota subregion,
-with cells selected at random. This was meant to assess the ability
-of the model to interpolate from a sparse set of cells/townships and
+\item The first experiment used a subregion containing most of Minnesota
+and a small amount of western Wisconsin, defined to be the cells whose
+x-coordinate was less than 300,000 m (this defines a north-south line
+that approximately goes through Duluth, Minnesota) and hereafter referred
+to as the ``Minnesota subregion''. We chose this subregion for \textcolor{red}{evaluation}
+because of its high data density, allowing us to experiment with the
+effects of increasing data sparsity on model performance. We held
+out all data from 95\% of the cells in this Minnesota subregion, with
+cells selected at random. This was meant to assess the ability of
+the model to interpolate from a sparse set of cells/townships and
 mimics the limited data in Illinois and Indiana.
 \item We held out 5\% of the trees from all of the trees in the dataset
-(leaving aside the held-out Minnesota subregion cells). This was meant
-to assess the ability of the model to estimate the composition in
-cells in which data were available. 
+for the midwestern subdomain (leaving aside the held-out Minnesota
+subregion cells). This was meant to assess the ability of the model
+to estimate the composition in cells in which data were available. 
 \end{enumerate}
@@
 metrics to use to compare the predictive ability of models; these
-metrics are referred to as scoring rules. In particular it is well-recognized
-that predictive distributions should maximize sharpness subject to
-calibration. That is, the predictive distribution should be as narrow
-as possible while being calibrated such that the observations are
-consistent with the distribution \citep{Gnei:etal:2007}. When thinking
-in terms of prediction intervals as summaries of the predictive distribution,
-we seek intervals that are as narrow as possible while still covering
-the truth the expected proportion of the time.
+metrics are referred to as scoring rules. A general conclusion from
+this work is that predictive distributions should maximize sharpness
+subject to calibration. That is, the predictive distribution should
+be as narrow as possible while being calibrated such that the observations
+are consistent with the distribution \citep{Gnei:etal:2007}. When
+thinking in terms of prediction intervals as summaries of the predictive
+distribution, we seek intervals that are as narrow as possible while
+still covering the truth the expected proportion (e.g., 95\% for a
+95\% prediction interval) of the time.
 
 Following the suggestions in \cite{Gnei:etal:2007}, we considered
-the metrics listed below. For experiment 1, we define $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$
-as the count of all trees in held-out cell $i$ and for experiment
-2, $Y_{i}$ is the count of held-out individual trees in the cell,
-while $y_{ijp}$ is an indicator variable taking value either 0 or
-1 depending on whether the $j$th held-out tree in the $i$th cell
-is of taxon $p$. We calculated each of the metrics below in two ways.
+the following metrics: Brier score, log predictive density, mean square
+prediction error, mean absolute error, and coverage and length of
+prediction intervals. Further details on each are given below. For
+experiment 1, we define $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$ as the count
+of all trees in held-out cell $i$ and for experiment 2, $Y_{i}$
+is the count of held-out individual trees in the cell, while $y_{ijp}$
+is an indicator variable taking value either 0 or 1 depending on whether
+the $j$th held-out tree in the $i$th cell is of taxon $p$. \textcolor{red}{$\hat{\theta}_{ip}=Y_{ip}/n_{i}$
+is the empirical proportion in category $p$ for the $n_{i}$ held-out
+trees in cell $i$.} We calculated each of the metrics in two ways.
 First, we used the posterior mean composition estimates (as an evaluation
@@
 $i$. 
-\item Log predictive density: This metric takes the log of the density of
-held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
+\item Log predictive density: This metric takes the log of the \textcolor{red}{probability
+}density of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
 summing on the log scale across all of the held-out data.
@@
 While in principle, this metric should be optimal \citep{Krnj:Drap:2014},
-it suffers from a lack of robustness in being very sensitive to small
-predictions near zero \citep{Gnei:etal:2007}. Even worse, our Monte
-Carlo estimation of $\theta$ used 10000 samples, so in some cases
-$\tilde{\theta}_{p}(s)=0$. When a tree is present in a cell but its
-corresponding proportion is 0, this gives a log density of $-\infty$,
-preventing use of the metric. As an informal solution to this we set
-$\tilde{\theta}_{p}(s)=\frac{1}{100000}$ in such cases, but given
-these issues we treat the log predictive density as a secondary measure.
-
-\item (Experiment 1 only) Weighted root mean square prediction error (RMSPE)
-and mean absolute error (MAE) of the estimated proportions relative
+it is very sensitive to small predictions near zero \citep{Gnei:etal:2007}.
+Even worse, our Monte Carlo estimation of $\theta$ used 10000 samples,
+so in some cases $\tilde{\theta}_{p}(s)=0$. When a tree is present
+in a cell but its corresponding proportion is 0, this gives a log
+density of $-\infty$, preventing use of the metric. As an informal
+solution to this we set $\tilde{\theta}_{p}(s)=\frac{1}{100000}$
+in such cases, but given these issues we treat the log predictive
+density as a secondary measure.
+
+\item (Experiment 1 only) Weighted root mean square prediction error (RMSPE),
+\textcolor{red}{
+\[
+\sqrt{\frac{1}{Pn}\sum_{i=1}^{m}\sum_{p=1}^{P}n_{i}(\hat{\theta}_{ip}-\tilde{\theta}_{p}(s))^{2}}
+\]
+}and mean absolute error (MAE)\textcolor{red}{
+\[
+\frac{1}{Pn}\sum_{i=1}^{m}\sum_{p=1}^{P}n_{i}|\hat{\theta}_{ip}-\tilde{\theta}_{p}(s)|:
+\]
+}These metrics calculate the error of the estimated proportions relative
 to the empirical proportions based on the held-out trees, averaging
@@
 \end{enumerate}
-Note that for the calculation of the metric applied to the individual
-posterior samples, we can estimate the posterior probability that
-one model has a lower (better) value of the metric than the other
-model by simply calculating the proportion of samples for which the
-model has a lower value of the metric.
+Note that all of the metrics except coverage and interval length can
+be applied to individual posterior samples and therefore allow us
+to estimate the posterior probability that one model has a lower (better)
+value of the metric than the other model by simply calculating the
+proportion of samples for which the model has a lower value of the
+metric. \textcolor{red}{Also note that in addition to allowing comparison
+between models the MAE and RMSPE metrics allow one to assess absolute
+performance of each model in predicting composition.}
 
@@
 
-Here we summarize the results of our cross-validation analyses that
-inform the choice between the CAR and SPDE models.
+Here we summarize the results of our analyses that inform the choice
+between the CAR and SPDE models.
 
@@
 
-Table \ref{tab:score_cell_fivepercent} shows predictive ability based
-on Experiment 1 for those cells in the Minnesota subregion held out
-of the fitting process, for both the mean of the metric applied to
-the individual posterior samples and for the metric applied to the
-posterior mean predictions. Table \ref{tab:coverage_cell_fivepercent}
-shows performance of the uncertainty estimates. For the posterior
-distribution over the predictive metric values, the CAR model outperforms
-the SPDE model, while for the posterior mean predictions, the SPDE
-model appears to outperform the CAR model to a lesser degree, but
-we do not have any uncertainty estimates for this comparison. Coverage
-and interval lengths are similar between the two models. From a practical
-perspective, based on the difference in mean absolute error, which
-is on a readily interpretable scale, the differences between the models
-are small.
-
+For Experiment 1 (full cells held out) for cells in the Minnesota
+subregion held out of the fitting process, the CAR model outperforms
+the SPDE model based on the posterior distribution over the predictive
+metric values (Table \ref{tab:score_cell_fivepercent}). For the posterior
+mean predictions, the SPDE model appears to outperform the CAR model
+to a lesser degree, but we do not have any uncertainty estimates for
+this comparison. Coverage and interval lengths are similar between
+the two models (Table \ref{tab:coverage_cell_fivepercent}). From
+a practical perspective, based on the difference in mean absolute
+error, the differences between the models are small (Table \ref{tab:score_cell_fivepercent}).
+
+%
 \begin{table}
@@
 
-Table \ref{tab:score_cell_20percent} shows the results when the proportion
-of cells that are held out decreases from 95\% to 80\%, indicating
-performance on less sparse data. Table \ref{tab:coverage_cell_20percent}
-shows performance of the uncertainty estimates. For denser data, unlike
-the results when data are more sparse, the SPDE model generally outperforms
-the CAR model, but again differences from a practical perspective,
-based on mean absolute error, are limited.
-
+The results for the variation on Experiment 1 in which the proportion
+of cells that are held out decreases from 95\% to 80\% show that the
+SPDE model generally outperforms the CAR model, but again differences
+from a practical perspective, based on mean absolute error, are limited
+(Tables \ref{tab:score_cell_20percent}-\ref{tab:coverage_cell_20percent}).
+
+%
 \begin{table}
@@
 
-Table \ref{tab:score_tree} shows the predictive ability based on
-Experiment 2. Here we have evidence (posterior probability of 0.93)
-that the SPDE model is better based on the Brier score, but the Brier
-score values for the two models are numerically almost the same.
-
+In Experiment 2 (individual trees held out), we have evidence (posterior
+probability of 0.93) that the SPDE model is better based on the Brier
+score, but the Brier score values for the two models are numerically
+almost the same (Table \ref{tab:score_tree}).
+
+%
 \begin{table}
@@
 case for Illinois and Indiana. Second, the model is simpler and easier
-to explain and computations can be done more quickly. Third, predictions
+to explain, and computations can be done more quickly. Third, predictions
 from the SPDE model showed boundary effects, with some taxa showing
@@
 
-For the final data product, we ran the model using the CAR specification
-for 150,000 iterations, discarding the first 25,000 iterations for
-burn-in and retaining 250 subsampled iterations (subsampled to limit
-storage needs and the size of the data product). Mixing was generally
-reasonable, but for some of the hyperparameters was relatively slow,
-particularly for less common taxa. Despite this, mixing for the variables
-of substantive interest -- the proportions -- was good.
-
-In Fig. \ref{fig:select_maps}, we show maps of estimated composition
-for the full domain for several taxa of substantive interest to illustrate
-the results. These maps contrast the raw data proportions, the posterior
-means, and posterior standard deviations as pointwise estimates of
-uncertainty. Fig. \ref{fig:all_predictions} shows posterior means
-for all 23 taxa.
+The final data product is a dataset that contains 250 posterior samples
+of the proportions of each of the 23 tree taxa at each grid cell in
+the states in our domain of the northeastern United States.
+
+For this final data product, we ran the model using the CAR specification
+\textcolor{red}{with all of the data (including the data held out
+in the model comparison analyses)} for 150,000 iterations with the
+same burn-in and subsampling details as described in Section \ref{sec:Model-comparison}.
+Based on graphical checks and calculation of effective sample size
+values, mixing was generally reasonable, but for some of the hyperparameters
+was relatively slow, particularly for less common taxa. Despite this,
+mixing for the variables of substantive interest -- the proportions
+-- was good, with effective sample sizes for the final product generally
+near 250.
+
+Maps of estimated composition for the full domain for several taxa
+of substantive interest illustrate the results, contrasting the raw
+data proportions, the posterior means, and posterior standard deviations
+as pointwise estimates of uncertainty (Fig. \ref{fig:select_maps}).
+We also present the posterior means for all 23 taxa (Fig. \ref{fig:all_predictions}).
 
@@
 
-The data product is publicly available at the NIS Data Portal (accession
-number to be inserted when available) under YY license as version
-0.4 as of August 2015. The product is in the form of a netCDF-4 file,
-with dimensions x, y, and MCMC iteration. There is one variable per
-taxon. The PalEON project (in particular the first author) will continue
+The data product is publicly available at the \textcolor{red}{NIS
+Data Portal as DOI:10.5072/FK2/dfc5e9748c8df4c3c275eab0f42d9744} under
+the CC BY 4.0 license as version 0.4 as of January 2016 \citep{paci:etal:data:2016}.
+The product is in the form of a netCDF-4 file, with dimensions x-coordinate,
+y-coordinate, and MCMC iteration. There is one variable per taxon.
+In addition, dynamic visualizations of the product using the Shiny
+tool are available at \href{https://www3.nd.edu/~paleolab/paleonproject/maps}{https://www3.nd.edu/$\sim$paleolab/paleonproject/maps}.
+The PalEON Project (in particular the first author) will continue
 to maintain this product, releasing new versions as additional data
@@
 additional data from Ohio is planned as well. As a result, at some
-point we expect to have complete data for the western half of the
-domain. There will remain some missing townships in the eastern half
-of the domain.
+point we expect to have complete data for the midwestern half of the
+domain.
 
@@
 
-Given the density of data and the limited differences seen between
-the CAR and SPDE models, we expect the data product to be reasonably
-robust to the choice of spatial model, particularly in those areas
-with complete data. However, additional investigation of other statistical
-representations is of interest, in particular nonstationary spatial
-models and use of covariates. The biggest shortcoming of the current
-model is its inability to account for local features such as rivers
-(note the Minnesota River floodplain in evidence in the raw data).
-The current model, by using a simple stationary spatial model that
-smooths as a function of Euclidean distance, does not account for
-topographic, soil, or other features. An additional drawback of the
-product is its focus on composition, which does not directly tell
-us about vegetation structure, in particular does not distinguish
-between closed forest, savanna, and prairie, of particular note in
-Minnesota, Wisconsin, Illinois, and into Indiana. The statistically-estimated
-biomass product mentioned next will directly inform questions about
-vegetation structure.
-
-In addition to the existing raw composition count data product discussed
-in Section \ref{sec:Data}, we have related data products that are
-under development:
-\begin{itemize}
-\item Gridded raw biomass estimates for Minnesota, Wisconsin, and Michigan
-based on the PLS data, with extension to Illinois and Indiana planned. 
-\item Statistically estimated biomass for Minnesota, Wisconsin, and Michigan
-using a statistical model applied to the raw biomass estimates, with
-extension to Illinois and Indiana planned.
-\end{itemize}
-We also plan to estimate basal area and stem density using a similar
-approach to that used for biomass.
+\textcolor{red}{In the parts of the modeled region with spatially
+complete data (in particular, Minnesota, Wisconsin, and Michigan),
+the statistical estimates of forest composition closely match the
+patterns apparent in the raw data (Fig. \ref{fig:select_maps}), as
+expected.  In these areas, the estimated tree composition from the
+model has the advantage of downweighting unusual or outlier values
+in the empirical proportions of individual grid cells, which are likely
+due to stochastic sampling variation within that grid cell (compare
+the first two columns in Fig. \ref{fig:select_maps}). Some stochastic
+variation is expected given that, even in the most spatially complete
+regions, each grid cell contains an average of 124 trees (120-135
+is typical) \citep{goring2015composition} and some cells contain
+many fewer trees. Hence, some smoothing of this stochastic variation
+is appropriate. This smoothing is based on information on data from
+nearby cells, and the estimates from the model reflect the smooth
+trends in forest composition across the spatial domain. A partial
+cost is that these maps can smooth out sharp ecotones or other forms
+of true spatial heterogeneity, particularly in areas with sparse data
+(including areas with low tree density). For example, the sharp increase
+in Elm along the Minnesota River (Fig. \ref{fig:select_maps}, first
+column) likely represents a real ecotone in the settlement-era vegetation.
+Vegetation gradients and ecotones were sharper in the settlement-era
+forests in the upper Midwest than in contemporary forests \citep{goring2015composition},
+and the modeled estimates may partially obscure this change. Users
+interested in using the original unsmoothed data are directed to the
+data product described earlier \citep{gori:etal:data:2016}. Additional
+investigation of other statistical representations to better capture
+sharp gradients is of interest, in particular nonstationary spatial
+models and use of covariates. Potential environmental covariates include
+soils, firebreaks, and topography \citep{grimm1984fire,shea2014reconstructing}.
+Here, however, we chose to limit our model to be exclusively a function
+of spatial distance in order to avoid dependence on the environmental
+drivers of pre-settlement forest composition that might lead to circular
+reasoning in subsequent inferences based on this dataset. Use of covariates
+could also lead to prediction that a taxa is present well beyond its
+range boundary in places where data are sparse. }
+
+\textcolor{red}{A key advance of this work over prior reconstructions
+of settlement-era vegetation lies in the estimates of uncertainty
+across the spatial domain. These estimates of uncertainty include
+the sampling uncertainty within grid cells (as do the within-grid
+cell estimates of uncertainty available from the raw proportions),
+but, because this is a spatial model, predictions and their associated
+uncertainty estimates are also informed by the information content
+of nearby cells. The maps of standard errors across species (Fig.
+\ref{fig:select_maps}, third column) highlight the advantages of
+this approach in areas of high data coverage (Minnesota, Wisconsin,
+Michigan) and in areas of sparse coverage (e.g., Illinois, Indiana,
+parts of Ohio). Where there are not large gaps in the data, the model
+provides low and fairly smooth estimates of uncertainty. Uncertainty
+is generally higher in the eastern subdomain than in the areas of
+the midwestern subdomain with high data coverage because of missing
+townships and lower sampling density even in townships with data.
+In areas of sparse coverage and in areas with low tree density (e.g.,
+southwestern Minnesota), the standard error of our estimates increases
+appropriately. Nevetheless, these uncertainties surround reasonable
+estimates of trends in composition. For example, the model does a
+good job of capturing the oak ecotone in Indiana and Illinois, representing
+a shift from oak savannas and woodlands to closed mesic forests (Fig.
+\ref{fig:select_maps}). Experiment 1 showed that both models predicted
+composition at cells with no data reasonably well, mimicking the case
+of sparsely sampled data and giving confidence in the broad spatial
+patterns predicted in more poorly sampled regions, particularly those
+with regular, but sparse sampling that mimic the experiment (Illinois
+and Indiana, but not Ohio). The apparent blockiness of uncertainty
+estimates in a few places such as Ohio is caused by spatial gaps and
+variations in sampling resolution. Absolute uncertainty generally
+increases with abundance for all taxa (Fig. \ref{fig:select_maps},
+column 3).}
+
+\textcolor{red}{The exploration of alternative approaches to spatial
+modeling of composition showed similar results for the SPDE and CAR
+models, both in terms of prediction accuracy and performance of prediction
+intervals. Small differences among the various metrics of goodness
+of fit favored each model in turn, but applied users of the models
+would find little pragmatic difference affecting scientific inference.
+Ultimately, we slightly favor the CAR model, because it avoids the
+boundary effects apparent in the SPDE model at the edges of the domain.}
+
+\textcolor{red}{The models presented here estimate only the relative
+abundance of tree taxa, which does not directly tell us about tree
+density or other aspects of vegetation structure. This becomes a particular
+limitation for interpreting vegetation where trees become sparse at
+the prairie-forest transition from northern Minnesota through southern
+Illinois \citep{transeau1935prairie}. Our model (correctly) predicts
+that tree composition there is dominated by oak, but this is less
+useful considering the sparseness of trees.} This limitation can be
+addressed by developing estimates of absolute abundance (e.g., biomass)
+rather than compositional estimates. A gridded dataset of biomass,
+stem density, and basal area is already available for Minnesota, Wisconsin,
+and northern Michigan \citep{goring2015composition}, based on the
+PLS data. An extension to southern Michigan, Illinois, and Indiana
+is planned. We are currently developing statistical estimates of biomass
+for Minnesota, Wisconsin, and Michigan using a statistical model applied
+to the gridded biomass dataset, with extension to Illinois and Indiana
+planned. We also plan to estimate stem density and basal area using
+a similar approach to that used for biomass.
 
@@
 
-CJP, AT, and JZ developed the statistical model and code. CJP carried
+CJP, ALT, and JZ developed the statistical model and code. CJP carried
 out the model comparison and created the data product. CJP wrote the
-paper based on an initial draft by AT and with feedback from SJG,
-CVC, JWW, DJM, JAP, JZ and JSM. CVC, DJM, SJG, JWW, and JAP provided
-data.
+paper based on an initial draft by ALT and with feedback and editing
+from SJG, CVC, JWW, DJM, JAP, JZ and JSM. SJG, JAP, CVC, DJM, JSM,
+and JWW led the processing and analysis of the PLS and TPS data and
+assisted with interpretation of results.
 
@@
 years who have preserved, collected, and digitized survey records,
-in particular Jim Dyer, Peter Marks, Robert McIntosh, Ronald Stuckey,
-Ed Schools, and the Ohio Biological Survey \{David/Simon please add
-names here\} We thank Madeline Ruid, Ben Seliger, Morgan Ross and
-Daniel Handel for processing of the southern Michigan data. Indiana
-and Illinois data were made possible through the hard work of many
-Notre Dame undergraduates in the McLachlan lab. This work was carried
-out by the PalEON Project with support from the National Science Foundation
-MacroSystems Program through grants EF-1065656, EF-1241868, DEB-1241874
-and DEB-1241868 and from the Notre Dame Environmental Change Initiative.
+in particular John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed
+Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
+We thank Madeline Ruid, Benjamin Seliger, Morgan Ripp and Daniel Handel
+for processing of the southern Michigan data. Indiana and Illinois
+data were made possible through the hard work of many Notre Dame undergraduates
+in the McLachlan lab. This work was carried out by the PalEON Project
+with support from the National Science Foundation MacroSystems Program
+through grants EF-1065702, EF-1065656, DEB-1241874 and DEB-1241868
+and from the Notre Dame Environmental Change Initiative.
 

### 2016-02-03 paperPlos.tex
revised and added for plos revision january 2016

@@
 \usepackage{rotating}
+
+\usepackage{color}
 
@@
 \\
-\bf{6} Department of Forestry and Wildlife Ecology, University of Wisconsin,
+\bf{6} Department of Forest and Wildlife Ecology, University of Wisconsin,
 Madison, Wisconsin, USA
@@
 consistency among points and the application of spatially-varying
-correction factors \cite{goring2015composition}
-to accurately assess tree stem density, basal area and biomass from
-the early settlement records, but the impact on composition estimates
-is limited \cite{liu2011broadscale}. Surveyors sometimes used ambiguous
-common names, which requires matching to scientific names and standardization
-\cite{mladenoff2002narrowing,goring2015composition}.
+correction factors \cite{goring2015composition} to accurately assess
+tree stem density, basal area and biomass from the early settlement
+records, but the impact on composition estimates is limited \cite{liu2011broadscale}.
+Surveyors sometimes used ambiguous common names, which requires matching
+to scientific names and standardization \cite{mladenoff2002narrowing,goring2015composition}.
 
@@
 locations. Second, there are data gaps: the available digitized data
-from Illinois and Indiana represent a small fraction of those states
+from Illinois and Indiana represent a small fraction of those states,
 and missing townships are common in the TPS data. Third, the TPS and
@@
 
-In \emph{Data} we describe the data sources, while in \emph{Statistical model}
-we describe the statistical model used to create the data product.
-In \emph{Model comparison} we quantitatively compare competing
-statistical specifications, and \emph{Data Product}
+In \emph{Data} we describe the data sources, while in \emph{Statistical model} we describe 
+the statistical model used to create the data product. In In \emph{Model comparison}
+ we quantitatively compare competing
+statistical specifications, and in \emph{Data Product}
 we describe the final data product. In \emph{Discussion}
-we discuss limitations of the statistical model, and we list related
-data products under development.
+we discuss the uncertainties estimated by and the
+limitations of the statistical model, and we list related data products
+under development.
 
@@
 and Illinois, at the moment trees are associated with township centroids
-and then assigned to 8 km grid cells based on the centroid but in
+and then assigned to 8 km grid cells based on the centroid, but in
 the near future we will have point locations available for each tree.)
@@
 six New England states (the eastern subdomain), data are aggregated
-at the township level. There are also data from a single township
-in Quebec and a single township in northern Delaware. Digitization
-of PLS data in Minnesota, Wisconsin and Michigan is essentially complete,
-with PLS data for nearly all 8 km grid cells, but data in Illinois
-and Indiana represent a sample of the full set of grid cells, with
-survey record transcription ongoing. Data for the eastern states are
-available for a subset of the full set of townships covering the domain;
-the TPS data for some townships were lost, incomplete, or have not
-been located \cite{Cogb:etal:2002}.
+at the township level. We make predictions for all
+of the states listed above; these constitute our core domain. There
+are also data from a single township in Quebec and a single township
+in northern Delaware; these data help inform predictions
+in nearby locations within our core domain, but predictions are not
+made for Quebec or Delaware. Digitization of PLS data in Minnesota,
+Wisconsin and Michigan is essentially complete, with PLS data for
+nearly all 8 km grid cells, but data in Illinois and Indiana represent
+a sample of the full set of grid cells, with survey record transcription
+ongoing. Data for the eastern states are available for a subset of
+the full set of townships covering the domain; the TPS data for some
+townships were lost, incomplete, or have not been located \cite{Cogb:etal:2002}.
 
@@
 are white (e.g., a few locations in the northwestern portion of the
-domain in the states of Minnesota and Wisconsin).}}
+domain in the states of Minnesota and Wisconsin).}
 \label{fig:Spatial-domain}
@@
 on the number of survey points in a cell, with an average of 124 trees
-per cell. The gridded data at the 8 km resolution for the upper Midwest
-are available through the NIS Data portal (accession number and version
-to be inserted when available). The TPS data were compiled by C.V.
-Cogbill from a myriad of archival sources representing land division
-surveys conducted in connection with local settlement.
+per cell. The gridded data at the 8 km resolution for the midwest
+subdomain are available through the NIS data portal as
+DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5 \cite{Gori:etal:data:2016}.
+The TPS data were compiled by C.V. Cogbill from a myriad of archival
+sources representing land division surveys conducted in connection
+with local settlement and are available through the
+NIS data portal as DOI:10.5072/FK2/4d44aebe71dbf65290d8ce1bde03598b
+and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}.
 
@@
 grouped into the other hardwood category are those for which fewer
-than roughly 2000 trees were present in the dataset; however we include
+than roughly 2000 trees were present in the dataset; however, we include
 Atlantic white cedar explicitly despite it only having 336 trees in
@@
 located at (-71000 m, 58000 m). For the midwestern subdomain we use
-the western-most 146 by 180 grid of cells. For the eastern subdomain
-we use the eastern-most 180 by 180 grid of cells and then omit 23
-rows of cells in the north and 17 rows of cells in the south, as these
-grid cells are outside of the states containing data.
+the western-most 146 by 180 grid of cells when fitting
+the statistical models. For the eastern subdomain we use the eastern-most
+180 by 180 grid of cells and then omit 23 rows of cells in the north
+and 17 rows of cells in the south, as these grid cells are outside
+of the states containing data.
 
@@
 that could be calculated using the full population of all the trees
-in grid cell or township.
+in a grid cell or township.
 \end{enumerate}
+At a high level, the Bayesian statistical model estimates
+composition across the domain, even in locations with sparse or no
+data, by combining the raw composition data with the assumption that
+composition varies in a smooth spatial fashion across the domain.
+The information in the data is quantified by the data model, also
+known as the likelihood. The assumption of smoothness is built into
+the model by representing the true unknown spatially-varying composition
+using a statistical spatial process representation that induces smoothing
+of estimates across nearby locations. This spatial process representation
+is a form of prior distribution and is a function of model parameters
+called hyperparameters that determine the correlation structure of
+the process and are also estimated based on the data.
+
 The result of fitting the Bayesian model via Markov chain Monte Carlo
@@
 cells. These samples are the data product (described further in the
-Data Product section) and can be used in subsequent analyses. The
-mean and standard deviation of the samples for each pair of cell and
-taxon represent our best estimate (i.e., prediction) of composition
+\emph{Data Product} section) and can be used in subsequent analyses.
+The mean and standard deviation of the samples for each pair of cell
+and taxon represent our best estimate (i.e., prediction) of composition
 and a Bayesian ``standard error'' quantifying the uncertainty in
 the estimate.
+
+In the remainder of this section we provide the technical
+specification of the model and of the computations involved in fitting
+the model.
 
@@
 The proportions, $\theta_{p}(s_{i}),\, p=1,\ldots,P$, are modeled
-spatially by a set of $P$ Gaussian spatial processes, one per taxon,
+spatially by a set of \emph{P} Gaussian spatial processes, one per taxon,
 $\alpha_{p}(s_{i}),\, p=1,\ldots,P$. This collection of processes
@@
 processes are defined on the 8 km grid, $\alpha_{p}=\{\alpha_{p}(s_{1}),\ldots,\alpha_{p}(s_{m})\}$
-for the $m$ grid cells. In \emph{Latent variable model}
+for the \emph{m} grid cells. In  \emph{Latent variable model}
 we introduce a multinomial probit model that relates the $\alpha_{p}(s)$
@@
 of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.
+A multinomial probit model is similar to logistic
+regression, used for modeling a binary outcome based on an underlying
+probability the outcome will occur, but generalizes to modeling a
+categorical variable based on probabilities for each category.
 
 The critical component of the statistical model is the representation
-of $\alpha_{p}(s)$ as a spatial process. This process is a prior
-structure that serves to smooth across noise in the observations and
-allows for interpolation to locations with no data. Apart from the
-sum-to-one constraint, the taxa are considered to be independent in
-the prior. We did not want to impose any structure that ties the different
-taxa together, as any correlation will likely vary across space.
+of $\alpha_{p}(s)$ as a spatial process. A spatial
+process is a statistical representation that models spatially-correlated
+values. It provides a prior structure that serves to smooth across
+noise in the observations and allows for prediction at locations based
+on information from nearby locations, including interpolation to locations
+with no data. Apart from the sum-to-one constraint, the taxa are considered
+to be independent in the prior. We did not want to impose any structure
+that ties the different taxa together, as any correlation will likely
+vary across space.
 
@@
 autoregressive model \cite{Bane:etal:2004} and a Gaussian Markov
-random field (MRF) approximation to a Gaussian process with Matern
-covariance \cite{Lind:etal:2011}.
+random field (MRF) approximation to a Gaussian process with \matern
+covariance \cite{Lind:etal:2011}. These models are
+specific statistical formulations of spatial processes that represent
+spatial correlation by defining neighborhoods around each location
+that are used to help inform predictions at the location.
 
@@
 
-MRF models work directly with the precision matrix of the values of
-the spatial process, so calculation of the prior density of $\alpha_{p}$
-is computationally simple \cite{Rue:Held:2005}, but in situations
-where the likelihood is not normal, it can be difficult to set up
-effective MCMC algorithms that are able to move in the high-dimensional
-space of $\alpha_{p}$. The latent variable representation helps to alleviate this problem.
-Next we describe two alternative spatial models that we considered;
-in \emph{Model comparison}, we use cross-validation to
-choose between these two representations.
+MRF models represent the neighborhood information
+by working directly with the precision matrix (the
+inverse of the covariance matrix) of the values of the spatial process,
+so calculation of the prior density of $\alpha_{p}$ is computationally
+simple \cite{Rue:Held:2005}. However, in situations
+in which the likelihood is not normal, such
+as our multinomial likelihood, it can be difficult to set up effective
+MCMC algorithms that are able to move in the high-dimensional space
+of $\alpha_{p}$. The latent variable representation
+helps to alleviate this problem. Next we describe two alternative
+spatial models that we considered; in \emph{Model comparison},
+we evaluate the models on held-out data to choose
+between the two.
 
@@
 
-Our first model is a standard conditional autoregressive (CAR) model
-\cite{Bane:etal:2004}. We use a standard form of this model, which
-treats the four cardinal neighbors of each grid cell as the neighbors
-of the grid cell. The corresponding precision matrix has diagonal
-elements, $Q_{ii}$, equal to the number of neighbors for the \emph{i}th
-area (i.e., four except for cells on the boundary of the domain),
-while $Q_{ik}=-1$ (the negative of a weight of one) when areas \emph{i}
-and \emph{k} are neighbors and $Q_{ik}=0$ when they are not. This gives
-the following model for the values of $\alpha_{p}(s_{i})$ collected
-as a vector across all of the grid cells, $i=1,\ldots,m$: 
+Our first model is a standard conditional autoregressive (CAR) model;
+technical details can be found in \cite{Bane:etal:2004}. We use
+a standard form of this model that treats the four
+cardinal neighbors of each grid cell as the neighbors of the grid
+cell. The corresponding precision matrix has diagonal elements, $Q_{ii}$,
+equal to the number of neighbors for the \emph{i}th area (i.e., four except
+for cells on the boundary of the domain), while $Q_{ik}=-1$ (the
+negative of a weight of one) when areas \emph{i} and \emph{k} are neighbors
+and $Q_{ik}=0$ when they are not. This gives the following model
+for the values of $\alpha_{p}(s_{i})$ collected as a vector across
+all of the grid cells, $i=1,\ldots,m$: 
 \[
-\alpha_{p}\sim\N(0,\sigma_{p}^{2}Q^{-}).
+\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-}).
 \]
@@
 not full-rank, but is of rank $m-1$; this gives an improper prior
-on an implicit overall mean for the process values. This specification
+on an implicit overall mean for the process values. Note
+that we specify an explicit mean of zero because a non-zero mean would
+not be identifiable in light of the implicit mean. This specification
 is called an \textit{intrinsic conditional autoregression (ICAR)}
-and we can write $Q=D-C$ where $C$ is the $m\times m$ adjacency
+and we can write $Q=D-C$ where \emph{C} is the $m\times m$ adjacency
 matrix defining the neighborhood relation of the locations; that is
 $C{}_{ik}=1$ if locations \emph{i} and \emph{k} are neighbors and zero otherwise.
-The matrix \emph{D} is an $m\times m$ diagonal matrix containing the
-row sums of matrix \emph{C} as the diagonal entries, $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$
+The matrix $D$ is an $m\times m$ diagonal matrix containing the
+row sums of matrix $C$ as the diagonal entries, $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$
 
@@
 GP models are computationally challenging for large datasets because
-of manipulations involving large covariance matrices. Given this,
-\cite{Lind:etal:2011} proposed a new framework for using Gaussian
-MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
-partial differential equations (SPDEs).
-
-We consider Gaussian processes in the \matern class, using the following
-parameterization of the \matern correlation function, 
+of computational manipulations involving large covariance
+matrices. Given this, \cite{Lind:etal:2011} proposed a new framework
+for using Gaussian MRFs (GMRFs) as approximations to GPs, based on
+the use of stochastic partial differential equations (SPDEs).
+
+Gaussian processes are generally constructed using
+one of a number of correlation functions that define how the strength
+of correlation between the values of the process at two locations
+decays as a function of the distance between the locations. We consider
+Gaussian processes in the commonly-used \matern
+class, using the following parameterization of the \matern correlation
+function, 
 \begin{equation}
-R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),
+R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
 \end{equation}
-where \emph{d} is Euclidean distance, $\rho$ is the spatial range parameter,
+where $d$ is Euclidean distance, $\rho$ is the spatial range parameter,
 and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
@@
 
-The  approach of \cite{Lind:etal:2011} allows us to consider MRF approximations
-to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second spatial
-model is the Lindgren approximation for \matern-based GPs with $\nu=1$.
-To implement the Lindgren model, one modifies the \emph{Q} matrix defined
-previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
+The approach of \cite{Lind:etal:2011} allows us to consider MRF
+approximations to the \matern-based GP for $\nu=1$ and $\nu=2$.
+Our second spatial model is this Lindgren approximation for \matern-based
+GPs with $\nu=1$. To implement this Lindgren model, one modifies
+the \emph{Q} matrix defined previously as follows based
+on the technical specification of the precision matrix provided in
+\cite{Lind:etal:2011}. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
 elements of \emph{Q} are $4+a^{2}$. The entries corresponding to cardinal
-neighbors are -2. Those for diagonal neighbors are 2, and those
-for 2nd-order cardinal neighbors are 1. This extends the neighborhood
+neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
+for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
 structure relative to the CAR model and parameterizes it as a function
@@
 $\rho$. In particular $\rho$ allows us to estimate the locality
-of the smoothing. As $\rho$ decreases, the model uses increasingly
-localized data to estimate the compositional proportions at a given
-location, effectively averaging the empirical proportions over smaller
-neighborhoods. In general, the model of \cite{Lind:etal:2011} will generally
-provide for a smoother estimate than the CAR model \cite{Paci:2013}.
+of the spatial smoothing. As $\rho$ decreases, the
+model uses increasingly localized data to estimate the compositional
+proportions at a given location, effectively averaging the empirical
+proportions over smaller neighborhoods. In general, the \cite{Lind:etal:2011}
+model will generally provide for a smoother estimate than the CAR
+model \cite{Paci:2013}.
 
@@
 \[
-\alpha_{p}\sim\N\left(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1}\right)
+\alpha_{p}\sim\mbox{N}\left(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1}\right)
 \]
@@
 we use a uniform distribution on each $\sigma_{p}$ parameter, with
-upper bound of 1000. For the SPDE model we also have parameters $\{\mu_{p}\}$,
-which we give flat, non-informative priors (truncated at 10),
-and $\{\rho_{p}\}$, which we give uniform priors on the interval
-$(0.1,\exp(5))$.
+upper bound of 1000. For the SPDE model we also have hyperparameters
+$\{\mu_{p}\}$, which we give flat, non-informative priors (truncated
+at $\pm10$), and $\{\rho_{p}\}$, which we give uniform priors on
+the interval $(0.1,\exp(5))$. These various hyperparameters
+are unknown parameters that control the spatial structure of the two
+spatial models and are estimated from both the data and the prior
+distributions just specified based on the Bayesian approach.
 
@@
 for the multinomial probit model \cite{McCu:Ross:1994}. The representation
-introduces latent variables that allow one to develop a MCMC sampling
-strategy that takes advantage of closed form full conditional distributions
+introduces latent variables that allow one to develop an MCMC sampling
+strategy that takes advantage of closed-form full conditional distributions
 (so-called Gibbs sampling steps) for $\alpha_{p}$.
@@
 in other words, the maximum of the set of latent variables $\{W_{ijp}\}{\displaystyle _{p=1}^{P}}$
-determines the category of observation \emph{j} at location \emph{i}. The
+determines the category of observation $j$ at location \emph{i}. The
 final piece of the latent variable representation is the relationship
@@
 \[
-W_{ijp}\sim\N(\alpha_{p}(s_{i}),1)
+W_{ijp}\sim\mbox{N}(\alpha_{p}(s_{i}),1)
 \]
@@
 Each tree \emph{j} at location \emph{i} is associated with two variables $W_{ij1}$
-and $W_{ij2}$, governed by the latent variables $\alpha_{1i}$ and
-$\alpha_{2i}$, respectively. Suppose that $\alpha_{1i}>\alpha_{2i}$
-for a given location \emph{i}. Then this model implies that any tree \emph{j}
+and $W_{ij2}$, governed by the latent variables $\alpha_{1}(s_{i})$
+and $\alpha_{2}(s_{i})$, respectively. Suppose that $\alpha_{1}(s_{i})>\alpha_{2}(s_{i})$
+for a given location \emph{i}. Then this model implies that any tree $j$
 is more likely to be labeled 1 than 2 at location \emph{i}. The difference
-between $\alpha_{1i}$ and $\alpha_{2i}$ explains the \textit{difference}
-in probability of \textit{categories} 1 and 2 at location \emph{i}, and
-the similarity between $\alpha_{p1}$ and $\alpha_{p2}$ explains
-the \textit{correlation} between the probabilities at \textit{locations}
-1 and 2 for category \emph{p}.
+between $\alpha_{1}(s_{i})$ and $\alpha_{2}(s_{i})$ explains the
+\textit{difference} in probability of \textit{categories} 1 and 2
+at location \emph{i}, and the similarity between $\alpha_{p}(s_{1})$
+and $\alpha_{p}(s_{2})$ explains the \textit{correlation} between
+the probabilities at \textit{locations} 1 and 2 for category \emph{p}.
 
@@
 grid. This extension introduces a new set of latent variables, one
-per tree, that indicate the grid cell in which the tree is located
-and can be sampled within the MCMC as additional unknown parameters.
+per tree, that indicate the grid cells in which the trees are located
+and that can be sampled within the MCMC as additional unknown parameters.
 Specifically, $c_{tj}$ is the latent ``membership'' variable for
@@
 
-Using the latent variable representation, we have that $W_{tjp}\sim\N(\alpha_{p}(s_{c_{tj}}),1)$
+Using the latent variable representation, we have that $W_{tjp}\sim\mbox{N}(\alpha_{p}(s_{c_{tj}}),1)$
 for tree \emph{j} in township \emph{t}. In updating the other parameters in
@@
 
-The representation of \cite{McCu:Ross:1994} is convenient for MCMC sampling,
+The \cite{McCu:Ross:1994} representation is convenient for MCMC sampling,
 particularly in this high-dimensional spatial context, as it allows
@@
 
-The full description of the MCMC sampling steps is provided in
+The full description of the MCMC sampling steps is provided in the
 \nameref{SI}. In addition, in the latent variable representation, $\theta_{p}(s)$
@@
 we use Monte Carlo integration over $W_{ijp},\, p=1,\ldots,P$ to
-estimate $\theta_{p}(s_{i})$, also described in \nameref{SI}.
+estimate $\theta_{p}(s_{i})$, also described in the \nameref{SI}.
 
@@
 is available on Github, including pre- and post-processing code, at
-https://github.com/Paleon-Project/composition.
+\href{https://github.com/PalEON-Project/composition}{https://github.com/PalEON-Project/composition}.
 
@@
 
-We compared the CAR and SPDE models using cross-validation by holding
-out data from the fitting process and assessing the fit of the model
-on the held-out data. We used two cross-validation experiments:
+We compared the CAR and SPDE models by holding out
+data from the fitting process and assessing the fit of the model
+on the held-out data. We used two experiments with
+held-out data:
 \begin{enumerate}
@@
 that approximately goes through Duluth, Minnesota) and hereafter referred
-to as the ``Minnesota subregion''. We chose this subregion for cross-validation
+to as the ``Minnesota subregion''. We chose this subregion for evaluation
 because of its high data density, allowing us to experiment with the
@@
 is an indicator variable taking value either 0 or 1 depending on whether
-the \emph{j}th held-out tree in the \emph{i}th cell is of taxon \emph{p}. We calculated
-each of the metrics in two ways. First, we used the posterior mean
-composition estimates (as an evaluation of our core predictions),
-with $\tilde{\theta}_{p}(s)$ being the posterior mean. Second, we
-averaged the metric over the posterior samples (as an evaluation of
-our full data product, including uncertainty), taking $\tilde{\theta}_{p}(s)$
-to be an individual MCMC sample and then averaging the metric over
-all the posterior samples.
-
+the \emph{j}th held-out tree in the \emph{i}th cell is of taxon \emph{p}. $\hat{\theta}_{ip}=Y_{ip}/n_{i}$
+is the empirical proportion in category \emph{p} for the $n_{i}$ held-out
+trees in cell \emph{i}. We calculated each of the metrics in two ways.
+First, we used the posterior mean composition estimates (as an evaluation
+of our core predictions), with $\tilde{\theta}_{p}(s)$ being the
+posterior mean. Second, we averaged the metric over the posterior
+samples (as an evaluation of our full data product, including uncertainty),
+taking $\tilde{\theta}_{p}(s)$ to be an individual MCMC sample and
+then averaging the metric over all the posterior samples. 
 \begin{enumerate}
@@
 \emph{i}. 
-\item Log predictive density: This metric takes the log of the density of
-held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
+\item Log predictive density: This metric takes the log of the probability
+density of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
 summing on the log scale across all of the held-out data.
@@
 
-\item (Experiment 1 only) Weighted root mean square prediction error (RMSPE)
-and mean absolute error (MAE): These metrics calculate the error of
-the estimated proportions relative to the empirical proportions based
-on the held-out trees, averaging over cells and taxa. We weight by
-the number of held-out trees in each cell to account for the greater
-variability in the empirical proportions in locations with few held-out
-trees. 
+\item (Experiment 1 only) Weighted root mean square prediction error (RMSPE),
+
+\[
+\sqrt{\frac{1}{Pn}\sum_{i=1}^{m}\sum_{p=1}^{P}n_{i}(\hat{\theta}_{ip}-\tilde{\theta}_{p}(s))^{2}}
+\]
+and mean absolute error (MAE)
+\[
+\frac{1}{Pn}\sum_{i=1}^{m}\sum_{p=1}^{P}n_{i}|\hat{\theta}_{ip}-\tilde{\theta}_{p}(s)|:
+\]
+These metrics calculate the error of the estimated proportions relative
+to the empirical proportions based on the held-out trees, averaging
+over cells and taxa. We weight by the number of held-out trees in
+each cell to account for the greater variability in the empirical
+proportions in locations with few held-out trees. 
 \item (Experiment 1 only) Coverage and length of 95\% prediction intervals
@@
 proportion of samples for which the model has a lower value of the
-metric.
+metric. Also note that in addition to allowing comparison
+between models the MAE and RMSPE metrics allow one to assess absolute
+performance of each model in predicting composition.
 
@@
 
-Here we summarize the results of our cross-validation analyses that
-inform the choice between the CAR and SPDE models.
-
-For Experiment 1 (full cells held out) for cells in the Minnesota subregion held out of
-the fitting process, the CAR model outperforms the SPDE model based
-on the posterior distribution over the predictive metric values (Table
-\ref{tab:score_cell_fivepercent}). For the posterior mean predictions,
-the SPDE model appears to outperform the CAR model to a lesser degree,
-but we do not have any uncertainty estimates for this comparison.
-Coverage and interval lengths are similar between the two models (Table
-\ref{tab:coverage_cell_fivepercent}). From a practical perspective,
-based on the difference in mean absolute error, the differences between
-the models are small (Table \ref{tab:score_cell_fivepercent}).
+Here we summarize the results of our analyses that inform the choice
+between the CAR and SPDE models.
+
+For Experiment 1 (full cells held out) for cells in the Minnesota
+subregion held out of the fitting process, the CAR model outperforms
+the SPDE model based on the posterior distribution over the predictive
+metric values (Table \ref{tab:score_cell_fivepercent}). For the posterior
+mean predictions, the SPDE model appears to outperform the CAR model
+to a lesser degree, but we do not have any uncertainty estimates for
+this comparison. Coverage and interval lengths are similar between
+the two models (Table \ref{tab:coverage_cell_fivepercent}). From
+a practical perspective, based on the difference in mean absolute
+error, the differences between the models are small (Table \ref{tab:score_cell_fivepercent}).
 
@@
 
-In Experiment 2 (individual trees held out), we have evidence (posterior probability of 0.93)
-that the SPDE model is better based on the Brier score, but the Brier
-score values for the two models are numerically almost the same (Table
-\ref{tab:score_tree}).
+In Experiment 2 (individual trees held out), we have evidence (posterior
+probability of 0.93) that the SPDE model is better based on the Brier
+score, but the Brier score values for the two models are numerically
+almost the same (Table \ref{tab:score_tree}).
 
@@
 For this final data product, we ran the model using the CAR specification
-for 150,000 iterations, discarding the first 25,000 iterations for
-burn-in and retaining 250 subsampled iterations from the remaining
-125,000 iterations (subsampled to limit storage needs and the size
-of the data product). Based on graphical checks and calculation of
-effective sample size values, mixing was generally reasonable, but
-for some of the hyperparameters was relatively slow, particularly
-for less common taxa. Despite this, mixing for the variables of substantive
-interest -- the proportions -- was good, with effective sample sizes
-for the final product generally near 250.
+with all of the data (including the data held out
+in the model comparison analyses) for 150,000 iterations with the
+same burn-in and subsampling details as described in Section \ref{sec:Model-comparison}.
+Based on graphical checks and calculation of effective sample size
+values, mixing was generally reasonable, but for some of the hyperparameters
+was relatively slow, particularly for less common taxa. Despite this,
+mixing for the variables of substantive interest -- the proportions
+-- was good, with effective sample sizes for the final product generally
+near 250.
 
@@
 
-The data product is being made publicly available at the NIS Data
-Portal (accession number to be inserted when available) under the
-CC BY license as version 0.4 as of August 2015. The product is in
-the form of a netCDF-4 file, with dimensions x-coordinate, y-coordinate,
-and MCMC iteration. There is one variable per taxon. In addition,
-dynamic visualizations of the product using the Shiny tool are available
-at http://gandalf.berkeley.edu:3838/paciorek/setVegComp [this URL
-is temporary and will be updated]. The PalEON Project (in particular
-the first author) will continue to maintain this product, releasing
-new versions as additional data in Illinois, Indiana and Ohio are
-digitized. Note that digitization of data from Illinois and Indiana
-is ongoing, and digitization of additional data from Ohio is planned
-as well. As a result, at some point we expect to have complete data
-for the midwestern half of the domain.
+The data product is publicly available at the NIS
+Data Portal as DOI:10.5072/FK2/dfc5e9748c8df4c3c275eab0f42d9744 under
+the CC BY 4.0 license as version 0.4 as of January 2016 \cite{paci:etal:data:2016}.
+The product is in the form of a netCDF-4 file, with dimensions x-coordinate,
+y-coordinate, and MCMC iteration. There is one variable per taxon.
+In addition, dynamic visualizations of the product using the Shiny
+tool are available at \href{https://www3.nd.edu/~paleolab/paleonproject/maps}{https://www3.nd.edu/$\sim$paleolab/paleonproject/maps}.
+The PalEON Project (in particular the first author) will continue
+to maintain this product, releasing new versions as additional data
+in Illinois, Indiana and Ohio are digitized. Note that digitization
+of data from Illinois and Indiana is ongoing, and digitization of
+additional data from Ohio is planned as well. As a result, at some
+point we expect to have complete data for the midwestern half of the
+domain.
 
@@
 
-Given the density of data and the limited differences seen between
-the CAR and SPDE models, we expect the data product to be reasonably
-robust to the choice of spatial model, particularly in those areas
-with complete data. However, additional investigation of other statistical
-representations is of interest, in particular nonstationary spatial
-models and use of covariates. The biggest shortcoming of the current
-model is its inability to account for local features such as rivers
-(e.g., the Minnesota River riparian corridor can be seen in the raw
-data but is missing in the statistical estimates). The current model,
-by using a simple stationary spatial model that smooths as a function
-of Euclidean distance, does not account for topographic, soil, or
-other features. An additional drawback of the product is its focus
-on composition, which does not directly tell us about tree density
-or other aspects of vegetation structure, and in particular does not
-distinguish between closed forest, savanna, and prairie. This limitation
-is particularly critical at the prairie-forest transition in Minnesota,
-Wisconsin, Illinois, and into Indiana \cite{transeau1935prairie}.
-
-This latter limitation can be addressed by developing estimates of
-absolute abundance (e.g., biomass) rather than the relative abundance
-compositional data used here. A gridded dataset of biomass, stem density,
-and basal area is already available for Minnesota, Wisconsin, and
-northern Michigan \cite{goring2015composition}, based on the PLS
-data. An extension to southern Michigan, Illinois, and Indiana is
-planned. We are currently developing statistical estimates of biomass
+In the parts of the modeled region with spatially
+complete data (in particular, Minnesota, Wisconsin, and Michigan),
+the statistical estimates of forest composition closely match the
+patterns apparent in the raw data (Fig. \ref{fig:select_maps}), as
+expected. In these areas, the estimated tree composition from the
+model has the advantage of downweighting unusual or outlier values
+in the empirical proportions of individual grid cells, which are likely
+due to stochastic sampling variation within that grid cell (compare
+the first two columns in Fig. \ref{fig:select_maps}). Some stochastic
+variation is expected given that, even in the most spatially complete
+regions, each grid cell contains an average of 124 trees (120-135
+is typical) \cite{goring2015composition} and some cells contain
+many fewer trees. Hence, some smoothing of this stochastic variation
+is appropriate. This smoothing is based on information on data from
+nearby cells, and the estimates from the model reflect the smooth
+trends in forest composition across the spatial domain. A partial
+cost is that these maps can smooth out sharp ecotones or other forms
+of true spatial heterogeneity, particularly in areas with sparse data
+(including areas with low tree density). For example, the sharp increase
+in Elm along the Minnesota River (Fig. \ref{fig:select_maps}, first
+column) likely represents a real ecotone in the settlement-era vegetation.
+Vegetation gradients and ecotones were sharper in the settlement-era
+forests in the upper Midwest than in contemporary forests \cite{goring2015composition},
+and the modeled estimates may partially obscure this change. Users
+interested in using the original unsmoothed data are directed to the
+data product described earlier \cite{Gori:etal:data:2016}. Additional
+investigation of other statistical representations to better capture
+sharp gradients is of interest, in particular nonstationary spatial
+models and use of covariates. Potential environmental covariates include
+soils, firebreaks, and topography \cite{grimm1984fire,shea2014reconstructing}.
+Here, however, we chose to limit our model to be exclusively a function
+of spatial distance in order to avoid dependence on the environmental
+drivers of pre-settlement forest composition that might lead to circular
+reasoning in subsequent inferences based on this dataset. Use of covariates
+could also lead to prediction that a taxa is present well beyond its
+range boundary in places where data are sparse.
+
+A key advance of this work over prior reconstructions
+of settlement-era vegetation lies in the estimates of uncertainty
+across the spatial domain. These estimates of uncertainty include
+the sampling uncertainty within grid cells (as do the within-grid
+cell estimates of uncertainty available from the raw proportions),
+but, because this is a spatial model, predictions and their associated
+uncertainty estimates are also informed by the information content
+of nearby cells. The maps of standard errors across species (Fig.
+\ref{fig:select_maps}, third column) highlight the advantages of
+this approach in areas of high data coverage (Minnesota, Wisconsin,
+Michigan) and in areas of sparse coverage (e.g., Illinois, Indiana,
+parts of Ohio). Where there are not large gaps in the data, the model
+provides low and fairly smooth estimates of uncertainty. Uncertainty
+is generally higher in the eastern subdomain than in the areas of
+the midwestern subdomain with high data coverage because of missing
+townships and lower sampling density even in townships with data.
+In areas of sparse coverage and in areas with low tree density (e.g.,
+southwestern Minnesota), the standard error of our estimates increases
+appropriately. Nevetheless, these uncertainties surround reasonable
+estimates of trends in composition. For example, the model does a
+good job of capturing the oak ecotone in Indiana and Illinois, representing
+a shift from oak savannas and woodlands to closed mesic forests (Fig.
+\ref{fig:select_maps}). Experiment 1 showed that both models predicted
+composition at cells with no data reasonably well, mimicking the case
+of sparsely sampled data and giving confidence in the broad spatial
+patterns predicted in more poorly sampled regions, particularly those
+with regular, but sparse sampling that mimic the experiment (Illinois
+and Indiana, but not Ohio). The apparent blockiness of uncertainty
+estimates in a few places such as Ohio is caused by spatial gaps and
+variations in sampling resolution. Absolute uncertainty generally
+increases with abundance for all taxa (Fig. \ref{fig:select_maps},
+column 3).
+
+The exploration of alternative approaches to spatial
+modeling of composition showed similar results for the SPDE and CAR
+models, both in terms of prediction accuracy and performance of prediction
+intervals. Small differences among the various metrics of goodness
+of fit favored each model in turn, but applied users of the models
+would find little pragmatic difference affecting scientific inference.
+Ultimately, we slightly favor the CAR model, because it avoids the
+boundary effects apparent in the SPDE model at the edges of the domain.
+
+The models presented here estimate only the relative
+abundance of tree taxa, which does not directly tell us about tree
+density or other aspects of vegetation structure. This becomes a particular
+limitation for interpreting vegetation where trees become sparse at
+the prairie-forest transition from northern Minnesota through southern
+Illinois \cite{transeau1935prairie}. Our model (correctly) predicts
+that tree composition there is dominated by oak, but this is less
+useful considering the sparseness of trees. This limitation can be
+addressed by developing estimates of absolute abundance (e.g., biomass)
+rather than compositional estimates. A gridded dataset of biomass,
+stem density, and basal area is already available for Minnesota, Wisconsin,
+and northern Michigan \cite{goring2015composition}, based on the
+PLS data. An extension to southern Michigan, Illinois, and Indiana
+is planned. We are currently developing statistical estimates of biomass
 for Minnesota, Wisconsin, and Michigan using a statistical model applied
@@
 {\bf Define $\bar{w}_{ip}=\frac{1}{n_{i}}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
-be the average of the $W$ values for the \emph{p}th taxon in the \emph{i}th
+as the average of the $W$ values for the \emph{p}th taxon in the \emph{i}th
 grid cell and $\bar{w}_{p}=\{\bar{w}_{ip}\},$ $i=1,\ldots,m$.} Let
@@
 Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
-We thank Madeline Ruid, Ben Seliger, Morgan Ross and Daniel Handel
+We thank Madeline Ruid, Benjamin Seliger, Morgan Ripp and Daniel Handel
 for processing of the southern Michigan data. Indiana and Illinois
@@
 in the McLachlan lab. 
-% This work was carried out by the PalEON Project
+%This work was carried out by the PalEON Project
 %with support from the National Science Foundation MacroSystems Program
@@
 
-\section*{References}
+%\section*{References}
 % Either type in your references using
@@
 
-\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
-
-%\begin{thebibliography}{10}
-
-%\end{thebibliography}
+\begin{thebibliography}{10}
+
+\bibitem{Cogb:etal:2002}
+Cogbill CV, Burk J, Motzkin G.
+\newblock The forests of presettlement {N}ew {E}ngland, {USA}: spatial and
+  compositional patterns based on town proprietor surveys.
+\newblock Journal of Biogeography. 2002;29:1279--1304.
+
+\bibitem{thompson2013four}
+Thompson JR, Carpenter DN, Cogbill CV, Foster DR.
+\newblock Four centuries of change in northeastern {United States} forests.
+\newblock PLOS ONE. 2013;8(9):e72540.
+
+\bibitem{bourdo1956review}
+Bourdo EA.
+\newblock A review of the {G}eneral {L}and {O}ffice survey and of its use in
+  quantitative studies of former forests.
+\newblock Ecology. 1956;37:754--768.
+
+\bibitem{pattison1957beginnings}
+Pattison WD, et~al.
+\newblock Beginnings of the American rectangular land survey system, 1784-1800.
+\newblock Chicago: University of Chicago; 1957.
+
+\bibitem{schulte2001original}
+Schulte LA, Mladenoff DJ.
+\newblock The original {US} public land survey records: their use and
+  limitations in reconstructing presettlement vegetation.
+\newblock Journal of Forestry. 2001;99(10):5--10.
+
+\bibitem{goring2015composition}
+Goring SJ, Williams JW, Mladenoff DJ, Cogbill CV, Record S, Paciorek CJ, et~al.
+\newblock Changes in Forest Composition, Stem Density, and Biomass from the
+  Settlement Era to Present in the upper {M}idwestern {U}nited {S}tates.
+\newblock PLOS ONE. 2016;p. under revision.
+
+\bibitem{liu2011broadscale}
+Liu F, Mladenoff DJ, Keuler NS, Moore LS.
+\newblock Broadscale variability in tree data of the historical Public Land
+  Survey and its consequences for ecological studies.
+\newblock Ecological Monographs. 2011;81(2):259--275.
+
+\bibitem{mladenoff2002narrowing}
+Mladenoff DJ, Dahir SE, Nordheim EV, Schulte LA, Guntenspergen GG.
+\newblock Narrowing Historical Uncertainty: Probabilistic Classification of
+  Ambiguously Identified Tree Species in Historical Forest Survey Data.
+\newblock Ecosystems. 2002;5(6):539--553.
+
+\bibitem{foster1998land}
+Foster DR, Motzkin G, Slater B.
+\newblock Land-use history as long-term broad-scale disturbance: regional
+  forest dynamics in central {N}ew {England}.
+\newblock Ecosystems. 1998;1(1):96--119.
+
+\bibitem{rhemtulla2009legacies}
+Rhemtulla JM, Mladenoff DJ, Clayton MK.
+\newblock Legacies of historical land use on regional forest composition and
+  structure in {W}isconsin, {USA} (mid-1800{s}-1930{s}-2000{s}).
+\newblock Ecological Applications. 2009;19(4):1061--1078.
+
+\bibitem{friedman2005regional}
+Friedman SK, Reich PB.
+\newblock Regional legacies of logging: departure from presettlement forest
+  conditions in northern {M}innesota.
+\newblock Ecological Applications. 2005;15(2):726--744.
+
+\bibitem{rhemtulla2009historical}
+Rhemtulla JM, Mladenoff DJ, Clayton MK.
+\newblock Historical forest baselines reveal potential for continued carbon
+  sequestration.
+\newblock Proceedings of the National Academy of Sciences.
+  2009;106(15):6082--6087.
+
+\bibitem{gray2012forest}
+Gray AN, Brandeis TJ, Shaw JD, McWilliams WH, Miles PD, et~al.
+\newblock Forest Inventory and Analysis Database of the {United States} of
+  {America} ({FIA}).
+\newblock Vegetation databases for the 21st century--Biodiversity \& Ecology.
+  2012;4:255--264.
+
+\bibitem{Whit:1996}
+Whitney GG.
+\newblock From Coastal Wilderness to Fruited Plain: a History of Environmental
+  Change in Temperate {N}orth {A}merica from 1500 to the Present.
+\newblock Cambridge University Press; 1996.
+
+\bibitem{Blac:etal:2006}
+Black BA, Ruffner CM, Abrams MD.
+\newblock {N}ative {A}merican influences on the forest composition of the
+  {A}llegheny {P}lateau, northwest {P}ennsylvania.
+\newblock Canadian Journal of Forest Research. 2006;36(5):1266--1275.
+
+\bibitem{munoz2014defining}
+Munoz SE, Mladenoff DJ, Schroeder S, Williams JW.
+\newblock Defining the spatial patterns of historical land use associated with
+  the indigenous societies of eastern {N}orth {A}merica.
+\newblock Journal of Biogeography. 2014;41(12):2195--2210.
+
+\bibitem{Gori:etal:data:2016}
+Goring SJ, {University of Wisconsin}.
+\newblock Settlement-era gridded tree composition, midwestern {US}: Level 1.
+\newblock Long Term Ecological Research Network Data Portal; 2016. Paleon.2.0.
+
+\bibitem{Cogb:dataNE:2016}
+Cogbill CV.
+\newblock Settlement-era tree composition, eastern {US}: level 1.
+\newblock Long Term Ecological Research Network Data Portal; 2016. Paleon.4.11.
+
+\bibitem{Cogb:dataOH:2016}
+Cogbill CV.
+\newblock Settlement-era tree composition, {O}hio: level 1.
+\newblock Long Term Ecological Research Network Data Portal; 2016. Paleon.5.11.
+
+\bibitem{Bane:etal:2004}
+Banerjee S, Carlin BP, Gelfand AE.
+\newblock Hierarchical Modeling and Analysis for Spatial Data.
+\newblock Boca Raton, Florida: Chapman \& Hall; 2004.
+
+\bibitem{Lind:etal:2011}
+Lindgren F, Rue H, Lindstr{\"o}m J.
+\newblock An explicit link between {G}aussian fields and {G}aussian {M}arkov
+  random fields: the stochastic partial differential equation approach.
+\newblock Journal of the Royal Statistical Society B. 2011;73:423--498.
+
+\bibitem{Rue:Held:2005}
+Rue H, Held L.
+\newblock Gaussian {M}arkov Random Fields: Theory and Applications.
+\newblock Boca Raton: Chapman \& Hall; 2005.
+
+\bibitem{Paci:2013}
+Paciorek CJ.
+\newblock Spatial models for point and areal data using {M}arkov random fields
+  on a fine grid.
+\newblock Electronic Journal of Statistics. 2013;7:946--972.
+
+\bibitem{Gelm:2006}
+Gelman A.
+\newblock Prior distributions for variance parameters in hierarchical models
+  (comment on article by {B}rowne and {D}raper).
+\newblock Bayesian Analysis. 2006;1(3):515--534.
+
+\bibitem{Chri:etal:2006}
+Christensen OF, Roberts GO, Sk{\"o}ld M.
+\newblock Robust {M}arkov chain {M}onte {C}arlo methods for spatial generalized
+  linear mixed models.
+\newblock Journal of Computational and Graphical Statistics. 2006;15:1--17.
+
+\bibitem{Tan:Nott:2013}
+Tan LSL, Nott DJ.
+\newblock Variational inference for generalized linear mixed models using
+  partially noncentered parametrizations.
+\newblock Statistical Science. 2013;28(2):168--188.
+
+\bibitem{McCu:Ross:1994}
+McCulloch R, Rossi PE.
+\newblock An exact likelihood analysis of the multinomial probit model.
+\newblock Journal of Econometrics. 1994;64(1):207--240.
+
+\bibitem{Shab:Well:2011}
+Shaby B, Wells M.
+\newblock Exploring an adaptive {M}etropolis algorithm.
+\newblock Department of Statistics, Duke University; 2011. 2011-14.
+
+\bibitem{R:2014}
+{R Core Team}. R: A Language and Environment for Statistical Computing.
+\newblock Vienna, Austria; 2014.
+\newblock Available from: \url{http://www.R-project.org/}.
+
+\bibitem{Edde:Fran:2011}
+Eddelbuettel D, Francois R.
+\newblock Rcpp: {S}eamless {R} and {C}++ integration.
+\newblock Journal of Statistical Software. 2011;40:1--18.
+\newblock Available from: \url{http://www.jstatsoft.org/v40/i08}.
+
+\bibitem{Furr:Sain:2010}
+Furrer R, Sain SR.
+\newblock {spam}: A Sparse Matrix {R} Package with Emphasis on {MCMC} Methods
+  for {G}aussian {M}arkov Random Fields.
+\newblock Journal of Statistical Software. 2010;36(10):1--25.
+\newblock Available from: \url{http://www.jstatsoft.org/v36/i10/}.
+
+\bibitem{Gnei:etal:2007}
+Gneiting T, Balabdaoui F, Raftery AE.
+\newblock Probabilistic forecasts, calibration and sharpness.
+\newblock Journal of the Royal Statistical Society: Series B (Statistical
+  Methodology). 2007;69(2):243--268.
+
+\bibitem{Krnj:Drap:2014}
+Krnjaji\'{c} M, Draper D.
+\newblock Bayesian model comparison: {L}og scores and {DIC}.
+\newblock Statistics and Probability Letters. 2014;88:9--14.
+
+\bibitem{paci:etal:data:2016}
+Paciorek CJ, Goring SJ, Thurman AL, Cogbill CV, Williams JW, Mladenoff DJ,
+  et~al.
+\newblock Settlement-era gridded tree composition, northeastern {US}: level 2.
+\newblock Long Term Ecological Research Network Data Portal; 2016. Paleon.1.2.
+
+\bibitem{grimm1984fire}
+Grimm EC.
+\newblock Fire and other factors controlling the {B}ig {W}oods vegetation of
+  {M}innesota in the mid-nineteenth century.
+\newblock Ecological Monographs. 1984;54(3):291--311.
+
+\bibitem{shea2014reconstructing}
+Shea ME, Schulte LA, Palik BJ.
+\newblock Reconstructing vegetation past: pre-{E}uro-{A}merican vegetation for
+  the midwest driftless area, {USA}.
+\newblock Ecological Restoration. 2014;32(4):417--433.
+
+\bibitem{transeau1935prairie}
+Transeau EN.
+\newblock The prairie peninsula.
+\newblock Ecology. 1935;16(3):423--437.
+
+\end{thebibliography}
 

### 2016-02-03 paperPlos_track.tex
revised and added for plos revision january 2016

--- 2016-02-03 paperPlos.tex
@@
 we describe the final data product. In \emph{Discussion}
-we discuss the uncertainties estimated by and the
+we discuss \textcolor{red}{the uncertainties estimated by }and the
 limitations of the statistical model, and we list related data products
@@
 six New England states (the eastern subdomain), data are aggregated
-at the township level. We make predictions for all
-of the states listed above; these constitute our core domain. There
+at the township level. \textcolor{red}{We make predictions for all
+of the states listed above; these constitute our core domain. }There
 are also data from a single township in Quebec and a single township
-in northern Delaware; these data help inform predictions
+in northern Delaware\textcolor{red}{; these data help inform predictions
 in nearby locations within our core domain, but predictions are not
-made for Quebec or Delaware. Digitization of PLS data in Minnesota,
+made for Quebec or Delaware}. Digitization of PLS data in Minnesota,
 Wisconsin and Michigan is essentially complete, with PLS data for
@@
 per cell. The gridded data at the 8 km resolution for the midwest
-subdomain are available through the NIS data portal as
-DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5 \cite{Gori:etal:data:2016}.
+subdomain are available through the NIS data portal \textcolor{red}{as
+DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5} \textcolor{red}{\cite{Gori:etal:data:2016}}.
 The TPS data were compiled by C.V. Cogbill from a myriad of archival
 sources representing land division surveys conducted in connection
-with local settlement and are available through the
+with local settlement\textcolor{red}{{} and are available through the
 NIS data portal as DOI:10.5072/FK2/4d44aebe71dbf65290d8ce1bde03598b
-and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}.
+and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}}.
 
@@
 located at (-71000 m, 58000 m). For the midwestern subdomain we use
-the western-most 146 by 180 grid of cells when fitting
-the statistical models. For the eastern subdomain we use the eastern-most
+the western-most 146 by 180 grid of cells\textcolor{red}{{} when fitting
+the statistical models}. For the eastern subdomain we use the eastern-most
 180 by 180 grid of cells and then omit 23 rows of cells in the north
@@
 \end{enumerate}
-At a high level, the Bayesian statistical model estimates
+\textcolor{red}{At a high level, the Bayesian statistical model estimates
 composition across the domain, even in locations with sparse or no
@@
 called hyperparameters that determine the correlation structure of
-the process and are also estimated based on the data.
+the process and are also estimated based on the data. }
 
@@
 
-In the remainder of this section we provide the technical
+\textcolor{red}{In the remainder of this section we provide the technical
 specification of the model and of the computations involved in fitting
-the model.
+the model.}
 
@@
 of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.
-A multinomial probit model is similar to logistic
+\textcolor{red}{A multinomial probit model is similar to logistic
 regression, used for modeling a binary outcome based on an underlying
 probability the outcome will occur, but generalizes to modeling a
-categorical variable based on probabilities for each category.
+categorical variable based on probabilities for each category. }
 
 The critical component of the statistical model is the representation
-of $\alpha_{p}(s)$ as a spatial process. A spatial
+of $\alpha_{p}(s)$ as a spatial process. \textcolor{red}{A spatial
 process is a statistical representation that models spatially-correlated
-values. It provides a prior structure that serves to smooth across
+values. It provides} a prior structure that serves to smooth across
 noise in the observations and allows for prediction at locations based
@@
 random field (MRF) approximation to a Gaussian process with \matern
-covariance \cite{Lind:etal:2011}. These models are
+covariance \cite{Lind:etal:2011}. \textcolor{red}{These models are
 specific statistical formulations of spatial processes that represent
 spatial correlation by defining neighborhoods around each location
-that are used to help inform predictions at the location.
+that are used to help inform predictions at the location. }
 
@@
 
-MRF models represent the neighborhood information
-by working directly with the precision matrix (the
-inverse of the covariance matrix) of the values of the spatial process,
+MRF models \textcolor{red}{represent the neighborhood information
+}by working directly with the precision matrix \textcolor{red}{(the
+inverse of the covariance matrix)} of the values of the spatial process,
 so calculation of the prior density of $\alpha_{p}$ is computationally
-simple \cite{Rue:Held:2005}. However, in situations
-in which the likelihood is not normal, such
-as our multinomial likelihood, it can be difficult to set up effective
+simple \cite{Rue:Held:2005}. \textcolor{red}{However,} in situations
+\textcolor{red}{in which} the likelihood is not normal, \textcolor{red}{such
+as our multinomial likelihood}, it can be difficult to set up effective
 MCMC algorithms that are able to move in the high-dimensional space
@@
 spatial models that we considered; in \emph{Model comparison},
-we evaluate the models on held-out data to choose
+\textcolor{red}{we evaluate the models on held-out data} to choose
 between the two.
@@
 
-Our first model is a standard conditional autoregressive (CAR) model;
-technical details can be found in \cite{Bane:etal:2004}. We use
-a standard form of this model that treats the four
+Our first model is a standard conditional autoregressive (CAR) model\textcolor{red}{;
+technical details can be found in} \cite{Bane:etal:2004}. We use
+a standard form of this model \textcolor{red}{that} treats the four
 cardinal neighbors of each grid cell as the neighbors of the grid
@@
 not full-rank, but is of rank $m-1$; this gives an improper prior
-on an implicit overall mean for the process values. Note
+on an implicit overall mean for the process values. \textcolor{red}{Note
 that we specify an explicit mean of zero because a non-zero mean would
-not be identifiable in light of the implicit mean. This specification
+not be identifiable in light of the implicit mean.} This specification
 is called an \textit{intrinsic conditional autoregression (ICAR)}
@@
 GP models are computationally challenging for large datasets because
-of computational manipulations involving large covariance
+of \textcolor{red}{computational} manipulations involving large covariance
 matrices. Given this, \cite{Lind:etal:2011} proposed a new framework
@@
 
-Gaussian processes are generally constructed using
+\textcolor{red}{Gaussian processes are generally constructed using
 one of a number of correlation functions that define how the strength
 of correlation between the values of the process at two locations
-decays as a function of the distance between the locations. We consider
-Gaussian processes in the commonly-used \matern
+decays as a function of the distance between the locations. }We consider
+Gaussian processes in the \textcolor{red}{commonly-used} \matern
 class, using the following parameterization of the \matern correlation
@@
 GPs with $\nu=1$. To implement this Lindgren model, one modifies
-the \emph{Q} matrix defined previously as follows based
+the \emph{Q} matrix defined previously as follows \textcolor{red}{based
 on the technical specification of the precision matrix provided in
-\cite{Lind:etal:2011}. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
+\cite{Lind:etal:2011}}. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
 elements of \emph{Q} are $4+a^{2}$. The entries corresponding to cardinal
@@
 $\rho$. In particular $\rho$ allows us to estimate the locality
-of the spatial smoothing. As $\rho$ decreases, the
+of the \textcolor{red}{spatial} smoothing. As $\rho$ decreases, the
 model uses increasingly localized data to estimate the compositional
@@
 at $\pm10$), and $\{\rho_{p}\}$, which we give uniform priors on
-the interval $(0.1,\exp(5))$. These various hyperparameters
+the interval $(0.1,\exp(5))$. \textcolor{red}{These various hyperparameters
 are unknown parameters that control the spatial structure of the two
 spatial models and are estimated from both the data and the prior
-distributions just specified based on the Bayesian approach.
+distributions just specified based on the Bayesian approach.}
 
@@
 
-We compared the CAR and SPDE models by holding out
-data from the fitting process and assessing the fit of the model
-on the held-out data. We used two experiments with
-held-out data:
+We compared the CAR and SPDE models \textcolor{red}{by holding out
+data} from the fitting process and assessing the fit of the model
+on the held-out data. We used two experiments\textcolor{red}{{} with
+held-out data}:
 \begin{enumerate}
@@
 that approximately goes through Duluth, Minnesota) and hereafter referred
-to as the ``Minnesota subregion''. We chose this subregion for evaluation
+to as the ``Minnesota subregion''. We chose this subregion for \textcolor{red}{evaluation}
 because of its high data density, allowing us to experiment with the
@@
 is an indicator variable taking value either 0 or 1 depending on whether
-the \emph{j}th held-out tree in the \emph{i}th cell is of taxon \emph{p}. $\hat{\theta}_{ip}=Y_{ip}/n_{i}$
+the \emph{j}th held-out tree in the \emph{i}th cell is of taxon \emph{p}. \textcolor{red}{$\hat{\theta}_{ip}=Y_{ip}/n_{i}$
 is the empirical proportion in category \emph{p} for the $n_{i}$ held-out
-trees in cell \emph{i}. We calculated each of the metrics in two ways.
+trees in cell \emph{i}.} We calculated each of the metrics in two ways.
 First, we used the posterior mean composition estimates (as an evaluation
@@
 \emph{i}. 
-\item Log predictive density: This metric takes the log of the probability
-density of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
+\item Log predictive density: This metric takes the log of the \textcolor{red}{probability
+}density of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
 summing on the log scale across all of the held-out data.
@@
 \item (Experiment 1 only) Weighted root mean square prediction error (RMSPE),
-
+\textcolor{red}{
 \[
@@
 \]
-and mean absolute error (MAE)
+}and mean absolute error (MAE)\textcolor{red}{
 \[
@@
 \]
-These metrics calculate the error of the estimated proportions relative
+}These metrics calculate the error of the estimated proportions relative
 to the empirical proportions based on the held-out trees, averaging
@@
 proportion of samples for which the model has a lower value of the
-metric. Also note that in addition to allowing comparison
+metric. \textcolor{red}{Also note that in addition to allowing comparison
 between models the MAE and RMSPE metrics allow one to assess absolute
-performance of each model in predicting composition.
+performance of each model in predicting composition.}
 
@@
 
-Here we summarize the results of our analyses that inform the choice
+Here we summarize the results of our \textcolor{red}{analyses} that inform the choice
 between the CAR and SPDE models.
@@
 For this final data product, we ran the model using the CAR specification
-with all of the data (including the data held out
-in the model comparison analyses) for 150,000 iterations with the
+\textcolor{red}{with all of the data (including the data held out
+in the model comparison analyses)} for 150,000 iterations with the
 same burn-in and subsampling details as described in Section \ref{sec:Model-comparison}.
@@
 
-The data product is publicly available at the NIS
-Data Portal as DOI:10.5072/FK2/dfc5e9748c8df4c3c275eab0f42d9744 under
+The data product is publicly available at the \textcolor{red}{NIS
+Data Portal as DOI:10.5072/FK2/dfc5e9748c8df4c3c275eab0f42d9744} under
 the CC BY 4.0 license as version 0.4 as of January 2016 \cite{paci:etal:data:2016}.
@@
 
-In the parts of the modeled region with spatially
+\textcolor{red}{In the parts of the modeled region with spatially
 complete data (in particular, Minnesota, Wisconsin, and Michigan),
@@
 could also lead to prediction that a taxa is present well beyond its
-range boundary in places where data are sparse.
-
-A key advance of this work over prior reconstructions
+range boundary in places where data are sparse.}
+
+\textcolor{red}{A key advance of this work over prior reconstructions
 of settlement-era vegetation lies in the estimates of uncertainty
@@
 increases with abundance for all taxa (Fig. \ref{fig:select_maps},
-column 3).
-
-The exploration of alternative approaches to spatial
+column 3).}
+
+\textcolor{red}{The exploration of alternative approaches to spatial
 modeling of composition showed similar results for the SPDE and CAR
@@
 Ultimately, we slightly favor the CAR model, because it avoids the
-boundary effects apparent in the SPDE model at the edges of the domain.
-
-The models presented here estimate only the relative
+boundary effects apparent in the SPDE model at the edges of the domain.}
+
+\textcolor{red}{The models presented here estimate only the relative
 abundance of tree taxa, which does not directly tell us about tree
@@
 that tree composition there is dominated by oak, but this is less
-useful considering the sparseness of trees. This limitation can be
+useful considering the sparseness of trees.} This limitation can be
 addressed by developing estimates of absolute abundance (e.g., biomass)
@@
 years who have preserved, collected, and digitized survey records,
-in particular John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed
+in particular \textcolor{red}{John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed
 Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
-We thank Madeline Ruid, Benjamin Seliger, Morgan Ripp and Daniel Handel
+We thank Madeline Ruid, Benjamin Seliger, Morgan Ripp and Daniel Handel}
 for processing of the southern Michigan data. Indiana and Illinois

### 2016-02-03 plos_original/paper.tex
revised and added for plos revision january 2016

--- 2015-09-05 arxiv/paperArxiv.tex
@@
-\documentclass[12pt]{article}
-\usepackage[]{graphicx}
-\usepackage[]{color}
+\batchmode
+\makeatletter
+\def\input@path{{/accounts/gen/vis/paciorek/research/jmac/composition_paper//}}
+\makeatother
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
 %% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-
-%\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-%\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-%\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-%\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-%\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-%\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-%\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-%\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-%\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-%\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-%\usepackage{framed}
-%\makeatletter
-%\newenvironment{kframe}{%
-% \def\at@end@of@kframe{}%
-% \ifinner\ifhmode%
-%  \def\at@end@of@kframe{\end{minipage}}%
-%  \begin{minipage}{\columnwidth}%
-% \fi\fi%
-% \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
-% \colorbox{shadecolor}{##1}\hskip-\fboxsep
-%     % There is no \\@totalrightmargin, so:
-%     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
-% \MakeFramed {\advance\hsize-\width
-%   \@totalleftmargin\z@ \linewidth\hsize
-%   \@setminipage}}%
-% {\par\unskip\endMakeFramed%
-% \at@end@of@kframe}
-%\makeatother
-
-%\definecolor{shadecolor}{rgb}{.97, .97, .97}
-%\definecolor{messagecolor}{rgb}{0, 0, 0}
-%\definecolor{warningcolor}{rgb}{1, 0, 1}
-%\definecolor{errorcolor}{rgb}{1, 0, 0}
-%\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
 
@@
 \makeatother
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-\lyxaddress{\textsuperscript{1}Department of Statistics, University of California, Berkeley, California, USA}
-
-\lyxaddress{\textsuperscript{2}Department of Geography, University of Wisconsin, Madison, Wisconsin, USA}
-
-\lyxaddress{\textsuperscript{3}Department of Statistics, University of Iowa, Iowa City, Iowa, USA}
-
-\lyxaddress{\textsuperscript{4}Harvard Forest, Harvard University, Petersham, Massachusetts, USA}
+\lyxaddress{\textsuperscript{1}Department of Statistics, University of California,
+Berkeley, California, USA}
+
+\lyxaddress{\textsuperscript{2}Department of Geography, University of Wisconsin,
+Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{3}Department of Statistics, University of Iowa,
+Iowa City, Iowa, USA}
+
+\lyxaddress{\textsuperscript{4}Harvard Forest, Harvard University, Petersham,
+Massachusetts, USA}
 
@@
 \newpage}
+\begin{quote}
+
+\end{quote}
 
@@
 \begin{figure}
-\label{fig:domain}\includegraphics[scale=1.3]{fig1}
+\label{fig:domain}\includegraphics[scale=1.3]{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}
 
@@
 \begin{figure}
-\includegraphics[scale=0.75]{fig2}
-
-\hspace{4mm}\includegraphics[scale=0.16]{legendRaw}\hspace{4mm}\includegraphics[scale=0.16]{legendProp}\hspace{3.5mm}\includegraphics[scale=0.16]{legendSD}
+\includegraphics[scale=0.75]{1_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig2.pdf}
+
+\hspace{4mm}\includegraphics[scale=0.16]{2_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendRaw.pdf}\hspace{4mm}\includegraphics[scale=0.16]{3_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendProp.pdf}\hspace{3.5mm}\includegraphics[scale=0.16]{4_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendSD.pdf}
 
@@
 \begin{figure}
-\includegraphics[scale=0.8]{fig3}
+\includegraphics[scale=0.8]{5_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig3.pdf}
 
@@
 
-% \bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
-\bibliography{paperArxiv}
+\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
 

### 2016-02-03 plos_original/paperPlos.tex
revised and added for plos revision january 2016

--- 2015-09-05 paperPlos.tex
@@
 one-half mile (800 m) spacing, for Ohio and westward during the period
-1785 to 1907 \cite{bourdo1956review,pattison1957beginnings,schulte2001original,goring2015composition}.
+1785 to 1907 \cite{bourdo1956review,pattison1957beginnings,schulte2001original,goring2015tr}.
 At each point 2-4 trees were identified, and the common name, diameter
@@
 consistency among points and the application of spatially-varying
-correction factors \cite{goring2015composition}
+correction factors \cite{goring2015tr}
 to accurately assess tree stem density, basal area and biomass from
@@
 common names, which requires matching to scientific names and standardization
-\cite{mladenoff2002narrowing,goring2015composition}.
+\cite{mladenoff2002narrowing,goring2015tr}.
 
 Logging, agriculture, and land abandonment have left an indelible
-mark on forests in the northeastern United States \cite{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
+mark on forests in the northeastern United States \cite{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015tr}.
 However most studies have assessed these effects in individual states
@@
 and with various spatial resolutions, from townships (36 square miles)
-to forest zones of hundreds or thousands of square miles. \cite{goring2015composition}
+to forest zones of hundreds or thousands of square miles. \cite{goring2015tr}
 provide a new dataset of forest composition, biomass, and stem density
@@
 In \emph{Model comparison} we quantitatively compare competing
-statistical specifications, and \emph{Data Product}
+statistical specifications, and in \emph{Data Product}
 we describe the final data product. In \emph{Discussion}
@@
 Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
-data and processing steps are available \cite{goring2015composition};
+data and processing steps are available \cite{goring2015tr};
 key elements include the use of only corner points, the use of only
@@
 because the set of trees in the dataset is roughly uniformly sampled
-across the cells or townships \cite{goring2015composition}.
+across the cells or townships \cite{goring2015tr}.
 
@@
 autoregressive model \cite{Bane:etal:2004} and a Gaussian Markov
-random field (MRF) approximation to a Gaussian process with Matern
+random field (MRF) approximation to a Gaussian process with \matern
 covariance \cite{Lind:etal:2011}.
@@
 to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second spatial
-model is the Lindgren approximation for \matern-based GPs with $\nu=1$.
+model is this Lindgren approximation for \matern-based GPs with $\nu=1$.
 To implement the Lindgren model, one modifies the \emph{Q} matrix defined
@@
 grid. This extension introduces a new set of latent variables, one
-per tree, that indicate the grid cell in which the tree is located
+per tree, that indicate the grid cells in which the trees are located
 and can be sampled within the MCMC as additional unknown parameters.
@@
 Portal (accession number to be inserted when available) under the
-CC BY license as version 0.4 as of August 2015. The product is in
+CC BY license as version 0.4. The product is in
 the form of a netCDF-4 file, with dimensions x-coordinate, y-coordinate,
@@
 and basal area is already available for Minnesota, Wisconsin, and
-northern Michigan \cite{goring2015composition}, based on the PLS
+northern Michigan \cite{goring2015tr}, based on the PLS
 data. An extension to southern Michigan, Illinois, and Indiana is
@@
 years who have preserved, collected, and digitized survey records,
-in particular John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed
-Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
-We thank Madeline Ruid, Ben Seliger, Morgan Ross and Daniel Handel
-for processing of the southern Michigan data. Indiana and Illinois
+in particular Ted Sickley, the US Forest Service, the Michigan Natural Features Inventory, the Minnesota Department of Natural Resources, and the Ohio Biological Survey.
+We thank Benjamin Seliger and Morgan Ripp for processing of the southern Michigan data.
+Indiana and Illinois
 data were made possible through the hard work of many Notre Dame undergraduates
 in the McLachlan lab. 
-% This work was carried out by the PalEON Project
+% John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed Schools, Ted Sickley, Ronald Stuckey
+%Madeline Ruid, Benjamin Seliger, Morgan Ripp and Daniel Handel% This work was carried out by the PalEON Project
 %with support from the National Science Foundation MacroSystems Program
@@
 
-\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
-
-%\begin{thebibliography}{10}
-
-%\end{thebibliography}
+\begin{thebibliography}{10}
+
+\bibitem{Cogb:etal:2002}
+Cogbill CV, Burk J, Motzkin G.
+\newblock The forests of presettlement {N}ew {E}ngland, {USA}: spatial and
+  compositional patterns based on town proprietor surveys.
+\newblock Journal of Biogeography. 2002;29:1279--1304.
+
+\bibitem{thompson2013four}
+Thompson JR, Carpenter DN, Cogbill CV, Foster DR.
+\newblock Four centuries of change in northeastern {United States} forests.
+\newblock PLOS ONE. 2013;8(9):e72540.
+
+\bibitem{bourdo1956review}
+Bourdo EA.
+\newblock A review of the General Land Office survey and of its use in
+  quantitative studies of former forests.
+\newblock Ecology. 1956;37:754--768.
+
+\bibitem{pattison1957beginnings}
+Pattison WD, et~al.
+\newblock Beginnings of the American rectangular land survey system, 1784-1800.
+\newblock Chicago: University of Chicago; 1957.
+
+\bibitem{schulte2001original}
+Schulte LA, Mladenoff DJ.
+\newblock The original {US} public land survey records: their use and
+  limitations in reconstructing presettlement vegetation.
+\newblock Journal of Forestry. 2001;99(10):5--10.
+
+\bibitem{goring2015tr}
+Goring S, Mladenoff DJ, Cogbill CV, Record S, Paciorek CJ, Jackson ST, et~al.
+\newblock Changes in Forest Composition, Stem Density, and Biomass from the
+  Settlement Era (1800s) to Present in the Upper Midwestern United States.
+\newblock bioRxiv; 2015.
+
+\bibitem{liu2011broadscale}
+Liu F, Mladenoff DJ, Keuler NS, Moore LS.
+\newblock Broadscale variability in tree data of the historical Public Land
+  Survey and its consequences for ecological studies.
+\newblock Ecological Monographs. 2011;81(2):259--275.
+
+\bibitem{mladenoff2002narrowing}
+Mladenoff DJ, Dahir SE, Nordheim EV, Schulte LA, Guntenspergen GG.
+\newblock Narrowing Historical Uncertainty: Probabilistic Classification of
+  Ambiguously Identified Tree Species in Historical Forest Survey Data.
+\newblock Ecosystems. 2002;5(6):539--553.
+
+\bibitem{foster1998land}
+Foster DR, Motzkin G, Slater B.
+\newblock Land-use history as long-term broad-scale disturbance: regional
+  forest dynamics in central {N}ew {England}.
+\newblock Ecosystems. 1998;1(1):96--119.
+
+\bibitem{rhemtulla2009legacies}
+Rhemtulla JM, Mladenoff DJ, Clayton MK.
+\newblock Legacies of historical land use on regional forest composition and
+  structure in {W}isconsin, {USA} (mid-1800{s}-1930{s}-2000{s}).
+\newblock Ecological Applications. 2009;19(4):1061--1078.
+
+\bibitem{friedman2005regional}
+Friedman SK, Reich PB.
+\newblock Regional legacies of logging: departure from presettlement forest
+  conditions in northern {M}innesota.
+\newblock Ecological Applications. 2005;15(2):726--744.
+
+\bibitem{rhemtulla2009historical}
+Rhemtulla JM, Mladenoff DJ, Clayton MK.
+\newblock Historical forest baselines reveal potential for continued carbon
+  sequestration.
+\newblock Proceedings of the National Academy of Sciences.
+  2009;106(15):6082--6087.
+
+\bibitem{gray2012forest}
+Gray AN, Brandeis TJ, Shaw JD, McWilliams WH, Miles PD, et~al.
+\newblock Forest Inventory and Analysis Database of the {United States} of
+  {America} ({FIA}).
+\newblock Vegetation databases for the 21st century--Biodiversity \& Ecology.
+  2012;4:255--264.
+
+\bibitem{Whit:1996}
+Whitney GG.
+\newblock From Coastal Wilderness to Fruited Plain: a History of Environmental
+  Change in Temperate {N}orth {A}merica from 1500 to the Present.
+\newblock Cambridge University Press; 1996.
+
+\bibitem{Blac:etal:2006}
+Black BA, Ruffner CM, Abrams MD.
+\newblock {N}ative {A}merican influences on the forest composition of the
+  {A}llegheny {P}lateau, northwest {P}ennsylvania.
+\newblock Canadian Journal of Forest Research. 2006;36(5):1266--1275.
+
+\bibitem{munoz2014defining}
+Munoz SE, Mladenoff DJ, Schroeder S, Williams JW.
+\newblock Defining the spatial patterns of historical land use associated with
+  the indigenous societies of eastern {N}orth {A}merica.
+\newblock Journal of Biogeography. 2014;41(12):2195--2210.
+
+\bibitem{Bane:etal:2004}
+Banerjee S, Carlin BP, Gelfand AE.
+\newblock Hierarchical Modeling and Analysis for Spatial Data.
+\newblock Boca Raton, Florida: Chapman \& Hall; 2004.
+
+\bibitem{Lind:etal:2011}
+Lindgren F, Rue H, Lindstr{\"o}m J.
+\newblock An explicit link between {G}aussian fields and {G}aussian {M}arkov
+  random fields: the stochastic partial differential equation approach.
+\newblock Journal of the Royal Statistical Society B. 2011;73:423--498.
+
+\bibitem{Rue:Held:2005}
+Rue H, Held L.
+\newblock Gaussian {M}arkov Random Fields: Theory and Applications.
+\newblock Boca Raton: Chapman \& Hall; 2005.
+
+\bibitem{Paci:2013}
+Paciorek CJ.
+\newblock Spatial models for point and areal data using {M}arkov random fields
+  on a fine grid.
+\newblock Electronic Journal of Statistics. 2013;7:946--972.
+
+\bibitem{Gelm:2006}
+Gelman A.
+\newblock Prior distributions for variance parameters in hierarchical models
+  (comment on article by {B}rowne and {D}raper).
+\newblock Bayesian Analysis. 2006;1(3):515--534.
+
+\bibitem{Chri:etal:2006}
+Christensen OF, Roberts GO, Sk{\"o}ld M.
+\newblock Robust {M}arkov chain {M}onte {C}arlo methods for spatial generalized
+  linear mixed models.
+\newblock Journal of Computational and Graphical Statistics. 2006;15:1--17.
+
+\bibitem{Tan:Nott:2013}
+Tan LSL, Nott DJ.
+\newblock Variational inference for generalized linear mixed models using
+  partially noncentered parametrizations.
+\newblock Statistical Science. 2013;28(2):168--188.
+
+\bibitem{McCu:Ross:1994}
+McCulloch R, Rossi PE.
+\newblock An exact likelihood analysis of the multinomial probit model.
+\newblock Journal of Econometrics. 1994;64(1):207--240.
+
+\bibitem{Shab:Well:2011}
+Shaby B, Wells M.
+\newblock Exploring an adaptive {M}etropolis algorithm.
+\newblock Department of Statistics, Duke University; 2011. 2011-14.
+
+\bibitem{R:2014}
+{R Core Team}. R: A Language and Environment for Statistical Computing.
+\newblock Vienna, Austria; 2014.
+\newblock Available from: \url{http://www.R-project.org/}.
+
+\bibitem{Edde:Fran:2011}
+Eddelbuettel D, Francois R.
+\newblock Rcpp: {S}eamless {R} and {C}++ integration.
+\newblock Journal of Statistical Software. 2011;40:1--18.
+\newblock Available from: \url{http://www.jstatsoft.org/v40/i08}.
+
+\bibitem{Furr:Sain:2010}
+Furrer R, Sain SR.
+\newblock {spam}: A Sparse Matrix {R} Package with Emphasis on {MCMC} Methods
+  for {G}aussian {M}arkov Random Fields.
+\newblock Journal of Statistical Software. 2010;36(10):1--25.
+\newblock Available from: \url{http://www.jstatsoft.org/v36/i10/}.
+
+\bibitem{Gnei:etal:2007}
+Gneiting T, Balabdaoui F, Raftery AE.
+\newblock Probabilistic forecasts, calibration and sharpness.
+\newblock Journal of the Royal Statistical Society: Series B (Statistical
+  Methodology). 2007;69(2):243--268.
+
+\bibitem{Krnj:Drap:2014}
+Krnjaji\'{c} M, Draper D.
+\newblock Bayesian model comparison: {L}og scores and {DIC}.
+\newblock Statistics and Probability Letters. 2014;88:9--14.
+
+\bibitem{transeau1935prairie}
+Transeau EN.
+\newblock The prairie peninsula.
+\newblock Ecology. 1935;16(3):423--437.
+
+\end{thebibliography}
+
+% \bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
 

### 2016-02-03 plos_revision/coverLetter.tex
revised and added for plos revision january 2016

--- 2015-04-13 coverLetter.tex
@@
 \begin{letter}{\mbox{}}
-\opening{To whom it may concern,}
+\opening{Dear Editor,}
 
-Please consider our submission as a research article.
+We are responding to reviewer comments for PONE-D-15-41521.
 
-This submission presents a new statistically-derived data product and the methods used to create the product.
-The product provides an estimate of forest composition, with uncertainty, for much of the northeastern United States at the time of European settlement.
-As such, it is valuable resource for ecologists and for climate change work as it provides baseline information about forests before widespread European land use change.
-
-The raw data underlying this product have not been collected and made available in one place to date as they represent the work of a number of researchers and in some cases data are still unpublished.
-Our product provides a publicly available resource.
-
-With regard to possible editors, either someone with expertise in statistics applied to environmental applications or someone with expertise in biogeography or landscape-scale ecology would be appropriate.
-There are a number listed on your website with expertise in biogeography, community ecology, global change ecology, or spatial and landscape ecology.
-In terms of editors with expertise in statistics, potential editors include Alex Cook, Thomas Nichols, Nicholas Reich.
+Thank you for your consideration of our manuscript.
 

### 2016-02-03 paper.R

@@ -46,14 +46,14 @@ presence_long <- presence_long[presence_long$value == 1, ]
 col = gray(c(.2,.5,.8))
 d <- ggplot(presence_long, aes(X, Y, group = id)) + geom_polygon(aes(fill = as.factor(value))) + 
   scale_fill_manual(values = col, guide = FALSE) 
-d <- add_map_albers(plot_obj = d, map_data = usFortified, dat = presence_long)
+d <- add_map_albers(plot_obj = d, map_data = paleonFortified, dat = presence_long)
 d <- d + scale_y_continuous(limits = c(70000, 1490000)) 
 d <- theme_clean(d)
 
 setwd(paperDir)
 
-pdf('fig1.pdf', width=4.85, height=3)
-#postscript('fig1.eps', width=4.85, height=3)
+#pdf('fig1.pdf', width=4.85, height=3)
+postscript('fig1.eps', width=4.85, height=3)
 print(d)
 dev.off()
 
@@ -72,6 +72,10 @@ figWthIndiv = 22*.5
 psd[psd > .25] = 0.25
 psdBreaks = c(0, 0.01, 0.03, 0.05, 0.075, 0.10, 0.15, 0.2, 0.25)
 
+cv = psd/pm
+cv[cv > 1.4] = 1.4
+cvBreaks = seq(0, 1.4, by = 0.2)
+
 
 
 
@@ -79,25 +83,27 @@ figs <- list()
 length(figs) <- length(focalTaxa) * 3
 cnt <- 1
 for( taxon in focalTaxa ) {
-  figs[[cnt]] <- make_areal_map(data = taxon_dat_long, variables = taxon, breaks = propBreaks, legendName = 'raw proportions', map_data = usFortified, facet = FALSE, ncol = 1, legend = FALSE, title = FALSE) + theme(plot.margin = unit(rep(0,4), 'lines'))
-  figs[[cnt + 1]] <- make_veg_map(data = pm[ , taxon, drop = FALSE], breaks = propBreaks, coords = coord, legendName = 'fitted proportions', map_data = usFortified, facet = FALSE, ncol = 1, legend = FALSE) + theme(plot.margin = unit(rep(0,4), 'lines'))
[…]

### 2016-02-03 plos_original/paper.R

@@ -1,0 +1,155 @@
+# code to generate figures for paper
+# run this code from ~/research/jmac/composition so use packrat controlled packages
+
+paperDir <- '~/research/jmac/composition_paper'
+
+source("config")
+
+source(file.path(codeDir, "plot.R"))
+source(file.path(codeDir, "prep_plot_data.R"))
+source(file.path(codeDir, "set_domain.R"))
+
+require(ncdf4)
+require(ggplot2)
+require(maptools)
+require(rgdal)
+require(raster)
+require(gridExtra)
+
+
+# for distinguishing no trees from no data
+west_presence <- read.csv(file.path(dataDir, 'western.csv'))
+west_presence <- west_presence[west_presence$x <= easternLimitOfWesternDomain, ]
+west_presence$data <- as.numeric(rowSums(west_presence[ , c(3:ncol(west_presence))]) > 0)
+# get in same order as data for fitting (from far N, W to E in rows)
+tmp <- matrix(west_presence$data, length(westernDomainY), length(westernDomainX))
+west_presence <- c(t(tmp[rev(seq_along(westernDomainY)), ]))
+
+west_presence_fort <- west_fort
+west_presence_fort$id <- as.numeric(west_presence_fort$id)
+west_presence_fort <- cbind(west_presence_fort, west_presence[west_presence_fort$id])
+names(west_presence_fort)[ncol(west_presence_fort)] <- "data"
+
+east_presence_fort <- east_fort
+east_presence_fort$data <- as.numeric(rowSums(east_presence_fort[ , 8:ncol(east_presence_fort)], na.rm = TRUE) > 0)
+
+fort_presence <- rbind(west_presence_fort[ , c(1:7, ncol(west_presence_fort))], east_presence_fort[ , c(1:7, ncol(east_presence_fort))])
+
+  
+#########################################################################################
+## Fig. 1
+#########################################################################################
+
+presence_long <- melt(fort_presence, c('X', 'Y', 'order', 'hole', 'piece', 'group', 'id'))
[…]

### 2016-02-03 run.sh

@@ -7,7 +7,7 @@ git clone https://github.com/PalEON-Project/composition
 cd composition
 
 # create final data product
-cp config_0.3-0 config
-./master.sh
+cp config_0.4-0 config
+./master.sh  # probably best to run individual components of this file and not expect full script to run without issues (plus the fitting takes two weeks)
 
 # now do CV analyses

### 2016-02-04 paper.R
cleaned up paper.R code

@@ -1,4 +1,5 @@
 # code to generate figures for paper
+
 # run this code from ~/research/jmac/composition so use packrat controlled packages
 
 paperDir <- '~/research/jmac/composition_paper'
@@ -52,8 +53,8 @@ d <- theme_clean(d)
 
 setwd(paperDir)
 
-#pdf('fig1.pdf', width=4.85, height=3)
-postscript('fig1.eps', width=4.85, height=3)
+pdf('fig1.pdf', width=4.85, height=3)
+#postscript('fig1.eps', width=4.85, height=3)
 print(d)
 dev.off()
 
@@ -92,18 +93,18 @@ for( taxon in focalTaxa ) {
 
 setwd(paperDir)
 
-#pdf('fig2.pdf', width=7, height=10)
+pdf('fig2.pdf', width=7, height=10)
 #postscript('fig2.eps', width=7, height=10)
-#do.call(grid.arrange, c(figs, nrow = length(focalTaxa), ncol = 3))
-#dev.off()
+do.call(grid.arrange, c(figs, nrow = length(focalTaxa), ncol = 3))
+dev.off()
 
 scaling = .5
 
 ix=1
 iy=seq(0,1,len=300)
 iz=matrix(iy,nc=300)
-#postscript(file='legendRaw.pdf',width=5/scaling,height=.7/scaling,paper="special")
-#pdf(file='legendRaw.eps',width=5/scaling,height=.7/scaling,paper="special")
+pdf(file='legendRaw.pdf',width=5/scaling,height=.7/scaling,paper="special")
+#postscript(file='legendRaw.eps',width=5/scaling,height=.7/scaling,paper="special")
 par(fig=c(0,1,0,1),mai=c(0.6,.2,0,.3),mgp=c(1.8,.7,0),cex.axis=1.5)
 cols <- rev(terrain.colors(length(propBreaks)-1))
 cols[1] <- terrain.colors(40)[39]
@@ -112,12 +113,12 @@ image(iy,ix,t(iz), yaxt = "n", xlab = "", xaxt='n',
 axis(1, at = seq(0,1, len = length(propBreaks)), labels = as.character(propBreaks),
      cex.axis = 1.8, tick = FALSE)
 box()
-#dev.off()
+dev.off()
 
 ix=1
 iy=seq(0,1,len=300)
 iz=matrix(iy,nc=300)
[…]

### 2016-02-12 paperPlos.tex
final minor typo changes requested by PLOS ONE to paperPlos

@@
 Jun Zhu\textsuperscript{8},
-Jason S. McLachlan\textsuperscript{7,\ddag}
+Jason S. McLachlan\textsuperscript{7}
 \\
@@
 % Also use this double-dagger symbol for special authorship notes, such as senior authorship.
- \ddag PI of the PalEON Project
 
@@
 In \emph{Data} we describe the data sources, while in \emph{Statistical model} we describe 
-the statistical model used to create the data product. In In \emph{Model comparison}
+the statistical model used to create the data product. In \emph{Model comparison}
  we quantitatively compare competing
@@
 subdomain are available through the NIS data portal as
-DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5 \cite{Gori:etal:data:2016}.
+DOI:10.6073/pasta/aa0ef9828d41569a96651b056ad89fb3 \cite{Gori:etal:data:2016}.
 The TPS data were compiled by C.V. Cogbill from a myriad of archival
@@
 with local settlement and are available through the
-NIS data portal as DOI:10.5072/FK2/4d44aebe71dbf65290d8ce1bde03598b
-and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}.
+NIS data portal as DOI:10.6073/pasta/e40c1ad172882d5113844296611451f6
+and DOI:10.6073/pasta/d09bb56c6af8ef0783cd7e76f8113b34 \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}.
 
@@
 \begin{equation}
-R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
+R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\nonumber
 \end{equation}
@@
 
-The full description of the MCMC sampling steps is provided in the
-\nameref{SI}. In addition, in the latent variable representation, $\theta_{p}(s)$
+The full description of the MCMC sampling steps is provided in \emph{Appendix}. In addition, in the latent variable representation, $\theta_{p}(s)$
 never appears explicitly and cannot be calculated in close form. Instead
 we use Monte Carlo integration over $W_{ijp},\, p=1,\ldots,P$ to
-estimate $\theta_{p}(s_{i})$, also described in the \nameref{SI}.
+estimate $\theta_{p}(s_{i})$, also described in \emph{Appendix}.
 
@@
 in the model comparison analyses) for 150,000 iterations with the
-same burn-in and subsampling details as described in Section \ref{sec:Model-comparison}.
+same burn-in and subsampling details as described in \emph{Model comparison}.
 Based on graphical checks and calculation of effective sample size
@@
 The data product is publicly available at the NIS
-Data Portal as DOI:10.5072/FK2/dfc5e9748c8df4c3c275eab0f42d9744 under
+Data Portal as DOI:10.6073/pasta/8544e091b64db26fdbbbafd0699fa4f9 under
 the CC BY 4.0 license as version 0.4 as of January 2016 \cite{paci:etal:data:2016}.
@@
 
-\section*{S1 Appendix}
+\section*{Appendix}
 \label{SI}
@@
 
-{\bf Define $\bar{w}_{ip}=\frac{1}{n_{i}}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
+Define $\bar{w}_{ip}=\frac{1}{n_{i}}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
 as the average of the $W$ values for the \emph{p}th taxon in the \emph{i}th
-grid cell and $\bar{w}_{p}=\{\bar{w}_{ip}\},$ $i=1,\ldots,m$.} Let
+grid cell and $\bar{w}_{p}=\{\bar{w}_{ip}\},$ $i=1,\ldots,m$. Let
 $A$ be a diagonal matrix where $A_{ii}$ is the number of trees in
@@
 \mbox{TN}\big(-\infty,w_{ijy_{ij}},\alpha_{p}(s_{i}),1\big), & \mbox{if }p\mbox{ \ensuremath{\neq y_{ij}}}
-\end{cases}
+\end{cases} \nonumber
 \end{align}
@@
 \begin{align}
-\alpha_{p} & \sim\N\bigg(\Big(A+Q_{p}\Big)^{-1}A\bar{w}_{p},\Big(A+Q_{p}\Big)^{-1}\bigg).
+\alpha_{p} & \sim\N\bigg(\Big(A+Q_{p}\Big)^{-1}A\bar{w}_{p},\Big(A+Q_{p}\Big)^{-1}\bigg). \nonumber
 \end{align}
@@
 \label{SI1}
-{\bf In the latent variable representation, $\theta_{p}(s)$ never appears
-explicitly and cannot be calculated in closed form.} Instead we use
+In the latent variable representation, $\theta_{p}(s)$ never appears
+explicitly and cannot be calculated in closed form. Instead we use
 Monte Carlo integration over $W_{ijp},\, p=1,\ldots,P$ to estimate
@@
 \bibitem{pattison1957beginnings}
-Pattison WD, et~al.
+Pattison WD.
 \newblock Beginnings of the American rectangular land survey system, 1784-1800.
@@
 \bibitem{gray2012forest}
-Gray AN, Brandeis TJ, Shaw JD, McWilliams WH, Miles PD, et~al.
+Gray AN, Brandeis TJ, Shaw JD, McWilliams WH, Miles PD.
 \newblock Forest Inventory and Analysis Database of the {United States} of
@@
 \newblock Settlement-era gridded tree composition, midwestern {US}: Level 1.
-\newblock Long Term Ecological Research Network Data Portal; 2016. Paleon.2.0.
+\newblock Long Term Ecological Research Network Data Portal; 2016. http://dx.doi.org/10.6073/pasta/aa0ef9828d41569a96651b056ad89fb3.
 
@@
 \newblock Settlement-era tree composition, eastern {US}: level 1.
-\newblock Long Term Ecological Research Network Data Portal; 2016. Paleon.4.11.
+\newblock Long Term Ecological Research Network; 2016. http://dx.doi.org/10.6073/pasta/e40c1ad172882d5113844296611451f6.
 
@@
 \newblock Settlement-era tree composition, {O}hio: level 1.
-\newblock Long Term Ecological Research Network Data Portal; 2016. Paleon.5.11.
+\newblock Long Term Ecological Research Network; 2016. http://dx.doi.org/10.6073/pasta/d09bb56c6af8ef0783cd7e76f8113b3.
 
@@
 \newblock Settlement-era gridded tree composition, northeastern {US}: level 2.
-\newblock Long Term Ecological Research Network Data Portal; 2016. Paleon.1.2.
+\newblock Long Term Ecological Research Network; 2016. http://dx.doi.org/10.6073/pasta/8544e091b64db26fdbbbafd0699fa4f9.
 

### 2016-04-03 paperPlos.tex
final changes to version numbers and urls for PLOS (committing a couple months late...)

@@
 \\
-\bf{3} Department of Statistics, University of Iowa, Iowa City, Iowa, USA
+\bf{3} VA Office of Rural Health, Veterans Rural Health Resource Center, Iowa City VAMC, Iowa City, Iowa, USA
 \\
@@
 change after intensive settlement. The data product is being made
-available at the NIS data portal as version 0.4.
+available at the NIS data portal as version 1.0.
 
@@
 per cell. The gridded data at the 8 km resolution for the midwest
-subdomain are available through the NIS data portal as
-DOI:10.6073/pasta/aa0ef9828d41569a96651b056ad89fb3 \cite{Gori:etal:data:2016}.
+subdomain are available through the NIS data portal \cite{Gori:etal:data:2016}.
 The TPS data were compiled by C.V. Cogbill from a myriad of archival
@@
 with local settlement and are available through the
-NIS data portal as DOI:10.6073/pasta/e40c1ad172882d5113844296611451f6
-and DOI:10.6073/pasta/d09bb56c6af8ef0783cd7e76f8113b34 \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}.
+NIS data portal \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}.
 
@@
 for the composition in the 23 taxonomic groupings at each of the grid
-cells. These samples are the data product (described further in the
-\emph{Data Product} section) and can be used in subsequent analyses.
+cells. These samples are the data product (described further in
+\emph{Data Product}) and can be used in subsequent analyses.
 The mean and standard deviation of the samples for each pair of cell
@@
 The data product is publicly available at the NIS
-Data Portal as DOI:10.6073/pasta/8544e091b64db26fdbbbafd0699fa4f9 under
-the CC BY 4.0 license as version 0.4 as of January 2016 \cite{paci:etal:data:2016}.
+Data Portal under
+the CC BY 4.0 license as version 1.0 as of February 2016 \cite{paci:etal:data:2016}.
 The product is in the form of a netCDF-4 file, with dimensions x-coordinate,
@@
 Goring SJ, {University of Wisconsin}.
-\newblock Settlement-era gridded tree composition, midwestern {US}: Level 1.
-\newblock Long Term Ecological Research Network Data Portal; 2016. http://dx.doi.org/10.6073/pasta/aa0ef9828d41569a96651b056ad89fb3.
+\newblock Settlement-era gridded tree composition, midwestern {US}: level 1.
+\newblock Long Term Ecological Research Network; 2016. http://dx.doi.org/10.6073/pasta/aa0ef9828d41569a96651b056ad89fb3.
 

### 2016-04-03 arxiv/paperArxiv.tex
adding updated arxiv materials

@@
 \usepackage[authoryear]{natbib}
+\usepackage[unicode=true]{hyperref}
 
@@
 
-\lyxaddress{\textsuperscript{1}Department of Statistics, University of California, Berkeley, California, USA}
-
-\lyxaddress{\textsuperscript{2}Department of Geography, University of Wisconsin, Madison, Wisconsin, USA}
-
-\lyxaddress{\textsuperscript{3}Department of Statistics, University of Iowa, Iowa City, Iowa, USA}
-
-\lyxaddress{\textsuperscript{4}Harvard Forest, Harvard University, Petersham, Massachusetts, USA}
+\lyxaddress{\textsuperscript{1}Department of Statistics, University of California,
+Berkeley, California, USA}
+
+\lyxaddress{\textsuperscript{2}Department of Geography, University of Wisconsin,
+Madison, Wisconsin, USA}
+
+\lyxaddress{\textsuperscript{3}VA Office of Rural Health, Veterans Rural Health
+Resource Center, Iowa City VAMC, Iowa City, Iowa, USA}
+
+\lyxaddress{\textsuperscript{4}Harvard Forest, Harvard University, Petersham,
+Massachusetts, USA}
 
@@
 
-\lyxaddress{\textsuperscript{6}Department of Forest and Wildlife Ecology,
-University of Wisconsin, Madison, Wisconsin, USA}
+\lyxaddress{\textsuperscript{6}Department of Forest and Wildlife Ecology, University
+of Wisconsin, Madison, Wisconsin, USA}
 
@@
 change after intensive settlement. The data product is being made
-available at the NIS data portal as version 0.4 under the CC BY license.
+available at the NIS data portal as version 1.0.
 
@@
 consistency among points and the application of spatially-varying
-correction factors \citep{cogbill2015corrections,goring2015composition}
-to accurately assess tree stem density, basal area and biomass from
-the early settlement records, but the impact on composition estimates
-is limited \citep{liu2011broadscale}. Surveyors sometimes used ambiguous
-common names, which requires matching to scientific names and standardization
-\citep{mladenoff2002narrowing,goring2015composition}.
+correction factors \citet{goring2015composition} to accurately assess
+tree stem density, basal area and biomass from the early settlement
+records, but the impact on composition estimates is limited \citep{liu2011broadscale}.
+Surveyors sometimes used ambiguous common names, which requires matching
+to scientific names and standardization \citep{mladenoff2002narrowing,goring2015composition}.
 
@@
 we describe the final data product. In Section \ref{sec:Discussion}
-we discuss limitations of the statistical model, and we list related
-data products under development.
+we discuss the uncertainties estimated by and the
+limitations of the statistical model, and we list related data products
+under development.
 
@@
 six New England states (the eastern subdomain), data are aggregated
-at the township level. There are also data from a single township
-in Quebec and a single township in northern Delaware. Digitization
-of PLS data in Minnesota, Wisconsin and Michigan is essentially complete,
-with PLS data for nearly all 8 km grid cells, but data in Illinois
-and Indiana represent a sample of the full set of grid cells, with
-survey record transcription ongoing. Data for the eastern states are
-available for a subset of the full set of townships covering the domain;
-the TPS data for some townships were lost, incomplete, or have not
-been located \citep{Cogb:etal:2002}.
+at the township level. We make predictions for all
+of the states listed above; these constitute our core domain. There
+are also data from a single township in Quebec and a single township
+in northern Delaware; these data help inform predictions
+in nearby locations within our core domain, but predictions are not
+made for Quebec or Delaware. Digitization of PLS data in Minnesota,
+Wisconsin and Michigan is essentially complete, with PLS data for
+nearly all 8 km grid cells, but data in Illinois and Indiana represent
+a sample of the full set of grid cells, with survey record transcription
+ongoing. Data for the eastern states are available for a subset of
+the full set of townships covering the domain; the TPS data for some
+townships were lost, incomplete, or have not been located \citep{Cogb:etal:2002}.
 
@@
 Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
-data and processing steps are available in \cite{goring2015composition};
+data and processing steps are available \cite{goring2015composition};
 key elements include the use of only corner points, the use of only
@@
 on the number of survey points in a cell, with an average of 124 trees
-per cell. The gridded data at the 8 km resolution for the upper Midwest
-are available through the NIS Data portal (accession number and version
-to be inserted when available). The TPS data were compiled by C.V.
-Cogbill from a myriad of archival sources representing land division
-surveys conducted in connection with local settlement.
+per cell. The gridded data at the 8 km resolution for the midwest
+subdomain are available through the NIS data portal \citep{Gori:etal:data:2016}.
+The TPS data were compiled by C.V. Cogbill from a myriad of archival
+sources representing land division surveys conducted in connection
+with local settlement and are available through the
+NIS data portal \citep{Cogb:dataNE:2016,Cogb:dataOH:2016}.
 
@@
 \end{enumerate}
+At a high level, the Bayesian statistical model estimates
+composition across the domain, even in locations with sparse or no
+data, by combining the raw composition data with the assumption that
+composition varies in a smooth spatial fashion across the domain.
+The information in the data is quantified by the data model, also
+known as the likelihood. The assumption of smoothness is built into
+the model by representing the true unknown spatially-varying composition
+using a statistical spatial process representation that induces smoothing
+of estimates across nearby locations. This spatial process representation
+is a form of prior distribution and is a function of model parameters
+called hyperparameters that determine the correlation structure of
+the process and are also estimated based on the data.
+
 The result of fitting the Bayesian model via Markov chain Monte Carlo
@@
 cells. These samples are the data product (described further in the
-Data Product section) and can be used in subsequent analyses. The
-mean and standard deviation of the samples for each pair of cell and
-taxon represent our best estimate (i.e., prediction) of composition
+Section \ref{sec:Data-product}) and can be used in subsequent analyses.
+The mean and standard deviation of the samples for each pair of cell
+and taxon represent our best estimate (i.e., prediction) of composition
 and a Bayesian ``standard error'' quantifying the uncertainty in
 the estimate.
+
+In the remainder of this section we provide the technical
+specification of the model and of the computations involved in fitting
+the model.
 
@@
 random field (MRF) approximation to a Gaussian process with \matern
-covariance \citep{Lind:etal:2011}.
+covariance \citep{Lind:etal:2011}. These models are
+specific statistical formulations of spatial processes that represent
+spatial correlation by defining neighborhoods around each location
+that are used to help inform predictions at the location.
 
@@
 
-MRF models work directly with the precision matrix of the values of
-the spatial process, so calculation of the prior density of $\alpha_{p}$
-is computationally simple \citep{Rue:Held:2005}, but in situations
-where the likelihood is not normal, it can be difficult to set up
-effective MCMC algorithms that are able to move in the high-dimensional
-space of $\alpha_{p}$. The latent variable representation of Section
-\ref{sub:Latent-Variable-Model} helps to alleviate this problem.
-Next we describe two alternative spatial models that we considered;
-in Section \ref{sec:Model-comparison}, we use cross-validation to
-choose between these two representations.
+MRF models represent the neighborhood information
+by working directly with the precision matrix (the
+inverse of the covariance matrix) of the values of the spatial process,
+so calculation of the prior density of $\alpha_{p}$ is computationally
+simple \citep{Rue:Held:2005}. However, in situations
+in which the likelihood is not normal, such
+as our multinomial likelihood, it can be difficult to set up effective
+MCMC algorithms that are able to move in the high-dimensional space
+of $\alpha_{p}$. The latent variable representation of Section \ref{sub:Latent-Variable-Model}
+helps to alleviate this problem. Next we describe two alternative
+spatial models that we considered; in Section \ref{sec:Model-comparison},
+we evaluate the models on held-out data to choose
+between the two.
 
@@
 
-Our first model is a standard conditional autoregressive (CAR) model
-\citep{Bane:etal:2004}. We use a standard form of this model, which
-treats the four cardinal neighbors of each grid cell as the neighbors
-of the grid cell. The corresponding precision matrix has diagonal
-elements, $Q_{ii}$, equal to the number of neighbors for the $i$th
-area (i.e., four except for cells on the boundary of the domain),
-while $Q_{ik}=-1$ (the negative of a weight of one) when areas $i$
-and $k$ are neighbors and $Q_{ik}=0$ when they are not. This gives
-the following model for the values of $\alpha_{p}(s_{i})$ collected
-as a vector across all of the grid cells, $i=1,\ldots,m$: 
+Our first model is a standard conditional autoregressive (CAR) model;
+technical details can be found in \citep{Bane:etal:2004}. We use
+a standard form of this model that treats the four
+cardinal neighbors of each grid cell as the neighbors of the grid
+cell. The corresponding precision matrix has diagonal elements, $Q_{ii}$,
+equal to the number of neighbors for the $i$th area (i.e., four except
+for cells on the boundary of the domain), while $Q_{ik}=-1$ (the
+negative of a weight of one) when areas $i$ and $k$ are neighbors
+and $Q_{ik}=0$ when they are not. This gives the following model
+for the values of $\alpha_{p}(s_{i})$ collected as a vector across
+all of the grid cells, $i=1,\ldots,m$: 
 \[
@@
 not full-rank, but is of rank $m-1$; this gives an improper prior
-on an implicit overall mean for the process values. This specification
+on an implicit overall mean for the process values. Note
+that we specify an explicit mean of zero because a non-zero mean would
+not be identifiable in light of the implicit mean. This specification
 is called an \textit{intrinsic conditional autoregression (ICAR)}
@@
 GP models are computationally challenging for large datasets because
-of manipulations involving large covariance matrices. Given this,
-\cite{Lind:etal:2011} proposed a new framework for using Gaussian
-MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
-partial differential equations (SPDEs).
-
-We consider Gaussian processes in the \matern class, using the following
-parameterization of the \matern correlation function, 
+of computational manipulations involving large covariance
+matrices. Given this, \cite{Lind:etal:2011} proposed a new framework
+for using Gaussian MRFs (GMRFs) as approximations to GPs, based on
+the use of stochastic partial differential equations (SPDEs).
+
+Gaussian processes are generally constructed using
+one of a number of correlation functions that define how the strength
+of correlation between the values of the process at two locations
+decays as a function of the distance between the locations. We consider
+Gaussian processes in the commonly-used \matern
+class, using the following parameterization of the \matern correlation
+function, 
 \begin{equation}
-R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
+R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\nonumber
 \end{equation}
@@
 
-The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
-to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second spatial
-model is this Lindgren approximation for \matern-based GPs with $\nu=1$.
-To implement the Lindgren model, one modifies the $Q$ matrix defined
-previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
+The approach of \citet{Lind:etal:2011} allows us to consider MRF
+approximations to the \matern-based GP for $\nu=1$ and $\nu=2$.
+Our second spatial model is this Lindgren approximation for \matern-based
+GPs with $\nu=1$. To implement this Lindgren model, one modifies
+the $Q$ matrix defined previously as follows based
+on the technical specification of the precision matrix provided in
+\citet{Lind:etal:2011}. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
 elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
@@
 $\rho$. In particular $\rho$ allows us to estimate the locality
-of the smoothing. As $\rho$ decreases, the model uses increasingly
-localized data to estimate the compositional proportions at a given
-location, effectively averaging the empirical proportions over smaller
-neighborhoods. In general, the \cite{Lind:etal:2011} model will generally
-provide for a smoother estimate than the CAR model \citep{Paci:2013}.
+of the spatial smoothing. As $\rho$ decreases, the
+model uses increasingly localized data to estimate the compositional
+proportions at a given location, effectively averaging the empirical
+proportions over smaller neighborhoods. In general, the \cite{Lind:etal:2011}
+model will generally provide for a smoother estimate than the CAR
+model \citep{Paci:2013}.
 
@@
 we use a uniform distribution on each $\sigma_{p}$ parameter, with
-upper bound of 1000. For the SPDE model we also have parameters $\{\mu_{p}\}$,
-which we give flat, non-informative priors (truncated at $\pm10$),
-and $\{\rho_{p}\}$, which we give uniform priors on the interval
-$(0.1,\exp(5))$.
+upper bound of 1000. For the SPDE model we also have hyperparameters
+$\{\mu_{p}\}$, which we give flat, non-informative priors (truncated
+at $\pm10$), and $\{\rho_{p}\}$, which we give uniform priors on
+the interval $(0.1,\exp(5))$. These various hyperparameters
+are unknown parameters that control the spatial structure of the two
+spatial models and are estimated from both the data and the prior
+distributions just specified based on the Bayesian approach.
 
@@
 per tree, that indicate the grid cells in which the trees are located
-and can be sampled within the MCMC as additional unknown parameters.
+and that can be sampled within the MCMC as additional unknown parameters.
 Specifically, $c_{tj}$ is the latent ``membership'' variable for
@@
 using the \emph{spam} package in R \citep{Furr:Sain:2010}. All code
-is available on Github, including pre- and post-processing code, at
-https://github.com/PalEON-Project/composition.
+is available on Github, including pre- and post-processing code, at\\\url{https://github.com/PalEON-Project/composition}.
 
@@
 
-We compared the CAR and SPDE models using cross-validation by holding
-out data from the fitting process and assessing the fit of the model
-on the held-out data. We used two cross-validation experiments:
+We compared the CAR and SPDE models by holding out
+data from the fitting process and assessing the fit of the model
+on the held-out data. We used two experiments\textcolor{red} with
+held-out data:
 \begin{enumerate}
@@
 is an indicator variable taking value either 0 or 1 depending on whether
-the $j$th held-out tree in the $i$th cell is of taxon $p$. We calculated
-each of the metrics in two ways. First, we used the posterior mean
-composition estimates (as an evaluation of our core predictions),
-with $\tilde{\theta}_{p}(s)$ being the posterior mean. Second, we
-averaged the metric over the posterior samples (as an evaluation of
-our full data product, including uncertainty), taking $\tilde{\theta}_{p}(s)$
-to be an individual MCMC sample and then averaging the metric over
-all the posterior samples. 
+the $j$th held-out tree in the $i$th cell is of taxon $p$. $\hat{\theta}_{ip}=Y_{ip}/n_{i}$
+is the empirical proportion in category $p$ for the $n_{i}$ held-out
+trees in cell $i$. We calculated each of the metrics in two ways.
+First, we used the posterior mean composition estimates (as an evaluation
+of our core predictions), with $\tilde{\theta}_{p}(s)$ being the
+posterior mean. Second, we averaged the metric over the posterior
+samples (as an evaluation of our full data product, including uncertainty),
+taking $\tilde{\theta}_{p}(s)$ to be an individual MCMC sample and
+then averaging the metric over all the posterior samples. 
 \begin{enumerate}
@@
 $i$. 
-\item Log predictive density: This metric takes the log of the density of
-held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
+\item Log predictive density: This metric takes the log of the probability
+density of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
 summing on the log scale across all of the held-out data.
@@
 
-For Experiment 1 for cells in the Minnesota subregion held out of
-the fitting process, the CAR model outperforms the SPDE model based
-on the posterior distribution over the predictive metric values (Table
-\ref{tab:score_cell_fivepercent}). For the posterior mean predictions,
-the SPDE model appears to outperform the CAR model to a lesser degree,
-but we do not have any uncertainty estimates for this comparison.
-Coverage and interval lengths are similar between the two models (Table
-\ref{tab:coverage_cell_fivepercent}). From a practical perspective,
-based on the difference in mean absolute error, the differences between
-the models are small (Table \ref{tab:score_cell_fivepercent}).
+For Experiment 1 (full cells held out) for cells in the Minnesota
+subregion held out of the fitting process, the CAR model outperforms
+the SPDE model based on the posterior distribution over the predictive
+metric values (Table \ref{tab:score_cell_fivepercent}). For the posterior
+mean predictions, the SPDE model appears to outperform the CAR model
+to a lesser degree, but we do not have any uncertainty estimates for
+this comparison. Coverage and interval lengths are similar between
+the two models (Table \ref{tab:coverage_cell_fivepercent}). From
+a practical perspective, based on the difference in mean absolute
+error, the differences between the models are small (Table \ref{tab:score_cell_fivepercent}).
 
@@
 
-In Experiment 2, we have evidence (posterior probability of 0.93)
-that the SPDE model is better based on the Brier score, but the Brier
-score values for the two models are numerically almost the same (Table
-\ref{tab:score_tree}).
+In Experiment 2 (individual trees held out), we have evidence (posterior
+probability of 0.93) that the SPDE model is better based on the Brier
+score, but the Brier score values for the two models are numerically
+almost the same (Table \ref{tab:score_tree}).
 
@@
 \includegraphics[scale=0.75]{fig2}
-
-\hspace{4mm}\includegraphics[scale=0.16]{legendRaw}\hspace{4mm}\includegraphics[scale=0.16]{legendProp}\hspace{3.5mm}\includegraphics[scale=0.16]{legendSD}
+\vspace{2mm}
+\hspace{2mm}\includegraphics[scale=0.16]{legendRaw}\hspace{4mm}\includegraphics[scale=0.16]{legendProp}\hspace{4mm}\includegraphics[scale=0.16]{legendSD}
 
@@
 
-The data product is being made publicly available at the NIS Data
-Portal (accession number to be inserted when available) under the
-CC BY license as version 0.4 as of August 2015. The product is in
-the form of a netCDF-4 file, with dimensions x-coordinate, y-coordinate,
-and MCMC iteration. There is one variable per taxon. In addition,
-dynamic visualizations of the product using the Shiny tool are available
-at http://gandalf.berkeley.edu:3838/paciorek/setVegComp {[}this URL
-is temporary and will be updated{]}. The PalEON Project (in particular
-the first author) will continue to maintain this product, releasing
-new versions as additional data in Illinois, Indiana and Ohio are
-digitized. Note that digitization of data from Illinois and Indiana
-is ongoing, and digitization of additional data from Ohio is planned
-as well. As a result, at some point we expect to have complete data
-for the midwestern half of the domain.
+The data product is publicly available at the NIS
+Data Portal under the CC BY 4.0 license as version 1.0 as of January
+2016 \citep{paci:etal:data:2016}. The product is in the form of a
+netCDF-4 file, with dimensions x-coordinate, y-coordinate, and MCMC
+iteration. There is one variable per taxon. In addition, dynamic visualizations
+of the product using the Shiny tool are available at \href{https://www3.nd.edu/~paleolab/paleonproject/maps}{https://www3.nd.edu/$\sim$paleolab/paleonproject/maps}.
+The PalEON Project (in particular the first author) will continue
+to maintain this product, releasing new versions as additional data
+in Illinois, Indiana and Ohio are digitized. Note that digitization
+of data from Illinois and Indiana is ongoing, and digitization of
+additional data from Ohio is planned as well. As a result, at some
+point we expect to have complete data for the midwestern half of the
+domain.
 
@@
 
-Given the density of data and the limited differences seen between
-the CAR and SPDE models, we expect the data product to be reasonably
-robust to the choice of spatial model, particularly in those areas
-with complete data. However, additional investigation of other statistical
-representations is of interest, in particular nonstationary spatial
-models and use of covariates. The biggest shortcoming of the current
-model is its inability to account for local features such as rivers
-(e.g., the Minnesota River riparian corridor can be seen in the raw
-data but is missing in the statistical estimates). The current model,
-by using a simple stationary spatial model that smooths as a function
-of Euclidean distance, does not account for topographic, soil, or
-other features. An additional drawback of the product is its focus
-on composition, which does not directly tell us about tree density
-or other aspects of vegetation structure, and in particular does not
-distinguish between closed forest, savanna, and prairie. This limitation
-is particularly critical at the prairie-forest transition in Minnesota,
-Wisconsin, Illinois, and into Indiana \citep{transeau1935prairie}.
-
-This latter limitation can be addressed by developing estimates of
-absolute abundance (e.g., biomass) rather than the relative abundance
-compositional data used here. A gridded dataset of biomass, stem density,
-and basal area is already available for Minnesota, Wisconsin, and
-northern Michigan \citep{goring2015composition}, based on the PLS
-data. An extension to southern Michigan, Illinois, and Indiana is
-planned. We are currently developing statistical estimates of biomass
+In the parts of the modeled region with spatially
+complete data (in particular, Minnesota, Wisconsin, and Michigan),
+the statistical estimates of forest composition closely match the
+patterns apparent in the raw data (Fig. \ref{fig:select_maps}), as
+expected.  In these areas, the estimated tree composition from the
+model has the advantage of downweighting unusual or outlier values
+in the empirical proportions of individual grid cells, which are likely
+due to stochastic sampling variation within that grid cell (compare
+the first two columns in Fig. \ref{fig:select_maps}). Some stochastic
+variation is expected given that, even in the most spatially complete
+regions, each grid cell contains an average of 124 trees (120-135
+is typical) \citep{goring2015composition} and some cells contain
+many fewer trees. Hence, some smoothing of this stochastic variation
+is appropriate. This smoothing is based on information on data from
+nearby cells, and the estimates from the model reflect the smooth
+trends in forest composition across the spatial domain. A partial
+cost is that these maps can smooth out sharp ecotones or other forms
+of true spatial heterogeneity, particularly in areas with sparse data
+(including areas with low tree density). For example, the sharp increase
+in Elm along the Minnesota River (Fig. \ref{fig:select_maps}, first
+column) likely represents a real ecotone in the settlement-era vegetation.
+Vegetation gradients and ecotones were sharper in the settlement-era
+forests in the upper Midwest than in contemporary forests \citep{goring2015composition},
+and the modeled estimates may partially obscure this change. Users
+interested in using the original unsmoothed data are directed to the
+data product described earlier \citep{Gori:etal:data:2016}. Additional
+investigation of other statistical representations to better capture
+sharp gradients is of interest, in particular nonstationary spatial
+models and use of covariates. Potential environmental covariates include
+soils, firebreaks, and topography \citep{grimm1984fire,shea2014reconstructing}.
+Here, however, we chose to limit our model to be exclusively a function
+of spatial distance in order to avoid dependence on the environmental
+drivers of pre-settlement forest composition that might lead to circular
+reasoning in subsequent inferences based on this dataset. Use of covariates
+could also lead to prediction that a taxa is present well beyond its
+range boundary in places where data are sparse.
+
+A key advance of this work over prior reconstructions
+of settlement-era vegetation lies in the estimates of uncertainty
+across the spatial domain. These estimates of uncertainty include
+the sampling uncertainty within grid cells (as do the within-grid
+cell estimates of uncertainty available from the raw proportions),
+but, because this is a spatial model, predictions and their associated
+uncertainty estimates are also informed by the information content
+of nearby cells. The maps of standard errors across species (Fig.
+\ref{fig:select_maps}, third column) highlight the advantages of
+this approach in areas of high data coverage (Minnesota, Wisconsin,
+Michigan) and in areas of sparse coverage (e.g., Illinois, Indiana,
+parts of Ohio). Where there are not large gaps in the data, the model
+provides low and fairly smooth estimates of uncertainty. Uncertainty
+is generally higher in the eastern subdomain than in the areas of
+the midwestern subdomain with high data coverage because of missing
+townships and lower sampling density even in townships with data.
+In areas of sparse coverage and in areas with low tree density (e.g.,
+southwestern Minnesota), the standard error of our estimates increases
+appropriately. Nevetheless, these uncertainties surround reasonable
+estimates of trends in composition. For example, the model does a
+good job of capturing the oak ecotone in Indiana and Illinois, representing
+a shift from oak savannas and woodlands to closed mesic forests (Fig.
+\ref{fig:select_maps}). Experiment 1 showed that both models predicted
+composition at cells with no data reasonably well, mimicking the case
+of sparsely sampled data and giving confidence in the broad spatial
+patterns predicted in more poorly sampled regions, particularly those
+with regular, but sparse sampling that mimic the experiment (Illinois
+and Indiana, but not Ohio). The apparent blockiness of uncertainty
+estimates in a few places such as Ohio is caused by spatial gaps and
+variations in sampling resolution. Absolute uncertainty generally
+increases with abundance for all taxa (Fig. \ref{fig:select_maps},
+column 3).
+
+The exploration of alternative approaches to spatial
+modeling of composition showed similar results for the SPDE and CAR
+models, both in terms of prediction accuracy and performance of prediction
+intervals. Small differences among the various metrics of goodness
+of fit favored each model in turn, but applied users of the models
+would find little pragmatic difference affecting scientific inference.
+Ultimately, we slightly favor the CAR model, because it avoids the
+boundary effects apparent in the SPDE model at the edges of the domain.
+
+The models presented here estimate only the relative
+abundance of tree taxa, which does not directly tell us about tree
+density or other aspects of vegetation structure. This becomes a particular
+limitation for interpreting vegetation where trees become sparse at
+the prairie-forest transition from northern Minnesota through southern
+Illinois \citep{transeau1935prairie}. Our model (correctly) predicts
+that tree composition there is dominated by oak, but this is less
+useful considering the sparseness of trees. This limitation can be
+addressed by developing estimates of absolute abundance (e.g., biomass)
+rather than compositional estimates. A gridded dataset of biomass,
+stem density, and basal area is already available for Minnesota, Wisconsin,
+and northern Michigan \citep{goring2015composition}, based on the
+PLS data. An extension to southern Michigan, Illinois, and Indiana
+is planned. We are currently developing statistical estimates of biomass
 for Minnesota, Wisconsin, and Michigan using a statistical model applied
@@
 
-CJP, AT, and JZ developed the statistical model and code. CJP carried
+CJP, ALT, and JZ developed the statistical model and code. CJP carried
 out the model comparison and created the data product. CJP wrote the
-paper based on an initial draft by AT and with feedback and editing
+paper based on an initial draft by ALT and with feedback and editing
 from SJG, CVC, JWW, DJM, JAP, JZ and JSM. SJG, JAP, CVC, DJM, JSM,
@@
 Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
-We thank Madeline Ruid, Ben Seliger, Morgan Ripp and Daniel Handel
+We thank Madeline Ruid, Benjamin Seliger, Morgan Ripp and Daniel Handel
 for processing of the southern Michigan data. Indiana and Illinois
@@
 
-% \bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
+%\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
 \bibliography{paperArxiv}
@@
 \mbox{TN}\big(-\infty,w_{ijy_{ij}},\alpha_{p}(s_{i}),1\big), & \mbox{if }p\mbox{ \ensuremath{\neq y_{ij}}}
-\end{cases}\label{sampW}
+\end{cases} \nonumber
 \end{align}
@@
 \begin{align}
-\alpha_{p} & \sim\mbox{N}\bigg(\Big(A+Q_{p}\Big)^{-1}A\bar{w}_{p},\Big(A+Q_{p}\Big)^{-1}\bigg).\label{sampalpha}
+\alpha_{p} & \sim\mbox{N}\bigg(\Big(A+Q_{p}\Big)^{-1}A\bar{w}_{p},\Big(A+Q_{p}\Big)^{-1}\bigg). \nonumber
 \end{align}

### 2016-04-03 paper.tex
adding updated arxiv materials

@@
 
-\lyxaddress{\textsuperscript{3}Department of Statistics, University of Iowa,
-Iowa City, Iowa, USA}
+\lyxaddress{\textsuperscript{3}VA Office of Rural Health, Veterans Rural Health
+Resource Center, Iowa City VAMC, Iowa City, Iowa, USA}
 
@@
 
-%
 \begin{abstract}
@@
 per cell. The gridded data at the 8 km resolution for the midwest
-subdomain are available through the NIS data portal \textcolor{red}{as
-DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5} \textcolor{red}{\citep{Gori:etal:data:2016}}.
+subdomain are available through the NIS data portal \textcolor{red}{\citep{Gori:etal:data:2016}}.
 The TPS data were compiled by C.V. Cogbill from a myriad of archival
@@
 with local settlement\textcolor{red}{{} and are available through the
-NIS data portal as DOI:10.5072/FK2/4d44aebe71dbf65290d8ce1bde03598b
-and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \citep{Cogb:dataNE:2016,Cogb:dataOH:2016}}.
+NIS data portal \citep{Cogb:dataNE:2016,Cogb:dataOH:2016}}.
 
@@
 The data product is publicly available at the \textcolor{red}{NIS
-Data Portal as DOI:10.5072/FK2/dfc5e9748c8df4c3c275eab0f42d9744} under
-the CC BY 4.0 license as version 0.4 as of January 2016 \citep{paci:etal:data:2016}.
-The product is in the form of a netCDF-4 file, with dimensions x-coordinate,
-y-coordinate, and MCMC iteration. There is one variable per taxon.
-In addition, dynamic visualizations of the product using the Shiny
-tool are available at \href{https://www3.nd.edu/~paleolab/paleonproject/maps}{https://www3.nd.edu/$\sim$paleolab/paleonproject/maps}.
+Data Portal }under the CC BY 4.0 license as version 1.0 as of January
+2016 \citep{paci:etal:data:2016}. The product is in the form of a
+netCDF-4 file, with dimensions x-coordinate, y-coordinate, and MCMC
+iteration. There is one variable per taxon. In addition, dynamic visualizations
+of the product using the Shiny tool are available at \href{https://www3.nd.edu/~paleolab/paleonproject/maps}{https://www3.nd.edu/$\sim$paleolab/paleonproject/maps}.
 The PalEON Project (in particular the first author) will continue
